---
title: "Lab 4: Categorical Variables and Planned Comparisons"
---

{{< include ../_common.qmd >}}

## Goals

::: {.callout-note}
## What You'll Learn in This Lab
- Understand why `factor()` is essential for categorical predictors
- Interpret dummy-coded regression coefficients with a three-level predictor
- Use `emmeans` for planned comparisons and custom contrasts
- Apply the Sidak correction for pairwise comparisons
- Change the reference group and see how coefficients change
- **Lecture Notes Reference:** Chapter 3: *One-Way ANOVA*
:::

::: {.callout-tip}
## Stata → R Translation
| Stata | R | Notes |
|:------|:--|:------|
| `regress salary i.rank` | `lm(salary ~ factor(rank))` | Use `factor()` for categorical variables |
| `lincom full-assoc` | `emmeans::contrast()` | Linear combinations via estimated marginal means |
| `oneway salary rank, sidak` | `emmeans::pairs(adjust = "sidak")` | Pairwise comparisons with multiplicity correction |
:::

**Dataset:** `faculty` from the `regdatasets` package


## Part 1: Quick Review --- ANOVA with the Faculty Data

In Lab 3, we explored the `faculty` dataset and tested whether salary differs by gender. Let's reload the data and briefly revisit that result before moving on to the main topic: using **rank** as a categorical predictor.

```{r}
#| label: lab04-setup

library(regdatasets)
data(faculty)
glimpse(faculty)
```

A quick reminder of what we found last time --- a one-way ANOVA of salary by gender:

```{r}
#| label: review-gender-anova

model_gender <- lm(salary ~ male, data = faculty)
summary(model_gender)
```

> The coefficient for `male` is the mean salary difference between men and women. We established in Lab 3 that ANOVA and regression with a dummy predictor give identical results. Today, we extend this idea to a predictor with **three** levels: academic rank.


## Part 2: Regression with a Categorical Predictor (Rank)

The variable `rank` has three levels:

- **1** = Assistant Professor
- **2** = Associate Professor
- **3** = Full Professor

We want to test whether salary differs across these three ranks. Sounds straightforward --- but there is a trap waiting for us.

### The Wrong Way: Treating Rank as Continuous

Watch what happens if we simply plug `rank` into `lm()` without telling R it is categorical:

```{r}
#| label: rank-wrong

model_wrong <- lm(salary ~ rank, data = faculty)
summary(model_wrong)
```

> This model treats rank as a **continuous number** and fits a single straight line through ranks 1, 2, and 3. It forces the salary jump from Assistant to Associate to be **exactly the same** as from Associate to Full. That is a very strong assumption --- and almost certainly wrong.

### The Right Way: Using `factor()`

When we wrap `rank` in `factor()`, R creates dummy variables automatically:

```{r}
#| label: rank-factor

faculty <- faculty |>
  mutate(rank_f = factor(rank,
                         levels = c(1, 2, 3),
                         labels = c("Assistant", "Associate", "Full")))
```

Now let's fit the model properly:

```{r}
#| label: rank-correct

model_rank <- lm(salary ~ rank_f, data = faculty)
summary(model_rank)
```

> With `factor()`, R creates **two dummy variables** automatically:
>
> - `rank_fAssociate`: compares Associate to the reference group (Assistant)
> - `rank_fFull`: compares Full to the reference group (Assistant)
>
> The **intercept** is the mean salary for **Assistant Professors** (the reference group). Each coefficient is a **difference from that reference group**.

::: {.callout-important}
## Always Use `factor()` for Categorical Variables
When a categorical variable is coded as numbers (1, 2, 3), R treats it as continuous by default. You **must** use `factor()` to tell R these are group labels, not a numeric scale. This is one of the most common mistakes in applied regression.
:::

::: {.callout-warning}
## Factor Level Ordering: R vs. Stata
R orders factor levels **alphabetically** by default (e.g., "Assistant" < "Associate" < "Full"). Stata orders them by their **numeric code** (e.g., 1, 2, 3). This means the reference group may differ between the two programs. Always check with `levels(your_factor)` to confirm which level is the reference.
:::

### Verify with Group Means

Let's confirm the coefficients by computing group means directly:

```{r}
#| label: rank-means

faculty |>
  group_by(rank_f) |>
  summarise(
    n = n(),
    mean_salary = mean(salary),
    sd_salary   = sd(salary),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !n, \(x) round(x, 2)))
```

> Check: the intercept should match the Assistant mean. The Associate coefficient should equal the Associate mean minus the Assistant mean. The Full coefficient should equal the Full mean minus the Assistant mean.

### Equivalence with ANOVA

The regression F-test and the ANOVA F-test are identical:

::: {.panel-tabset}

#### Simple

```{r}
#| label: anova-rank-aov

anova_rank <- aov(salary ~ rank_f, data = faculty)
summary(anova_rank)
```

#### Advanced

```{r}
#| label: anova-rank-lm

anova(model_rank)
```

```{r}
#| label: anova-rank-tidy

tidy_model(model_rank)
```

:::

> The F-statistic from `aov()` and `anova()` is the same. This confirms: one-way ANOVA is just regression with dummy variables.


## Part 3: Planned Comparisons with `emmeans`

The regression output gives us comparisons to the **reference group** (Assistant). But what if we want other comparisons? For example:

- Associate vs. Full?
- The average of Associate and Full vs. Assistant?

This is where `emmeans` (Estimated Marginal Means) comes in. It lets us extract group means and test **any** comparison we want.

### Step 1: Estimated Marginal Means

```{r}
#| label: emmeans-rank

emm_rank <- emmeans(model_rank, "rank_f")
emm_rank
```

> These are the predicted means for each rank. Because we have no covariates, they are identical to the raw group means.

::: {.callout-note}
## Two Post-Estimation Packages
The Lecture Notes use the `marginaleffects` package for predicted values and comparisons. In this lab, we use the `emmeans` package, which provides a particularly intuitive interface for **custom contrasts** and **planned comparisons**. Both packages produce the same results for standard tasks like pairwise comparisons. Think of `marginaleffects` as the "general-purpose" tool and `emmeans` as the "contrast specialist."
:::

### Step 2: Associate vs. Assistant

This comparison is already in the regression output (the `rank_fAssociate` coefficient). But let's see how to get it from `emmeans`:

::: {.panel-tabset}

#### Simple

```{r}
#| label: contrast-assoc-assist

contrast(emm_rank, method = list("Associate - Assistant" = c(-1, 1, 0)))
```

#### Advanced

The weights `c(-1, 1, 0)` mean: subtract the first group (Assistant) from the second (Associate), and ignore the third (Full). The estimate should match the `rank_fAssociate` coefficient from the regression output.

```{r}
#| label: contrast-assoc-assist-confint

contrast(emm_rank, method = list("Associate - Assistant" = c(-1, 1, 0))) |>
  confint()
```

:::

> The estimate and p-value should match the `rank_fAssociate` coefficient from `summary(model_rank)`. The `emmeans` approach is more explicit about what we are comparing.

### Step 3: Full vs. Associate

This comparison is **not** directly available from the regression output (which only compares to the reference group). With `emmeans`, it is easy:

```{r}
#| label: contrast-full-assoc

contrast(emm_rank, method = list("Full - Associate" = c(0, -1, 1)))
```

> The weights `c(0, -1, 1)` subtract the Associate mean from the Full mean. Is the salary jump from Associate to Full statistically significant?

### Step 4: A Custom Contrast --- Higher Ranks vs. Assistant

Suppose we want to compare the **average salary of Associate and Full Professors** to the salary of Assistant Professors. This is a custom planned comparison:

```{r}
#| label: contrast-higher-vs-assist

contrast(emm_rank,
         method = list("Avg(Assoc, Full) - Assistant" = c(-1, 0.5, 0.5)))
```

> The weights `c(-1, 0.5, 0.5)` take the average of Associate and Full (each weighted 0.5) and subtract the Assistant mean (weighted -1). Notice the weights sum to zero --- this is a requirement for a valid contrast.

::: {.callout-tip}
## Contrast Weights Must Sum to Zero
For any valid contrast, the weights must sum to zero. This ensures we are comparing one group (or combination of groups) **against** another, rather than just computing a weighted average.
:::

### Step 5: Multiple Planned Contrasts at Once

We can test several planned contrasts in a single call:

```{r}
#| label: contrast-all-planned

my_contrasts <- list(
  "Associate - Assistant"          = c(-1,  1,   0),
  "Full - Associate"               = c( 0, -1,   1),
  "Avg(Assoc, Full) - Assistant"   = c(-1,  0.5, 0.5)
)

contrast(emm_rank, method = my_contrasts)
```

> All three planned comparisons are tested simultaneously. Each row gives the estimated difference, standard error, t-statistic, and p-value.


## Part 4: Pairwise Comparisons with Sidak Correction

Sometimes you want to compare **all pairs** of groups. With three ranks, there are three pairwise comparisons:

- Assistant vs. Associate
- Assistant vs. Full
- Associate vs. Full

The danger of testing all pairs is **inflated Type I error** --- the more tests you run, the greater the chance of a false positive. The **Sidak correction** adjusts p-values to control this.

### All Pairwise Comparisons

```{r}
#| label: pairwise-sidak

pairs(emm_rank, adjust = "sidak")
```

> The `adjust = "sidak"` argument applies the Sidak correction to the p-values. Compare these adjusted p-values to the unadjusted ones from our planned contrasts above. The Sidak-corrected p-values are larger (more conservative) because they account for the multiple testing problem.

### Unadjusted vs. Sidak-Adjusted

Let's see the difference side by side:

```{r}
#| label: pairwise-comparison

bind_rows(
  pairs(emm_rank, adjust = "none") |>
    as_tibble() |>
    mutate(correction = "None"),
  pairs(emm_rank, adjust = "sidak") |>
    as_tibble() |>
    mutate(correction = "Sidak")
) |>
  select(contrast, estimate, SE, p.value, correction) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

> Notice how the Sidak correction increases the p-values. In this case, with large salary differences across ranks, all comparisons remain significant even after correction. In other datasets with smaller effects, the correction can change your conclusions.

::: {.callout-note}
## Other Corrections
The Sidak correction is just one option. Common alternatives include:

- **Bonferroni**: multiply each p-value by the number of tests. Slightly more conservative than Sidak.
- **Tukey**: specifically designed for all pairwise comparisons. Use `adjust = "tukey"`.
- **None**: no correction. Only appropriate for a small number of **pre-planned** comparisons.

For most exploratory pairwise comparisons, Sidak or Tukey are good default choices.
:::

### Connection to the Overall F-Test

The overall ANOVA F-test tells us whether **any** group differs from any other. The pairwise comparisons tell us **which specific** pairs differ. They complement each other:

```{r}
#| label: overall-f-test

anova(model_rank)
```

> The significant F-test tells us that rank is related to salary overall. The pairwise comparisons (above) tell us that all three rank pairs differ significantly --- even after the Sidak correction.


## Part 5: Changing the Reference Group

So far, R has used **Assistant Professor** (the first level) as the reference group. Every regression coefficient is a comparison to Assistant. But what if we want **Full Professor** as the reference? Then the coefficients would tell us how much less the other ranks earn compared to Full Professors.

### Relevel the Factor

```{r}
#| label: relevel-rank

faculty <- faculty |>
  mutate(rank_f2 = relevel(rank_f, ref = "Full"))

# Check the new level order
levels(faculty$rank_f2)
```

### Re-run the Regression

```{r}
#| label: model-relevel

model_rank2 <- lm(salary ~ rank_f2, data = faculty)
summary(model_rank2)
```

> Now the **intercept** is the mean salary for **Full Professors**. The coefficients for Assistant and Associate are **negative** --- they tell us how much less each group earns compared to Full Professors.

### Side-by-Side Comparison

::: {.panel-tabset}

#### Reference = Assistant

```{r}
#| label: compare-ref-assistant

tidy_model(model_rank)
```

#### Reference = Full

```{r}
#| label: compare-ref-full

tidy_model(model_rank2)
```

:::

> Compare the two tables. The **model is the same** --- same F-statistic, same $R^2$, same residual standard error. Only the **parameterization** changes. Changing the reference group does not change what the model says about the data; it changes how the story is told through the coefficients.

::: {.callout-tip}
## Choosing a Reference Group
Pick the reference group that makes interpretation easiest. Common choices:

- **The largest group** (for stability of the intercept estimate)
- **A control or baseline group** (for treatment comparisons)
- **The group of theoretical interest** (the group others are compared against)

There is no "correct" choice --- it depends on your research question.
:::

### The `emmeans` Results Do Not Change

Regardless of which reference group we use, `emmeans` gives the same estimated means and contrasts:

```{r}
#| label: emmeans-relevel

emmeans(model_rank2, "rank_f2")
```

> The estimated marginal means are identical to what we got earlier. The reference group affects only the regression coefficients, not the underlying group means or contrasts. This is one reason `emmeans` is so useful --- it gives you interpretable results that do not depend on the coding scheme.


## Key Takeaways

::: {.callout-tip}
## What We Learned

1. **Always use `factor()` for categorical predictors.** Without it, R treats numbered categories (1, 2, 3) as a continuous scale, fitting a single slope instead of separate group means.

2. **Reference group determines coefficient interpretation.** The intercept equals the reference group mean, and each coefficient is a difference from that reference. Use `relevel()` to change the reference group.

3. **`emmeans` is your best friend for group comparisons.** It extracts estimated marginal means and lets you test any contrast --- planned comparisons, pairwise comparisons, or custom contrasts with user-defined weights.

4. **Contrast weights must sum to zero.** This ensures you are comparing one group (or combination) against another.

5. **The Sidak correction protects against multiple testing.** When you test all pairwise comparisons, use `adjust = "sidak"` (or `"tukey"`) to control the family-wise error rate.

6. **Changing the reference group changes the coefficients but not the model.** The F-statistic, $R^2$, and all group means remain identical. Only the interpretation of the intercept and slopes changes.

**Next:** In Lab 5, we will add continuous covariates alongside categorical predictors to build multiple regression models --- and learn how to interpret coefficients when several variables are in the model simultaneously.
:::
