---
title: "Lab 4: One-Way ANOVA and Categorical Predictors"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Create a three-level factor for a one-way ANOVA
- See why numbered categories must be treated as factors
- Fit ANOVA as a regression model with `lm()`
- Interpret the intercept and dummy-variable coefficients
- Use `modelbased::estimate_means()` to recover group means from the model
- Use `anova()`, `performance::r2()`, and `effectsize::eta_squared()` to summarize the omnibus group effect
- Compare unadjusted and adjusted pairwise contrasts
- Run a small set of planned comparisons with `emmeans`
- **Lecture Notes Reference:** Chapter 3: *One-Way ANOVA*

::: {.meta}
Dataset: `instruction`. Estimated time: 70-85 minutes.
:::
:::


## Part 1: Prepare the ANOVA Dataset

In Chapter 3, the core idea is:

> One-way ANOVA is regression with a categorical predictor.

We will use the `instruction` dataset. The outcome is `mathgain`, a student's math achievement gain. The grouping variable is teacher math preparation, which we will turn into three levels.

### Step 1: Load the Data

```{r}
#| label: lab04-load-data

data(instruction)
instruction_raw <- instruction
```

We keep the original dataset in `instruction_raw` and create a separate analysis dataset.

### Step 2: Create the Factor

```{r}
#| label: lab04-create-prep-factor

instruction_anova <- instruction_raw |>
  mutate(
    prep_code = case_when(
      mathprep <= 3 ~ 1,
      mathprep <= 4 ~ 2,
      mathprep > 4 ~ 3
    ),
    prep_level = factor(
      prep_code,
      levels = c(1, 2, 3),
      labels = c("Low", "Medium", "High")
    )
  ) |>
  filter(!is.na(mathgain), !is.na(prep_level))
```

`prep_code` is a numbered category. `prep_level` is the factor version that we want for ANOVA.

### Step 3: Check the Analysis Data

```{r}
#| label: lab04-preflight

instruction_anova |>
  summarise(
    students = n(),
    groups = n_distinct(prep_level),
    missing_mathgain = sum(is.na(mathgain)),
    missing_prep = sum(is.na(prep_level))
  )
```

### Step 4: Group Summary

```{r}
#| label: lab04-group-summary

prep_summary <- instruction_anova |>
  group_by(prep_level) |>
  summarise(
    n = n(),
    mean_gain = mean(mathgain),
    sd_gain = sd(mathgain),
    .groups = "drop"
  )

prep_summary |>
  round_numeric(digits = 2)
```

This table gives the raw mean math gain for each preparation group.

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the summary table to answer two questions:

1. Which preparation group has the highest raw mean math gain?
2. Which group has the largest sample size?

Write one sentence explaining why both mean differences and sample sizes matter.
:::


## Part 2: Visualize the Group Means

```{r}
#| label: lab04-boxplot-prep-level
#| fig-cap: "Math gains by teacher preparation level."
#| fig-alt: "Boxplots with jittered points showing math gain for Low, Medium, and High teacher preparation groups. The Medium and High groups have slightly higher central values than the Low group, but all three distributions overlap substantially."

ggplot(instruction_anova, aes(x = prep_level, y = mathgain, fill = prep_level)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.18, size = 0.9, show.legend = FALSE) +
  ber640_fill +
  labs(x = "Teacher Preparation Level", y = "Math Gain")
```

::: {.interpretation}
The group means differ, but the distributions also overlap a lot.

ANOVA asks whether the differences among group means are large relative to the within-group variation.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Look at the boxplots and jittered points.
Do the groups look completely separated, or do they overlap?
Write one short sentence.
:::


## Part 3: Fit ANOVA as Regression

### Step 1: The Numeric-Coding Trap

```{r}
#| label: lab04-numeric-trap

model_numeric <- lm(mathgain ~ prep_code, data = instruction_anova)
model_params(model_numeric)
```

This model treats preparation level as a one-unit numeric scale. It forces the Low-to-Medium difference to be the same kind of step as the Medium-to-High difference.

::: {.pitfall}
If a variable is a category, make it a factor.

Numbered labels like 1, 2, and 3 are still categories unless you truly want to assume a straight numeric trend.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Compare `model_numeric` to the factor model below.
What assumption does the numeric model make that the factor model does not make?
:::

### Step 2: Fit the Factor Model

```{r}
#| label: lab04-fit-anova-model

levels(instruction_anova$prep_level)

model_anova <- lm(mathgain ~ prep_level, data = instruction_anova)
model_params(model_anova)
```

R uses the first factor level, `Low`, as the reference group.

::: {.interpretation}
The intercept is the estimated mean math gain for the Low preparation group.

`prep_levelMedium` is the Medium minus Low difference.

`prep_levelHigh` is the High minus Low difference.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Interpret the `prep_levelMedium` coefficient in one sentence.
Your sentence should say "Medium minus Low" or "compared with Low."
:::


## Part 4: Recover the Group Means

In a one-way ANOVA with no covariates, the model-estimated means are the same as the raw group means.

```{r}
#| label: lab04-estimate-means

estimated_means <- modelbased::estimate_means(model_anova, by = "prep_level")
estimated_means
```

Now compare the raw means and model means side by side.

```{r}
#| label: lab04-raw-vs-model-means

estimated_means_tbl <- estimated_means |>
  as_tibble() |>
  select(prep_level, model_mean = Mean, SE, CI_low, CI_high)

prep_summary |>
  select(prep_level, n, raw_mean = mean_gain, sd_gain) |>
  left_join(estimated_means_tbl, by = "prep_level") |>
  round_numeric(digits = 2)
```

### Plot the Estimated Means

```{r}
#| label: lab04-estimated-means-plot
#| fig-cap: "Model-estimated mean math gains with 95% confidence intervals."
#| fig-alt: "Point-range plot showing estimated mean math gain and 95 percent confidence intervals for Low, Medium, and High teacher preparation groups. Medium and High are above Low, but the confidence intervals overlap."

ggplot(estimated_means_tbl,
       aes(x = prep_level, y = model_mean, color = prep_level)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high),
                  linewidth = 0.8, size = 1.1, show.legend = FALSE) +
  ber640_palette +
  labs(x = "Teacher Preparation Level", y = "Estimated Mean Math Gain")
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the raw-vs-model-means table.
Why are the raw means and model means identical in this one-way ANOVA?
:::


## Part 5: Omnibus ANOVA Test and Effect Size

The coefficient table compares each non-reference group to the Low group.
The ANOVA table answers a different question:

> Is there evidence that at least one group mean differs from at least one other group mean?

### Step 1: Omnibus F-Test

```{r}
#| label: lab04-omnibus-anova

anova(model_anova)
```

::: {.interpretation}
A significant F-test does not tell us that every pair of groups differs.

It only tells us that the set of group means is not perfectly equal.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Write one sentence interpreting the omnibus F-test.
Your sentence should include the phrase "at least one group mean."
:::

### Step 2: R-Squared

```{r}
#| label: lab04-r2

performance::r2(model_anova)
```

In one-way ANOVA, $R^2$ is the proportion of outcome variance explained by group membership.

### Step 3: Eta-Squared

```{r}
#| label: lab04-eta-squared

effectsize::eta_squared(
  model_anova,
  partial = FALSE,
  alternative = "two.sided",
  ci = 0.90
)
```

::: {.interpretation}
For one-way ANOVA, $R^2$ and $\eta^2$ are the same quantity: the share of total variation explained by group membership.

A statistically significant F-test can still have a small practical effect size.
:::


## Part 6: Pairwise Group Contrasts

The omnibus F-test says whether group means differ somewhere.
Contrasts ask which differences we want to inspect.

### Step 1: Unadjusted Pairwise Contrasts

```{r}
#| label: lab04-pairwise-preview

modelbased::estimate_contrasts(model_anova, contrast = "prep_level")
```

These are unadjusted pairwise contrasts. They answer specific group-difference questions, but they do not yet account for the fact that we are testing three differences.

### Step 2: Sidak-Adjusted Pairwise Contrasts

```{r}
#| label: lab04-pairwise-sidak

modelbased::estimate_contrasts(
  model_anova,
  contrast = "prep_level",
  p_adjust = "sidak"
)
```

::: {.pitfall}
The more pairwise tests you run, the easier it is to get at least one false positive.

For exploratory "compare everything" work, use an adjustment such as Sidak, Holm, Bonferroni, or Tukey.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Compare the unadjusted and Sidak-adjusted tables.
Which p-values changed, and did the substantive conclusion change?
:::


## Part 7: Planned Comparisons

Planned comparisons are contrasts chosen before looking at the results.
They should come from the research question, not from chasing the biggest observed difference.

For custom planned contrasts, `emmeans` gives a direct way to specify contrast weights.
The weights below follow the order of the levels:

```{r}
#| label: lab04-level-order-for-contrasts

levels(instruction_anova$prep_level)
```

### Step 1: Estimated Means for Planned Contrasts

```{r}
#| label: lab04-emmeans-prep

prep_emm <- emmeans::emmeans(model_anova, "prep_level")
prep_emm
```

### Step 2: Define Planned Contrasts

```{r}
#| label: lab04-planned-contrasts

stopifnot(identical(
  levels(instruction_anova$prep_level),
  c("Low", "Medium", "High")
))

planned_contrasts <- list(
  "Medium - Low" = c(-1, 1, 0),
  "High - Low" = c(-1, 0, 1),
  "Avg(Medium, High) - Low" = c(-1, 0.5, 0.5)
)

emmeans::contrast(
  prep_emm,
  method = planned_contrasts,
  adjust = "none"
)
```

::: {.interpretation}
The contrast weights must match the level order: Low, Medium, High.

For `Avg(Medium, High) - Low`, the weights `c(-1, 0.5, 0.5)` subtract Low and add the average of Medium and High.
The weights sum to zero, so this is a valid contrast.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Pick one planned contrast from the table and interpret it in one sentence.
Your sentence should name the two groups or group combinations being compared.
:::


## Part 8: Change the Reference Group

Changing the reference group changes the coefficient table, but it does not change the fitted group means or the omnibus F-test.

```{r}
#| label: lab04-relevel-high

instruction_anova <- instruction_anova |>
  mutate(prep_level_high_ref = relevel(prep_level, ref = "High"))

model_high_ref <- lm(mathgain ~ prep_level_high_ref, data = instruction_anova)
model_params(model_high_ref)
```

::: {.interpretation}
With High as the reference, the intercept is the High group mean.

The coefficients compare Low and Medium to High.
The underlying fitted means are the same as before.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Compare `model_anova` and `model_high_ref`.
Which quantities changed, and which quantities stayed the same?
:::


::: {.exercises}
## What to Submit

1. The group summary table for `prep_level`. [core]{.chip .easy}
2. The factor-model coefficient table from `model_params(model_anova)`. [core]{.chip .easy}
3. One sentence interpreting the `prep_levelMedium` coefficient. [core]{.chip .med}
4. The omnibus ANOVA table and one sentence explaining what the F-test does and does not tell us. [core]{.chip .med}
5. The `performance::r2()` and `effectsize::eta_squared()` outputs, plus one sentence about practical effect size. [core]{.chip .med}
6. The unadjusted and Sidak-adjusted pairwise contrast tables, with one sentence comparing them. [core]{.chip .hard}
7. One planned contrast table and one planned-contrast interpretation. [core]{.chip .hard}
:::


::: {.summary}
## Lab Summary

1. One-way ANOVA is regression with a categorical predictor.

2. Use `factor()` for numbered categories so R creates dummy variables.

3. In treatment coding, the intercept is the reference group mean.

4. Each dummy coefficient is a difference from the reference group.

5. `modelbased::estimate_means()` recovers the model-estimated mean for every group.

6. `anova()` gives the omnibus F-test, while `modelbased::estimate_contrasts()` gives pairwise group differences.

7. In one-way ANOVA, $R^2$ equals $\eta^2$.

8. Planned contrasts should be chosen from research questions before looking at the results.

9. Exploratory pairwise comparisons need multiple-comparison adjustment.

**Next:** Lab 5 moves into multiple regression with several predictors in the same model.
:::
