---
title: "Lab 6: Interactions in Regression"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Fit an interaction model with `*`
- Compare additive and interaction models with `anova()`
- Interpret interaction terms as conditional effects
- Probe a continuous x categorical interaction with `modelbased::estimate_slopes()`
- Plot fitted interaction lines with `ggeffects::predict_response()`
- Probe a categorical x categorical interaction with `modelbased::estimate_means()` and `modelbased::estimate_contrasts()`
- **Lecture Notes Reference:** Chapter 5: *Interactions*

::: {.meta}
Datasets: `gcse` and `faculty`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: What Is an Interaction?

An interaction means "it depends."

The effect of one predictor depends on the value or category of another predictor.
In this lab, we practice two common cases:

1. Continuous x categorical: does the LRT-GCSE slope differ by school?
2. Categorical x categorical: does the gender salary gap differ by academic rank?

::: {.pitfall}
After you add an interaction, the main effects are conditional.

They do not usually mean "the average effect overall."
They mean the effect when the other interacting variable is at zero or at its reference category.
:::


## Part 2: Continuous x Categorical Interaction

We use the `gcse` data from Chapter 5 and compare two schools:

- Hackney: school 10
- Chelsea: school 6

The question is:

> Does the relationship between prior achievement (`lrt`) and exam performance (`gcse`) differ between these two schools?

### Step 1: Prepare the Data

```{r}
#| label: lab06-load-gcse

data(gcse)

two_schools <- gcse |>
  filter(school %in% c(6, 10)) |>
  mutate(
    school_name = factor(
      if_else(school == 10, "Hackney", "Chelsea"),
      levels = c("Hackney", "Chelsea")
    )
  )
```

```{r}
#| label: lab06-gcse-preflight

two_schools |>
  count(school_name)
```

### Step 2: Fit the Additive Model

The additive model assumes parallel regression lines.
It allows the two schools to have different intercepts, but the same LRT slope.

```{r}
#| label: lab06-fit-parallel-model

model_parallel <- lm(gcse ~ lrt + school_name, data = two_schools)
model_params(model_parallel)
```

### Step 3: Fit the Interaction Model

The interaction model allows the LRT slope to differ by school.
The `*` operator expands to both main effects plus their product term.

```{r}
#| label: lab06-fit-cont-cat-interaction

model_interact <- lm(gcse ~ lrt * school_name, data = two_schools)
model_params(model_interact)
```

::: {.interpretation}
In `gcse ~ lrt * school_name`, the interaction term is `lrt:school_nameChelsea`.

It tells us how much the LRT slope for Chelsea differs from the LRT slope for Hackney.
If this coefficient is positive, the fitted LRT-GCSE slope is steeper for Chelsea.
:::

### Step 4: Test the Interaction

Use a nested model comparison to ask whether the interaction improves fit enough to justify the extra parameter.

```{r}
#| label: lab06-test-cont-cat-interaction

anova(model_parallel, model_interact)
```

::: {.interpretation}
The interaction test compares the parallel-lines model against the non-parallel-lines model.

In this example, the interaction is not statistically significant, so the additive model is the more cautious reporting model.
We still probe the interaction to learn how the mechanics work.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `anova(model_parallel, model_interact)` to answer the research question.
Is there evidence that the LRT-GCSE slope differs by school?
Write two short sentences: one about the interaction test and one naming the more cautious reporting model.
:::

### Step 5: Probe the School-Specific Slopes

`modelbased::estimate_slopes()` gives the LRT slope separately for each school.

```{r}
#| label: lab06-slopes-by-school

slopes_by_school <- modelbased::estimate_slopes(
  model_interact,
  trend = "lrt",
  by = "school_name"
) |>
  as_tibble() |>
  select(school_name, Slope, SE, CI_low, CI_high, p)

slopes_by_school |>
  round_numeric(digits = 3)
```

::: {.interpretation}
The slope for Hackney is the `lrt` coefficient.

The slope for Chelsea is the `lrt` coefficient plus the `lrt:school_nameChelsea` coefficient.
The difference between those two slopes is the interaction coefficient.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `slopes_by_school`.
Which school has the steeper conditional LRT slope?
Write one sentence, then explain in words how the Chelsea slope combines the Hackney slope and the interaction coefficient.
:::

### Step 6: Plot the Additive and Interaction Models

`ggeffects::predict_response()` creates predicted values and confidence intervals from a model.
We can plot those predictions with ordinary `ggplot2`.

```{r}
#| label: lab06-cont-cat-prediction-data

parallel_predictions <- ggeffects::predict_response(
  model_parallel,
  terms = c("lrt", "school_name")
) |>
  as_tibble() |>
  mutate(Model = "Additive: parallel lines")

interaction_predictions <- ggeffects::predict_response(
  model_interact,
  terms = c("lrt", "school_name")
) |>
  as_tibble() |>
  mutate(Model = "Interaction: non-parallel lines")

school_predictions <- bind_rows(
  parallel_predictions,
  interaction_predictions
)
```

```{r}
#| label: lab06-cont-cat-interaction-plot
#| fig-cap: "Additive and interaction models for LRT and school."
#| fig-alt: "Two-panel plot of predicted GCSE scores by LRT score for Hackney and Chelsea. The additive model panel shows parallel fitted lines. The interaction model panel shows non-parallel fitted lines, allowing the LRT slope to differ by school."
#| fig-width: 10
#| fig-height: 4.8

ggplot(
  school_predictions,
  aes(x = x, y = predicted, color = group, fill = group)
) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ Model) +
  scale_color_manual(values = c(ber640_colors[["primary"]], ber640_colors[["secondary"]])) +
  scale_fill_manual(values = c(ber640_colors[["primary"]], ber640_colors[["secondary"]])) +
  labs(
    x = "LRT Score",
    y = "Predicted GCSE Score",
    color = "School",
    fill = "School"
  )
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Look at the additive and interaction panels.
Which panel has parallel lines, and which panel allows school-specific slopes?
Write one sentence explaining why non-parallel lines represent a continuous x categorical interaction.
:::


## Part 3: Categorical x Categorical Interaction

Now we ask whether the gender salary gap differs by academic rank in the `faculty` data.

The question is:

> Is the male-female salary difference the same for assistant, associate, and full professors?

### Step 1: Prepare the Data

```{r}
#| label: lab06-load-faculty

data(faculty)

faculty_df <- faculty |>
  mutate(
    gender = factor(
      if_else(male == 1, "Male", "Female"),
      levels = c("Female", "Male")
    ),
    rank_f = factor(
      rank,
      levels = 1:3,
      labels = c("Assistant", "Associate", "Full")
    )
  )
```

```{r}
#| label: lab06-faculty-preflight

faculty_df |>
  count(gender, rank_f)
```

### Step 2: Fit Additive and Interaction Models

The additive model assumes one gender gap that is constant across ranks.
The interaction model allows the gender gap to be different at each rank.

```{r}
#| label: lab06-fit-cat-cat-models

model_rank_add <- lm(salary ~ gender + rank_f, data = faculty_df)
model_rank_int <- lm(salary ~ gender * rank_f, data = faculty_df)

model_params(model_rank_int)
```

### Step 3: Test the Overall Interaction

For a 2 x 3 interaction, the interaction has two coefficients.
The nested F-test gives one overall test of whether those two coefficients improve model fit.

```{r}
#| label: lab06-test-cat-cat-interaction

anova(model_rank_add, model_rank_int)
```

::: {.interpretation}
The `gender:rank_f` interaction asks whether the gender gap differs across ranks.

In this example, the interaction is not statistically significant, so the additive model is the more cautious reporting model.
The conditional comparisons below show how we would describe the pattern if we needed to probe it.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `anova(model_rank_add, model_rank_int)`.
Does the overall test suggest that the gender salary gap differs by academic rank?
Write one sentence that starts: "The gender-by-rank interaction tests whether..."
:::

### Step 4: Estimate the Six Cell Means

`modelbased::estimate_means()` gives the predicted salary for each gender-by-rank cell.

```{r}
#| label: lab06-rank-gender-cell-means

rank_gender_means <- modelbased::estimate_means(
  model_rank_int,
  by = c("gender", "rank_f")
) |>
  as_tibble() |>
  select(gender, rank_f, Mean, SE, CI_low, CI_high)

rank_gender_means |>
  round_numeric(digits = 0)
```

### Step 5: Estimate the Gender Gap at Each Rank

`modelbased::estimate_contrasts()` compares Male and Female faculty within each rank.

```{r}
#| label: lab06-gender-gap-by-rank

gender_gap_by_rank <- modelbased::estimate_contrasts(
  model_rank_int,
  contrast = "gender",
  by = "rank_f"
) |>
  as_tibble() |>
  select(rank_f, Level1, Level2, Difference, SE, CI_low, CI_high, p)

gender_gap_by_rank |>
  mutate(
    across(c(Difference, SE, CI_low, CI_high), \(x) round(x, 0)),
    p = round(p, 3)
  )
```

::: {.interpretation}
The `Difference` column is the estimated Male - Female salary gap within each rank.

If those differences were clearly different across ranks, that would be evidence of a categorical x categorical interaction.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `gender_gap_by_rank`.
Which rank has the largest estimated Male - Female salary gap?
Write one cautious sentence connecting the three conditional gender gaps to the overall interaction test.
:::

### Step 6: Plot the Cell Means

```{r}
#| label: lab06-cat-cat-interaction-plot
#| fig-cap: "Predicted faculty salary by gender and academic rank."
#| fig-alt: "Line plot of predicted salary by academic rank for female and male faculty. Points and error bars show model-estimated cell means and confidence intervals. The vertical male-female gap varies somewhat by rank, but the overall interaction test is not statistically significant."
#| fig-width: 8
#| fig-height: 5

ggplot(
  rank_gender_means,
  aes(x = rank_f, y = Mean, color = gender, group = gender)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.1) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.12) +
  scale_color_manual(values = c(ber640_colors[["secondary"]], ber640_colors[["accent"]])) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Academic Rank",
    y = "Predicted Salary",
    color = "Gender"
  )
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Look at the cell-means plot.
Do the gender gaps look exactly constant across ranks?
Then explain why the overall interaction test still matters before making a strong claim.
:::


::: {.exercises}
## What to Submit

1. The continuous x categorical interaction test from `anova(model_parallel, model_interact)` and one sentence answering whether the LRT-GCSE slope differs by school. [core]{.chip .med}
2. The `slopes_by_school` table and one sentence comparing the school-specific LRT slopes. [core]{.chip .med}
3. The additive-vs-interaction prediction plot and one sentence explaining what parallel vs. non-parallel lines mean. [core]{.chip .med}
4. The categorical x categorical interaction test from `anova(model_rank_add, model_rank_int)` and one sentence answering whether the gender salary gap differs by rank. [core]{.chip .med}
5. The `rank_gender_means` table or the cell-means plot. [core]{.chip .easy}
6. The `gender_gap_by_rank` table and one sentence naming the largest conditional gender gap. [core]{.chip .hard}
7. A 3-4 sentence mini-report that tests each interaction first, then summarizes the conditional effects without causal language. [core]{.chip .hard}
:::


::: {.summary}
## Lab Summary

1. The `*` operator fits both main effects and the interaction term.

2. A continuous x categorical interaction means the slope of the continuous predictor differs across groups.

3. A categorical x categorical interaction means the group difference on one factor differs across levels of the other factor.

4. `anova(additive_model, interaction_model)` gives a simple nested test of whether the interaction improves model fit.

5. `modelbased::estimate_slopes()` is useful for probing continuous x categorical interactions.

6. `modelbased::estimate_means()` and `modelbased::estimate_contrasts()` are useful for probing categorical x categorical interactions.

7. `ggeffects::predict_response()` gives simple prediction data for interaction plots.

**Next:** Lab 7 turns to nonlinear relationships.
:::
