---
title: "Lab 5: Multiple Regression and Partial Correlation"
---

{{< include ../_common.qmd >}}

## Goals

::: {.callout-note}
## What You'll Learn in This Lab
- Fit a multiple regression model with two continuous predictors
- Interpret coefficients as effects "controlling for" other variables
- Compute and interpret partial correlations
- Understand the "what's left" logic behind partial correlation using residuals
- Create added-variable plots to visualize partial relationships
- Compute standardized regression coefficients to compare predictor importance
- **Lecture Notes Reference:** Chapter 4: *Continuous Predictors*
:::

::: {.callout-tip}
## Stata → R Translation
| Stata | R | Notes |
|:------|:--|:------|
| `regress salary marketc yearsdg, beta` | `parameters::standardize_parameters()` | Standardized (beta) coefficients |
| `pcorr salary marketc yearsdg` | `correlation::correlation(partial = TRUE)` | Partial correlations |
| `avplots` | `car::avPlots()` | Added-variable (partial) plots |
:::

**Dataset:** `faculty` from the `regdatasets` package


## Part 1: Brief Review --- Single-Predictor Models

In Lab 4, we explored categorical predictors and planned comparisons. Before we move to multiple regression, let's reload the `faculty` data and set up the variables we need.

```{r}
#| label: lab05-setup

library(regdatasets)
data(faculty)
glimpse(faculty)
```

Recall that `market` captures the discipline-specific market demand for faculty in each department. Since zero market demand is not meaningful, we center it:

```{r}
#| label: center-market

faculty <- faculty |>
  mutate(market_c = market - mean(market))
```

Let's quickly fit two single-predictor models to set the stage for what comes next. First, salary predicted by centered market demand:

```{r}
#| label: review-market

model_market <- lm(salary ~ market_c, data = faculty)
summary(model_market)
```

And salary predicted by years since degree:

```{r}
#| label: review-yearsdg

model_years <- lm(salary ~ yearsdg, data = faculty)
summary(model_years)
```

> Take note of the $R^2$ values from each model. We will compare them to the multiple regression model shortly.


## Part 2: Multiple Regression with Two Continuous Predictors

Now the key question: what happens when we put **both** predictors in the model at the same time?

```{r}
#| label: multiple-regression

model_multi <- lm(salary ~ market_c + yearsdg, data = faculty)
summary(model_multi)
```

This is the heart of multiple regression. Each coefficient now has a specific interpretation:

- **`market_c`**: the expected change in salary for a one-unit increase in market demand, **holding years since degree constant**.
- **`yearsdg`**: the expected change in salary for one additional year since degree, **holding market demand constant**.

The phrase "holding constant" (or "controlling for") is critical. In a single-predictor model, the coefficient captures **everything** associated with that predictor --- including its overlap with other variables. In a multiple regression, each coefficient isolates the **unique** contribution of that predictor after removing the influence of the others.

::: {.callout-important}
## The "Controlling For" Interpretation
In multiple regression, the coefficient for $X_1$ answers:
*"If two faculty members have the same value of $X_2$, how much does salary differ for a one-unit difference in $X_1$?"*

This is why multiple regression is so powerful --- it lets us separate the effects of correlated predictors.
:::

### Comparing $R^2$ Values

Let's see how much explanatory power we gain by using both predictors:

```{r}
#| label: compare-r-squared

tibble(
  Model   = c("market_c only", "yearsdg only", "market_c + yearsdg"),
  R_squared = c(
    summary(model_market)$r.squared,
    summary(model_years)$r.squared,
    summary(model_multi)$r.squared
  )
) |>
  mutate(R_squared = round(R_squared, 4))
```

> Notice that the combined $R^2$ is **less than the sum** of the two individual $R^2$ values. This is because the two predictors share some explanatory power --- they are correlated with each other. Multiple regression partitions the variance in salary into the unique contribution of each predictor plus their shared overlap.


## Part 3: Understanding Partial Correlation

We just saw that each regression coefficient represents a "controlled" effect. Partial correlation captures the same idea in correlation form. The **partial correlation** between salary and market demand, controlling for years since degree, tells us: *how strongly are salary and market demand related, after we remove the influence of years since degree from both?*

### Step 1: Regular Pearson Correlations

First, let's see the zero-order (simple) correlations among our three variables:

```{r}
#| label: pearson-correlations

faculty |>
  select(salary, market_c, yearsdg) |>
  cor() |>
  round(4)
```

These are the standard Pearson correlations --- each one ignores all other variables.

### Step 2: Partial Correlations

Now let's compute partial correlations using the `correlation` package:

```{r}
#| label: partial-correlations

# Partial correlations using easystats correlation package
faculty |>
  select(salary, market_c, yearsdg) |>
  correlation::correlation(partial = TRUE)
```

> Compare the partial correlations to the zero-order correlations above. How do they differ? The partial correlation between salary and `market_c` is the correlation **after removing the linear effect of `yearsdg` from both variables**. Similarly, the partial correlation between salary and `yearsdg` controls for `market_c`.

### Step 3: The "What's Left" Explanation

Partial correlation might seem abstract, so let's build it from scratch. The logic is simple:

1. **Remove yearsdg from salary:** Regress salary on yearsdg, save the residuals. These residuals represent the part of salary that has nothing to do with yearsdg.
2. **Remove yearsdg from market_c:** Regress market_c on yearsdg, save the residuals. These residuals represent the part of market demand that has nothing to do with yearsdg.
3. **Correlate the two sets of residuals:** This correlation is the partial correlation.

Let's do it step by step:

```{r}
#| label: residual-salary

# Step 1: Remove yearsdg from salary
resid_salary <- residuals(lm(salary ~ yearsdg, data = faculty))
```

```{r}
#| label: residual-market

# Step 2: Remove yearsdg from market_c
resid_market <- residuals(lm(market_c ~ yearsdg, data = faculty))
```

```{r}
#| label: partial-from-residuals

# Step 3: Correlate the residuals
cor(resid_salary, resid_market) |> round(4)
```

> This should match the partial correlation between salary and `market_c` (controlling for `yearsdg`) from the `correlation::correlation(partial = TRUE)` output above. This is the "what's left" interpretation: after we strip out everything that yearsdg explains in both salary and market demand, how strongly are the leftovers related?

::: {.callout-tip}
## Why This Matters
The residual-based explanation is not just a mathematical trick. It reveals the core logic of multiple regression: **each coefficient reflects the relationship between a predictor and the outcome after the other predictors have had their say.** When you report that market demand is significantly related to salary "controlling for years since degree," you are saying exactly this --- the leftovers are still correlated.
:::

### Step 4: Added-Variable Plots

The `car` package provides a beautiful visualization of this residual logic. An **added-variable plot** (also called a partial regression plot) shows the residual-vs-residual scatterplot for each predictor:

```{r}
#| label: avplots
#| fig-alt: "Two-panel added-variable plot showing the partial relationship of each predictor with salary after controlling for the other: left panel shows a positive slope for centered marketability and right panel shows a positive slope for years since degree"
#| fig-width: 10
#| fig-height: 5

car::avPlots(model_multi)
```

> Each panel shows the relationship between one predictor and salary **after removing the effect of the other predictor from both**. The slope of the line in each panel equals the corresponding regression coefficient from the multiple regression model. These plots are a powerful diagnostic tool --- they show you the partial relationship that each coefficient is capturing.

::: {.callout-note}
## Reading Added-Variable Plots

- **X-axis:** Residuals from regressing the predictor on all *other* predictors (what's left of the predictor).
- **Y-axis:** Residuals from regressing salary on all *other* predictors (what's left of salary).
- **Slope:** Equals the multiple regression coefficient for that predictor.
- **Spread:** Shows how much variation in salary is explained uniquely by that predictor.

If you see a clear linear trend, the predictor has a strong unique contribution. If the points are scattered with no trend, the predictor adds little after controlling for the others.
:::


## Part 4: Standardized Coefficients

In our model, `market_c` and `yearsdg` are measured on different scales. Market demand might range from 0 to 10, while years since degree might range from 1 to 40. This makes it hard to compare their raw coefficients directly --- a one-unit change means very different things for each variable.

**Standardized coefficients** (beta weights) solve this problem by putting both predictors on the same scale: standard deviations.

::: {.panel-tabset}

#### Simple

```{r}
#| label: standardized-simple

parameters::standardize_parameters(model_multi)
```

> The standardized coefficients (the `Std_Coefficient` column) tell us: "When market demand increases by **1 standard deviation**, salary increases by $\beta$ standard deviations, holding yearsdg constant." This lets us compare the relative importance of market demand and years since degree on the same scale.

#### Advanced

Standardized coefficients are obtained by fitting the model to z-scored (standardized) versions of all variables. We can verify this manually:

```{r}
#| label: standardized-manual

faculty_z <- faculty |>
  mutate(
    salary_z   = scale(salary)   |> as.numeric(),
    market_c_z = scale(market_c) |> as.numeric(),
    yearsdg_z  = scale(yearsdg)  |> as.numeric()
  )

model_z <- lm(salary_z ~ market_c_z + yearsdg_z, data = faculty_z)
tidy_model(model_z)
```

> The coefficients from the z-scored model should match the standardized parameters above. Notice that the intercept is essentially zero --- when all predictors are at their mean (z = 0), the predicted outcome is also at its mean (z = 0).

:::

### When to Use Standardized Coefficients

Standardized coefficients are useful when:

- **Predictors are on different scales** and you want to compare their relative importance.
- **You want to communicate effect sizes** in a scale-free way.

However, keep in mind:

- **Unstandardized coefficients** are easier to interpret in practical terms ("$1,000 more salary per year since degree").
- **Standardized coefficients depend on the sample variability.** If one predictor has restricted range in your sample, its standardized coefficient will be smaller even if the underlying relationship is strong.

::: {.callout-tip}
## Reporting Both
In practice, many researchers report **unstandardized coefficients** in the main results (for practical interpretation) and include **standardized coefficients** in a supplementary table (for comparing relative importance). This gives readers both perspectives.
:::


## Key Takeaways

::: {.callout-tip}
## What We Learned

1. **Multiple regression isolates unique effects.** Each coefficient represents the relationship between a predictor and the outcome, controlling for all other predictors in the model.

2. **Combined $R^2$ is less than the sum of individual $R^2$ values.** This happens because correlated predictors share explanatory power. Multiple regression partitions variance into unique and shared components.

3. **Partial correlation quantifies controlled relationships.** The partial correlation between $X$ and $Y$ controlling for $Z$ is the correlation between the residuals from regressing $X$ on $Z$ and the residuals from regressing $Y$ on $Z$. It is the "what's left" after removing $Z$'s influence from both.

4. **Added-variable plots visualize partial relationships.** Each panel shows the residual-vs-residual scatterplot, and the slope equals the regression coefficient. Use `car::avPlots()` to create them.

5. **Standardized coefficients allow cross-predictor comparison.** They express effects in standard deviation units, making it possible to compare the relative importance of predictors measured on different scales.

**Next:** In Lab 6, we will introduce **interaction terms** --- allowing the effect of one predictor to depend on the value of another.
:::
