---
title: "Lab 2: Simple Linear Regression"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Prepare a clean analysis dataset without overwriting the raw data
- Fit a simple linear regression model with `lm()`
- Inspect coefficients with `model_params()`
- Inspect model fit with `model_fit()` and `performance::r2()`
- Use the model to make simple predictions
- Compute fitted values and residuals with minimal code
- Use `performance::check_model()` for a quick visual diagnostic panel
- **Lecture Notes Reference:** Chapter 1: *Simple Linear Regression*

::: {.meta}
Dataset: `instruction`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Data Preparation

In Lab 1, we learned the basic R workflow: load data, inspect it, create a clean working dataset, and then analyze. We will use the same pattern here.

### Step 1: Load the Data

The course setup has already loaded the packages we need. The first code you run is the data-loading code:

```{r}
#| label: lab02-load-data

data(instruction)
instruction_raw <- instruction
```

We keep the original dataset in `instruction_raw`. All changes for analysis will go into a separate object.

### Step 2: Create a Clean Working Dataset

For readable tables and plots, we will rename `girl` to `female`, add labeled group variables, and create a first-grade math score.

```{r}
#| label: lab02-clean-data

instruction_clean <- instruction_raw |>
  rename(female = girl) |>
  mutate(
    gender = factor(female,
                    levels = c(0, 1),
                    labels = c("Male", "Female")),
    minority_group = factor(minority,
                            levels = c(0, 1),
                            labels = c("Non-minority", "Minority")),
    newteach = if_else(yearstea < 5, 1, 0),
    mathfirst = mathkind + mathgain
  )
```

> `instruction_clean` is the dataset we will use for analysis. `instruction_raw` remains unchanged.

### Step 3: Focus on One School

To learn the mechanics of simple regression, we will start with a small subset: students in School 11.

```{r}
#| label: lab02-create-school11

school11 <- instruction_clean |>
  filter(schoolid == 11)

tibble(
  students = nrow(school11),
  classrooms = n_distinct(school11$classid)
)
```

Working with one school keeps the first regression example small and concrete. Later in the lab, we will return to the full dataset.

### Step 4: Preflight Check

Before fitting a model, check the variables that will enter the model:

```{r}
#| label: lab02-preflight

school11 |>
  summarise(
    n = n(),
    missing_mathgain = sum(is.na(mathgain)),
    missing_mathkind = sum(is.na(mathkind)),
    min_mathkind = min(mathkind, na.rm = TRUE),
    max_mathkind = max(mathkind, na.rm = TRUE),
    min_mathgain = min(mathgain, na.rm = TRUE),
    max_mathgain = max(mathgain, na.rm = TRUE)
  )
```

The outcome will be `mathgain`, and the predictor will be `mathkind`.

### Step 5: Descriptive Statistics

Here are simple summaries for the key continuous variables:

```{r}
#| label: lab02-descriptives

school11 |>
  summarise(
    mean_ses = round(mean(ses, na.rm = TRUE), 2),
    sd_ses = round(sd(ses, na.rm = TRUE), 2),
    mean_mathkind = round(mean(mathkind, na.rm = TRUE), 2),
    sd_mathkind = round(sd(mathkind, na.rm = TRUE), 2),
    mean_mathgain = round(mean(mathgain, na.rm = TRUE), 2),
    sd_mathgain = round(sd(mathgain, na.rm = TRUE), 2)
  )
```

And here are the same regression variables summarized by gender:

```{r}
#| label: lab02-descriptives-by-gender

school11 |>
  group_by(gender) |>
  summarise(
    n = n(),
    mean_mathkind = round(mean(mathkind, na.rm = TRUE), 2),
    mean_mathgain = round(mean(mathgain, na.rm = TRUE), 2),
    .groups = "drop"
  )
```

```{r}
#| label: lab02-count-gender-minority

school11 |>
  count(gender, minority_group)
```


## Part 2: Explore the Relationship

Regression starts with a visual question: do the two variables appear to move together?

### Boxplot

```{r}
#| label: lab02-boxplot-mathkind
#| fig-cap: "Distribution of kindergarten math scores by gender in School 11"
#| fig-alt: "Boxplot comparing kindergarten math scores for male and female students in School 11, with similar medians and overlapping interquartile ranges"

ggplot(school11, aes(x = gender, y = mathkind, fill = gender)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  ber640_fill +
  labs(x = "Gender", y = "Math Score at Kindergarten Entry")
```

### Histogram

```{r}
#| label: lab02-histogram-ses
#| fig-cap: "Distribution of SES in School 11"
#| fig-alt: "Histogram of socioeconomic status in School 11, showing values distributed around the center with moderate spread"

ggplot(school11, aes(x = ses)) +
  geom_histogram(bins = 10, fill = ber640_colors["accent"],
                 color = "white", alpha = 0.8) +
  labs(x = "Socioeconomic Status (SES)", y = "Count")
```

### Scatterplot

The most important plot for simple regression is a scatterplot of the outcome and predictor.

```{r}
#| label: lab02-scatter-mathkind-mathgain
#| fig-cap: "Math gain vs. kindergarten math score in School 11"
#| fig-alt: "Scatterplot of math gain versus kindergarten math score in School 11, showing a negative relationship where students with higher initial scores tend to gain fewer points"

ggplot(school11, aes(x = mathkind, y = mathgain)) +
  geom_point(alpha = 0.65, size = 2, color = ber640_colors["primary"]) +
  labs(x = "Math Score at Kindergarten Entry",
       y = "Math Achievement Gain")
```

We can color the same relationship by gender:

```{r}
#| label: lab02-scatter-by-gender
#| fig-cap: "Math gain vs. kindergarten math score by gender"
#| fig-alt: "Scatterplot of math gain versus kindergarten math score colored by gender, showing similar negative trends for male and female students"

ggplot(school11, aes(x = mathkind, y = mathgain, color = gender)) +
  geom_point(alpha = 0.65, size = 2) +
  ber640_palette +
  labs(x = "Math Score at Kindergarten Entry",
       y = "Math Achievement Gain",
       color = "Gender")
```

Finally, add a fitted regression line:

```{r}
#| label: lab02-scatter-with-line
#| fig-cap: "Math gain vs. kindergarten math score with fitted regression line"
#| fig-alt: "Scatterplot of math gain versus kindergarten math score with a downward-sloping fitted regression line and confidence band"

ggplot(school11, aes(x = mathkind, y = mathgain)) +
  geom_point(alpha = 0.65, size = 2, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = TRUE,
              color = ber640_colors["secondary"], linewidth = 1) +
  labs(x = "Math Score at Kindergarten Entry",
       y = "Math Achievement Gain")
```

The downward line suggests a **negative relationship**: students who entered with higher math scores tended to gain fewer points.


## Part 3: Fit a Simple Linear Regression

Now we fit the same relationship formally:

$$
\text{mathgain}_i = \beta_0 + \beta_1 \cdot \text{mathkind}_i + \varepsilon_i
$$

### Step 1: Fit the Model

```{r}
#| label: lab02-fit-model1

model1 <- lm(mathgain ~ mathkind, data = school11)
```

The formula `mathgain ~ mathkind` reads as: "predict math gain from kindergarten math score."

### Step 2: Inspect the Coefficients

```{r}
#| label: lab02-model1-parameters

model_params(model1)
```

> `model_params()` prints the coefficient estimate, standard error, confidence interval, test statistic, and p-value in a compact table.

::: {.interpretation}
The intercept is the predicted math gain for a student with `mathkind = 0`.
That value is outside the observed range here, so it is mostly a mathematical anchor for the line.

The slope is the main result: for each 1-point higher kindergarten math score, the model predicts about `r abs(round(coef(model1)[2], 2))` fewer math-gain points.
Use "predicts" or "is associated with" here; this simple model alone does not prove a causal effect.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Write one plain-language sentence interpreting the slope for `mathkind`.
Your sentence should include the direction, the unit of the predictor, and the unit of the outcome.

<details><summary>Hint</summary>
Start with: "In School 11, each 1-point higher kindergarten math score is associated with..."
</details>
:::

::: {.pitfall}
Do not write that a higher kindergarten score **causes** lower math gain.
At this point, the model describes an association in one school.
:::

### Step 3: Inspect Model Fit

```{r}
#| label: lab02-model1-fit

model_fit(model1)
performance::r2(model1)
```

> `model_fit()` gives common model-fit statistics. `performance::r2()` focuses on $R^2$, the proportion of outcome variation explained by the model.

::: {.interpretation}
$R^2$ is about `r round(performance::r2(model1)$R2, 3)`.
This means the model explains about `r round(100 * performance::r2(model1)$R2, 1)`% of the variation in math gain among students in School 11.
Do not read $R^2$ as "percent correct"; it is about variation explained, not classification accuracy.
:::

### Step 4: Write the Regression Equation

Based on the fitted model, the estimated regression equation is:

$$
\widehat{\text{mathgain}} = `r round(coef(model1)[1], 2)` + (`r round(coef(model1)[2], 2)`) \times \text{mathkind}
$$

The slope is negative. Students with kindergarten math scores 1 point higher are predicted to have math gain about `r abs(round(coef(model1)[2], 2))` points lower, on average.


## Part 4: Predictions and Residuals

After fitting a model, we often ask two practical questions:

- What does the model predict for specific predictor values?
- How far away are the observed values from the fitted line?

### Step 1: Make Simple Predictions

```{r}
#| label: lab02-prediction-grid

prediction_grid <- tibble(mathkind = c(450, 500, 550))

prediction_grid |>
  mutate(
    predicted_mathgain = round(predict(model1, newdata = prediction_grid), 2)
  )
```

This table answers: "What math gain does the model predict for students with these kindergarten math scores?"

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Add `600` to `prediction_grid` and rerun the prediction code.
What happens to predicted math gain as the kindergarten score gets larger?

<details><summary>Hint</summary>
Change the first line to `prediction_grid <- tibble(mathkind = c(450, 500, 550, 600))`.
</details>
:::

### Step 2: Add Fitted Values and Residuals

```{r}
#| label: lab02-add-fitted-residuals

school11_aug <- school11 |>
  mutate(
    mathgain_hat = predict(model1),
    resid = residuals(model1)
  )
```

The fitted value is the model's prediction. The residual is:

$$
e_i = Y_i - \widehat{Y}_i
$$

### Step 3: Check the Residual Calculation

```{r}
#| label: lab02-check-residuals

school11_aug |>
  select(mathgain, mathgain_hat, resid) |>
  mutate(check = mathgain - mathgain_hat) |>
  head(8)
```

The `resid` column and the `check` column should match, up to small rounding differences.

### Step 4: Summarize Residuals

```{r}
#| label: lab02-residual-summary

school11_aug |>
  summarise(
    mean_resid = round(mean(resid), 4),
    sd_resid = round(sd(resid), 4),
    min_resid = round(min(resid), 4),
    max_resid = round(max(resid), 4)
  )
```

In an ordinary least squares regression with an intercept, the mean residual is essentially zero.

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Find one student with a positive residual and one student with a negative residual.
In one sentence for each student, describe whether the observed math gain was above or below the model prediction.

<details><summary>Hint</summary>
Use `arrange(desc(resid))` to find a large positive residual and `arrange(resid)` to find a large negative residual.
</details>
:::

### Step 5: Visualize Residuals

```{r}
#| label: lab02-scatter-fitted-residuals
#| fig-cap: "Observed data, fitted line, and residuals"
#| fig-alt: "Scatterplot with observed math gain values, a fitted regression line, and vertical grey line segments showing residuals from each observed value to the fitted line"

ggplot(school11_aug, aes(x = mathkind, y = mathgain)) +
  geom_segment(aes(xend = mathkind, yend = mathgain_hat),
               color = "grey60", linewidth = 0.4) +
  geom_point(alpha = 0.7, size = 2, color = ber640_colors["primary"]) +
  geom_line(data = school11_aug |> arrange(mathkind),
            aes(y = mathgain_hat),
            color = ber640_colors["secondary"], linewidth = 1) +
  labs(x = "Math Score at Kindergarten Entry",
       y = "Math Achievement Gain",
       subtitle = "Vertical grey lines show residuals")
```

Each grey line is one residual. The regression line is the line that minimizes the sum of squared residuals.


## Part 5: Quick Model Diagnostics

After fitting a model, inspect whether the model's assumptions look reasonable enough for the classroom analysis.
For linear regression, start with a visual diagnostic panel:

```{r}
#| label: lab02-check-model
#| fig-cap: "Quick diagnostic checks for the simple regression model"
#| fig-alt: "Four diagnostic plots for the simple regression model. The residual smooth is mostly near zero with some waviness, the variance smooth rises in the middle, one labeled point has the largest standardized residual, and the normality panel shows most points near the reference line with one upper-tail point."
#| fig-width: 10
#| fig-height: 8
#| out-width: "100%"

performance::check_model(
  model1,
  check = c("linearity", "homogeneity", "outliers", "qq"),
  base_size = 9
)
```

::: {.assumption-check}
Use the diagnostic panel to make a decision:

1. **Linearity:** Is the smooth line roughly flat, or does it curve strongly?
2. **Equal variance:** Does the spread of residuals look fairly even, or does it fan out?
3. **Influential observations:** Are any points marked as unusually influential?
4. **Normality:** Do most points in the Q-Q plot follow the reference line?

For this lab, write a short diagnostic note rather than trying to "fix" the model.
Later labs will show targeted tools for nonlinearity, robust standard errors, and influential observations.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Look at the diagnostic panel and choose one label: **acceptable**, **questionable**, or **serious concern**.
Write one sentence naming the plot that most influenced your decision.

<details><summary>Hint</summary>
A good diagnostic sentence has this shape: "I would label the model as ___ because the ___ plot shows ___."
</details>
:::


## Part 6: Additional Simple Regression Examples

The same workflow applies to other simple regression questions. The pattern is:

1. Fit the model with `lm()`.
2. Inspect coefficients with `model_params()`.
3. Inspect fit with `model_fit()` when fit matters.
4. Interpret the slope in substantive units.

### Example 1: A Binary Predictor

Is there a difference in kindergarten math scores between minority and non-minority students?

```{r}
#| label: lab02-example-minority-ttest

t.test(mathkind ~ minority_group,
       data = instruction_clean,
       var.equal = TRUE)
```

```{r}
#| label: lab02-example-minority-regression

model2 <- lm(mathkind ~ minority, data = instruction_clean)
model_params(model2)
```

When the predictor is binary, the slope is the mean difference between the two groups.

### Example 2: A Continuous Predictor

Does SES predict math gain in the full dataset?

```{r}
#| label: lab02-example-ses-regression

model3 <- lm(mathgain ~ ses, data = instruction_clean)
model_params(model3)
model_fit(model3)
```

```{r}
#| label: lab02-example-ses-plot
#| fig-cap: "Math gain vs. SES with fitted regression line"
#| fig-alt: "Scatterplot of math gain versus socioeconomic status for the full dataset with a nearly flat fitted regression line and confidence band, showing little association"

ggplot(instruction_clean, aes(x = ses, y = mathgain)) +
  geom_point(alpha = 0.3, size = 1.5, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = TRUE,
              color = ber640_colors["secondary"], linewidth = 1) +
  labs(x = "Socioeconomic Status (SES)",
       y = "Math Achievement Gain")
```

### Example 3: A Gender Difference

Does average math gain differ by gender?

```{r}
#| label: lab02-example-gender-regression

model4 <- lm(mathgain ~ female, data = instruction_clean)
model_params(model4)
```

```{r}
#| label: lab02-example-gender-means

instruction_clean |>
  group_by(gender) |>
  summarise(
    n = n(),
    mean_mathgain = round(mean(mathgain, na.rm = TRUE), 2),
    sd_mathgain = round(sd(mathgain, na.rm = TRUE), 2),
    .groups = "drop"
  )
```

The intercept is the mean for boys (`female = 0`). The slope is the difference in mean math gain between girls and boys.


## Part 7: Mini Regression Report

A compact regression report should answer the research question, name the model, interpret the main coefficient, summarize fit, and mention one diagnostic check.

::: {.interpretation}
Example report:

> In School 11, kindergarten math score was negatively associated with math gain.
> A simple linear regression estimated that each 1-point higher kindergarten score predicted about `r abs(round(coef(model1)[2], 2))` fewer gain points.
> The model explained about `r round(performance::r2(model1)$R2, 3)` of the variation in math gain.
> The quick diagnostic panel should be checked for curvature, unequal residual spread, influential observations, and Q-Q plot departures before treating this as a final analysis.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Write your own 4-sentence mini report for `model1`.
Use the example above as a template, but use your own words.
:::

::: {.exercises}
## What to Submit

1. A coefficient table for `model1` from `model_params(model1)`. [core]{.chip .easy}
2. A prediction table with at least four `mathkind` values. [core]{.chip .easy}
3. One residual check: one positive residual and one negative residual, with a sentence interpreting each. [core]{.chip .med}
4. One diagnostic note using the `performance::check_model()` panel. [core]{.chip .med}
5. A 4-sentence mini regression report answering the School 11 research question. [core]{.chip .med}
:::

::: {.summary}
## Lab Summary

1. Simple linear regression starts with a clear outcome, one predictor, and an `lm()` formula.

2. `model_params()` is the default coefficient table for this course.

3. `model_fit()` and `performance::r2()` summarize how well the model fits.

4. `predict()` turns a fitted model into predicted values for new or existing cases.

5. Residuals are the differences between observed outcomes and fitted values.

6. The same `lm()` workflow works for continuous predictors and binary predictors.

7. `performance::check_model()` gives a quick visual diagnostic panel; your job is to make a reasoned decision, not just paste the plot.

**Next:** Lab 3 moves from simple regression to ANCOVA: comparing groups while adjusting for another variable.
:::
