---
title: "Lab 3: ANCOVA and Statistical Control"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Prepare a two-group analysis dataset from `gcse`
- Compare raw group means with a simple regression model
- Fit an ANCOVA model with one categorical predictor and one continuous covariate
- Interpret "holding a covariate constant" in plain language
- Compute adjusted means with `modelbased::estimate_means()`
- Compare unadjusted and adjusted group differences
- Check collinearity with `performance::check_collinearity()`
- **Lecture Notes Reference:** Chapter 2: *ANCOVA*

::: {.meta}
Dataset: `gcse`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Data Preparation

In Chapter 2, the motivating question is:

> Do two schools differ in GCSE scores after accounting for students' earlier LRT scores?

We will focus on two schools from the `gcse` dataset:

- **Battersea**: school code `15`
- **Chelsea**: school code `6`

### Step 1: Load the Data

```{r}
#| label: lab03-load-data

data(gcse)
gcse_raw <- gcse
```

As in the earlier labs, we keep the original data in a raw object and create a clean analysis object separately.

### Step 2: Create the Two-School Dataset

```{r}
#| label: lab03-create-two-schools

two_schools <- gcse_raw |>
  filter(school %in% c(15, 6)) |>
  mutate(
    school_name = factor(
      if_else(school == 15, "Battersea", "Chelsea"),
      levels = c("Battersea", "Chelsea")
    )
  )
```

The reference group is Battersea because it is the first factor level.

### Step 3: Preflight Check

```{r}
#| label: lab03-preflight

two_schools |>
  summarise(
    students = n(),
    schools = n_distinct(school_name),
    missing_gcse = sum(is.na(gcse)),
    missing_lrt = sum(is.na(lrt)),
    min_lrt = round(min(lrt, na.rm = TRUE), 2),
    max_lrt = round(max(lrt, na.rm = TRUE), 2),
    mean_lrt = round(mean(lrt, na.rm = TRUE), 2)
  )
```

The outcome is `gcse`, the age-16 exam score. The covariate is `lrt`, an earlier reading test score measured at age 11.

### Step 4: Raw Means by School

```{r}
#| label: lab03-raw-school-summary

two_schools |>
  group_by(school_name) |>
  summarise(
    n = n(),
    mean_gcse = round(mean(gcse, na.rm = TRUE), 2),
    mean_lrt = round(mean(lrt, na.rm = TRUE), 2),
    .groups = "drop"
  )
```

This table shows the core confounding pattern: Chelsea has a higher raw GCSE mean, but Chelsea also has a higher mean LRT score at intake.

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Use the raw summary table to write two short sentences:

1. Which school has the higher raw GCSE mean?
2. Which school has the higher mean LRT score?

<details><summary>Hint</summary>
The second sentence is the clue that LRT may be a confounder.
</details>
:::


## Part 2: Visualize the Confounding Pattern

### Raw GCSE Scores

```{r}
#| label: lab03-boxplot-gcse-school
#| fig-cap: "Raw GCSE scores by school"
#| fig-alt: "Boxplot comparing raw GCSE scores for Battersea and Chelsea, with Chelsea having a higher median GCSE score"

ggplot(two_schools, aes(x = school_name, y = gcse, fill = school_name)) +
  geom_boxplot(alpha = 0.75, show.legend = FALSE) +
  ber640_fill +
  labs(x = "School", y = "GCSE Score")
```

### GCSE and LRT Together

```{r}
#| label: lab03-scatter-lrt-gcse-by-school
#| fig-cap: "GCSE vs. LRT by school"
#| fig-alt: "Scatterplot of GCSE versus LRT for Battersea and Chelsea, with separate upward-sloping regression lines; Chelsea students tend to have higher LRT scores and higher GCSE scores"

ggplot(two_schools, aes(x = lrt, y = gcse, color = school_name)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  ber640_palette +
  labs(x = "LRT Score at Age 11",
       y = "GCSE Score at Age 16",
       color = "School")
```

The two lines are roughly parallel. That makes ANCOVA a reasonable first model: one common slope for LRT, plus one adjusted school difference.

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Look at the scatterplot and decide whether the lines look roughly parallel.
Write one sentence saying whether the ANCOVA equal-slopes assumption seems reasonable as a starting point.
:::


## Part 3: Fit the Models

ANCOVA compares schools while holding LRT constant. To see what adjustment changes, we fit two models:

1. An **unadjusted** model with school only.
2. An **adjusted** ANCOVA model with school and LRT.

### Step 1: Unadjusted School Difference

```{r}
#| label: lab03-fit-unadjusted

model_unadjusted <- lm(gcse ~ school_name, data = two_schools)
model_params(model_unadjusted)
```

The Chelsea coefficient is the raw mean difference in GCSE scores between Chelsea and Battersea.

### Step 2: Adjusted School Difference

```{r}
#| label: lab03-fit-ancova

model_ancova <- lm(gcse ~ lrt + school_name, data = two_schools)
model_params(model_ancova)
```

::: {.interpretation}
In the ANCOVA model, `lrt` is the covariate.

The `school_nameChelsea` coefficient compares Chelsea and Battersea students with the **same LRT score**.
That is what "holding LRT constant" means in this model.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Interpret the `school_nameChelsea` coefficient from `model_ancova` in one sentence.
Your sentence should use the phrase "after accounting for LRT" or "holding LRT constant."

<details><summary>Hint</summary>
Start with: "After accounting for LRT, Chelsea students are predicted to..."
</details>
:::

::: {.pitfall}
Do not write that Chelsea **causes** higher GCSE scores.
This is observational school data, so the adjusted difference is a model-adjusted comparison.
It is not automatic proof of a causal school effect.
:::

### Step 3: Model Fit

```{r}
#| label: lab03-ancova-fit

model_fit(model_ancova)
performance::r2(model_ancova)
```

The ANCOVA model should explain more variation in GCSE than the school-only model because LRT is strongly related to later GCSE performance.

### Step 4: Compare the School Coefficient

```{r}
#| label: lab03-gap-comparison

gap_comparison <- bind_rows(
  tidy_model(model_unadjusted) |>
    filter(term == "school_nameChelsea") |>
    mutate(model = "Unadjusted"),
  tidy_model(model_ancova) |>
    filter(term == "school_nameChelsea") |>
    mutate(model = "Adjusted for LRT")
) |>
  select(model, estimate, std.error, conf.low, conf.high, p.value) |>
  mutate(
    across(c(estimate, std.error, conf.low, conf.high), \(x) round(x, 3)),
    p.value = if_else(p.value < .001, "< .001", as.character(round(p.value, 3)))
  )

gap_comparison
```

::: {.interpretation}
The adjusted estimate is smaller than the unadjusted estimate.
This tells us that part of the raw Chelsea-Battersea GCSE gap reflects differences in LRT scores at intake.

This is a model-adjusted comparison, not automatic proof that school attendance caused the remaining gap.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Use `gap_comparison` to compute how much the Chelsea-Battersea estimate changed after adding LRT.
Then write one sentence explaining what the change suggests.

<details><summary>Hint</summary>
Subtract the adjusted estimate from the unadjusted estimate. A positive difference means the raw gap was larger than the adjusted gap.
</details>
:::


## Part 4: Adjusted Means

Adjusted means answer this question:

> What GCSE score does the model predict for each school if both schools are compared at the same representative LRT value?

### Step 1: Estimate Adjusted Means

```{r}
#| label: lab03-adjusted-means

adjusted_means <- modelbased::estimate_means(model_ancova, by = "school_name")
adjusted_means
```

The adjusted means are model-predicted means for each school after accounting for LRT.
Here, both schools are evaluated at the same representative LRT value: the sample mean, about `r round(mean(two_schools$lrt, na.rm = TRUE), 2)`.

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Compare the adjusted means with the raw means.
Which school moved upward or downward after adjustment?

<details><summary>Hint</summary>
Use the raw school summary from Part 1 and the `adjusted_means` output above; compare `mean_gcse` with the adjusted mean within each school.
</details>
:::

### Step 2: Estimate the Adjusted Contrast

```{r}
#| label: lab03-adjusted-contrast

modelbased::estimate_contrasts(model_ancova, contrast = "school_name")
```

The adjusted contrast equals the `school_nameChelsea` coefficient from `model_ancova`.
Because `Level1` is Chelsea and `Level2` is Battersea, the contrast is Chelsea minus Battersea.

### Step 3: Raw vs. Adjusted Means

```{r}
#| label: lab03-raw-adjusted-table

raw_means <- two_schools |>
  group_by(school_name) |>
  summarise(raw_mean = mean(gcse, na.rm = TRUE), .groups = "drop")

adjusted_means_tbl <- adjusted_means |>
  as_tibble() |>
  select(school_name, adjusted_mean = Mean)

left_join(raw_means, adjusted_means_tbl, by = "school_name") |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

::: {.interpretation}
The raw means are observed school averages.
The adjusted means are model predictions at the same representative LRT value.
They answer a comparison question, not a causal question.
:::

### Step 4: Plot Raw and Adjusted Means

```{r}
#| label: lab03-raw-adjusted-plot
#| fig-cap: "Raw and LRT-adjusted mean GCSE scores"
#| fig-alt: "Dot and line plot comparing raw and LRT-adjusted mean GCSE scores for Battersea and Chelsea, showing that the school gap becomes smaller after adjustment"

means_plot <- bind_rows(
  raw_means |>
    rename(mean_score = raw_mean) |>
    mutate(type = "Raw mean"),
  adjusted_means_tbl |>
    rename(mean_score = adjusted_mean) |>
    mutate(type = "Adjusted mean")
) |>
  mutate(type = factor(type, levels = c("Raw mean", "Adjusted mean")))

ggplot(means_plot, aes(x = type, y = mean_score,
                       color = school_name, group = school_name)) +
  geom_line(linetype = "dashed", linewidth = 0.8, alpha = 0.7) +
  geom_point(size = 4) +
  ber640_palette +
  labs(x = NULL, y = "Mean GCSE Score", color = "School")
```


## Part 5: Model Comparison and Collinearity

### Step 1: Compare Model Performance

```{r}
#| label: lab03-compare-performance

compare_models(model_unadjusted, model_ancova)
```

The ANCOVA model should have higher $R^2$ and lower AIC/BIC than the school-only model.

### Step 2: Nested Model Test

```{r}
#| label: lab03-nested-test

anova(model_unadjusted, model_ancova)
```

This test asks whether adding LRT improves the model beyond school alone.

### Step 3: Check the Parallel-Slopes Assumption

The ANCOVA model assumes that the LRT slope is the same in both schools.
We can check that assumption by comparing the ANCOVA model with an interaction model:

```{r}
#| label: lab03-parallel-slopes-test

model_interaction <- lm(gcse ~ lrt * school_name, data = two_schools)
anova(model_ancova, model_interaction)
```

This test asks whether allowing the LRT slope to differ by school improves the model.
A non-significant result does not prove the slopes are exactly equal, but it supports using the simpler parallel-lines ANCOVA model for this lab.

### Step 4: Check Collinearity

```{r}
#| label: lab03-check-collinearity

performance::check_collinearity(model_ancova)
```

::: {.assumption-check}
Low VIF values mean that `lrt` and `school_name` are not so redundant that they make the coefficients unstable.
In later labs with more predictors, collinearity checks will become more important.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Use the parallel-slopes test and the collinearity table to write one diagnostic note.
Your note should mention whether the interaction test and VIF table support the simple ANCOVA model as a reasonable working model.
:::

::: {.exercises}
## What to Submit

1. The raw school summary table with `mean_gcse` and `mean_lrt`. [core]{.chip .easy}
2. The coefficient table for `model_ancova`. [core]{.chip .easy}
3. The `gap_comparison` table comparing the unadjusted and LRT-adjusted school coefficient. [core]{.chip .med}
4. The raw vs. adjusted means table or plot. [core]{.chip .med}
5. One diagnostic note using the parallel-slopes test and `performance::check_collinearity()` output. [core]{.chip .med}
6. A 4-sentence interpretation answering the research question. Use model-adjusted language and avoid causal claims. [core]{.chip .hard}
:::


::: {.summary}
## Lab Summary

1. ANCOVA is a regression model with both a categorical predictor and a continuous covariate.

2. The unadjusted school coefficient is the raw group mean difference.

3. The adjusted school coefficient compares schools at the same LRT level.

4. `modelbased::estimate_means()` gives adjusted means, and `modelbased::estimate_contrasts()` gives adjusted group differences.

5. `compare_models()` and `anova()` show whether the covariate improves model fit.

6. A quick interaction test checks whether the parallel-slopes ANCOVA is a reasonable working model.

7. `performance::check_collinearity()` checks whether predictors are too redundant for stable coefficient interpretation.

**Next:** Lab 4 moves the one-way ANOVA and categorical-predictor material into its own focused workflow.
:::
