---
title: "Lab 7: Model Building and Nonlinear Relationships"
---

{{< include ../_common.qmd >}}

## Goals

::: {.callout-note}
## What You'll Learn in This Lab
- Compare nested models using F-tests (`anova()` and `car::linearHypothesis()`)
- Detect nonlinearity with LOWESS smoothing curves
- Apply log and square-root transformations to address skewness
- Fit polynomial regression models (quadratic, cubic)
- Model piecewise constant relationships using `cut()`
- **Lecture Notes Reference:** Chapter 6: *Nonlinear Relationships* and Chapter 7: *Model Building*
:::

**Dataset:** `hsbs1` from the `regdatasets` package


## Setup

```{r}
#| label: lab07-setup

library(regdatasets)
data(hsbs1)

hsbs1 <- hsbs1 |>
  mutate(
    prog_f = factor(prog,
                    levels = 1:3,
                    labels = c("General", "Academic", "Vocational")),
    race_f = factor(race,
                    levels = 1:4,
                    labels = c("Hispanic", "Asian", "African-Am", "White"))
  )

glimpse(hsbs1)
```


## Part 1: Model Comparison --- Testing Sets of Predictors

So far, every time we wanted to test whether a predictor matters, we looked at its individual *t*-test. But what about categorical variables with more than two levels? A 3-category variable like `prog` uses **two** dummy indicators, and we need a **joint** test that asks: "Do these two dummies, taken together, improve the model?"

The tool for this is the **incremental F-test** --- compare a reduced model (without the predictors in question) to a full model (with them), and see whether the improvement in $R^2$ is statistically significant.

### Fitting the Reduced and Full Models

```{r}
#| label: reduced-full-models

model_reduced <- lm(write ~ read, data = hsbs1)
model_full    <- lm(write ~ read + prog_f, data = hsbs1)
```

The reduced model predicts writing from reading alone. The full model adds program type (`prog_f`), which requires two dummy variables (Academic vs. General, Vocational vs. General).

### The F-Test with `anova()`

```{r}
#| label: anova-comparison

anova(model_reduced, model_full)
```

> The null hypothesis is $H_0$: all coefficients for `prog_f` equal zero --- meaning program type adds nothing beyond reading score. A significant *p*-value tells us the full model fits meaningfully better.

Let's also look at what the full model actually estimates:

```{r}
#| label: full-model-output

tidy_model(model_full)
```

### Alternative: `car::linearHypothesis()`

The `linearHypothesis()` function from the `car` package lets you test the same hypothesis directly on the full model, without fitting the reduced model separately. You specify which coefficients should equal zero:

```{r}
#| label: linear-hypothesis-prog

linearHypothesis(model_full, c("prog_fAcademic = 0", "prog_fVocational = 0"))
```

> You should get the same F-statistic and *p*-value as `anova()`. The two approaches are equivalent --- `linearHypothesis()` is simply more convenient when you want to test a subset of coefficients without re-fitting a reduced model.

### Another Example: Testing Race

Let's test whether race contributes beyond reading and program:

```{r}
#| label: test-race

model_no_race   <- lm(write ~ read + prog_f, data = hsbs1)
model_with_race <- lm(write ~ read + prog_f + race_f, data = hsbs1)

anova(model_no_race, model_with_race)
```

::: {.callout-tip}
## When to Use Incremental F-Tests
Use them whenever you want to test a **set** of predictors simultaneously: a multi-category factor, a block of control variables, or an interaction that involves multiple terms. Individual *t*-tests only test one coefficient at a time.
:::


## Part 2: Detecting Nonlinearity

Before we fit more complex models, we need a way to *see* whether the linear assumption holds. The key visual tool is the **LOWESS** (Locally Weighted Scatterplot Smoother) curve --- a flexible, data-driven curve that reveals the shape of the relationship without assuming linearity.

### Linear Fit vs. LOWESS: Read and Concept

```{r}
#| label: lowess-linear-concept
#| fig-alt: "Scatterplot with linear and LOWESS curves showing reading score versus self-concept, where both curves nearly overlap indicating a linear relationship"
#| fig-cap: "Reading vs. self-concept: the linear fit tracks the LOWESS curve well."
#| fig-width: 8
#| fig-height: 5

ggplot(hsbs1, aes(x = concept, y = read)) +
  geom_point(alpha = 0.3, size = 2, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = FALSE, color = ber640_colors["secondary"],
              linewidth = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = ber640_colors["accent"],
              linewidth = 1.2, linetype = "dashed") +
  labs(
    x = "Self-Concept", y = "Reading Score",
    title = "Read vs. Concept: Linear Looks Fine",
    subtitle = "Solid red = linear fit | Dashed blue = LOWESS"
  )
```

Here, the LOWESS curve and the straight line nearly overlap --- a good sign that linearity is reasonable.

### Linear Fit vs. LOWESS: Write and Read

```{r}
#| label: lowess-linear-write
#| fig-alt: "Scatterplot with linear and LOWESS curves showing reading score versus writing score, with slight curvature visible at the extremes of the LOWESS curve"
#| fig-cap: "Writing vs. reading: is there curvature the linear fit misses?"
#| fig-width: 8
#| fig-height: 5

ggplot(hsbs1, aes(x = write, y = read)) +
  geom_point(alpha = 0.3, size = 2, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = FALSE, color = ber640_colors["secondary"],
              linewidth = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = ber640_colors["accent"],
              linewidth = 1.2, linetype = "dashed") +
  labs(
    x = "Writing Score", y = "Reading Score",
    title = "Read vs. Write: Any Curvature?",
    subtitle = "Solid red = linear fit | Dashed blue = LOWESS"
  )
```

> Look carefully: if the LOWESS curve bends away from the straight line at the extremes, that is a clue that a nonlinear specification (polynomial, transformation) might be warranted. We will address this in Parts 4 and 5.


## Part 3: Data Transformations

When a variable is heavily skewed, a transformation can make the distribution more symmetric and the relationship with other variables more linear. The two most common transformations are:

- **Log transformation** `log(x)`: best for strong positive skew (long right tail).
- **Square-root transformation** `sqrt(x)`: a gentler option for moderate skew.

### Creating a Skewed Variable

The `hsbs1` variables are fairly well-behaved, so let's simulate a positively skewed variable to see transformations in action. We exponentiate `career` (which is roughly normal) to create something that mimics income or reaction time:

```{r}
#| label: create-skewed

hsbs1 <- hsbs1 |>
  mutate(career_exp = exp(career / 10))

summary(hsbs1$career_exp)
```

### Before and After: Histograms

```{r}
#| label: transformation-histograms
#| fig-alt: "Three-panel histogram showing original right-skewed distribution, log-transformed distribution approaching symmetry, and square-root-transformed distribution with moderate correction"
#| fig-cap: "Left: original skewed variable. Middle: log-transformed. Right: sqrt-transformed."
#| fig-width: 10
#| fig-height: 4

p_raw <- ggplot(hsbs1, aes(x = career_exp)) +
  geom_histogram(bins = 25, fill = ber640_colors["primary"], color = "white") +
  labs(title = "Original (Skewed)", x = "career_exp", y = "Count")

p_log <- ggplot(hsbs1, aes(x = log(career_exp))) +
  geom_histogram(bins = 25, fill = ber640_colors["accent"], color = "white") +
  labs(title = "Log-Transformed", x = "log(career_exp)", y = "Count")

p_sqrt <- ggplot(hsbs1, aes(x = sqrt(career_exp))) +
  geom_histogram(bins = 25, fill = ber640_colors["success"], color = "white") +
  labs(title = "Sqrt-Transformed", x = "sqrt(career_exp)", y = "Count")

p_raw + p_log + p_sqrt
```

> The log transformation pulls in the long right tail dramatically. The square-root transformation is more moderate. Choose whichever produces a roughly symmetric distribution.

::: {.callout-important}
## When to Transform
- **Log:** the variable spans several orders of magnitude or has strong positive skew. Common for income, population, reaction times. Requires all values to be positive.
- **Square root:** moderate positive skew, or count data. Also requires non-negative values.
- **Interpretation changes:** a one-unit increase in log(x) corresponds to a multiplicative change in x (roughly a percentage change), not an additive change.
:::


## Part 4: Polynomial Regression

When the LOWESS curve suggests a bend, one approach is to add polynomial terms --- $x^2$ for a quadratic, $x^3$ for a cubic. This keeps the model within the linear regression framework (polynomial terms are just additional predictors).

### Quadratic Model

Let's fit a quadratic model predicting reading from writing:

```{r}
#| label: quadratic-model

model_linear    <- lm(read ~ write, data = hsbs1)
model_quadratic <- lm(read ~ write + I(write^2), data = hsbs1)

tidy_model(model_quadratic)
```

> The `I()` function tells R to treat `write^2` as an arithmetic operation rather than a formula operator. Alternatively, you can use `poly(write, 2, raw = TRUE)` to get the same result.

### Testing the Quadratic Term

Is the quadratic term significant? We compare the linear and quadratic models:

```{r}
#| label: test-quadratic

anova(model_linear, model_quadratic)
```

If the *p*-value is small, the curvature is statistically meaningful. If not, the straight line was good enough.

### Adding a Cubic Term

We can go further and test whether a cubic term helps:

```{r}
#| label: cubic-model

model_cubic <- lm(read ~ write + I(write^2) + I(write^3), data = hsbs1)

anova(model_quadratic, model_cubic)
```

> If the cubic term is not significant, stop at the quadratic. This is the standard strategy: start simple, add complexity only when the data demand it.

::: {.callout-important}
## The Hierarchy Principle
If a higher-order term (e.g., $x^2$) is in the model, **always keep all lower-order terms** ($x$) regardless of their individual significance. Removing $x$ while keeping $x^2$ changes the meaning of the quadratic term and makes the model uninterpretable.
:::

### Visualizing the Polynomial Fit

```{r}
#| label: polynomial-plot
#| fig-alt: "Scatterplot of reading versus writing scores with three overlaid curves: linear, quadratic, and cubic fits that are nearly indistinguishable, suggesting minimal curvature"
#| fig-cap: "Comparing linear, quadratic, and cubic fits."
#| fig-width: 8
#| fig-height: 5

# Generate predictions on a grid
write_grid <- tibble(write = seq(min(hsbs1$write), max(hsbs1$write), length.out = 200))

write_grid <- write_grid |>
  mutate(
    linear    = predict(model_linear,    newdata = write_grid),
    quadratic = predict(model_quadratic, newdata = write_grid),
    cubic     = predict(model_cubic,     newdata = write_grid)
  ) |>
  pivot_longer(cols = c(linear, quadratic, cubic),
               names_to = "model", values_to = "predicted") |>
  mutate(model = factor(model, levels = c("linear", "quadratic", "cubic")))

ggplot(hsbs1, aes(x = write, y = read)) +
  geom_point(alpha = 0.3, size = 2, color = "grey60") +
  geom_line(data = write_grid, aes(y = predicted, color = model),
            linewidth = 1.2) +
  scale_color_manual(
    values = c("linear" = ber640_colors["secondary"],
               "quadratic" = ber640_colors["accent"],
               "cubic" = ber640_colors["success"]),
    labels = c("Linear", "Quadratic", "Cubic")
  ) +
  labs(
    x = "Writing Score", y = "Reading Score", color = "Model",
    title = "Polynomial Regression: Read ~ Write",
    subtitle = "Does adding curvature improve the fit?"
  )
```

> Look at whether the curves separate meaningfully. If the quadratic and linear lines are nearly identical, the extra complexity was unnecessary.


## Part 5: Piecewise Constant Relationships

Sometimes the relationship is not smooth at all --- instead, the outcome jumps up at certain thresholds. A **piecewise constant** model captures this by cutting a continuous predictor into groups and using those groups as dummy variables.

### Cutting Reading into Quartile Groups

```{r}
#| label: create-quartile-groups

hsbs1 <- hsbs1 |>
  mutate(
    read_q = cut(read,
                 breaks = quantile(read, probs = c(0, .25, .50, .75, 1)),
                 include.lowest = TRUE,
                 labels = c("Q1", "Q2", "Q3", "Q4"))
  )

hsbs1 |> count(read_q)
```

### Regression with Quartile Dummies

```{r}
#| label: piecewise-model

model_piecewise <- lm(write ~ read_q, data = hsbs1)
tidy_model(model_piecewise)
```

> The intercept is the mean writing score for Q1 (the reference group). Each coefficient is the difference in mean writing between that quartile and Q1.

### Comparing All Approaches Visually

Let's put the linear, polynomial, and piecewise fits side by side:

```{r}
#| label: comparison-plot
#| fig-alt: "Three-panel comparison showing linear fit as a straight line, quadratic fit with slight curvature, and piecewise constant fit as horizontal segments at quartile group means"
#| fig-cap: "Three ways to model the read-write relationship."
#| fig-width: 10
#| fig-height: 5

# Piecewise predictions (step function)
piecewise_means <- hsbs1 |>
  group_by(read_q) |>
  summarise(
    read_min  = min(read),
    read_max  = max(read),
    predicted = mean(write),
    .groups   = "drop"
  )

p_linear <- ggplot(hsbs1, aes(x = read, y = write)) +
  geom_point(alpha = 0.3, size = 1.5, color = "grey60") +
  geom_smooth(method = "lm", se = FALSE,
              color = ber640_colors["secondary"], linewidth = 1.2) +
  labs(title = "Linear", x = "Reading", y = "Writing")

p_poly <- ggplot(hsbs1, aes(x = read, y = write)) +
  geom_point(alpha = 0.3, size = 1.5, color = "grey60") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE,
              color = ber640_colors["accent"], linewidth = 1.2) +
  labs(title = "Quadratic", x = "Reading", y = "Writing")

p_piece <- ggplot(hsbs1, aes(x = read, y = write)) +
  geom_point(alpha = 0.3, size = 1.5, color = "grey60") +
  geom_segment(data = piecewise_means,
               aes(x = read_min, xend = read_max,
                   y = predicted, yend = predicted),
               color = ber640_colors["success"], linewidth = 1.5) +
  labs(title = "Piecewise Constant", x = "Reading", y = "Writing")

p_linear + p_poly + p_piece
```

> The linear model assumes a constant slope everywhere. The quadratic allows curvature. The piecewise model makes no shape assumption at all --- it just estimates a separate mean for each group. The tradeoff: piecewise is the most flexible but throws away information about *where within* each quartile a student falls.

::: {.callout-tip}
## Choosing a Nonlinear Strategy

| Strategy | Best When | Drawback |
|---|---|---|
| **Polynomial** | Smooth curvature visible in LOWESS | Unstable at extremes (high-order) |
| **Transformation** | Skewed predictors or outcomes | Changes the scale of interpretation |
| **Piecewise constant** | Known meaningful thresholds | Loses within-group variation |

Start with LOWESS to *see* the relationship, then choose the simplest specification that captures the pattern.
:::


## Key Takeaways

::: {.callout-note}
## What to Remember

1. **Incremental F-tests** let you test whether a *set* of predictors (e.g., all dummies for a multi-category factor) jointly improve the model. Use `anova(reduced, full)` or `car::linearHypothesis()`.

2. **LOWESS curves** are your best diagnostic for nonlinearity. Overlay them on scatterplots and compare with the linear fit --- if they diverge, consider a nonlinear specification.

3. **Log and sqrt transformations** address skewness. Log is stronger; sqrt is gentler. Both require positive values and change the interpretation of coefficients.

4. **Polynomial regression** adds $x^2$, $x^3$, etc. as predictors. Always follow the **hierarchy principle**: keep lower-order terms when higher-order terms are significant.

5. **Piecewise constant models** use `cut()` to turn a continuous variable into groups. Maximum flexibility, but they discard within-group information.

6. **Model building is iterative:** visualize first, then formalize with a test. Let the data guide complexity --- add it when needed, not by default.
:::
