---
title: "Lab 7: Nonlinear Relationships"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Use LOWESS curves to detect nonlinearity
- Fit log-transformed regression models
- Center a predictor before adding polynomial terms
- Compare linear, quadratic, and cubic models with `anova()` and `model_fit()`
- Fit a piecewise constant model with `cut()`
- Fit a natural spline with `splines::ns()`
- Plot nonlinear fitted values with `ggeffects::predict_response()`
- **Lecture Notes Reference:** Chapter 6: *Nonlinear Relationships*

::: {.meta}
Datasets: `gcse` and `crime`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Prepare the Data

This lab uses two examples from the lecture notes:

1. `gcse`: LRT score and GCSE exam performance.
2. `crime`: crime rate and urbanization.

```{r}
#| label: lab07-load-data

data(gcse)
data(crime)

gcse_lab <- gcse |>
  filter(!is.na(gcse), !is.na(lrt)) |>
  mutate(
    lrt_c = lrt - mean(lrt, na.rm = TRUE)
  )

crime_lab <- crime |>
  rename(
    crime_rate = c,
    high_school = hs,
    urbanization = u
  ) |>
  filter(crime_rate > 0, urbanization > 0)
```

```{r}
#| label: lab07-preflight

tibble(
  Dataset = c("gcse", "crime"),
  Rows = c(nrow(gcse_lab), nrow(crime_lab)),
  Main_outcome = c("gcse", "crime_rate"),
  Main_predictor = c("lrt", "urbanization")
)
```


## Part 2: Detect Nonlinearity with LOWESS

Before fitting a nonlinear model, plot the data.
The LOWESS curve is a flexible smooth curve that helps us see whether a straight line misses the shape of the relationship.

```{r}
#| label: lab07-lowess-gcse-lrt
#| fig-cap: "Linear and LOWESS fits for the GCSE-LRT relationship."
#| fig-alt: "Scatter plot of GCSE score versus LRT score with a straight linear fit and a dashed LOWESS curve. The two curves mostly overlap, with small departures near the extremes."
#| fig-width: 8
#| fig-height: 5

ggplot(gcse_lab, aes(x = lrt, y = gcse)) +
  geom_point(alpha = 0.05, color = "grey50") +
  geom_smooth(
    method = "lm", formula = y ~ x, se = FALSE,
    color = ber640_colors[["primary"]], linewidth = 1.1
  ) +
  geom_smooth(
    method = "loess", formula = y ~ x, se = FALSE,
    color = ber640_colors[["secondary"]], linewidth = 1.1,
    linetype = "dashed"
  ) +
  labs(
    x = "LRT Score",
    y = "GCSE Score",
    color = NULL
  )
```

::: {.interpretation}
If the LOWESS curve closely follows the straight line, a linear model may be enough.

If the LOWESS curve bends away from the straight line, consider a transformation, polynomial, piecewise model, or spline.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Compare the straight line and LOWESS curve in the GCSE-LRT plot.
Does the relationship look mostly linear, or is there meaningful curvature?
Write two short sentences: one describing the visual pattern and one naming a reasonable next model to try.
:::


## Part 3: Log Transformations

Log transformations are useful when a relationship is multiplicative or when the variables have strong positive skew.

The `crime` example compares four common specifications.

```{r}
#| label: lab07-fit-log-models

model_level_level <- lm(crime_rate ~ urbanization, data = crime_lab)
model_log_level <- lm(log(crime_rate) ~ urbanization, data = crime_lab)
model_level_log <- lm(crime_rate ~ log(urbanization), data = crime_lab)
model_log_log <- lm(log(crime_rate) ~ log(urbanization), data = crime_lab)
```

```{r}
#| label: lab07-log-coefficient-table

log_model_slopes <- bind_rows(
  tidy_model(model_level_level) |>
    filter(term == "urbanization") |>
    mutate(Model = "Level-level"),
  tidy_model(model_log_level) |>
    filter(term == "urbanization") |>
    mutate(Model = "Log-level"),
  tidy_model(model_level_log) |>
    filter(term == "log(urbanization)") |>
    mutate(Model = "Level-log"),
  tidy_model(model_log_log) |>
    filter(term == "log(urbanization)") |>
    mutate(Model = "Log-log")
) |>
  select(Model, term, estimate, std.error, conf.low, conf.high, p.value)

log_model_slopes |>
  mutate(p.value = format_p(p.value)) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
The coefficient meaning depends on which side is logged.

- Level-level: raw-unit change in `crime_rate`.
- Log-level: approximate percent change in `crime_rate`.
- Level-log: raw-unit change in `crime_rate` for a percent change in urbanization.
- Log-log: elasticity, or percent change in `crime_rate` for a percent change in urbanization.
:::

### Visual Check for Log Specifications

```{r}
#| label: lab07-log-specification-plot
#| fig-cap: "Four log transformation choices for crime rate and urbanization."
#| fig-alt: "Four-panel plot comparing level-level, log-level, level-log, and log-log specifications. The log-log panel has the closest overlap between the straight linear fit and dashed LOWESS curve across most of the urbanization range; the level-log panel bends upward most strongly at high urbanization."
#| fig-width: 10
#| fig-height: 8

p_level_level <- ggplot(crime_lab, aes(x = urbanization, y = crime_rate)) +
  geom_point(alpha = 0.5, color = "grey50") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = ber640_colors[["primary"]]) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = ber640_colors[["secondary"]], linetype = "dashed") +
  labs(title = "Level-level", x = "Urbanization", y = "Crime Rate")

p_log_level <- ggplot(crime_lab, aes(x = urbanization, y = log(crime_rate))) +
  geom_point(alpha = 0.5, color = "grey50") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = ber640_colors[["primary"]]) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = ber640_colors[["secondary"]], linetype = "dashed") +
  labs(title = "Log-level", x = "Urbanization", y = "log(Crime Rate)")

p_level_log <- ggplot(crime_lab, aes(x = log(urbanization), y = crime_rate)) +
  geom_point(alpha = 0.5, color = "grey50") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = ber640_colors[["primary"]]) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = ber640_colors[["secondary"]], linetype = "dashed") +
  labs(title = "Level-log", x = "log(Urbanization)", y = "Crime Rate")

p_log_log <- ggplot(crime_lab, aes(x = log(urbanization), y = log(crime_rate))) +
  geom_point(alpha = 0.5, color = "grey50") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = ber640_colors[["primary"]]) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = ber640_colors[["secondary"]], linetype = "dashed") +
  labs(title = "Log-log", x = "log(Urbanization)", y = "log(Crime Rate)")

(p_level_level + p_log_level) / (p_level_log + p_log_log)
```

::: {.pitfall}
Do not compare R2 across models with different outcome scales.

A model with `crime_rate` as the outcome and a model with `log(crime_rate)` as the outcome explain variance on different scales.
Use residual patterns, prediction error on the original scale, and substantive interpretability.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `log_model_slopes` and the four-panel log plot.

Pick one logged specification and interpret its slope in the correct units.
Then write one sentence about which plot makes the linear fit and LOWESS curve look most similar.
Your answer should also explain why R2 should not be the only criterion here.
:::


## Part 4: Centered Polynomial Regression

Polynomial regression adds powers of a predictor.
We center `lrt` first so the linear and squared terms are less redundant and the lower-order terms are easier to interpret.

```{r}
#| label: lab07-fit-polynomial-models

model_linear <- lm(gcse ~ lrt_c, data = gcse_lab)
model_quadratic <- lm(gcse ~ lrt_c + I(lrt_c^2), data = gcse_lab)
model_cubic <- lm(gcse ~ lrt_c + I(lrt_c^2) + I(lrt_c^3), data = gcse_lab)

model_params(model_quadratic) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  round_numeric(digits = 3)
```

### Test Added Polynomial Terms

```{r}
#| label: lab07-polynomial-anova

polynomial_tests <- anova(model_linear, model_quadratic, model_cubic) |>
  as_tibble() |>
  mutate(
    Model = c("Linear", "Quadratic", "Cubic"),
    p.value = format_p(`Pr(>F)`)
  ) |>
  select(Model, Res.Df, RSS, Df, `Sum of Sq`, F, p.value)

polynomial_tests |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab07-polynomial-fit-table

polynomial_fit <- bind_rows(
  model_fit(model_linear) |>
    as_tibble() |>
    mutate(Model = "Linear"),
  model_fit(model_quadratic) |>
    as_tibble() |>
    mutate(Model = "Quadratic"),
  model_fit(model_cubic) |>
    as_tibble() |>
    mutate(Model = "Cubic")
) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC)

polynomial_fit |>
  round_numeric(digits = 3)
```

::: {.interpretation}
Polynomial terms can be statistically significant even when the visual improvement is small.

Use the F-tests, fit statistics, plots, and theory together.
Do not choose a higher-degree polynomial just because R2 increases.
:::

### Plot Polynomial Predictions

```{r}
#| label: lab07-polynomial-prediction-data

lrt_mean <- mean(gcse_lab$lrt, na.rm = TRUE)

polynomial_predictions <- bind_rows(
  ggeffects::predict_response(model_linear, terms = "lrt_c [all]") |>
    as_tibble() |>
    mutate(Model = "Linear"),
  ggeffects::predict_response(model_quadratic, terms = "lrt_c [all]") |>
    as_tibble() |>
    mutate(Model = "Quadratic"),
  ggeffects::predict_response(model_cubic, terms = "lrt_c [all]") |>
    as_tibble() |>
    mutate(Model = "Cubic")
) |>
  mutate(lrt = x + lrt_mean)
```

```{r}
#| label: lab07-polynomial-prediction-plot
#| fig-cap: "Linear, quadratic, and cubic predictions for the GCSE-LRT relationship."
#| fig-alt: "Scatter plot of GCSE score versus LRT score with three fitted prediction curves. The linear, quadratic, and cubic curves are similar, with small differences near the extremes."
#| fig-width: 8
#| fig-height: 5

ggplot(gcse_lab, aes(x = lrt, y = gcse)) +
  geom_point(alpha = 0.04, color = "grey55") +
  geom_line(
    data = polynomial_predictions,
    aes(x = lrt, y = predicted, color = Model),
    linewidth = 1.1
  ) +
  scale_color_manual(values = c(
    "Linear" = ber640_colors[["primary"]],
    "Quadratic" = ber640_colors[["secondary"]],
    "Cubic" = ber640_colors[["success"]]
  )) +
  labs(
    x = "LRT Score",
    y = "Predicted GCSE Score",
    color = "Model"
  )
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `anova(model_linear, model_quadratic, model_cubic)`, `polynomial_fit`, and the prediction plot.

Which model would you choose for the GCSE-LRT relationship?
Write two sentences: one sentence about statistical or fit evidence, and one sentence about whether the added curve is substantively useful.
:::


## Part 5: Piecewise Constant Model

A piecewise constant model cuts a continuous predictor into groups.
This is easy to interpret, but it throws away within-group information.

```{r}
#| label: lab07-create-lrt-groups

gcse_piece <- gcse_lab |>
  mutate(
    lrt_group = cut(
      lrt,
      breaks = quantile(lrt, probs = c(0, .25, .50, .75, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Q1 Low", "Q2", "Q3", "Q4 High")
    )
  ) |>
  filter(!is.na(lrt_group))
```

```{r}
#| label: lab07-piecewise-model

model_piecewise <- lm(gcse ~ lrt_group, data = gcse_piece)
model_params(model_piecewise)
```

```{r}
#| label: lab07-piecewise-means

piecewise_means <- modelbased::estimate_means(
  model_piecewise,
  by = "lrt_group"
) |>
  as_tibble() |>
  select(lrt_group, Mean, SE, CI_low, CI_high)

piecewise_means |>
  round_numeric(digits = 2)
```

```{r}
#| label: lab07-piecewise-means-plot
#| fig-cap: "Predicted GCSE means by LRT quartile."
#| fig-alt: "Point-and-interval plot showing predicted GCSE means for four LRT quartile groups. Mean GCSE increases from the lowest LRT quartile to the highest LRT quartile."
#| fig-width: 7
#| fig-height: 4.5

ggplot(piecewise_means, aes(x = lrt_group, y = Mean)) +
  geom_point(size = 3, color = ber640_colors[["primary"]]) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.12,
                color = ber640_colors[["primary"]]) +
  labs(
    x = "LRT Quartile",
    y = "Predicted GCSE Mean"
  )
```

::: {.pitfall}
Cutting a continuous predictor into groups can make communication easier, but it loses information.

Two students just above and just below a cutoff may receive different predicted means even if their LRT scores are almost identical.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `piecewise_means` and the quartile plot.

Describe how predicted GCSE changes across LRT quartiles.
Then write one benefit and one cost of replacing continuous LRT with quartile groups.
:::


## Part 6: Natural Spline

A natural spline is a smooth flexible curve fit inside an ordinary linear model.
The coefficients are not directly interpretable, so we focus on predictions.

```{r}
#| label: lab07-fit-spline-model

model_spline <- lm(gcse ~ splines::ns(lrt_c, df = 4), data = gcse_lab)
model_params(model_spline)
```

```{r}
#| label: lab07-model-fit-comparison

nonlinear_fit <- bind_rows(
  model_fit(model_linear) |>
    as_tibble() |>
    mutate(Model = "Linear"),
  model_fit(model_quadratic) |>
    as_tibble() |>
    mutate(Model = "Quadratic"),
  model_fit(model_piecewise) |>
    as_tibble() |>
    mutate(Model = "Piecewise constant"),
  model_fit(model_spline) |>
    as_tibble() |>
    mutate(Model = "Natural spline")
) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC)

nonlinear_fit |>
  round_numeric(digits = 3)
```

### Compare Smooth Prediction Curves

```{r}
#| label: lab07-spline-comparison-data

smooth_predictions <- bind_rows(
  polynomial_predictions |>
    filter(Model %in% c("Linear", "Quadratic")) |>
    select(lrt, predicted, conf.low, conf.high, Model),
  ggeffects::predict_response(model_spline, terms = "lrt_c [all]") |>
    as_tibble() |>
    mutate(
      lrt = x + lrt_mean,
      Model = "Natural spline"
    ) |>
    select(lrt, predicted, conf.low, conf.high, Model)
)
```

```{r}
#| label: lab07-spline-comparison-plot
#| fig-cap: "Linear, quadratic, and natural spline fits compared with LOWESS."
#| fig-alt: "Scatter plot of GCSE score versus LRT score with linear, quadratic, and natural spline prediction curves. All three fitted curves nearly overlap through the middle of the LRT range; the quadratic line rises slightly above the others at the high end, while the natural spline remains close to the linear curve."
#| fig-width: 9
#| fig-height: 5

ggplot(gcse_lab, aes(x = lrt, y = gcse)) +
  geom_point(alpha = 0.035, color = "grey55") +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = "grey30", linetype = "dashed", linewidth = 0.9) +
  geom_line(
    data = smooth_predictions,
    aes(x = lrt, y = predicted, color = Model),
    linewidth = 1.1
  ) +
  scale_color_manual(values = c(
    "Linear" = ber640_colors[["primary"]],
    "Quadratic" = ber640_colors[["secondary"]],
    "Natural spline" = ber640_colors[["success"]]
  )) +
  labs(
    x = "LRT Score",
    y = "Predicted GCSE Score",
    color = "Model"
  )
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `nonlinear_fit` and the spline comparison plot.

Does the natural spline improve the substantive story enough to justify its extra flexibility?
Write two sentences comparing the linear, quadratic, and spline fits.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Choose one final model strategy for this lab: linear, quadratic, piecewise constant, or natural spline.

Use your comparison from the previous prompt as evidence.
Write a short 3--4 sentence recommendation.
Your recommendation should mention visual fit, fit statistics, interpretability, and model complexity.
:::

::: {.exercises}
## What to Submit

Submit the following items:

1. The GCSE-LRT plot with the linear fit and LOWESS curve, plus two sentences diagnosing nonlinearity. [core]{.chip .med}

2. The `log_model_slopes` table, plus one logged coefficient interpretation. [core]{.chip .med}

3. One sentence using the four-panel log plot to choose the most linear-looking crime specification. [core]{.chip .med}

4. The polynomial nested F-test output and `polynomial_fit` table, plus two sentences choosing a polynomial degree. [core]{.chip .hard}

5. The `piecewise_means` table or quartile plot, plus one benefit and one cost of grouping LRT. [core]{.chip .med}

6. The `nonlinear_fit` table and spline comparison plot, plus two sentences comparing the linear, quadratic, and natural spline models. [core]{.chip .hard}

7. The 3--4 sentence recommendation from the final Your turn block. [core]{.chip .hard}
:::

::: {.summary}
## Lab Summary

1. LOWESS is a visual diagnostic for nonlinearity.

2. Log transformations change coefficient interpretation and should be guided by theory and residual patterns.

3. Center predictors before adding polynomial terms.

4. Use nested F-tests and fit statistics to compare linear, quadratic, and cubic models.

5. Piecewise constant models are easy to explain but discard within-group variation.

6. Natural splines are flexible, so interpret them through prediction plots rather than individual coefficients.

**Next:** Lab 8 checks regression diagnostics and model-building decisions.
:::
