---
title: "Lab 5: Multiple Regression with Continuous Predictors"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Fit a multiple regression model with two continuous predictors
- Compare bivariate and adjusted coefficients
- Interpret coefficients with "holding constant" language
- Use `model_fit()` to compare fit across models
- Create a side-by-side model table with `modelsummary()`
- Visualize a partial relationship with residuals
- Compute standardized coefficients with `standardize_parameters()`
- Check collinearity with `performance::check_collinearity()`
- Compute partial correlations with `correlation::correlation(partial = TRUE)`
- **Lecture Notes Reference:** Chapter 4: *Continuous Predictors*

::: {.meta}
Dataset: `crime`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Prepare the Crime Data

In Chapter 4, we use Florida county data to study how education and urbanization relate to crime rates.

The motivating question is:

> Does the relationship between education and crime change after accounting for urbanization?

### Step 1: Load and Rename Variables

```{r}
#| label: lab05-load-crime-data

data(crime)

crime_df <- crime |>
  rename(
    crime_rate = c,
    income = i,
    high_school = hs,
    urbanization = u
  )
```

The original dataset uses short variable names. The renamed variables make the analysis code easier to read.

### Step 2: Preflight Check

```{r}
#| label: lab05-preflight

crime_df |>
  summarise(
    counties = n(),
    missing_crime = sum(is.na(crime_rate)),
    missing_education = sum(is.na(high_school)),
    missing_urbanization = sum(is.na(urbanization)),
    missing_income = sum(is.na(income))
  )
```

### Step 3: Descriptive Summary

```{r}
#| label: lab05-descriptive-summary

crime_df |>
  summarise(
    mean_crime = mean(crime_rate),
    mean_high_school = mean(high_school),
    mean_urbanization = mean(urbanization),
    mean_income = mean(income)
  ) |>
  round_numeric(digits = 2)
```

### Step 4: Correlation Preview

```{r}
#| label: lab05-correlation-preview

crime_df |>
  select(crime_rate, high_school, urbanization, income) |>
  correlation::correlation()
```

::: {.interpretation}
The zero-order correlation between `high_school` and `crime_rate` is positive.

That looks counterintuitive: counties with more high-school graduates seem to have higher crime.
The rest of this lab shows why the bivariate relationship can be misleading.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the correlation table to answer two questions:

1. Which predictor has the strongest zero-order correlation with `crime_rate`?
2. Why might the positive correlation between `high_school` and `crime_rate` be misleading?

Write two short sentences.
:::


## Part 2: Bivariate vs. Multiple Regression

### Step 1: Fit the Education-Only Model

```{r}
#| label: lab05-fit-bivariate

model_education <- lm(crime_rate ~ high_school, data = crime_df)
model_params(model_education)
```

This bivariate model ignores urbanization.

### Step 2: Fit the Adjusted Model

```{r}
#| label: lab05-fit-adjusted

model_adjusted <- lm(crime_rate ~ high_school + urbanization, data = crime_df)
model_params(model_adjusted)
```

::: {.interpretation}
The `high_school` coefficient now compares counties with the same urbanization level.

That is the multiple-regression meaning of "holding urbanization constant."
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Interpret the `high_school` coefficient from `model_adjusted` in one sentence.
Your sentence should use the phrase "holding urbanization constant" and should avoid causal language.
:::

### Step 3: Compare the Education Coefficient

```{r}
#| label: lab05-education-coefficient-comparison

education_comparison <- bind_rows(
  tidy_model(model_education) |>
    mutate(model = "Education only"),
  tidy_model(model_adjusted) |>
    mutate(model = "Education + urbanization")
) |>
  filter(term == "high_school") |>
  select(model, estimate, std.error, conf.low, conf.high, p.value)

education_comparison |>
  round_numeric(digits = 3)
```

::: {.interpretation}
The education coefficient changes from positive to negative after adding urbanization.

This is a sign reversal: the bivariate association and adjusted association point in different directions.
:::

::: {.pitfall}
A sign reversal does not mean the first model was "wrong."

It means the bivariate model and the adjusted model answer different questions.
The bivariate coefficient describes the raw education-crime association.
The adjusted coefficient describes the education-crime association among counties with the same urbanization level.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Use `education_comparison` to identify the direction of the `high_school` coefficient in each model.
Then write one sentence explaining what changed after urbanization entered the model.
:::


## Part 3: Compare Model Fit

```{r}
#| label: lab05-model-fit-comparison

fit_comparison <- bind_rows(
  model_fit(model_education) |>
    as_tibble() |>
    mutate(Model = "Education only"),
  model_fit(model_adjusted) |>
    as_tibble() |>
    mutate(Model = "Education + urbanization")
) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC)

fit_comparison |>
  round_numeric(digits = 3)
```

Adding urbanization substantially increases $R^2$ and lowers RMSE.
That tells us urbanization adds important explanatory information.

`fit_comparison` compares model fit.
`modelsummary()` gives a polished side-by-side coefficient table for the same two models.

```{r}
#| label: lab05-modelsummary-comparison

comparison_models <- list(
  "Education only" = model_education,
  "Education + urbanization" = model_adjusted
)

modelsummary::modelsummary(comparison_models, output = "markdown", fmt = 3)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Which model fits better: `model_education` or `model_adjusted`?
Name two fit statistics from `fit_comparison` that support your answer.
:::


## Part 4: Visualize the Sign Reversal

The bivariate plot shows the raw relationship between education and crime.
The partial plot shows the relationship between education and crime after removing urbanization from both variables.

```{r}
#| label: lab05-sign-reversal-plot
#| fig-cap: "Bivariate and partial relationships between education and crime."
#| fig-alt: "Two-panel plot. The left panel shows a positive bivariate relationship between high school graduation percentage and crime rate. The right panel shows a partial residual plot after controlling for urbanization, where the relationship between residualized high school percentage and residualized crime rate is negative."
#| fig-width: 10
#| fig-height: 4.8

partial_education <- tibble(
  education_resid = residuals(lm(high_school ~ urbanization, data = crime_df)),
  crime_resid = residuals(lm(crime_rate ~ urbanization, data = crime_df))
)

p_bivariate <- ggplot(crime_df, aes(x = high_school, y = crime_rate)) +
  geom_point(alpha = 0.65, color = ber640_colors[["primary"]]) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = ber640_colors[["secondary"]]) +
  labs(
    title = "Bivariate relationship",
    x = "High School Graduates (%)",
    y = "Crime Rate"
  )

p_partial <- ggplot(partial_education,
                    aes(x = education_resid, y = crime_resid)) +
  geom_point(alpha = 0.65, color = ber640_colors[["primary"]]) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = ber640_colors[["success"]]) +
  labs(
    title = "Partial relationship",
    x = "High School % | Urbanization",
    y = "Crime Rate | Urbanization"
  )

p_bivariate + p_partial
```

### Check the Partial Slope

```{r}
#| label: lab05-partial-slope-check

partial_slope <- lm(crime_resid ~ education_resid, data = partial_education)

tibble(
  Quantity = c(
    "Adjusted coefficient for high_school",
    "Slope in partial plot",
    "Partial correlation"
  ),
  Value = c(
    coef(model_adjusted)["high_school"],
    coef(partial_slope)["education_resid"],
    cor(partial_education$education_resid, partial_education$crime_resid)
  )
) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
The partial-plot slope matches the `high_school` coefficient in the multiple regression model.

That is the geometry behind "holding urbanization constant."
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

In the partial plot, what does `High School % | Urbanization` mean?
Write one sentence connecting the partial plot slope to the `high_school` coefficient in `model_adjusted`.
:::


## Part 5: Standardized Coefficients

Raw coefficients are in original units. Standardized coefficients put the outcome and predictors on a standard-deviation scale.

```{r}
#| label: lab05-standardized-coefficients

standardized_coefficients <- parameters::standardize_parameters(model_adjusted) |>
  as_tibble() |>
  filter(Parameter != "(Intercept)") |>
  select(Parameter, Std_Coefficient, CI_low, CI_high)

standardized_coefficients |>
  round_numeric(digits = 3)
```

::: {.interpretation}
Use unstandardized coefficients when you want practical units.

Use standardized coefficients when you want to compare predictors measured on different scales.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Use `standardized_coefficients` to decide which predictor has the stronger adjusted association with crime rate.
Then write one sentence explaining why standardized coefficients are useful here.
:::


## Part 6: Check Collinearity

Education and urbanization are correlated, so we should check whether the model can still separate their contributions.

```{r}
#| label: lab05-check-collinearity

performance::check_collinearity(model_adjusted, ci = FALSE) |>
  as_tibble() |>
  select(Term, VIF, SE_factor, Tolerance) |>
  round_numeric(digits = 3)
```

::: {.assumption-check}
VIF values near 1 mean little overlap among predictors.
Larger VIF values mean more overlap and wider standard errors.

As a rough rule of thumb, VIF above 5 deserves attention and VIF above 10 is serious.
:::


## Part 7: Partial Correlations

Partial correlations summarize adjusted relationships in correlation form.

```{r}
#| label: lab05-partial-correlations

crime_df |>
  select(crime_rate, high_school, urbanization) |>
  correlation::correlation(partial = TRUE)
```

::: {.interpretation}
The partial correlation between crime and education controls for urbanization.

It should match the direction of the adjusted `high_school` coefficient.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Write one diagnostic note using both the VIF table and the partial correlation table.
Your note should say whether collinearity looks acceptable and whether the partial correlation matches the adjusted coefficient direction.
:::


## Part 8: Optional Challenge: Add Income

The dataset also includes county income.
Use this extension only after you understand the two-predictor model.

```{r}
#| label: lab05-practice-income-model

model_income <- lm(
  crime_rate ~ high_school + urbanization + income,
  data = crime_df
)

tidy_model(model_income) |>
  select(term, estimate, std.error, conf.low, conf.high, p.value) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab05-practice-income-fit

income_fit_comparison <- bind_rows(
  model_fit(model_adjusted) |>
    as_tibble() |>
    mutate(Model = "Education + urbanization"),
  model_fit(model_income) |>
    as_tibble() |>
    mutate(Model = "Education + urbanization + income")
) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC)

income_fit_comparison |>
  round_numeric(digits = 3)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Does adding `income` meaningfully change the `high_school` coefficient or improve model fit?
Write two sentences: one about the coefficient table and one about `income_fit_comparison`.
:::


::: {.exercises}
## What to Submit

1. The correlation table for `crime_rate`, `high_school`, `urbanization`, and `income`. [core]{.chip .easy}
2. The coefficient table for `model_adjusted`. [core]{.chip .easy}
3. One sentence interpreting the `high_school` coefficient using "holding urbanization constant." [core]{.chip .med}
4. The `education_comparison` table and one sentence explaining the sign reversal. [core]{.chip .med}
5. The `fit_comparison` table and one sentence naming the better-fitting model. [core]{.chip .med}
6. The bivariate/partial plot and one sentence connecting the partial slope to the adjusted coefficient. [core]{.chip .hard}
7. The standardized coefficient table and one sentence comparing predictor strength. [core]{.chip .med}
8. One diagnostic note using `performance::check_collinearity()` and `correlation::correlation(partial = TRUE)`. [core]{.chip .hard}
:::

::: {.callout-tip}
## Optional Challenge

Submit the income model and fit comparison, with a two-sentence interpretation. [challenge]{.chip .hard}
:::


::: {.summary}
## Lab Summary

1. Multiple regression lets us estimate the association between one predictor and the outcome while holding other predictors constant.

2. In the crime data, the education coefficient changes sign after controlling for urbanization.

3. `model_fit()` helps compare model fit across nested or competing models.

4. `modelsummary()` is useful when you want a polished side-by-side regression table.

5. Partial plots visualize the residual-vs-residual relationship behind a multiple-regression coefficient.

6. `standardize_parameters()` puts coefficients on a standard-deviation scale.

7. `performance::check_collinearity()` checks whether predictors overlap enough to inflate standard errors.

8. `correlation::correlation(partial = TRUE)` reports adjusted relationships as partial correlations.

9. A third predictor should be added only when it helps answer the research question or improves the model enough to justify the extra complexity.

**Next:** Lab 6 extends multiple regression to interaction terms.
:::
