---
title: "Lab 8: Model Building and Diagnostics"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Build regression models in a confirmatory sequence
- Compare nested models with `anova()`
- Track change in R2 and adjusted R2
- Compare candidate models with `compare_models()`
- Extract diagnostic quantities with `broom::augment()`
- Use `performance::check_model()` as a diagnostic dashboard
- Check collinearity and heteroscedasticity with `performance`
- Compare OLS and HC3 robust standard errors with `parameters`
- **Lecture Notes Reference:** Chapter 7: *Model Building* and Chapter 8: *Model Diagnostics*

::: {.meta}
Dataset: `hsb_sub`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Research Question and Data

In this lab, we practice a complete regression workflow:

1. state a research question,
2. build models in a planned sequence,
3. compare models,
4. choose a working final model,
5. check diagnostics,
6. decide whether a remedy is needed.

**Research question:** Does socioeconomic status predict mathematics achievement differently in public and Catholic schools, after accounting for minority status and gender?

```{r}
#| label: lab08-load-data

data(hsb_sub)

hsb_lab <- hsb_sub |>
  mutate(
    sector_f = factor(sector, levels = c(0, 1), labels = c("Public", "Catholic")),
    minority_f = factor(minority, levels = c(0, 1), labels = c("Non-minority", "Minority")),
    female_f = factor(female, levels = c(0, 1), labels = c("Male", "Female")),
    size_100 = size / 100
  ) |>
  select(mathach, ses, sector_f, minority_f, female_f, size_100, schoolid)

model_data <- hsb_lab |>
  drop_na(mathach, ses, sector_f, minority_f, female_f, size_100)

```

::: {.callout-note}
## Teaching Simplification

The data include students nested in schools, but this lab treats the observations as a single-level regression example so we can focus on model-building and diagnostic tools.
`schoolid` is kept only as context.

This simplification matters: OLS residual diagnostics and HC3 robust standard errors do not solve school-level dependence.
Clustered or multilevel models are a later extension.
:::

```{r}
#| label: lab08-preflight

tibble(
  Rows = nrow(model_data),
  Outcome = "mathach",
  Focal_predictor = "ses",
  Planned_covariates = "sector_f, minority_f, female_f",
  Planned_refinement = "ses x sector_f interaction",
  Exploratory_candidate = "size_100"
)
```

::: {.pitfall}
For nested model comparisons, all models must use the same outcome and the same rows.

That is why we create `model_data` first and fit every model to that same complete-case data frame.
:::


## Part 2: Explore Before Modeling

Before fitting models, look at the outcome and the focal predictor.

```{r}
#| label: lab08-outcome-summary

model_data |>
  summarise(
    mean_mathach = mean(mathach),
    sd_mathach = sd(mathach),
    min_mathach = min(mathach),
    max_mathach = max(mathach),
    mean_ses = mean(ses),
    sd_ses = sd(ses)
  ) |>
  round_numeric(digits = 2)
```

```{r}
#| label: lab08-outcome-histogram
#| fig-cap: "Distribution of mathematics achievement scores."
#| fig-alt: "Histogram of mathematics achievement scores. The distribution is roughly unimodal with scores spread across the observed achievement scale."
#| fig-width: 7
#| fig-height: 4.5

ggplot(model_data, aes(x = mathach)) +
  geom_histogram(bins = 18, fill = ber640_colors[["primary"]], color = "white") +
  labs(
    x = "Mathematics Achievement",
    y = "Count"
  )
```

```{r}
#| label: lab08-ses-lowess
#| fig-cap: "Mathematics achievement by socioeconomic status with linear and LOWESS fits."
#| fig-alt: "Scatter plot of mathematics achievement versus socioeconomic status with a straight linear fit and a dashed LOWESS curve. The curves are broadly similar, with modest local departures."
#| fig-width: 8
#| fig-height: 5

ggplot(model_data, aes(x = ses, y = mathach)) +
  geom_point(alpha = 0.45, color = "grey55") +
  geom_smooth(
    method = "lm", formula = y ~ x, se = FALSE,
    color = ber640_colors[["primary"]], linewidth = 1.1
  ) +
  geom_smooth(
    method = "loess", formula = y ~ x, se = FALSE,
    color = ber640_colors[["secondary"]], linewidth = 1.1,
    linetype = "dashed"
  ) +
  labs(
    x = "Socioeconomic Status",
    y = "Mathematics Achievement"
  )
```

::: {.interpretation}
The linear fit gives the model's default straight-line summary.

The LOWESS curve is a flexible visual check.
If the LOWESS curve bends sharply away from the straight line, the model-building sequence should consider a nonlinear term before moving on.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the SES LOWESS plot.
Does the straight-line model look reasonable as a starting point?
Write two sentences: one sentence about the visual evidence, and one sentence about what model refinement you would consider if the curve looked strongly nonlinear.
:::


## Part 3: Build Models in a Planned Sequence

The confirmatory approach starts from the research question.
We add predictors because they have a role in the analysis, not because a search algorithm found a small p-value.

For this lab, we use three main confirmatory models:

- **M1: research model**: SES and school sector.
- **M2: adjusted model**: add minority status and gender as possible confounders.
- **M3: interaction model**: test whether the SES slope differs by school sector.

```{r}
#| label: lab08-fit-models

model_interest <- lm(mathach ~ ses + sector_f, data = model_data)
model_confounders <- lm(mathach ~ ses + sector_f + minority_f + female_f,
                        data = model_data)
model_interaction <- lm(mathach ~ ses * sector_f + minority_f + female_f,
                        data = model_data)
```

```{r}
#| label: lab08-model-parameters

model_params(model_confounders) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
`model_confounders` is the main confirmatory model.

It estimates the SES and sector associations after adjusting for minority status and gender.
We keep those adjustment variables because of the analysis plan, not because of a stepwise search.
:::


## Part 4: Compare Nested Models

Use `anova()` when one model is nested inside another.
Here, M1 is nested inside M2, and M2 is nested inside M3.

```{r}
#| label: lab08-format-anova

tidy_anova <- function(...) {
  anova(...) |>
    as_tibble() |>
    mutate(p.value = format_p(`Pr(>F)`)) |>
    select(Res.Df, RSS, Df, `Sum of Sq`, F, p.value) |>
    round_numeric(digits = 3)
}
```

```{r}
#| label: lab08-nested-confounders

tidy_anova(model_interest, model_confounders)
```

```{r}
#| label: lab08-nested-interaction

tidy_anova(model_confounders, model_interaction)
```

::: {.interpretation}
The first nested comparison asks whether the adjustment variables jointly improve the model.

The second comparison asks whether the SES slope differs by school sector.
If the interaction does not clearly improve the model, the simpler adjusted model is easier to explain.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `lab08-nested-confounders` and `lab08-nested-interaction`.
Which planned addition has stronger evidence: the confounder block or the SES-by-sector interaction?
Write two sentences using the nested F-test results.
:::


## Part 5: Track Fit Statistics

Model comparison is not only a p-value exercise.
Track R2, adjusted R2, AIC, BIC, and RMSE together.

```{r}
#| label: lab08-delta-r2

r2_interest <- performance::r2(model_interest)
r2_confounders <- performance::r2(model_confounders)
r2_interaction <- performance::r2(model_interaction)

fit_steps <- tibble(
  Model = c("M1: SES + sector", "M2: + confounders", "M3: + interaction"),
  R2 = c(r2_interest$R2, r2_confounders$R2, r2_interaction$R2),
  R2_adjusted = c(
    r2_interest$R2_adjusted,
    r2_confounders$R2_adjusted,
    r2_interaction$R2_adjusted
  )
) |>
  mutate(Delta_R2 = R2 - lag(R2))

fit_steps |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab08-compare-performance

performance_table <- compare_models(
  model_interest,
  model_confounders,
  model_interaction,
  rank = FALSE
) |>
  as_tibble() |>
  mutate(
    Model = c("M1: SES + sector", "M2: + confounders", "M3: + interaction"),
    .before = 1
  ) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC, AIC_wt, BIC_wt)

performance_table |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab08-criteria-plot
#| fig-cap: "Fit criteria across the model-building sequence."
#| fig-alt: "Four-panel line plot showing R-squared, adjusted R-squared, AIC, and BIC across three candidate models. R-squared increases with each added predictor, while adjusted R-squared and information criteria show the tradeoff between fit and complexity."
#| fig-width: 9
#| fig-height: 6

performance_table |>
  select(Model, R2, R2_adjusted, AIC, BIC) |>
  pivot_longer(-Model, names_to = "Criterion", values_to = "Value") |>
  mutate(
    Model = factor(Model, levels = c(
      "M1: SES + sector",
      "M2: + confounders",
      "M3: + interaction"
    )),
    Criterion = factor(
      Criterion,
      levels = c("R2", "R2_adjusted", "AIC", "BIC"),
      labels = c("R2", "Adjusted R2", "AIC", "BIC")
    )
  ) |>
  ggplot(aes(x = Model, y = Value, group = 1)) +
  geom_point(size = 3, color = ber640_colors[["primary"]]) +
  geom_line(color = ber640_colors[["primary"]], linewidth = 0.9) +
  facet_wrap(~ Criterion, scales = "free_y") +
  labs(x = NULL, y = "Value") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

::: {.pitfall}
Raw R2 cannot choose a model.

R2 almost always increases when you add predictors.
Adjusted R2, AIC, and BIC include a penalty for complexity, so they are more useful for model comparison.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `performance_table` and the criteria plot.
Which model would you report as the main model?
Write three short sentences: one about theory, one about fit statistics, and one about parsimony.
:::


## Part 6: Choose a Working Final Model

For this lab, we choose `model_confounders` as the working final model.

The reason is substantive: SES and sector answer the research question, minority status and gender are planned adjustment variables, and the interaction does not clearly improve the explanation-focused model enough to justify the extra complexity.

```{r}
#| label: lab08-final-model-table

final_model <- model_confounders

model_params(final_model) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
Choosing a model is a judgment, not a single command.

For a confirmatory analysis, the final model should follow the research question and analysis plan.
Fit statistics help you evaluate the tradeoff, but they do not replace the design logic.
:::

### Exploratory Sensitivity Check

School size is a plausible extra predictor, but it was not central to the research question.
If we inspect it, we label that move as exploratory.

```{r}
#| label: lab08-exploratory-size

model_size_exploratory <- lm(
  mathach ~ ses + sector_f + minority_f + female_f + size_100,
  data = model_data
)

compare_models(
  final_model,
  model_size_exploratory,
  rank = FALSE
) |>
  as_tibble() |>
  mutate(
    Model = c("Confirmatory final", "Exploratory + school size"),
    .before = 1
  ) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC) |>
  round_numeric(digits = 3)
```

::: {.pitfall}
Exploratory model checks are useful, but they change the strength of the claim.

If a predictor is added after looking at the data, describe the result as exploratory or hypothesis-generating.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Compare the confirmatory final model with the exploratory school-size model.
Does the exploratory addition change your main conclusion?
Write one sentence about fit and one sentence about how you would label the claim.
:::


## Part 7: Extract Diagnostics with `broom::augment()`

The `broom::augment()` function turns a fitted model into a row-level diagnostic data frame.
This is the easiest way to make custom diagnostic plots.

```{r}
#| label: lab08-augment-diagnostics

diagnostics_df <- broom::augment(final_model, data = model_data) |>
  mutate(
    obs = row_number(),
    stud_deleted = rstudent(final_model)
  )

diagnostics_df |>
  select(obs, mathach, .fitted, .resid, .std.resid, stud_deleted, .hat, .cooksd) |>
  slice_head(n = 8) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab08-residuals-fitted
#| fig-cap: "Studentized residuals versus fitted values for the final model."
#| fig-alt: "Scatter plot of studentized residuals versus fitted values with a horizontal zero line and a smooth curve. The smooth curve stays close to zero with a small dip and rise, residual spread is fairly similar across fitted values, and most points fall between about minus two and plus two."
#| fig-width: 8
#| fig-height: 5

ggplot(diagnostics_df, aes(x = .fitted, y = .std.resid)) +
  geom_point(alpha = 0.45, color = "grey50") +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = ber640_colors[["secondary"]]) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = ber640_colors[["primary"]], linewidth = 1.1) +
  labs(
    x = "Fitted Values",
    y = "Studentized Residuals"
  )
```

```{r}
#| label: lab08-influence-cases

cook_cutoff <- 4 / nrow(diagnostics_df)
hat_cutoff <- 2 * length(coef(final_model)) / nrow(diagnostics_df)

diagnostics_df |>
  mutate(
    high_residual = abs(stud_deleted) > 3,
    high_leverage = .hat > hat_cutoff,
    high_cook = .cooksd > cook_cutoff
  ) |>
  arrange(desc(.cooksd)) |>
  select(
    obs, mathach, ses, sector_f, minority_f, female_f,
    .fitted, stud_deleted, .hat, .cooksd,
    high_residual, high_leverage, high_cook
  ) |>
  slice_head(n = 8) |>
  round_numeric(digits = 3)
```

The Cook's D flag is a screening rule.
A `TRUE` value means "investigate this case," not "delete this case."

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the residuals-vs-fitted plot and `lab08-influence-cases`.
Do you see evidence of curvature, changing spread, or potentially influential cases?
Write three bullets: one for residual pattern, one for influence, and one for whether you would investigate any cases.
:::


## Part 8: Use the `performance::check_model()` Dashboard

`performance::check_model()` is a one-command diagnostic dashboard.
Use it after you understand what each panel is checking.

```{r}
#| label: lab08-check-model-dashboard
#| fig-cap: "Diagnostic dashboard for the final model."
#| fig-alt: "Six-panel diagnostic dashboard. The predicted density resembles the observed outcome density, the residual linearity curve is roughly flat, the variance curve rises then levels off, influential observations stay inside Cook's-distance contours, VIF values remain in the low collinearity region, and the normality panel shows mild tail departures."
#| fig-width: 10
#| fig-height: 10

performance::check_model(final_model)
```

::: {.interpretation}
Read the dashboard in layers.

First look for residual patterns that suggest misspecification.
Then check whether any observation has unusual influence.
Finally, check whether collinearity is inflating standard errors.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the `performance::check_model()` dashboard.
Choose two panels and write one sentence for each panel explaining whether it raises a concern.
Focus on the pattern, not on whether the plot looks perfectly clean.
:::


## Part 9: Focused Diagnostic Checks

After the dashboard, zoom in on specific questions with focused functions.

```{r}
#| label: lab08-check-collinearity

performance::check_collinearity(final_model) |>
  as_tibble() |>
  select(Term, VIF, SE_factor, Tolerance) |>
  round_numeric(digits = 2)
```

```{r}
#| label: lab08-check-heteroscedasticity

performance::check_heteroscedasticity(final_model)
```

::: {.interpretation}
`performance::check_collinearity()` summarizes VIF and tolerance.

`performance::check_heteroscedasticity()` gives a formal test for non-constant error variance.
The plot remains the primary diagnostic, but the test is useful supporting evidence.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `performance::check_collinearity()` and `performance::check_heteroscedasticity()`.
Write one sentence about VIF/collinearity and one sentence about whether robust standard errors are needed.
:::


## Part 10: Diagnostic Decision Checklist

Use this table when writing a short diagnostic memo.

| Diagnostic question | Evidence to use | If there is a concern |
|---|---|---|
| Is the mean function reasonable? | SES LOWESS plot, residuals-vs-fitted plot, `performance::check_model()` linearity panel | Add a nonlinear term or transformation, then re-check diagnostics |
| Is residual spread roughly constant? | Residuals-vs-fitted plot, `performance::check_heteroscedasticity()` | Use HC3 robust standard errors; consider transforming the outcome if the pattern is severe |
| Are residuals normal enough? | `performance::check_model()` normality panel | In large samples, mild departures are usually tolerable; severe skew may suggest transformation |
| Are any observations influential? | Cook's distance, leverage, studentized deleted residuals | Investigate data quality and report sensitivity; do not delete automatically |
| Is collinearity inflating uncertainty? | `performance::check_collinearity()` | Combine, remove, or rethink redundant predictors if VIF is high |
| Are observations independent enough for this model? | Study design, clustering variables such as `schoolid`, residual patterns by group | Treat the single-level model as a simplification; consider clustered or multilevel models in a fuller analysis |

::: {.pitfall}
Diagnostics are not a pass/fail exam.

The question is whether any problem is large enough to change the substantive conclusion.
Minor imperfections are normal in real data.
:::


## Part 11: Compare OLS and Robust Standard Errors

Robust standard errors change uncertainty estimates, not coefficient estimates.
HC3 is a good default for small-to-moderate samples.

```{r}
#| label: lab08-robust-parameters

ols_parameters <- model_params(final_model) |>
  as_tibble() |>
  mutate(SE_type = "OLS")

hc3_parameters <- model_params(final_model, vcov = "HC3") |>
  as_tibble() |>
  mutate(SE_type = "HC3 robust")

bind_rows(ols_parameters, hc3_parameters) |>
  mutate(p = format_p(p)) |>
  select(SE_type, Parameter, Coefficient, SE, CI_low, CI_high, p) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
If robust standard errors are close to the OLS standard errors, the usual inference is probably not very sensitive to heteroscedasticity.

If robust standard errors are much larger, report the robust results and explain why.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the OLS and HC3 robust tables.
Pick the SES coefficient and compare its standard error and p-value across the two rows.
Write one sentence saying whether the inference is sensitive to robust standard errors.
:::

::: {.exercises}
## What to Submit

Submit the following core items:

1. The SES LOWESS plot plus two sentences about whether a linear starting model is reasonable. [core]{.chip .med}

2. The two nested-model comparison tables plus two sentences choosing between the confounder block and the interaction refinement. [core]{.chip .med}

3. The `compare_models()` table or criteria plot plus a three-sentence model choice justification. [core]{.chip .med}

4. The residuals-vs-fitted plot and influence table plus three diagnostic bullets. [core]{.chip .hard}

5. Two `performance::check_model()` panel interpretations and the focused collinearity/heteroscedasticity results. [core]{.chip .hard}

Optional challenges:

6. The exploratory school-size comparison plus one sentence labeling the result as confirmatory or exploratory. [challenge]{.chip .med}

7. The OLS vs HC3 robust SE table plus a short diagnostic memo recommending whether to report ordinary or robust standard errors. [challenge]{.chip .hard}
:::

::: {.summary}
## Lab Summary

1. Start model building from a research question, not a search algorithm.

2. Use `anova()` for nested model comparisons.

3. Use `compare_models()` to compare fit, complexity, and prediction-error summaries.

4. Choose a working final model by combining theory, fit statistics, and parsimony.

5. Use `broom::augment()` to create row-level diagnostic data.

6. Use `performance::check_model()` as a dashboard, then follow up with focused checks.

7. Robust standard errors change the uncertainty, not the coefficients.

**Next:** Lab 9 introduces simple logistic regression.
:::
