---
title: "Lab 9: Introduction to Logistic Regression"
---

{{< include ../_common.qmd >}}

## Goals

::: {.callout-note}
## What You'll Learn in This Lab
- Understand why we need logistic regression for binary outcomes
- Compute probabilities, odds, and odds ratios from a cross-tabulation
- Fit a logistic regression using `glm()` with `family = binomial`
- Interpret coefficients on the log-odds, odds, and probability scales
- Fit logistic regression with a continuous predictor
- **Lecture Notes Reference:** Chapter 9: *Introduction to Logistic Regression*
:::

::: {.callout-tip}
## Stata → R Translation
| Stata | R | Notes |
|:------|:--|:------|
| `logit survived female` | `glm(survived ~ female, family = binomial)` | Log-odds coefficients |
| `logit survived female, or` | `exp(coef(model))` | Odds ratios |
| `logistic survived female` | `tbl_regression(exponentiate = TRUE)` | OR with pretty table |
| `cs survived female, or` | Manual calculation | Cohort study odds ratio |
:::

**Dataset:** `titanic` from the `regdatasets` package

## Setup

```{r}
#| label: lab09-setup

library(regdatasets)
data(titanic)
glimpse(titanic)
```


## Part 1: Meet the Titanic Data

The `titanic` dataset describes the survival status of individual passengers on the Titanic. Our goal today is to model **who survived** --- a binary outcome (0 = died, 1 = survived).

### Step 1: Explore the Variables

```{r}
#| label: explore-titanic

titanic |>
  summarise(
    n = n(),
    survived_mean = mean(survived, na.rm = TRUE),
    age_mean = mean(age, na.rm = TRUE),
    fare_mean = mean(fare, na.rm = TRUE)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

> The variable `survived` is coded 0/1, so its mean gives us the overall survival rate. What proportion of passengers survived?

### Step 2: Create the Female Dummy Variable

The `sex` variable is already coded as 0 (male) and 1 (female). We'll rename it to `female` for clarity.

```{r}
#| label: create-female

# The 'sex' variable is already coded as 0 = male, 1 = female
# Rename it to 'female' for clarity
titanic <- titanic |>
  rename(female = sex)
```

Let's verify the coding:

```{r}
#| label: verify-female

titanic |>
  count(female)
```

> We created `female` so that 1 = female and 0 = male. This makes males the reference group, which will be important when we interpret logistic regression coefficients.

### Step 3: Cross-Tabulation of Survival by Gender

Before fitting any model, let's look at the raw data. A cross-tabulation tells us everything we need.

```{r}
#| label: crosstab-survive-female

titanic |>
  count(female, survived) |>
  pivot_wider(names_from = survived, values_from = n, names_prefix = "survived_")
```

Now let's add row percentages to see survival rates within each gender:

```{r}
#| label: crosstab-proportions

titanic |>
  group_by(female) |>
  summarise(
    n = n(),
    n_survived = sum(survived),
    prop_survived = mean(survived),
    .groups = "drop"
  ) |>
  mutate(prop_survived = round(prop_survived, 4))
```

> Compare the survival rates. Females had a much higher survival rate than males. This is the pattern we will model with logistic regression.


## Part 2: Understanding Odds and Odds Ratios

Before jumping into logistic regression, let's build intuition about **probabilities**, **odds**, and **odds ratios**. These three quantities are all different ways of describing the same information, and converting between them is the key skill for this lab.

### Step 1: Probabilities from the Cross-Tab

Let's extract the survival probabilities we just computed:

```{r}
#| label: probs-from-table

# Store the survival proportions
surv_stats <- titanic |>
  group_by(female) |>
  summarise(prob_survived = mean(survived), .groups = "drop")

prob_male   <- surv_stats$prob_survived[surv_stats$female == 0]
prob_female <- surv_stats$prob_survived[surv_stats$female == 1]

cat("P(survived | male)   =", round(prob_male, 4), "\n")
cat("P(survived | female) =", round(prob_female, 4), "\n")
```

### Step 2: From Probabilities to Odds

**Odds** express the ratio of the probability of an event happening to the probability of it not happening:

$$\text{Odds} = \frac{p}{1 - p}$$

```{r}
#| label: compute-odds

odds_male   <- prob_male / (1 - prob_male)
odds_female <- prob_female / (1 - prob_female)

cat("Odds(survived | male)   =", round(odds_male, 4), "\n")
cat("Odds(survived | female) =", round(odds_female, 4), "\n")
```

> An odds of 0.24 for males means that for every male who survived, roughly 4 died. An odds greater than 1 for females means more females survived than died.

### Step 3: The Odds Ratio

The **odds ratio (OR)** compares the odds between two groups:

$$\text{OR} = \frac{\text{Odds(female)}}{\text{Odds(male)}}$$

```{r}
#| label: compute-or

or_manual <- odds_female / odds_male

cat("Odds Ratio (female vs. male) =", round(or_manual, 4), "\n")
```

> This tells us that the odds of survival for females were about `r round(or_manual, 1)` times the odds of survival for males. This is a huge difference --- and it is exactly what logistic regression will estimate for us.

::: {.callout-tip}
## The Three Scales
Keep this conversion chain in mind --- you will use it throughout the rest of the course:

**Log-odds** $\xrightarrow{\exp()}$ **Odds** $\xrightarrow{p = \text{odds}/(1+\text{odds})}$ **Probability**

Going back: **Probability** $\xrightarrow{p/(1-p)}$ **Odds** $\xrightarrow{\log()}$ **Log-odds**
:::


## Part 3: Your First Logistic Regression

Now let's fit the same analysis as a logistic regression. We use `glm()` with `family = binomial` (which uses the logit link by default).

### Step 1: Fit the Model

```{r}
#| label: logit-female

model_female <- glm(survived ~ female, data = titanic, family = binomial)
```

:::: {.panel-tabset}

#### Simple

```{r}
#| label: logit-female-summary

summary(model_female)
```

#### Advanced

```{r}
#| label: logit-female-gtsummary

model_female |>
  tbl_regression(exponentiate = TRUE) |>
  bold_labels() |>
  bold_p(t = 0.05)
```

> The `exponentiate = TRUE` option displays odds ratios instead of log-odds coefficients. This is typically what we report in applied research.

::::

### Step 2: Interpret the Log-Odds Coefficients

The default output is on the **log-odds scale**:

```{r}
#| label: logit-coefs

coef(model_female)
```

> - **Intercept** (`r round(coef(model_female)[1], 4)`): This is the log-odds of survival for the reference group (males, when `female = 0`).
> - **female** (`r round(coef(model_female)[2], 4)`): This is the difference in log-odds of survival between females and males.

Log-odds are hard to interpret directly. Let's convert them.

::: {.callout-warning}
## R Reports Log-Odds, Not Odds Ratios
R's `glm()` with `family = binomial` always reports coefficients on the **log-odds** scale. To get **odds ratios** (which is what Stata's `logistic` command reports by default), you need to exponentiate: `exp(coef(model))` or use `parameters::model_parameters(model, exponentiate = TRUE)`. Don't interpret log-odds coefficients as odds ratios!
:::

### Step 3: Transform to Odds

Apply `exp()` to convert from log-odds to odds:

```{r}
#| label: logit-to-odds

# Odds of survival for males (reference group)
odds_male_model <- exp(coef(model_female)[1])

# Odds of survival for females
odds_female_model <- exp(coef(model_female)[1] + coef(model_female)[2])

cat("Estimated odds (male)   =", round(odds_male_model, 4), "\n")
cat("Estimated odds (female) =", round(odds_female_model, 4), "\n")
```

> Compare these to the odds we calculated by hand in Part 2. They should match exactly!

### Step 4: Get the Odds Ratio Directly

The odds ratio is simply `exp()` of the slope coefficient:

```{r}
#| label: odds-ratio

exp(coef(model_female))
```

> The odds ratio for `female` is `r round(exp(coef(model_female)[2]), 2)`. This means the estimated odds of surviving for females are about `r round(exp(coef(model_female)[2]), 1)` times the odds for males --- exactly what we calculated by hand.

### Step 5: Transform to Probabilities

Finally, let's convert to the probability scale using the inverse logit function:

```{r}
#| label: logit-to-prob

# Predicted probability of survival for males
prob_male_model <- plogis(coef(model_female)[1])

# Predicted probability of survival for females
prob_female_model <- plogis(coef(model_female)[1] + coef(model_female)[2])

cat("P(survived | male)   =", round(prob_male_model, 4), "\n")
cat("P(survived | female) =", round(prob_female_model, 4), "\n")
```

> Again, these match our hand calculations from Part 2. With a single dummy predictor, logistic regression simply reproduces the group proportions from the cross-tabulation.

::: {.callout-important}
## Why Not Just Use Linear Regression?
With a binary outcome, linear regression (OLS) can predict probabilities below 0 or above 1, which makes no sense. Logistic regression constrains predicted probabilities to stay between 0 and 1 by modeling the **log-odds** as a linear function of the predictors.
:::


## Part 4: Logistic Regression with a Continuous Predictor

The real power of logistic regression emerges when we use continuous predictors. Let's model survival as a function of ticket fare.

### Step 1: Fit the Model

```{r}
#| label: logit-fare

model_fare <- glm(survived ~ fare, data = titanic, family = binomial)
```

:::: {.panel-tabset}

#### Simple

```{r}
#| label: logit-fare-summary

summary(model_fare)
```

#### Advanced

```{r}
#| label: logit-fare-gtsummary

model_fare |>
  tbl_regression(exponentiate = TRUE) |>
  bold_labels() |>
  bold_p(t = 0.05)
```

::::

### Step 2: Interpret the Odds Ratio

```{r}
#| label: fare-or

exp(coef(model_fare))
```

> The odds ratio for `fare` is approximately `r round(exp(coef(model_fare)[2]), 4)`. For each additional British pound of fare, the odds of survival increase by about `r round((exp(coef(model_fare)[2]) - 1) * 100, 2)`%. Passengers who paid more for their tickets had higher odds of surviving.

::: {.callout-tip}
## Interpreting OR for Continuous Predictors
- **OR > 1**: each unit increase in the predictor is associated with *higher* odds of the outcome.
- **OR < 1**: each unit increase is associated with *lower* odds.
- **OR = 1**: no association.

For small effects close to 1, the interpretation is: (OR - 1) x 100 = percentage change in odds per unit increase.
:::

### Step 3: Predicted Probability Curve

Unlike linear regression, the relationship between a continuous predictor and the predicted probability is **curved** (S-shaped):

```{r}
#| label: predicted-prob-curve
#| fig-alt: "S-shaped predicted probability curve showing survival probability increasing with ticket fare, with jittered observed outcomes at 0 and 1"
#| fig-cap: "Predicted probability of survival as a function of ticket fare"

# Create prediction data
pred_data <- tibble(fare = seq(0, 300, length.out = 200))

pred_data <- pred_data |>
  mutate(prob = predict(model_fare, newdata = pred_data, type = "response"))

ggplot() +
  geom_jitter(data = titanic, aes(x = fare, y = survived),
              height = 0.03, alpha = 0.2, size = 1, color = "#2C3E50") +
  geom_line(data = pred_data, aes(x = fare, y = prob),
            color = "#E74C3C", linewidth = 1.2) +
  labs(x = "Ticket Fare (British Pounds)",
       y = "Predicted P(Survived)",
       title = "Logistic Regression: Survival by Fare",
       subtitle = "S-shaped curve keeps predictions between 0 and 1") +
  scale_y_continuous(limits = c(0, 1))
```

> The jittered dots at 0 and 1 are the actual outcomes. The red S-shaped curve shows the predicted probability from our logistic model. Notice how the curve stays between 0 and 1 --- this is the advantage of logistic over linear regression for binary outcomes.


## Key Takeaways

::: {.callout-note}
## What We Learned

1. **Why logistic regression?** When the outcome is binary (0/1), linear regression can produce impossible predictions (below 0 or above 1). Logistic regression solves this by modeling the log-odds as a linear function of predictors.

2. **The log-odds to odds to probability pipeline:**
   - `glm()` estimates coefficients on the **log-odds** scale
   - `exp(coef)` converts to the **odds** scale (odds ratios)
   - `plogis(linear predictor)` converts to the **probability** scale

3. **Odds ratio interpretation:**
   - **OR > 1** means higher odds of the outcome
   - **OR < 1** means lower odds of the outcome
   - For a **dummy variable**: OR compares the odds between two groups
   - For a **continuous variable**: OR is the multiplicative change in odds for each one-unit increase

4. **In our Titanic example:**
   - Females had roughly 11 times the odds of survival compared to males
   - Each additional pound of fare increased the odds of survival by about 1.25%
:::
