---
title: "Lab 9: Simple Logistic Regression"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Use logistic regression for a binary outcome
- Convert among probability, odds, and log-odds
- Fit a logistic regression with `glm(..., family = binomial)`
- Read coefficients on the log-odds scale
- Report odds ratios with `model_params(..., exponentiate = TRUE)`
- Convert fitted values back to predicted probabilities
- Distinguish odds-ratio interpretations from probability interpretations
- Plot predicted probabilities with `ggeffects::predict_response()`
- **Lecture Notes Reference:** Chapter 9: *Simple Logistic Regression*

::: {.meta}
Dataset: `titanic`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Prepare the Data

We use the Titanic passenger data from the lecture notes.
The outcome is `survived`, coded 1 if the passenger survived and 0 otherwise.

```{r}
#| label: lab09-load-data

data(titanic)

titanic_lab <- titanic |>
  mutate(
    sex_f = factor(sex, levels = c(0, 1), labels = c("Male", "Female")),
    survived_f = factor(survived, levels = c(0, 1), labels = c("Died", "Survived")),
    first_class = if_else(pclass == 1, 1L, 0L)
  )

```

```{r}
#| label: lab09-preflight

tibble(
  Rows = nrow(titanic_lab),
  Survival_rate = mean(titanic_lab$survived),
  Female_rate = mean(titanic_lab$sex),
  Mean_age = mean(titanic_lab$age, na.rm = TRUE),
  Missing_age = sum(is.na(titanic_lab$age))
) |>
  round_numeric(digits = 3)
```

::: {.pitfall}
Logistic regression expects a binary outcome.

In this lab, `survived = 1` is the event we model.
Every coefficient describes how a predictor changes the log-odds, odds, or probability of survival.
:::


## Part 2: Start with a Cross-Tabulation

Before fitting a model, compute the observed survival rate by sex.
Because `survived` is coded 0/1, its mean is the survival probability.

```{r}
#| label: lab09-survival-by-sex

survival_by_sex <- titanic_lab |>
  group_by(sex_f) |>
  summarise(
    n = n(),
    n_survived = sum(survived),
    n_died = n - n_survived,
    probability = n_survived / n,
    odds = probability / (1 - probability),
    log_odds = qlogis(probability),
    .groups = "drop"
  )

stopifnot(all(is.finite(survival_by_sex$probability)))
stopifnot(all(survival_by_sex$probability >= 0, survival_by_sex$probability <= 1))

survival_by_sex |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab09-survival-bar
#| fig-cap: "Titanic survival probability by passenger sex."
#| fig-alt: "Bar chart showing the observed probability of survival for male and female Titanic passengers. Female passengers have a much higher survival probability than male passengers."
#| fig-width: 7
#| fig-height: 4.5

ggplot(survival_by_sex, aes(x = sex_f, y = probability, fill = sex_f)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = scales::percent(probability, accuracy = 0.1)),
    vjust = -0.4,
    fontface = "bold"
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(values = c(
    "Male" = ber640_colors[["primary"]],
    "Female" = ber640_colors[["secondary"]]
  )) +
  labs(
    x = "Passenger Sex",
    y = "Observed Survival Probability"
  ) +
  guides(fill = "none")
```

::: {.interpretation}
The table shows the same information on three scales.

The probability scale is intuitive.
The odds scale is what odds ratios compare.
The log-odds scale is where logistic regression is linear.
:::


## Part 3: Convert Between Scales

These three functions are the main conversion tools:

- Probability to odds: `p / (1 - p)`
- Probability to log-odds: `qlogis(p)`
- Log-odds to probability: `plogis(log_odds)`

```{r}
#| label: lab09-scale-conversions

scale_examples <- tibble(
  probability = c(.10, .25, .50, .75, .90)
) |>
  mutate(
    odds = probability / (1 - probability),
    log_odds = qlogis(probability),
    probability_check = plogis(log_odds)
  )

scale_examples |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab09-three-scales-plot
#| fig-cap: "Probability, odds, and log-odds scales."
#| fig-alt: "Three-panel line plot showing how probability maps to probability, odds, and log-odds. Probability is bounded between zero and one, odds are bounded below by zero, and log-odds are unbounded."
#| fig-width: 10
#| fig-height: 4.5

scale_plot_data <- tibble(probability = seq(.01, .99, by = .01)) |>
  mutate(
    Probability = probability,
    Odds = probability / (1 - probability),
    `Log-odds` = qlogis(probability)
  ) |>
  pivot_longer(-probability, names_to = "Scale", values_to = "Value")

ggplot(scale_plot_data, aes(x = probability, y = Value)) +
  geom_line(color = ber640_colors[["primary"]], linewidth = 1.1) +
  facet_wrap(~ Scale, scales = "free_y") +
  labs(
    x = "Probability",
    y = "Scale Value"
  )
```

::: {.pitfall}
Odds and probabilities are not the same.

An odds ratio of 2 does not usually mean "twice as likely."
That shortcut is only close when the event is rare.
:::


## Part 4: Fit a Logistic Regression Model

Use `glm()` with `family = binomial`.
The default link is the logit link, so the coefficients are log-odds coefficients.

```{r}
#| label: lab09-fit-sex-model

model_sex <- glm(
  survived ~ sex_f,
  data = titanic_lab,
  family = binomial
)
```

```{r}
#| label: lab09-log-odds-table

log_odds_table <- model_params(model_sex) |>
  as_tibble() |>
  mutate(p = format_p(p))

log_odds_table |>
  round_numeric(digits = 3)
```

::: {.interpretation}
The intercept is the log-odds of survival for the reference group: male passengers.

The `sex_fFemale` coefficient is the difference in log-odds between female and male passengers.
Most applied reports convert this difference to an odds ratio.
:::


## Part 5: Report Odds Ratios

Set `exponentiate = TRUE` to convert log-odds coefficients into odds ratios.

```{r}
#| label: lab09-odds-ratio-table

odds_ratio_table <- model_params(model_sex, exponentiate = TRUE) |>
  as_tibble() |>
  rename(Odds_ratio = Coefficient) |>
  mutate(p = format_p(p))

odds_ratio_table |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab09-manual-or-check

manual_or <- survival_by_sex |>
  summarise(odds_ratio = odds[sex_f == "Female"] / odds[sex_f == "Male"])

model_or <- odds_ratio_table |>
  filter(Parameter == "sex_fFemale") |>
  transmute(odds_ratio = Odds_ratio)

tibble(
  Source = c("Manual cross-tab", "Logistic model"),
  Odds_ratio = c(manual_or$odds_ratio, model_or$odds_ratio)
) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
With one binary predictor, the logistic regression odds ratio matches the cross-tab odds ratio.

The model is doing the same comparison, but it expresses it in the regression framework that we can extend to more predictors later.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the odds-ratio table to write one sentence interpreting `sex_fFemale`.

Use this structure:

> Female passengers had about ___ times the odds of survival compared with male passengers.

Then add one sentence explaining what the odds ratio does not mean.
:::


## Part 6: Predicted Probabilities for Groups

Odds ratios are compact, but predicted probabilities are easier to communicate.
`ggeffects::predict_response()` returns predicted probabilities on the response scale.

```{r}
#| label: lab09-predicted-sex-probabilities

sex_predictions <- ggeffects::predict_response(model_sex, terms = "sex_f") |>
  as_tibble() |>
  transmute(
    sex_f = x,
    predicted_probability = predicted,
    conf.low,
    conf.high
  )

sex_predictions |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab09-sex-probability-contrast

sex_probability_contrast <- sex_predictions |>
  select(sex_f, predicted_probability) |>
  pivot_wider(names_from = sex_f, values_from = predicted_probability) |>
  mutate(probability_difference = Female - Male)

sex_probability_contrast |>
  round_numeric(digits = 3)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Write two sentences on the probability scale.

1. State the predicted probability of survival for male passengers and female passengers.

2. State the difference in predicted probabilities using percentage points.
:::

```{r}
#| label: lab09-predicted-sex-plot
#| fig-cap: "Predicted probability of Titanic survival by passenger sex."
#| fig-alt: "Point-and-interval plot showing predicted survival probabilities for male and female Titanic passengers. Female passengers have a substantially higher predicted probability of survival."
#| fig-width: 7
#| fig-height: 4.5

ggplot(sex_predictions, aes(x = sex_f, y = predicted_probability)) +
  geom_point(size = 3, color = ber640_colors[["primary"]]) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.12,
    color = ber640_colors[["primary"]]
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    x = "Passenger Sex",
    y = "Predicted Survival Probability"
  )
```

::: {.interpretation}
The odds ratio and predicted probabilities answer related but different questions.

The odds ratio compares survival on the odds scale.
The predicted probabilities describe survival on the probability scale.
A large odds ratio can correspond to a smaller probability difference because probabilities are bounded between 0 and 1.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Explain why these two statements can both be true:

- Female passengers had much higher odds of survival than male passengers.
- The predicted probabilities are still numbers between 0 and 1.

Use the words **odds**, **probability**, and **bounded**.
:::


## Part 7: A Continuous Predictor

Logistic regression also works with continuous predictors.
Here we fit a simple model using passenger age.

```{r}
#| label: lab09-fit-age-model

titanic_age <- titanic_lab |>
  drop_na(age)

model_age <- glm(
  survived ~ age,
  data = titanic_age,
  family = binomial
)

model_params(model_age, exponentiate = TRUE) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
For a continuous predictor, the odds ratio is the multiplicative change in odds for a one-unit increase.

For `age`, that means the change in odds of survival for each additional year of age.
If a one-year unit feels too small, rescale the predictor before fitting the model.
:::

```{r}
#| label: lab09-age-prediction-data

age_predictions <- ggeffects::predict_response(model_age, terms = "age [all]") |>
  as_tibble()
```

```{r}
#| label: lab09-age-probability-plot
#| fig-cap: "Predicted probability of Titanic survival by passenger age."
#| fig-alt: "Scatter plot of Titanic survival status by passenger age with a fitted logistic regression curve. The predicted survival probability slopes downward from roughly the mid-40 percent range for children to about one-third by age 80, with the confidence band widest at older ages."
#| fig-width: 8
#| fig-height: 5

ggplot(titanic_age, aes(x = age, y = survived)) +
  geom_jitter(height = 0.035, width = 0, alpha = 0.18,
              color = "grey45") +
  geom_ribbon(
    data = age_predictions,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    inherit.aes = FALSE,
    fill = ber640_colors[["accent"]],
    alpha = 0.18
  ) +
  geom_line(
    data = age_predictions,
    aes(x = x, y = predicted),
    inherit.aes = FALSE,
    color = ber640_colors[["primary"]],
    linewidth = 1.1
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(-0.05, 1.05),
    breaks = seq(0, 1, by = .25)
  ) +
  labs(
    x = "Passenger Age",
    y = "Survival"
  )
```

```{r}
#| label: lab09-selected-age-probabilities

age_predictions_selected <- ggeffects::predict_response(
  model_age,
  terms = "age [10, 30, 50, 70]"
) |>
  as_tibble() |>
  transmute(
    age = x,
    predicted_probability = predicted,
    conf.low,
    conf.high
  )

age_predictions_selected |>
  round_numeric(digits = 3)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Choose two ages from the table.
Write one sentence comparing their predicted probabilities of survival.

Use probability language, not classification language.
:::

::: {.pitfall}
Logistic regression predicts probabilities, not classifications.

The curve gives `P(survived = 1)`.
Turning those probabilities into "survived" or "died" requires a separate decision threshold, which belongs to diagnostics and classification.
:::

::: {.exercises}
## What to Submit

Submit the following items:

1. The odds-ratio table for `model_sex`. [core]{.chip .easy}

2. One sentence interpreting the `sex_fFemale` odds ratio using odds language. [core]{.chip .med}

3. The predicted-probability table for `sex_f`. [core]{.chip .easy}

4. Two sentences interpreting the predicted probabilities and probability difference. [core]{.chip .med}

5. One short explanation of why an odds ratio and a probability difference are not the same quantity. [core]{.chip .med}

6. The selected-age predicted-probability table and one sentence comparing two ages. [core]{.chip .med}
:::

::: {.summary}
## Lab Summary

1. Logistic regression is for binary outcomes.

2. `glm(..., family = binomial)` estimates coefficients on the log-odds scale.

3. `exp(coef)` or `model_params(..., exponentiate = TRUE)` converts log-odds coefficients to odds ratios.

4. Probability, odds, and log-odds are different scales for the same event probability.

5. `plogis()` converts log-odds to probabilities.

6. `ggeffects::predict_response()` gives simple predicted probability tables and plots.

**Next:** Lab 10 extends logistic regression to multiple predictors and adjusted predictions.
:::
