---
title: "Lab 1: Introduction to R"
---

{{< include ../_common.qmd >}}

## Goals

::: {.callout-note}
## What You'll Learn in This Lab
- Load a dataset and explore its structure
- Understand what a data frame (tibble) is: rows, columns, variables
- Work with variables: rename, create, select, and filter
- Compute descriptive statistics for categorical and continuous variables
- Create basic visualizations: boxplots, histograms, scatterplots, bar plots
- Run your first hypothesis test (two-sample t-test)
- **Lecture Notes Reference:** No specific lecture chapter — this lab covers R fundamentals
:::

**Dataset:** `instruction` from the `regdatasets` package


## Part 1: Getting Started with R

Welcome to your first R lab! If this is your very first time using R, that is perfectly fine. We will walk through everything step by step.

::: {.callout-tip}
## A Quick Note on R and RStudio
R is the programming language. RStudio is the application you use to work with R --- think of R as the engine and RStudio as the car dashboard. If you have not yet installed R and RStudio, see the **Appendix** for instructions. For now, we assume you have both open and ready.
:::

### Step 1: Load Packages

In R, **packages** are collections of tools that extend what R can do. Think of them like apps on your phone: R comes with basic features, and packages add specialized capabilities.

We need two packages for this lab:

- `regdatasets` --- contains the datasets we use in this course
- `tidyverse` --- a collection of packages for data manipulation and visualization

These packages are already loaded for you behind the scenes (via the book's setup file), but in your own work you would type:

```r
library(regdatasets)
library(tidyverse)
```

### Step 2: Load the Data

Our dataset is called `instruction`. It contains math achievement data for elementary school students. Let's load it:

```{r}
#| label: lab01-load-data

data(instruction)
```

That's it! The dataset is now available in your R environment.

### Step 3: Take a First Look

The `glimpse()` function shows us the structure of a dataset --- how many rows, how many columns, and what each column looks like:

```{r}
#| label: lab01-glimpse

glimpse(instruction)
```

> Each **row** is one student. Each **column** is one variable (a piece of information measured for each student). This is what we call a **data frame** --- R's version of a spreadsheet.

Let's also check the dimensions directly:

```{r}
#| label: lab01-dimensions

nrow(instruction)
ncol(instruction)
```

> `nrow()` tells us the number of rows (students), and `ncol()` tells us the number of columns (variables).

### Step 4: View the First Few Rows

The `head()` function shows us the first few rows of the dataset, so we can see what the actual data values look like:

```{r}
#| label: lab01-head

head(instruction)
```

> Here is what each variable means:
>
> - `childid` --- a unique ID for each student
> - `classid` --- which classroom the student belongs to
> - `schoolid` --- which school the student belongs to
> - `girl` --- 1 if the student is a girl, 0 if a boy
> - `minority` --- 1 if the student is a minority, 0 if not
> - `ses` --- socioeconomic status (continuous)
> - `mathkind` --- math score at kindergarten entry
> - `mathgain` --- how much the student's math score increased
> - `yearstea` --- years of teaching experience
> - `mathknow` --- teacher's math knowledge score
> - `housepov` --- proportion of households in poverty (school level)
> - `mathprep` --- teacher's math preparation score

::: {.callout-note}
## What Is a Tibble?
When you see the word "tibble" in R output, it just means a modern version of a data frame. It behaves exactly like a data frame but prints more neatly on screen. You can treat "tibble" and "data frame" as the same thing.
:::


## Part 2: Working with Variables

Now that we can see our data, let's learn how to manipulate it. In R, we use **verbs** (functions) from the `dplyr` package to work with data. We will also use a special operator called the **pipe** (`|>`).

::: {.callout-tip}
## What Is the Pipe (`|>`)?
The pipe takes the result from the left side and passes it as input to the right side. Read it as **"and then."**

For example: `instruction |> head()` means "take the `instruction` dataset **and then** show the first few rows."

This lets us chain steps together in a readable, left-to-right sequence.
:::

### Step 1: Rename a Variable

The variable `girl` uses 1 for girls and 0 for boys. In research, it is more common to use the name `female` for this type of indicator. Let's rename it:

```{r}
#| label: lab01-rename

instruction <- instruction |>
  rename(female = girl)
```

> `rename(new_name = old_name)` changes the name of a column. We save the result back into `instruction` using `<-` (the assignment arrow). Now `female = 1` for girls and `female = 0` for boys.

Let's verify the rename worked:

```{r}
#| label: lab01-check-rename

names(instruction)
```

### Step 2: Create New Variables with `mutate()`

`mutate()` lets us add new columns to our dataset. Let's create two new variables.

**Example 1: A dummy variable for new teachers**

We will create a variable called `newteach` that equals 1 if the teacher has fewer than 5 years of experience, and 0 otherwise:

```{r}
#| label: lab01-mutate-newteach

instruction <- instruction |>
  mutate(newteach = if_else(yearstea < 5, 1, 0))
```

> `if_else(condition, value_if_true, value_if_false)` is like an "if-then" statement. For each row, R checks: is `yearstea` less than 5? If yes, `newteach` gets 1. If no, it gets 0.

Let's verify:

```{r}
#| label: lab01-check-newteach

instruction |>
  count(newteach)
```

**Example 2: A computed variable**

Let's create a variable `mathfirst` that represents each student's estimated first-grade math score --- the sum of their kindergarten score and their gain:

```{r}
#| label: lab01-mutate-mathfirst

instruction <- instruction |>
  mutate(mathfirst = mathkind + mathgain)
```

```{r}
#| label: lab01-check-mathfirst

instruction |>
  select(mathkind, mathgain, mathfirst) |>
  head()
```

> `select()` lets us pick specific columns to display. Here we show just the three columns that are relevant to verify our calculation.

### Step 3: Select Variables

Sometimes we want to work with only a few columns. `select()` lets us pick the ones we need:

```{r}
#| label: lab01-select

instruction |>
  select(childid, female, minority, ses, mathkind, mathgain) |>
  head()
```

> This does not change the original dataset. It just creates a temporary view with only the columns we listed.

### Step 4: Filter Rows

`filter()` lets us keep only the rows that meet a certain condition. For example, let's look at only the students in School 11:

```{r}
#| label: lab01-filter

school11 <- instruction |>
  filter(schoolid == 11)

nrow(school11)
```

> **Important:** Notice the **double equals sign** (`==`). In R, `==` means "is equal to" (a comparison). A single `=` is used for other things (like naming arguments inside functions). This is a common source of early confusion, so keep an eye on it.

> School 11 has `r nrow(school11)` students. We saved this subset as a new object called `school11`.


## Part 3: Data Cleaning — Identifying Suspicious Values

Before jumping into analysis, good practice is to check the data for suspicious values. Let's look at the `mathknow` variable (teacher's math knowledge score):

```{r}
#| label: lab01-check-mathknow

summary(instruction$mathknow)
```

> Notice that the maximum value is very high. Let's look more closely:

```{r}
#| label: lab01-suspicious-values

instruction |>
  filter(mathknow > 3) |>
  count(mathknow)
```

> There are observations coded as very large values that likely represent **missing data** in the original coding scheme. In many datasets, codes like `99`, `999`, or `-99` are used as placeholders for "not available." We need to recode these to `NA` (R's way of representing missing values).

```{r}
#| label: lab01-recode-na

instruction <- instruction |>
  mutate(mathknow = na_if(mathknow, max(mathknow, na.rm = TRUE)))

summary(instruction$mathknow)
```

> After recoding, the summary now shows `NA's` instead of the suspicious maximum value. The `na_if()` function replaces a specific value with `NA`.

::: {.callout-tip}
## Always Check Your Data
Before any analysis, look at `summary()` output for each variable. Watch for:

- Unusually large or small values (potential missing value codes)
- Values outside the expected range
- An unexpected number of `NA`s

A few minutes of data checking can save hours of confusion later.
:::


## Part 4: Descriptive Statistics

Before running any statistical model, it is always a good idea to compute descriptive statistics. Let's see how to summarize both categorical and continuous variables.

### Categorical Variables: `count()`

For categorical variables (variables with a small number of distinct values), we want to see **frequency tables** --- how many observations fall into each category.

```{r}
#| label: lab01-count-female

instruction |>
  count(female)
```

> This tells us how many boys (`female = 0`) and girls (`female = 1`) are in the dataset.

We can also cross-tabulate two categorical variables:

```{r}
#| label: lab01-count-cross

instruction |>
  count(female, minority)
```

> This shows us the breakdown by both gender and minority status. For example, we can see how many non-minority boys, minority boys, non-minority girls, and minority girls are in the data.

### Continuous Variables: `summary()`

For continuous variables, we want measures like the mean, median, minimum, and maximum. The simplest way is `summary()`:

```{r}
#| label: lab01-summary-ses

summary(instruction$ses)
```

> The `$` symbol is how we pull out a single column from a data frame. So `instruction$ses` means "the `ses` column from `instruction`."

### Custom Summaries with `summarise()`

For more control, we can use `summarise()` (or `summarize()` --- both spellings work):

```{r}
#| label: lab01-summarise

instruction |>
  summarise(
    mean_ses   = mean(ses, na.rm = TRUE),
    sd_ses     = sd(ses, na.rm = TRUE),
    mean_math  = mean(mathkind, na.rm = TRUE),
    sd_math    = sd(mathkind, na.rm = TRUE)
  )
```

> The `na.rm = TRUE` argument tells R to ignore missing values when computing the statistic. If you forget it and the data contains any `NA` (missing) values, R will return `NA` as the result --- which is R's cautious way of saying "I'm not sure because some data is missing."

### Grouped Summaries: `group_by()`

One of the most powerful features in `dplyr` is computing statistics **by group**. Let's compare mean SES and mean math scores between boys and girls:

```{r}
#| label: lab01-grouped-summary

instruction |>
  group_by(female) |>
  summarise(
    n          = n(),
    mean_ses   = mean(ses, na.rm = TRUE),
    mean_math  = mean(mathkind, na.rm = TRUE),
    mean_gain  = mean(mathgain, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric) & !n, \(x) round(x, 2)))
```

> `group_by(female)` tells R: "do the next calculation separately for each value of `female`." The `n()` function counts how many rows are in each group. The `.groups = "drop"` argument is just a technical detail that prevents a warning message.

> Compare the means for boys (`female = 0`) and girls (`female = 1`). Do you see any differences in SES, kindergarten math, or math gain?


## Part 5: Visualizing Data

Tables of numbers are useful, but visualizations let us **see** patterns in the data. We use the `ggplot2` package, which follows a consistent grammar: start with `ggplot()`, then **add layers** with `+`.

::: {.callout-tip}
## The ggplot2 Pattern
Every plot follows the same basic recipe:

1. `ggplot(data, aes(x = ..., y = ...))` --- set up the canvas and map variables to axes
2. `+ geom_*()` --- add a geometric layer (points, bars, boxes, etc.)
3. `+ labs(...)` --- add labels (optional but recommended)

Think of it like building a painting: first you set up the canvas, then you add shapes, then you add titles.
:::

### Boxplot: Comparing Groups

Boxplots are great for comparing the distribution of a continuous variable across groups:

```{r}
#| label: lab01-boxplot
#| fig-cap: "Distribution of kindergarten math scores by gender"
#| fig-alt: "Boxplot comparing kindergarten math scores by gender, showing similar medians for males and females with comparable spread and a few low outliers in both groups"

ggplot(instruction, aes(x = factor(female, labels = c("Male", "Female")),
                        y = mathkind)) +
  geom_boxplot(fill = c("#2C3E50", "#E74C3C"), alpha = 0.7) +
  labs(x = "Gender", y = "Math Score at Kindergarten Entry")
```

> **Reading a boxplot:** The thick line in the middle is the **median**. The box spans the **interquartile range** (IQR) --- the middle 50% of the data. The "whiskers" extend to the most extreme points that are not outliers, and individual dots are **outliers** (unusually high or low values).

### Histogram: Shape of a Distribution

A histogram shows how values of a single continuous variable are distributed:

```{r}
#| label: lab01-histogram
#| fig-cap: "Distribution of SES"
#| fig-alt: "Histogram of socioeconomic status showing an approximately normal distribution centered near zero with slight right skew"

ggplot(instruction, aes(x = ses)) +
  geom_histogram(bins = 20, fill = "#3498DB", color = "white", alpha = 0.8) +
  labs(x = "Socioeconomic Status (SES)", y = "Count")
```

> **What to look for:** Is the distribution roughly symmetric (bell-shaped), or is it skewed to one side? Are there any unusual gaps or clusters?

### Scatterplot: Relationship Between Two Variables

Scatterplots show the relationship between two continuous variables --- each dot is one student:

```{r}
#| label: lab01-scatter
#| fig-cap: "SES vs. kindergarten math score"
#| fig-alt: "Scatterplot of SES versus kindergarten math score showing a moderate positive relationship, with higher SES students tending to have higher math scores"

ggplot(instruction, aes(x = ses, y = mathkind)) +
  geom_point(alpha = 0.3, size = 1.5, color = "#2C3E50") +
  labs(x = "Socioeconomic Status (SES)",
       y = "Math Score at Kindergarten Entry")
```

> Each dot represents one student. When dots tend to go from lower-left to upper-right, we say there is a **positive relationship** --- as SES increases, math scores tend to increase too.

### Scatterplot by Group

We can color the points by a grouping variable to see if the pattern differs across groups:

```{r}
#| label: lab01-scatter-group
#| fig-cap: "SES vs. kindergarten math score, by gender"
#| fig-alt: "Scatterplot of SES versus kindergarten math score colored by gender, showing a similar positive trend for both males and females with largely overlapping point clouds"

ggplot(instruction, aes(x = ses, y = mathkind,
                        color = factor(female, labels = c("Male", "Female")))) +
  geom_point(alpha = 0.3, size = 1.5) +
  scale_color_manual(values = c("#2C3E50", "#E74C3C")) +
  labs(x = "Socioeconomic Status (SES)",
       y = "Math Score at Kindergarten Entry",
       color = "Gender")
```

> Do boys and girls show a similar pattern in the relationship between SES and math scores, or does the pattern differ?

### Bar Plot: Group Means

Bar plots can display the average value of a variable for each group. Here we use `stat_summary()` to compute the mean automatically:

```{r}
#| label: lab01-barplot
#| fig-cap: "Mean kindergarten math score by gender"
#| fig-alt: "Bar plot of mean kindergarten math scores by gender with standard error bars, showing males with a slightly higher mean than females but overlapping error bars"

ggplot(instruction, aes(x = factor(female, labels = c("Male", "Female")),
                        y = mathkind)) +
  stat_summary(fun = mean, geom = "bar",
               fill = c("#2C3E50", "#E74C3C"), alpha = 0.7, width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(x = "Gender", y = "Mean Math Score at Kindergarten Entry")
```

> The bars show the **mean** for each group, and the error bars show the **standard error of the mean**. Standard errors give us a sense of the precision of the group means.


## Part 6: Your First Hypothesis Test

You have explored the data with tables and graphs. Now let's do a formal statistical test. We will use a **two-sample t-test** to ask: Is there a statistically significant difference in kindergarten math scores between minority and non-minority students?

### Run the t-test

```{r}
#| label: lab01-ttest

t.test(mathkind ~ minority, data = instruction, var.equal = TRUE)
```

> The `~` symbol means "predicted by" or "as a function of." So `mathkind ~ minority` reads as "mathkind by minority group."

> The `var.equal = TRUE` argument tells R to assume equal variances in both groups (which matches the pooled t-test you may have learned in your introductory statistics course).

### Interpret the Output

The output gives us several important pieces of information:

> **Group means:** The mean kindergarten math score for non-minority students (group 0) and minority students (group 1). Compare these two numbers --- which group scored higher on average?

> **t-statistic:** This measures how many standard errors apart the two group means are. Larger absolute values indicate stronger evidence of a difference.

> **p-value:** This tells us the probability of seeing a difference this large (or larger) if the two groups actually had the same population mean. A small p-value (typically less than 0.05) suggests the difference is statistically significant.

> **95% confidence interval:** This gives a range of plausible values for the true difference in population means. If this interval does not contain 0, we conclude the difference is statistically significant at the 0.05 level.

### Connection to Regression

Here is a preview of what we will learn in Lab 2: the two-sample t-test is actually a **special case** of linear regression. When we use a binary predictor (like `minority`) in a regression model, the slope equals the difference in group means --- exactly the same value the t-test compares.

```{r}
#| label: lab01-ttest-as-regression

model_preview <- lm(mathkind ~ minority, data = instruction)
summary(model_preview)
```

> Notice that the t-statistic for the `minority` coefficient matches the t-statistic from the t-test above (the sign may differ). The slope tells us the mean difference between groups. We will explore this connection fully in Lab 2.


## Key Takeaways

::: {.callout-tip}
## What We Learned

1. **R uses packages** like apps on a phone. We loaded `regdatasets` (for data) and use `tidyverse` (for data manipulation and visualization).

2. **Data frames** are R's version of a spreadsheet. Rows are observations (students), columns are variables. Use `glimpse()`, `head()`, `nrow()`, and `ncol()` to explore.

3. **The pipe (`|>`)** chains steps together. Read it as "and then."

4. **Key dplyr verbs** for data manipulation:
   - `rename()` --- change column names
   - `mutate()` --- create or modify columns
   - `select()` --- pick specific columns
   - `filter()` --- keep specific rows
   - `count()` --- frequency tables for categorical variables
   - `summarise()` --- compute summary statistics
   - `group_by()` --- do calculations separately by group

5. **ggplot2** follows the pattern: `ggplot(data, aes(...)) + geom_*()`. We made boxplots, histograms, scatterplots, and bar plots.

6. **`t.test()`** compares the means of two groups. Look at the t-statistic, p-value, and confidence interval to draw conclusions.

7. A two-sample t-test is a special case of regression with a binary predictor --- we will see this connection in Lab 2.

**Next:** In Lab 2, we will fit our first linear regression model using `lm()`, learn to interpret the output, and check model assumptions.
:::
