---
title: "Lab 10: Multiple Logistic Regression"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Recognize Simpson's paradox as confounding with a binary outcome
- Fit unadjusted and adjusted logistic regression models
- Interpret adjusted odds ratios with `model_params(..., exponentiate = TRUE)`
- Create a polished logistic regression table with `gtsummary`
- Compare nested logistic models using the same analysis sample
- Use likelihood-ratio tests for nested logistic models
- Compute adjusted predicted probabilities with `ggeffects::predict_response()`
- **Lecture Notes Reference:** Chapter 10: *Multiple Logistic Regression*

::: {.meta}
Dataset: `berk_sub`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Prepare the Data

Chapter 10 uses a Berkeley admissions example.
The outcome is `admitted`, coded 1 if the applicant was admitted and 0 otherwise.

```{r}
#| label: lab10-load-data

data(berk_sub)

berk_lab <- berk_sub |>
  mutate(
    sex_f = factor(male, levels = c(0, 1), labels = c("Female", "Male")),
    dept_f = factor(
      engineering,
      levels = c(0, 1),
      labels = c("Psychology", "Engineering")
    ),
    admitted_f = factor(
      admitted,
      levels = c(0, 1),
      labels = c("Rejected", "Admitted")
    )
  )

```

```{r}
#| label: lab10-preflight

tibble(
  Rows = nrow(berk_lab),
  Admission_rate = mean(berk_lab$admitted),
  Male_rate = mean(berk_lab$male),
  Engineering_rate = mean(berk_lab$engineering),
  Missing_values = sum(is.na(berk_lab))
) |>
  round_numeric(digits = 3)
```

::: {.pitfall}
The event is `admitted = 1`.

Every odds ratio and predicted probability in this lab describes admission, not rejection.
Changing the event coding would change the interpretation.
:::


## Part 2: The Aggregate Picture

First ignore department and compare admission rates by sex.
This is the unadjusted view.

```{r}
#| label: lab10-aggregate-by-sex

aggregate_by_sex <- berk_lab |>
  group_by(sex_f) |>
  summarise(
    n = n(),
    n_admitted = sum(admitted),
    n_rejected = n - n_admitted,
    probability = n_admitted / n,
    odds = probability / (1 - probability),
    log_odds = qlogis(probability),
    .groups = "drop"
  )

aggregate_by_sex |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-aggregate-bar
#| fig-cap: "Aggregate Berkeley admission probability by applicant sex."
#| fig-alt: "Bar chart showing aggregate Berkeley admission probabilities for female and male applicants. The male bar is much taller than the female bar when department is ignored."
#| fig-width: 7
#| fig-height: 4.5

ggplot(aggregate_by_sex, aes(x = sex_f, y = probability, fill = sex_f)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = scales::percent(probability, accuracy = 0.1)),
    vjust = -0.4,
    fontface = "bold"
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .7)) +
  scale_fill_manual(values = c(
    "Female" = ber640_colors[["secondary"]],
    "Male" = ber640_colors[["primary"]]
  )) +
  labs(
    x = "Applicant Sex",
    y = "Observed Admission Probability"
  ) +
  guides(fill = "none")
```

::: {.interpretation}
The aggregate table suggests a large sex difference.

At this point, however, the comparison ignores department.
If department affects admission and is associated with applicant sex, this unadjusted comparison is confounded.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the aggregate table and plot.
Write two sentences:

1. Describe the unadjusted sex difference in admission probability.

2. Name the variable that might confound this comparison.
:::


## Part 3: Look Inside the Confounder

Now stratify by department.
This asks whether the aggregate sex difference still appears within each department.

```{r}
#| label: lab10-department-by-sex

department_by_sex <- berk_lab |>
  group_by(dept_f, sex_f) |>
  summarise(
    n = n(),
    n_admitted = sum(admitted),
    probability = n_admitted / n,
    odds = probability / (1 - probability),
    .groups = "drop"
  )

department_by_sex |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-stratified-bar
#| fig-cap: "Berkeley admission probability by sex within department."
#| fig-alt: "Faceted bar chart showing Berkeley admission probabilities by applicant sex within Psychology and Engineering. Within each department, the sex difference is much smaller than in the aggregate view."
#| fig-width: 9
#| fig-height: 4.8

ggplot(department_by_sex, aes(x = sex_f, y = probability, fill = sex_f)) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = scales::percent(probability, accuracy = 0.1)),
    vjust = -0.4,
    fontface = "bold",
    size = 3.8
  ) +
  facet_wrap(~ dept_f) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .8)) +
  scale_fill_manual(values = c(
    "Female" = ber640_colors[["secondary"]],
    "Male" = ber640_colors[["primary"]]
  )) +
  labs(
    x = "Applicant Sex",
    y = "Observed Admission Probability"
  ) +
  guides(fill = "none")
```

```{r}
#| label: lab10-application-pattern

application_pattern <- berk_lab |>
  count(sex_f, dept_f, name = "n") |>
  group_by(sex_f) |>
  mutate(
    total = sum(n),
    proportion = n / total
  ) |>
  ungroup()

application_pattern |>
  round_numeric(digits = 3)
```

::: {.interpretation}
Department is a confounder in this example.

It predicts admission because Engineering and Psychology have different admission rates.
It is also associated with applicant sex because male and female applicants applied to the two departments in different proportions.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use `department_by_sex` and `application_pattern`.
Write a three-bullet confounding check:

1. Does department predict admission?

2. Is department associated with applicant sex?

3. Does the aggregate sex difference look smaller within departments?
:::


## Part 4: Fit Unadjusted and Adjusted Models

The unadjusted model includes sex only.
The adjusted model adds department.

```{r}
#| label: lab10-fit-models

model_unadjusted <- glm(
  admitted ~ sex_f,
  data = berk_lab,
  family = binomial
)

model_department <- glm(
  admitted ~ dept_f,
  data = berk_lab,
  family = binomial
)

model_adjusted <- glm(
  admitted ~ sex_f + dept_f,
  data = berk_lab,
  family = binomial
)
```

```{r}
#| label: lab10-adjusted-or-table

unadjusted_sex_or <- model_params(model_unadjusted, exponentiate = TRUE) |>
  as_tibble() |>
  filter(Parameter == "sex_fMale") |>
  mutate(Model = "Unadjusted: sex only")

adjusted_sex_or <- model_params(model_adjusted, exponentiate = TRUE) |>
  as_tibble() |>
  filter(Parameter == "sex_fMale") |>
  mutate(Model = "Adjusted: sex + department")

bind_rows(unadjusted_sex_or, adjusted_sex_or) |>
  rename(Odds_ratio = Coefficient) |>
  mutate(p = format_p(p)) |>
  select(Model, Parameter, Odds_ratio, CI_low, CI_high, p) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-full-adjusted-or

model_params(model_adjusted, exponentiate = TRUE) |>
  as_tibble() |>
  rename(Odds_ratio = Coefficient) |>
  mutate(p = format_p(p)) |>
  select(Parameter, Odds_ratio, CI_low, CI_high, p) |>
  round_numeric(digits = 3)
```

For a report-style table, `gtsummary::tbl_regression()` can format the same adjusted model in one short call.

```{r}
#| label: lab10-gtsummary-adjusted-or

gtsummary::tbl_regression(model_adjusted, exponentiate = TRUE) |>
  gtsummary::bold_labels()
```

::: {.interpretation}
The unadjusted odds ratio for `sex_fMale` compares male and female applicants while ignoring department.

The adjusted odds ratio compares male and female applicants holding department constant.
The change from the unadjusted to adjusted odds ratio is the regression version of the Simpson's paradox pattern.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the unadjusted-vs-adjusted odds-ratio table.
Write two sentences:

1. Interpret the unadjusted `sex_fMale` odds ratio.

2. Interpret the adjusted `sex_fMale` odds ratio using the phrase "holding department constant."
:::


## Part 5: Compare Nested Models on the Same Sample

Likelihood-ratio tests compare nested logistic models.
The models must use the same rows.

```{r}
#| label: lab10-same-sample-check

model_samples <- tibble(
  Model = c("Sex only", "Department only", "Sex + department"),
  N = c(
    nobs(model_unadjusted),
    nobs(model_department),
    nobs(model_adjusted)
  )
)

stopifnot(length(unique(model_samples$N)) == 1)

model_samples
```

```{r}
#| label: lab10-fit-table

bind_rows(
  broom::glance(model_unadjusted) |>
    mutate(Model = "Sex only"),
  broom::glance(model_department) |>
    mutate(Model = "Department only"),
  broom::glance(model_adjusted) |>
    mutate(Model = "Sex + department")
) |>
  select(Model, nobs, AIC, BIC, logLik, deviance, df.residual) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-lr-test-sex

anova(model_department, model_adjusted, test = "Chisq") |>
  as_tibble() |>
  mutate(Model = c("Department only", "Department + sex"), .before = 1) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-lr-test-department

anova(model_unadjusted, model_adjusted, test = "Chisq") |>
  as_tibble() |>
  mutate(Model = c("Sex only", "Sex + department"), .before = 1) |>
  round_numeric(digits = 3)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the two likelihood-ratio test tables.
Write two sentences:

1. Does adding sex improve a model that already includes department? Cite the LR-test p-value.

2. Does adding department improve a model that already includes sex? Cite the LR-test p-value.
:::

::: {.pitfall}
Use likelihood-ratio tests only for nested models fit to the same sample.

If one model drops rows because of missing data, refit all compared models on a common analysis dataset before testing.
:::


## Part 6: Adjusted Predicted Probabilities

Adjusted odds ratios are useful, but predicted probabilities are easier to communicate.
Here we use the adjusted model to estimate admission probabilities for each sex-by-department combination.

```{r}
#| label: lab10-adjusted-probabilities

adjusted_probabilities <- ggeffects::predict_response(
  model_adjusted,
  terms = c("sex_f", "dept_f")
) |>
  as_tibble() |>
  transmute(
    sex_f = x,
    dept_f = group,
    predicted_probability = predicted,
    conf.low,
    conf.high
  )

adjusted_probabilities |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-adjusted-probability-differences

adjusted_probability_differences <- adjusted_probabilities |>
  select(dept_f, sex_f, predicted_probability) |>
  pivot_wider(
    names_from = sex_f,
    values_from = predicted_probability
  ) |>
  mutate(male_minus_female = Male - Female)

adjusted_probability_differences |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab10-adjusted-probability-plot
#| fig-cap: "Adjusted Berkeley admission probabilities by sex and department."
#| fig-alt: "Point-and-interval plot showing adjusted Berkeley admission probabilities for female and male applicants within Psychology and Engineering. Department differences are much larger than sex differences within each department."
#| fig-width: 8
#| fig-height: 4.8

ggplot(
  adjusted_probabilities,
  aes(x = dept_f, y = predicted_probability, color = sex_f)
) +
  geom_point(
    position = position_dodge(width = 0.35),
    size = 3
  ) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.35),
    width = 0.15,
    linewidth = 0.8
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .75)) +
  scale_color_manual(values = c(
    "Female" = ber640_colors[["secondary"]],
    "Male" = ber640_colors[["primary"]]
  )) +
  labs(
    x = "Department",
    y = "Adjusted Admission Probability",
    color = "Applicant Sex"
  )
```

::: {.interpretation}
Predicted probabilities translate the adjusted model back to the 0-to-1 probability scale.

In this example, department explains much more of the admission pattern than applicant sex.
The adjusted predictions make that comparison visible without requiring readers to think on the log-odds scale.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the adjusted probability table, probability-difference table, and plot.
Write a three-sentence interpretation:

1. Compare female and male applicants within Psychology.

2. Compare female and male applicants within Engineering.

3. State whether department or sex appears to explain more of the probability-scale pattern.
:::

::: {.exercises}
## What to Submit

Submit the following items:

1. The aggregate admission table and one sentence describing the unadjusted sex difference. [core]{.chip .easy}

2. The department-by-sex table or stratified plot plus the three-bullet confounding check. [core]{.chip .med}

3. The unadjusted-vs-adjusted odds-ratio table plus two interpretation sentences. [core]{.chip .med}

4. The same-sample check and both likelihood-ratio test tables plus two LR-test interpretation sentences. [core]{.chip .hard}

5. The adjusted probability table, probability-difference table, and plot plus a three-sentence probability-scale interpretation. [core]{.chip .hard}
:::

::: {.summary}
## Lab Summary

1. Multiple logistic regression extends simple logistic regression by adding predictors.

2. Confounding can make an aggregate logistic association misleading.

3. `model_params(..., exponentiate = TRUE)` reports adjusted odds ratios when the model contains multiple predictors.

4. `gtsummary::tbl_regression()` creates a polished regression table when you want report-ready formatting.

5. Likelihood-ratio tests compare nested logistic models fit to the same rows.

6. `ggeffects::predict_response()` converts adjusted logistic models into probability-scale summaries.

**Next:** Lab 11 turns to logistic diagnostics, classification, and model-performance checks.
:::
