---
title: "Lab 11: Logistic Diagnostics and Classification"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Explain why linear regression diagnostics do not directly transfer to logistic regression
- Compute grouped goodness-of-fit checks for categorical predictor patterns
- Report McFadden and Tjur pseudo R-squared measures
- Build a classification table and compute sensitivity/specificity
- Plot an ROC curve and report AUC with `pROC`
- Check calibration with grouped predicted probabilities
- Inspect residuals, Cook's distance, and possible separation
- **Lecture Notes Reference:** Chapter 11: *Logistic Regression Diagnostics*

::: {.meta}
Datasets: `penalty` and `titanic`. Estimated time: 75-90 minutes.
:::
:::


## Part 1: Prepare the Data

We begin with the `penalty` data.
The outcome is `death_pe`, coded 1 if the death penalty was imposed and 0 otherwise.

```{r}
#| label: lab11-load-penalty

data(penalty)

penalty_lab <- penalty |>
  mutate(
    def_f = factor(
      def_bl,
      levels = c(0, 1),
      labels = c("White defendant", "Black defendant")
    ),
    vic_f = factor(
      vic_bl,
      levels = c(0, 1),
      labels = c("White victim", "Black victim")
    ),
    death_f = factor(
      death_pe,
      levels = c(0, 1),
      labels = c("No death penalty", "Death penalty")
    )
  )

```

```{r}
#| label: lab11-penalty-preflight

tibble(
  Rows = nrow(penalty_lab),
  Death_penalty_rate = mean(penalty_lab$death_pe),
  Black_defendant_rate = mean(penalty_lab$def_bl),
  Black_victim_rate = mean(penalty_lab$vic_bl),
  Missing_values = sum(is.na(penalty_lab))
) |>
  round_numeric(digits = 3)
```

::: {.pitfall}
The event is `death_pe = 1`.

Classification, sensitivity, specificity, ROC, and calibration all refer to predicting the death penalty event.
:::


## Part 2: Fit the Diagnostic Model

Fit the logistic model from the lecture notes.
The predictors are defendant race and victim race.

```{r}
#| label: lab11-fit-penalty-model

model_penalty <- glm(
  death_pe ~ def_f + vic_f,
  data = penalty_lab,
  family = binomial
)
```

```{r}
#| label: lab11-penalty-or-table

model_params(model_penalty, exponentiate = TRUE) |>
  as_tibble() |>
  rename(Odds_ratio = Coefficient) |>
  mutate(p = format_p(p)) |>
  select(Parameter, Odds_ratio, CI_low, CI_high, p) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
Diagnostics start after the model is fit.

The odds-ratio table tells us the substantive association.
The rest of this lab asks whether the model predicts, calibrates, and behaves in a trustworthy way.
:::


## Part 3: Goodness of Fit for Grouped Data

For categorical predictors, we can group observations by covariate pattern.
Here there are four patterns: defendant race crossed with victim race.

```{r}
#| label: lab11-grouped-gof-table

penalty_gof <- penalty_lab |>
  mutate(pi_hat = fitted(model_penalty)) |>
  group_by(def_f, vic_f) |>
  summarise(
    n = n(),
    observed = sum(death_pe),
    pi_hat = first(pi_hat),
    .groups = "drop"
  ) |>
  mutate(
    expected = n * pi_hat,
    expected_nonevent = n * (1 - pi_hat),
    pearson_component = (observed - expected)^2 /
      (expected * (1 - pi_hat)),
    deviance_component = 2 * (
      if_else(observed > 0, observed * log(observed / expected), 0) +
        if_else(
          n - observed > 0,
          (n - observed) * log((n - observed) / expected_nonevent),
          0
        )
    )
  )

penalty_gof |>
  select(
    def_f, vic_f, n, observed,
    expected, expected_nonevent,
    pearson_component, deviance_component
  ) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab11-gof-summary

X2 <- sum(penalty_gof$pearson_component)
G2 <- sum(penalty_gof$deviance_component)
df_gof <- nrow(penalty_gof) - length(coef(model_penalty))

tibble(
  Statistic = c("Pearson X2", "Deviance G2"),
  Value = c(X2, G2),
  df = df_gof,
  p = 1 - pchisq(c(X2, G2), df_gof)
) |>
  mutate(p = format_p(p)) |>
  round_numeric(digits = 3)
```

::: {.pitfall}
Grouped goodness-of-fit tests require non-sparse grouped data.

If expected event counts are very small, or if continuous predictors create many one-person patterns, the chi-square p-values are not reliable.
Use these tests as one diagnostic, not as a final pass/fail rule.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the grouped goodness-of-fit table.
Write two sentences:

1. Identify the smallest expected event count.

2. Explain why small expected counts make the chi-square p-values fragile.
:::


## Part 4: Pseudo R-Squared

Logistic regression has no single equivalent of OLS R-squared.
Report the name of the measure whenever you report a pseudo R-squared value.

```{r}
#| label: lab11-pseudo-r2

null_model <- glm(
  death_pe ~ 1,
  data = penalty_lab,
  family = binomial
)

mcfadden_r2 <- (deviance(null_model) - deviance(model_penalty)) /
  deviance(null_model)

penalty_pred <- penalty_lab |>
  mutate(predicted_probability = fitted(model_penalty))

tjur_components <- penalty_pred |>
  group_by(death_pe) |>
  summarise(mean_probability = mean(predicted_probability), .groups = "drop")

tjur_r2 <- diff(tjur_components$mean_probability)

tibble(
  Measure = c("McFadden R2", "Tjur R2"),
  Value = c(mcfadden_r2, tjur_r2)
) |>
  round_numeric(digits = 4)
```

```{r}
#| label: lab11-performance-r2

performance::r2(model_penalty)
```

::: {.interpretation}
Pseudo R-squared values are usually smaller than OLS R-squared values.

McFadden R-squared describes deviance improvement over the null model.
Tjur R-squared describes how much the model separates event and non-event predicted probabilities.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the pseudo R-squared output.
Write two sentences:

1. Report McFadden R2 and Tjur R2 by name.

2. Explain why a small pseudo R2 does not automatically make an inference-focused model useless.
:::


## Part 5: Classification at a Cutoff

A logistic model returns probabilities.
To make classifications, we must choose a cutoff.
Here we use the common cutoff of 0.50.

```{r}
#| label: lab11-classification-table

penalty_pred <- penalty_pred |>
  mutate(
    predicted_class = if_else(predicted_probability >= .50, 1L, 0L),
    observed_f = factor(death_pe, levels = c(0, 1)),
    predicted_f = factor(predicted_class, levels = c(0, 1))
  )

conf_matrix <- table(
  Observed = penalty_pred$observed_f,
  Predicted = penalty_pred$predicted_f
)

conf_matrix |>
  addmargins()
```

```{r}
#| label: lab11-classification-metrics

TP <- conf_matrix["1", "1"]
TN <- conf_matrix["0", "0"]
FP <- conf_matrix["0", "1"]
FN <- conf_matrix["1", "0"]

classification_metrics <- tibble(
  Metric = c(
    "Accuracy",
    "Base rate accuracy",
    "Sensitivity",
    "Specificity"
  ),
  Value = c(
    (TP + TN) / sum(conf_matrix),
    max(table(penalty_pred$death_pe)) / nrow(penalty_pred),
    TP / (TP + FN),
    TN / (TN + FP)
  )
)

classification_metrics |>
  round_numeric(digits = 3)
```

::: {.interpretation}
High accuracy can be misleading when the event is rare.

In this example, a 0.50 cutoff predicts no death penalty for every case.
Accuracy looks high because the event is rare, but sensitivity is zero.
:::

```{r}
#| label: lab11-threshold-comparison

make_cutoff_row <- function(cutoff) {
  predicted <- if_else(penalty_pred$predicted_probability >= cutoff, 1L, 0L)

  TP <- sum(penalty_pred$death_pe == 1 & predicted == 1)
  TN <- sum(penalty_pred$death_pe == 0 & predicted == 0)
  FP <- sum(penalty_pred$death_pe == 0 & predicted == 1)
  FN <- sum(penalty_pred$death_pe == 1 & predicted == 0)

  tibble(
    cutoff = cutoff,
    TP = TP,
    FP = FP,
    FN = FN,
    TN = TN,
    Accuracy = (TP + TN) / nrow(penalty_pred),
    Sensitivity = TP / (TP + FN),
    Specificity = TN / (TN + FP)
  )
}

threshold_comparison <- map_dfr(
  c(.05, .10, .20, .50),
  make_cutoff_row
)

threshold_comparison |>
  round_numeric(digits = 3)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the threshold comparison table.
Choose one cutoff that would be better if missing death-penalty cases were very costly.
Then choose one cutoff that would be better if false alarms were very costly.

Write two sentences and mention sensitivity and specificity.
:::


## Part 6: ROC Curve and AUC

The ROC curve summarizes classification performance across all possible cutoffs.
AUC summarizes the ranking ability of the model.

```{r}
#| label: lab11-roc-data

roc_penalty <- pROC::roc(
  response = penalty_pred$death_pe,
  predictor = penalty_pred$predicted_probability,
  quiet = TRUE
)

auc_penalty <- as.numeric(pROC::auc(roc_penalty))

roc_data <- tibble(
  specificity = roc_penalty$specificities,
  sensitivity = roc_penalty$sensitivities
) |>
  mutate(false_positive_rate = 1 - specificity) |>
  arrange(false_positive_rate, sensitivity)

tibble(AUC = auc_penalty) |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab11-roc-plot
#| fig-cap: "ROC curve for the death penalty logistic model."
#| fig-alt: "ROC curve plotting sensitivity against false positive rate for the death penalty logistic model. The curve is above the diagonal reference line but not close to the upper-left corner, indicating modest discrimination."
#| fig-width: 6.5
#| fig-height: 5

ggplot(roc_data, aes(x = false_positive_rate, y = sensitivity)) +
  geom_abline(linetype = "dashed", color = "grey55") +
  geom_path(color = ber640_colors[["primary"]], linewidth = 1.2) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "False Positive Rate",
    y = "Sensitivity",
    subtitle = paste("AUC =", round(auc_penalty, 3))
  )
```

::: {.pitfall}
AUC does not choose a classification cutoff.

It describes how well the model ranks event cases above non-event cases.
Calibration and cutoff choice are separate questions.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the AUC value and ROC plot.
Write one sentence interpreting discrimination.
Then write one sentence explaining why AUC does not choose a cutoff for you.
:::


## Part 7: Calibration with Continuous Predictors

Calibration asks whether predicted probabilities match observed event rates.
For a calibration plot, the Titanic data give a more useful example because the model has a continuous predictor.
This section is a worked extension on a different model and dataset; do not use the Titanic calibration results as evidence about the death-penalty model.

```{r}
#| label: lab11-load-titanic-diagnostics

data(titanic)

titanic_diag <- titanic |>
  drop_na(age) |>
  mutate(
    sex_f = factor(sex, levels = c(0, 1), labels = c("Male", "Female")),
    pclass_f = factor(pclass, levels = c(1, 2, 3), labels = c("1st", "2nd", "3rd"))
  )

model_titanic <- glm(
  survived ~ sex_f + age + pclass_f,
  data = titanic_diag,
  family = binomial
)

titanic_diag <- titanic_diag |>
  mutate(
    pi_hat = fitted(model_titanic),
    linear_pred = predict(model_titanic, type = "link"),
    resid_deviance = residuals(model_titanic, type = "deviance"),
    resid_pearson = residuals(model_titanic, type = "pearson"),
    cooks_d = cooks.distance(model_titanic),
    obs_id = row_number()
  )
```

```{r}
#| label: lab11-calibration-data

calibration_table <- titanic_diag |>
  mutate(decile = ntile(pi_hat, 10)) |>
  group_by(decile) |>
  summarise(
    mean_predicted = mean(pi_hat),
    observed_probability = mean(survived),
    n = n(),
    se_observed = sqrt(
      observed_probability * (1 - observed_probability) / n
    ),
    .groups = "drop"
  )

calibration_table |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab11-calibration-plot
#| fig-cap: "Calibration plot for the Titanic logistic model."
#| fig-alt: "Calibration plot comparing mean predicted survival probability and observed survival probability across ten groups of Titanic passengers. A dashed diagonal reference line marks perfect calibration."
#| fig-width: 6.5
#| fig-height: 5

ggplot(
  calibration_table,
  aes(x = mean_predicted, y = observed_probability)
) +
  geom_abline(linetype = "dashed", color = "grey55") +
  geom_errorbar(
    aes(
      ymin = observed_probability - 2 * se_observed,
      ymax = observed_probability + 2 * se_observed
    ),
    width = 0,
    color = ber640_colors[["accent"]]
  ) +
  geom_point(size = 2.8, color = ber640_colors[["primary"]]) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Mean Predicted Probability",
    y = "Observed Survival Probability"
  )
```

::: {.interpretation}
Points near the diagonal indicate good calibration.

Points above the line mean the model under-predicts the event.
Points below the line mean the model over-predicts the event.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the calibration table and plot.
Choose one decile where the model appears to under-predict or over-predict survival.
Write two sentences using predicted probability and observed probability language.
:::


## Part 8: Residuals and Influence

Logistic residual plots do not look like OLS residual plots.
With binary outcomes, residuals naturally form bands.
The goal is to look for systematic patterns and influential observations.
As in the calibration section, the Titanic model is used here because it gives a clearer teaching example for continuous-predictor residual and influence displays.

```{r}
#| label: lab11-residual-influence-plot
#| fig-cap: "Deviance residuals and Cook's distance for the Titanic logistic model."
#| fig-alt: "Two-panel diagnostic plot for the Titanic logistic model. The first panel plots deviance residuals against the linear predictor and shows the characteristic banded pattern of binary outcomes. The second panel plots Cook's distance by observation index and highlights a small number of more influential observations."
#| fig-width: 10
#| fig-height: 4.8

residual_plot <- ggplot(
  titanic_diag,
  aes(x = linear_pred, y = resid_deviance)
) +
  geom_point(alpha = 0.25, color = ber640_colors[["primary"]]) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey55") +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    color = ber640_colors[["secondary"]],
    linewidth = 0.9
  ) +
  labs(
    x = "Linear Predictor",
    y = "Deviance Residual",
    title = "Residuals"
  )

cook_plot <- ggplot(
  titanic_diag,
  aes(x = obs_id, y = cooks_d)
) +
  geom_point(alpha = 0.3, color = ber640_colors[["primary"]]) +
  geom_hline(
    yintercept = 4 / nrow(titanic_diag),
    linetype = "dashed",
    color = ber640_colors[["secondary"]]
  ) +
  labs(
    x = "Observation Index",
    y = "Cook's Distance",
    title = "Influence"
  )

residual_plot + cook_plot
```

```{r}
#| label: lab11-top-influence

titanic_diag |>
  slice_max(cooks_d, n = 5) |>
  select(
    obs_id, survived, sex_f, age, pclass_f,
    pi_hat, resid_pearson, cooks_d
  ) |>
  round_numeric(digits = 3)
```

::: {.interpretation}
Influential cases are not automatically wrong.

They are cases worth investigating because they have a larger effect on the fitted model.
For a serious report, compare the main odds ratios with and without the most influential cases.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the residual/influence plot and the top-influence table.
Write two diagnostic bullets:

1. Describe the overall residual pattern.

2. Name one influential case characteristic you would investigate before changing the model.
:::


## Part 9: Separation Warning

Separation occurs when a predictor pattern perfectly predicts an outcome.
In the penalty data, one pattern has no death penalty events.

```{r}
#| label: lab11-separation-cell-check

penalty_lab |>
  count(def_f, vic_f, death_pe) |>
  group_by(def_f, vic_f) |>
  mutate(total = sum(n), proportion = n / total) |>
  ungroup() |>
  round_numeric(digits = 3)
```

```{r}
#| label: lab11-separation-model

penalty_sep <- penalty_lab |>
  mutate(separation_pattern = as.integer(def_bl == 0 & vic_bl == 1))

model_separation <- suppressWarnings(
  glm(
    death_pe ~ separation_pattern + def_bl,
    data = penalty_sep,
    family = binomial
  )
)

model_params(model_separation) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  select(Parameter, Coefficient, SE, CI_low, CI_high, p) |>
  round_numeric(digits = 3)
```

::: {.pitfall}
The `glm()` warning is suppressed above only to keep the rendered lab clean.
In an interactive R session, that warning is itself useful diagnostic evidence.

A huge coefficient with a huge standard error is a separation warning sign.

The model is not discovering a precise effect.
It is struggling because one predictor pattern nearly or perfectly separates the outcome.
Possible remedies include combining sparse categories, collecting more data, or using penalized logistic regression.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the separation cell table and separation-model output.
Write two sentences:

1. Identify the cell pattern with no events.

2. Identify the coefficient/SE pattern that signals separation trouble.
:::


## Part 10: OLS Diagnostics Do Not Transfer Directly

The diagnostic mindset carries over from OLS, but the tools change.

```{r}
#| label: lab11-diagnostic-map

tibble::tribble(
  ~OLS_question, ~Logistic_analogue,
  "How much variation is explained?", "Pseudo R-squared; deviance improvement",
  "Are residuals centered with constant spread?", "Binned residuals; calibration",
  "Are predictions accurate?", "Classification metrics; ROC/AUC",
  "Are probabilities numerically trustworthy?", "Calibration plot",
  "Do a few cases drive the model?", "Cook's distance; DFBETAS",
  "Are estimates stable?", "Sparse cells and separation checks"
)
```

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Write a short diagnostic memo for the penalty model.
Use 5-7 sentences and address grouped goodness of fit, pseudo R-squared, classification threshold, ROC/AUC, and separation.
Mention that calibration and residual/influence plots were demonstrated with the Titanic model, so they should not be cited as direct evidence about the penalty model unless you compute those diagnostics for the penalty model.
End with one recommendation about how cautiously the model should be used.
:::

::: {.exercises}
## What to Submit

Submit the following core items:

1. The grouped GOF table and two-sentence sparse-cell warning. [core]{.chip .med}

2. The pseudo R-squared table plus two sentences reporting McFadden R2 and Tjur R2. [core]{.chip .med}

3. The classification table, classification metrics, and threshold comparison table plus two cutoff-choice sentences. [core]{.chip .hard}

4. The ROC plot and AUC value plus two sentences about discrimination and cutoff choice. [core]{.chip .med}

5. The Titanic calibration table and plot plus two sentences about one over-predicted or under-predicted decile. Label this as a worked example, not evidence about the penalty model. [core]{.chip .hard}

6. One diagnostic extension: either the Titanic residual/influence plot and top-influence table plus two diagnostic bullets, or the penalty separation cell table and separation-model output plus two sentences diagnosing separation risk. [core]{.chip .hard}

7. A short diagnostic memo that states whether the model is mainly useful for inference, classification, calibration, or none of these. [core]{.chip .hard}
:::

::: {.summary}
## Lab Summary

1. Logistic diagnostics focus on fit, classification, calibration, influence, and separation.

2. Grouped goodness-of-fit tests work best for categorical predictor patterns with adequate expected counts.

3. Pseudo R-squared values are useful only when the specific measure is named.

4. Classification requires a cutoff; accuracy can be misleading when the event is rare.

5. ROC/AUC evaluates ranking, while calibration evaluates predicted probability accuracy.

6. Residual and influence diagnostics exist for logistic regression, but they do not look like OLS residual diagnostics.

**Next:** Lab 12 extends generalized linear models beyond the logit link.
:::
