---
title: "Lab 12: GLM Extensions"
---

{{< include ../_common.qmd >}}

::: {.objectives}
## What You'll Learn in This Lab

- Fit logit and probit models with the same `glm()` syntax
- Compare logit and probit coefficients using the 1.7 scaling rule
- Compare binary-response models on the probability scale with `marginaleffects`
- Identify the three GLM components: family, linear predictor, and link
- Fit a Poisson regression model for count outcomes
- Check Poisson overdispersion with a short diagnostic workflow
- **Lecture Notes Reference:** Chapter 12: *Latent Response Models and GLMs*

::: {.meta}
Datasets: `titanic` from `regdatasets` and a small simulated count dataset. Estimated time: 75-90 minutes.
:::
:::

## Part 1: One Function, Different GLMs

In Labs 9-11, we used logistic regression for binary outcomes. This lab zooms out: logistic regression is one member of the larger generalized linear model (GLM) family. The good news is that the R workflow barely changes. We keep the same formula interface and change the `family` argument.

::: {.callout-tip}
## Stata to R Translation

| Stata | R | Notes |
|:--|:--|:--|
| `logit survived i.sex age` | `glm(survived ~ sex_f + age, data = titanic_glm, family = binomial(link = "logit"))` | Actual lab logit model |
| `probit survived i.sex age` | `glm(survived ~ sex_f + age, data = titanic_glm, family = binomial(link = "probit"))` | Actual lab probit model |
| `margins, at(...) predict(pr)` | `marginaleffects::predictions(..., type = "response")` | Probability-scale predictions |
| `poisson publications experience` | `glm(publications ~ experience, data = publication_data, family = poisson(link = "log"))` | Actual lab count model |
| `poisson publications experience, irr` | `model_params(poisson_model, exponentiate = TRUE)` | Incidence rate ratios |
| `estat gof` / dispersion checks | `performance::check_overdispersion(poisson_model)` | Quick Poisson overdispersion check |
:::

### Setup

```{r}
#| label: lab12-setup

data(titanic)

titanic_glm <- titanic |>
  mutate(
    sex_f = factor(if_else(sex == 1, "Female", "Male")),
    pclass_f = factor(pclass, labels = c("1st", "2nd", "3rd"))
  ) |>
  drop_na(survived, sex_f, age)

glimpse(titanic_glm)
```

The core GLM pattern is:

```r
glm(outcome ~ predictors, data = my_data, family = some_family)
```

The formula stays familiar. The `family` argument tells R what kind of outcome you have and what link function connects the model to the outcome scale.

```{r}
#| label: glm-family-examples

gaussian_model <- glm(
  age ~ fare + sex_f,
  data = titanic_glm,
  family = gaussian(link = "identity")
)

logit_model <- glm(
  survived ~ sex_f + age,
  data = titanic_glm,
  family = binomial(link = "logit")
)

probit_model <- glm(
  survived ~ sex_f + age,
  data = titanic_glm,
  family = binomial(link = "probit")
)
```

```{r}
#| label: glm-components-table

tibble::tribble(
  ~Model, ~Outcome, ~Family, ~Link, ~`Linear predictor`,
  "Linear regression", "Continuous", "Gaussian", "Identity", "fare + sex",
  "Logit regression", "Binary 0/1", "Binomial", "Logit", "sex + age",
  "Probit regression", "Binary 0/1", "Binomial", "Probit", "sex + age",
  "Poisson regression", "Count", "Poisson", "Log", "experience"
) |>
  knitr::kable()
```

::: {.callout-note}
## The GLM Checklist

Every GLM asks three questions:

1. What is the outcome distribution? This is the **family**.
2. What predictors enter the model? This is the **linear predictor**.
3. How does the mean outcome connect to the linear predictor? This is the **link**.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Choose one model from the GLM table and identify its three components:

1. Family
2. Link
3. Linear predictor

Then write one sentence explaining why that family matches the outcome type.
:::

## Part 2: Logit vs Probit

Logit and probit models answer the same kind of binary-outcome question. They differ in the assumed link function:

- Logit uses the logistic CDF.
- Probit uses the normal CDF.

In R, the only code difference is `link = "logit"` versus `link = "probit"`.

```{r}
#| label: logit-probit-fit

logit_model <- glm(
  survived ~ sex_f + age,
  data = titanic_glm,
  family = binomial(link = "logit")
)

probit_model <- glm(
  survived ~ sex_f + age,
  data = titanic_glm,
  family = binomial(link = "probit")
)
```

```{r}
#| label: logit-probit-coefficients

coefficient_comparison <- tibble(
  term = names(coef(logit_model)),
  logit = coef(logit_model),
  probit = coef(probit_model),
  ratio = logit / probit
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

coefficient_comparison |>
  knitr::kable(
    col.names = c("Term", "Logit beta", "Probit beta", "Logit / Probit")
  )
```

The logit coefficients are larger because logit and probit use different latent scales. A useful rule of thumb is:

$$
\hat{\beta}_{logit} \approx 1.7 \times \hat{\beta}_{probit}
$$

::: {.callout-warning}
## Do Not Compare Raw Coefficients Too Literally

The logit and probit coefficients are not on the same scale. If your audience cares about substantive conclusions, compare predicted probabilities or marginal effects instead.
:::

::: {.your-turn}
[Your turn]{.yt-title} [Write]{.badge-r}

Use the coefficient comparison table.

1. Find the `Logit / Probit` ratio for `sex_fMale`.
2. State whether it is close to the 1.7 rule of thumb.
3. Explain why this does not mean the two models disagree.
:::

## Part 3: Compare on the Probability Scale

The `marginaleffects` package lets us compare models where people actually interpret them: predicted probabilities.

```{r}
#| label: logit-probit-prediction-table

pred_logit <- marginaleffects::predictions(
  logit_model,
  newdata = marginaleffects::datagrid(
    sex_f = "Female",
    age = c(10, 30, 50, 70)
  ),
  type = "response"
) |>
  as.data.frame() |>
  mutate(link = "Logit")

pred_probit <- marginaleffects::predictions(
  probit_model,
  newdata = marginaleffects::datagrid(
    sex_f = "Female",
    age = c(10, 30, 50, 70)
  ),
  type = "response"
) |>
  as.data.frame() |>
  mutate(link = "Probit")

bind_rows(pred_logit, pred_probit) |>
  select(link, sex_f, age, estimate, conf.low, conf.high) |>
  mutate(across(c(estimate, conf.low, conf.high), \(x) round(x, 3))) |>
  knitr::kable(
    col.names = c("Link", "Sex", "Age", "Predicted probability", "CI low", "CI high")
  )
```

The probabilities are nearly identical. This is why logit and probit often lead to the same applied conclusion even though their coefficients differ.

```{r}
#| label: logit-probit-prediction-plot
#| fig-cap: "Logit and probit predicted probabilities for female Titanic passengers by age."
#| fig-alt: "Line plot comparing logit and probit predicted survival probabilities for female Titanic passengers over age. The two lines almost completely overlap, showing very similar predictions from the two link functions."
#| fig-width: 8
#| fig-height: 5

age_grid <- tibble(
  age = seq(1, 75, length.out = 120),
  sex_f = factor("Female", levels = levels(titanic_glm$sex_f))
)

curve_logit <- marginaleffects::predictions(
  logit_model,
  newdata = age_grid,
  type = "response"
) |>
  as.data.frame() |>
  mutate(link = "Logit")

curve_probit <- marginaleffects::predictions(
  probit_model,
  newdata = age_grid,
  type = "response"
) |>
  as.data.frame() |>
  mutate(link = "Probit")

bind_rows(curve_logit, curve_probit) |>
  ggplot(aes(x = age, y = estimate, color = link, fill = link)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.12, color = NA) +
  geom_line(linewidth = 1.1) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(values = c("Logit" = ber640_colors[["accent"]],
                                "Probit" = ber640_colors[["secondary"]])) +
  scale_fill_manual(values = c("Logit" = ber640_colors[["accent"]],
                               "Probit" = ber640_colors[["secondary"]])) +
  labs(
    x = "Age",
    y = "Predicted probability of survival",
    color = "Link",
    fill = "Link"
  )
```

::: {.callout-tip}
## Practical Rule

Use logit when odds ratios are useful for your audience. Use probit when your theory is explicitly about a normally distributed latent response. If neither reason is strong, compare predicted probabilities and choose the model your field expects.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Modify the prediction table code.

1. Change `sex_f = "Female"` to `sex_f = "Male"`.
2. Keep the same ages: 10, 30, 50, and 70.
3. Compare the logit and probit predicted probabilities for male passengers.

Write two sentences: one describing the age pattern and one describing whether the link choice changes the substantive conclusion.
:::

## Part 4: Poisson Regression for Counts

Poisson regression is for count outcomes: 0, 1, 2, 3, and so on. In this section we use a small simulated dataset so the data-generating story is transparent.

```{r}
#| label: poisson-data

set.seed(640)

publication_data <- tibble(
  experience = runif(200, 1, 30),
  publications = rpois(200, lambda = exp(0.5 + 0.04 * experience))
)

publication_data |>
  summarise(
    n = n(),
    mean_publications = mean(publications),
    var_publications = var(publications),
    min_publications = min(publications),
    max_publications = max(publications)
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The raw mean and variance provide a quick first look at the counts.
In a regression model, however, the key Poisson assumption is conditional on the predictors.

Now fit the model. The log link keeps predicted counts positive.

```{r}
#| label: poisson-fit

poisson_model <- glm(
  publications ~ experience,
  data = publication_data,
  family = poisson(link = "log")
)

model_params(poisson_model, exponentiate = TRUE)
```

The exponentiated coefficient is an incidence rate ratio (IRR). If the IRR for `experience` is 1.04, each additional year of experience is associated with about a 4% higher expected publication count.

```{r}
#| label: poisson-prediction-plot
#| fig-cap: "Poisson regression fitted curve for simulated publication counts by years of experience."
#| fig-alt: "Scatter plot of simulated publication counts against years of experience with an upward-sloping Poisson fitted curve and a confidence band. The fitted expected count remains positive across the full range."
#| fig-width: 8
#| fig-height: 5

poisson_predictions <- ggeffects::predict_response(
  poisson_model,
  terms = "experience [all]"
) |>
  as.data.frame()

ggplot(publication_data, aes(x = experience, y = publications)) +
  geom_point(alpha = 0.45, color = ber640_colors[["primary"]]) +
  geom_ribbon(
    data = poisson_predictions,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    inherit.aes = FALSE,
    fill = ber640_colors[["accent"]],
    alpha = 0.18
  ) +
  geom_line(
    data = poisson_predictions,
    aes(x = x, y = predicted),
    inherit.aes = FALSE,
    color = ber640_colors[["secondary"]],
    linewidth = 1.2
  ) +
  labs(
    x = "Years of experience",
    y = "Publication count"
  )
```

## Part 5: Check Overdispersion

The basic Poisson assumption is:

$$
Var(Y \mid X) = E(Y \mid X)
$$

In real data, the conditional variance is often larger than the conditional mean. This is called **overdispersion**.
The deviance ratio below is a quick manual screen. `performance::check_overdispersion()` uses a Pearson-based diagnostic, so the two numbers do not have to be identical.

```{r}
#| label: poisson-overdispersion-ratio

overdispersion_table <- tibble(
  residual_deviance = deviance(poisson_model),
  residual_df = df.residual(poisson_model),
  dispersion_ratio = residual_deviance / residual_df
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

overdispersion_table |>
  knitr::kable(
    col.names = c("Residual deviance", "Residual df", "Dispersion ratio")
  )
```

```{r}
#| label: poisson-check-overdispersion

performance::check_overdispersion(poisson_model)
```

::: {.callout-warning}
## What If the Ratio Is Large?

If the dispersion ratio is much larger than 1, the Poisson standard errors may be too small. A common next step is:

```r
glm(publications ~ experience, data = publication_data, family = quasipoisson)
```

For serious overdispersion, analysts often consider negative binomial regression.
:::

::: {.your-turn}
[Your turn]{.yt-title} [R]{.badge-r}

Use the Poisson output and overdispersion checks.

1. Interpret the exponentiated `experience` coefficient as an incidence rate ratio.
2. Create a deliberately overdispersed version of the publication dataset.
3. Compare the overdispersed diagnostic to the original Poisson model.

Start with this code:

```r
set.seed(640)

overdispersed_data <- tibble(
  experience = runif(200, 1, 30),
  person_rate = rgamma(200, shape = 2, rate = 2),
  publications = rpois(200, lambda = person_rate * exp(0.5 + 0.04 * experience))
)

overdispersed_model <- glm(
  publications ~ experience,
  data = overdispersed_data,
  family = poisson(link = "log")
)

performance::check_overdispersion(overdispersed_model)
```

Write three sentences:

1. One IRR interpretation.
2. One overdispersion comparison.
3. One possible next step if overdispersion is serious.
:::

## Part 6: GLM Workflow Summary

```{r}
#| label: glm-workflow-summary

tibble::tribble(
  ~Question, ~Recommended_code, ~Interpretation_scale,
  "Binary outcome, odds ratios", "glm(..., family = binomial(link = 'logit'))", "Odds ratios or probabilities",
  "Binary outcome, latent-normal theory", "glm(..., family = binomial(link = 'probit'))", "Predicted probabilities",
  "Count outcome", "glm(..., family = poisson(link = 'log'))", "Incidence rate ratios or expected counts",
  "Predictions", "marginaleffects::predictions()", "Outcome scale",
  "Model parameters", "model_params()", "Coefficient or exponentiated scale",
  "Poisson diagnostics", "performance::check_overdispersion()", "Dispersion ratio"
) |>
  knitr::kable()
```

::: {.exercises}
## What to Submit

Submit the following items:

1. A three-component GLM description for one model in the workflow table. [core]{.chip .easy}

2. The logit/probit coefficient comparison table and two sentences interpreting the 1.7 scaling rule. [core]{.chip .med}

3. A probability table for male Titanic passengers at ages 10, 30, 50, and 70, plus two sentences comparing logit and probit predictions. [core]{.chip .med}

4. The Poisson IRR output, the original and overdispersed `check_overdispersion()` outputs, and a three-sentence diagnostic interpretation. [core]{.chip .hard}
:::

::: {.summary}
## Lab Summary

1. GLMs reuse the familiar regression formula syntax.

2. `family` and `link` tell R what outcome scale you are modeling.

3. Logit and probit coefficients differ in scale, but predicted probabilities are usually very similar.

4. Poisson regression models expected counts with a log link.

5. For Poisson models, always check overdispersion before trusting standard errors.

**Next:** Lab 13 extends the GLM idea to ordered categories with ordinal regression.
:::
