# ANCOVA {#sec-ch2}

{{< include ../_common.qmd >}}

## Introduction {#sec-ch2-intro}

In @sec-ch1, we modeled the relationship between a single predictor and a continuous outcome. But in education research, we rarely have just one variable of interest. Consider our GCSE example: Chelsea school has higher mean GCSE scores than Battersea. But Chelsea also has students with higher LRT scores at intake. Is the difference in GCSE scores a genuine "school effect," or is it simply explained by the pre-existing differences in student ability?

This question --- **how to make a fair comparison by statistically controlling for other variables** --- motivates the Analysis of Covariance (ANCOVA) model. ANCOVA extends simple regression by including both a categorical predictor (e.g., school) and a continuous covariate (e.g., LRT), allowing us to estimate the "adjusted" difference between groups.

More broadly, this chapter introduces the **multiple linear regression** model, which can include any number of predictors --- both continuous and categorical. Along the way, we encounter some of the most important concepts in applied statistics: omitted variable bias, confounding, and collinearity.

::: {.callout-note}
## Learning Objectives
After completing this chapter, you will be able to:

1. Specify and interpret an ANCOVA model with a continuous covariate and a categorical factor
2. Distinguish between unadjusted and adjusted group means
3. Compute and interpret $R^2$ and the multiple correlation in multiple regression
4. Conduct sequential and partial $F$-tests (Type I vs Type III SS)
5. Explain omitted variable bias (OVB) and identify confounders
6. Recognize the consequences of collinearity for regression estimates
:::


## Motivating Example: Controlling for Prior Achievement {#sec-ch2-example}

Let us continue with the GCSE data from @sec-ch1. Recall that we have data from two London schools --- Battersea and Chelsea --- with both GCSE scores (outcome, age 16) and LRT scores (covariate, age 11).

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-data-setup
#| code-fold: true
#| code-summary: "Show R Code"

# Subset to two schools
two_schools <- gcse |>
  filter(school %in% c(1, 2)) |>
  mutate(school_name = factor(ifelse(school == 1, "Battersea", "Chelsea")))

# Compare raw means
two_schools |>
  group_by(school_name) |>
  summarise(
    n = n(),
    mean_gcse = mean(gcse, na.rm = TRUE),
    mean_lrt  = mean(lrt, na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(digits = 2, col.names = c("School", "n", "Mean GCSE", "Mean LRT"))
```

### Output Interpretation
```{r}
#| ref.label: "ch2-data-setup"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Mean GCSE**: The raw (unadjusted) average GCSE score for each school. Any difference here may reflect genuine school effects, pre-existing ability differences, or both
- **Mean LRT**: The average intake ability for each school. If this differs between schools, the raw GCSE comparison is confounded --- Chelsea students may score higher simply because they started with higher ability
- The gap in Mean LRT motivates the need for ANCOVA: we need to "adjust" the school comparison for pre-existing ability differences (see @sec-ch2-estimation)

### Code Explanation
```{r}
#| ref.label: "ch2-data-setup"
#| eval: false
```
- We filter to schools 1 (Battersea) and 2 (Chelsea) and create a factor for the school name
- The table reveals that Chelsea has a higher mean GCSE score **and** a higher mean LRT score
- This raises the question: is the GCSE difference due to the school, or due to pre-existing differences in student ability?


:::

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-parallel
#| fig-cap: "Regression of GCSE on LRT for Battersea and Chelsea schools. Separate regression lines suggest similar slopes but different intercepts, motivating the ANCOVA (parallel lines) model."
#| fig-alt: "Scatter plot with LRT on the horizontal axis and GCSE on the vertical axis, with points colored by school. Battersea and Chelsea each have a separate regression line. The two lines have similar positive slopes but different intercepts, with Chelsea shifted higher, motivating the parallel lines ANCOVA model that estimates a constant group difference after adjusting for LRT."
#| code-fold: true
#| code-summary: "Show R Code"

ggplot(two_schools, aes(x = lrt, y = gcse, color = school_name)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  scale_color_manual(values = c(ber640_colors["primary"], ber640_colors["accent"])) +
  labs(
    title = "GCSE vs LRT by School",
    subtitle = "Are the regression lines parallel?",
    x = "LRT Score (age 11)",
    y = "GCSE Score (age 16)",
    color = "School"
  )
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-parallel"
#| code-fold: true
#| code-summary: "Show R Code"
```
- The two regression lines appear **roughly parallel**, supporting the ANCOVA assumption of equal slopes across schools
- Chelsea's regression line sits **above** Battersea's, suggesting Chelsea has higher GCSE scores even at the same LRT level
- The vertical gap between the lines is the **adjusted school difference** --- the quantity ANCOVA estimates
- Note the overlap in LRT ranges: both schools have students with similar intake scores, which allows meaningful comparison at common LRT values

### Code Explanation
```{r}
#| ref.label: "fig-ch2-parallel"
#| eval: false
```
- `geom_smooth(method = "lm")` fits separate regression lines for each school
- If the slopes are similar (lines are roughly parallel), the ANCOVA model is appropriate
- The vertical distance between the two parallel lines is the **LRT-adjusted** school difference


:::


## The ANCOVA Model {#sec-ch2-model}

The ANCOVA model includes both a continuous covariate ($x_i$ = LRT) and a dummy variable ($d_{Ci}$ = Chelsea indicator):

$$
y_i = \alpha + \beta_1 x_i + \beta_2 d_{Ci} + \epsilon_i
$$ {#eq-ch2-ancova}

For students in **Battersea** ($d_{Ci} = 0$):

$$
\text{E}(y_i \mid x_i, d_{Ci} = 0) = \alpha + \beta_1 x_i
$$

For students in **Chelsea** ($d_{Ci} = 1$):

$$
\text{E}(y_i \mid x_i, d_{Ci} = 1) = (\alpha + \beta_2) + \beta_1 x_i
$$

The key parameter is $\beta_2$: the adjusted difference between Chelsea and Battersea, **holding LRT constant**:

$$
\underbrace{(\alpha + \beta_1 x_i + \beta_2)}_{\text{Chelsea}} - \underbrace{(\alpha + \beta_1 x_i)}_{\text{Battersea}} = \beta_2
$$

::: {.callout-tip}
## Interpreting ANCOVA Coefficients
- $\alpha$: Expected GCSE score for a Battersea student with LRT = 0
- $\beta_1$: Expected change in GCSE per one-unit increase in LRT, **controlling for school**
- $\beta_2$: Expected difference in GCSE between Chelsea and Battersea, **controlling for LRT**

This "controlling for" language is the hallmark of multiple regression. Each coefficient is interpreted *holding all other variables constant*.
:::

::: {.callout-tip}
## Concept Check: What Does "Controlling For" Actually Mean?
Consider two students --- one from Battersea, one from Chelsea --- **both with LRT = 5**. The ANCOVA model predicts their GCSE scores differ by exactly $\hat{\beta}_2$, regardless of which LRT value we choose (as long as it is the same for both).

This is what "holding LRT constant" means: we compare groups *at the same level* of the covariate. It is a *statistical* operation, not a physical one --- we did not assign students to schools. Whether this adjusted difference has a **causal** interpretation depends on whether all confounders are accounted for (see @sec-ch2-ovb).
:::

### The General Multiple Regression Model

More generally, with $k$ predictors:

$$
y_i = \alpha + \beta_1 x_{1i} + \beta_2 x_{2i} + \cdots + \beta_k x_{ki} + \epsilon_i
$$ {#eq-ch2-mlr}

Each $\beta_j$ represents the change in $\text{E}(y)$ per one-unit increase in $x_j$, **holding all other predictors constant**. The error term $\epsilon_i$ has the same assumptions as in simple regression: mean zero, constant variance $\sigma^2$, independence, and (for inference) normality.


## Estimation and Interpretation {#sec-ch2-estimation}

### Fitting the ANCOVA Model

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-ancova-fit
#| code-fold: true
#| code-summary: "Show R Code"

# ANCOVA model: GCSE predicted by LRT and School
model_ancova <- lm(gcse ~ lrt + school_name, data = two_schools)

# Parameters with easystats
model_parameters(model_ancova) |> print_md()
```

### Output Interpretation
```{r}
#| ref.label: "ch2-ancova-fit"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Intercept** ($a$): Expected GCSE for a Battersea student with LRT = 0. Since LRT is standardized around 0, this is roughly the expected GCSE for an average-ability Battersea student
- **lrt** ($b_1$): Each additional LRT point is associated with a $b_1$-point increase in GCSE, controlling for school. This is the common slope shared by both schools under the parallel lines assumption
- **school_nameChelsea** ($b_2$): Comparing two students with the **same** LRT score, a Chelsea student is expected to score $b_2$ points higher (or lower, if negative) on GCSE than a Battersea student. This is the key ANCOVA quantity --- the "adjusted" school difference
- **SE and 95% CI**: The standard error and confidence interval for each coefficient. Narrower intervals indicate more precisely estimated effects
- Compare the Chelsea coefficient here to the unadjusted difference (next section) to see how controlling for LRT changes the estimated school effect

### Code Explanation
```{r}
#| ref.label: "ch2-ancova-fit"
#| eval: false
```
- `lm(gcse ~ lrt + school_name)`: Fits the ANCOVA model with LRT (continuous) and school (categorical) as predictors
- R automatically creates a dummy variable for `school_name`, with Battersea as the reference group
- The `school_nameChelsea` coefficient is $\hat{\beta}_2$: the LRT-adjusted school difference
- `print_md()` formats the output as a clean markdown table


:::

### Checking the Parallel Slopes Assumption

Before fitting the ANCOVA model, we should verify that the effect of LRT on GCSE is similar for both schools --- the **parallel slopes** assumption. We test this by fitting an interaction model and checking whether the interaction term is significant:

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-parallel-test
#| code-fold: true
#| code-summary: "Show R Code"

# Interaction model: allows different slopes per school
model_interact <- lm(gcse ~ lrt * school_name, data = two_schools)

# Compare ANCOVA (parallel) vs interaction (non-parallel) model
anova(model_ancova, model_interact)
```

### Output Interpretation
```{r}
#| ref.label: "ch2-parallel-test"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **F-statistic and p-value**: If $p > .05$, we fail to reject the null that the slopes are equal, supporting the parallel lines assumption
- This is consistent with the visual evidence in @fig-ch2-parallel, where the two school regression lines appear roughly parallel
- If the interaction were significant, we would need to use the interaction model from @sec-ch5 instead of ANCOVA

### Code Explanation
```{r}
#| ref.label: "ch2-parallel-test"
#| eval: false
```
- `lrt * school_name` creates both main effects and the LRT $\times$ School interaction
- The interaction term ($\beta_3$) represents the **difference in slopes** between Chelsea and Battersea
- `anova()` performs an $F$-test comparing the restricted (parallel) and unrestricted (interaction) models
- A **non-significant** $F$-test supports the simpler parallel-slopes ANCOVA model


:::

### Comparing Unadjusted vs Adjusted Differences

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-adjusted-means-compare
#| code-fold: true
#| code-summary: "Show R Code"

# Unadjusted model (school only)
model_unadj <- lm(gcse ~ school_name, data = two_schools)

# Compare coefficients side by side
bind_rows(
  tidy_model(model_unadj) |> mutate(model = "Unadjusted"),
  tidy_model(model_ancova) |> mutate(model = "Adjusted (ANCOVA)")
) |>
  filter(term == "school_nameChelsea") |>
  select(model, estimate, std.error, statistic, p.value) |>
  kable(digits = 3, col.names = c("Model", "Estimate", "SE", "t", "p-value"))
```

### Output Interpretation
```{r}
#| ref.label: "ch2-adjusted-means-compare"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Unadjusted estimate**: The raw difference in mean GCSE between Chelsea and Battersea, ignoring LRT. This confounds the school effect with intake ability differences
- **Adjusted estimate**: The school difference after controlling for LRT. This is the ANCOVA estimate --- the "fair" comparison of schools as if both had the same average intake ability
- The reduction from unadjusted to adjusted illustrates omitted variable bias in action (see @sec-ch2-ovb). The unadjusted model overestimates the school effect because it attributes to the school what is actually due to student ability

### Code Explanation
```{r}
#| ref.label: "ch2-adjusted-means-compare"
#| eval: false
```
- `tidy_model()` extracts coefficient estimates from each model in a tidy format
- The unadjusted difference (from a model with only `school_name`) is larger than the adjusted difference (from the ANCOVA model with both `school_name` and `lrt`)
- The reduction occurs because part of the school difference is "explained" by differences in LRT scores at intake


:::

### Adjusted Group Means

The LRT-adjusted mean GCSE score for each school is the predicted GCSE score at the overall mean LRT:

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-marginal-means
#| code-fold: true
#| code-summary: "Show R Code"

# Adjusted means using marginaleffects
avg_predictions(model_ancova, by = "school_name") |>
  as.data.frame() |>
  select(school_name, estimate, std.error, conf.low, conf.high) |>
  kable(digits = 2, col.names = c("School", "Adjusted Mean", "SE", "CI Lower", "CI Upper"))
```

### Output Interpretation
```{r}
#| ref.label: "ch2-marginal-means"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Estimate**: The LRT-adjusted (predicted) mean GCSE for each school. These represent "fair" comparisons because they account for differences in intake ability
- **Std. Error and 95% CI**: The precision of the adjusted mean estimates. Narrower intervals indicate more precisely estimated adjusted means
- The difference between the two adjusted means equals the `school_nameChelsea` coefficient from the ANCOVA model
- Compare these adjusted means to the raw means from the earlier table to see how controlling for LRT changes the picture

### Code Explanation
```{r}
#| ref.label: "ch2-marginal-means"
#| eval: false
```
- `avg_predictions()` from `marginaleffects` computes model-predicted means for each school, averaging over the observed distribution of LRT
- These are the "adjusted means" --- what each school's average GCSE would be if both schools had the same LRT distribution


:::

::: {.callout-tip}
## How Adjusted Means Are Computed
The adjusted mean for school $g$ is the predicted GCSE at the **grand mean** of LRT:

$$\hat{\mu}_g^{\text{adj}} = \hat{\alpha} + \hat{\beta}_1 \bar{x}_{\cdot} + \hat{\beta}_2 \cdot \mathbf{1}[g = \text{Chelsea}]$$

Both schools are evaluated at the same LRT value --- this is the source of "fairness" in the adjusted comparison.
:::

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-adjusted-means
#| fig-cap: "Unadjusted (raw) and LRT-adjusted mean GCSE scores. Arrows show how adjusting for intake ability changes the estimated school difference."
#| fig-alt: "Dot plot comparing unadjusted and LRT-adjusted mean GCSE scores for Battersea and Chelsea schools. Arrows connect each school's raw mean to its adjusted mean, showing how controlling for LRT intake ability shifts the estimated group means. The gap between schools changes after adjustment, illustrating that the unadjusted difference partly reflects pre-existing differences in intake scores rather than a pure school effect."
#| fig-width: 7
#| fig-height: 5
#| code-fold: true
#| code-summary: "Show R Code"

raw_means <- two_schools |>
  group_by(school_name) |>
  summarise(estimate = mean(gcse, na.rm = TRUE), .groups = "drop") |>
  mutate(type = "Unadjusted (raw)")

adj_means <- avg_predictions(model_ancova, by = "school_name") |>
  as.data.frame() |>
  select(school_name, estimate) |>
  mutate(type = "Adjusted (ANCOVA)")

means_df <- bind_rows(raw_means, adj_means) |>
  mutate(type = factor(type, levels = c("Unadjusted (raw)", "Adjusted (ANCOVA)")))

ggplot(means_df, aes(x = type, y = estimate, color = school_name, group = school_name)) +
  geom_line(linetype = "dashed", linewidth = 0.7, alpha = 0.6) +
  geom_point(size = 5) +
  geom_text(aes(label = round(estimate, 1)), vjust = -1, size = 3.8,
            fontface = "bold", show.legend = FALSE) +
  scale_color_manual(values = c(ber640_colors["primary"], ber640_colors["accent"])) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.15))) +
  labs(title = "How ANCOVA Adjusts the School Comparison",
       subtitle = "Controlling for LRT reduces the estimated Chelsea advantage",
       x = NULL, y = "Mean GCSE Score", color = "School")
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-adjusted-means"
#| code-fold: true
#| code-summary: "Show R Code"
```
- The **unadjusted** means reflect the raw school difference, which confounds the school effect with intake ability differences
- The **adjusted** means show the ANCOVA-estimated school difference after controlling for LRT
- The convergence of the two points (lines moving closer together) illustrates that part of the raw school difference was due to LRT differences, not the school itself
- The remaining gap between adjusted means equals the `school_nameChelsea` coefficient from the ANCOVA model

### Code Explanation
```{r}
#| ref.label: "fig-ch2-adjusted-means"
#| eval: false
```
- `raw_means` computes simple group averages (ignoring LRT); `adj_means` uses `avg_predictions()` from the ANCOVA model
- The dashed lines connecting each school's raw and adjusted mean show how the estimates shift after controlling for LRT
- `scale_color_manual()` applies the course color palette for consistency


:::

### Interactive Exploration: Adjusted vs. Unadjusted Means {.unnumbered}

::: {.callout-tip collapse="false"}
## Try It: How Covariates Confound Group Comparisons

Adjust the confounding strength to see how a covariate correlated with group membership biases unadjusted mean comparisons. The ANCOVA adjustment removes this bias by accounting for group differences on the covariate.
:::

```{ojs}
//| echo: false
import { ber640 as ancova_ber640, withAlpha as ancova_withAlpha,
         themeColors as ancova_themeColors, plotDefaults as ancova_plotDefaults,
         normalRandom as ancova_normalRandom, seq as ancova_seq,
         formatCoef as ancova_formatCoef } from "../ojs/shared.js"
```

```{ojs}
//| echo: false
viewof ancova_confounding = Inputs.range([0, 0.9], {
  value: 0.5, step: 0.05, label: "Confounding strength"
})

viewof ancova_trueEffect = Inputs.range([-2, 2], {
  value: 0.5, step: 0.1, label: "True group effect"
})
```

```{ojs}
//| echo: false
ancova_nPerGroup = 100
ancova_betaX = 0.8
ancova_confoundShift = ancova_confounding * 3

ancova_xA = ancova_normalRandom(ancova_nPerGroup, 0, 1, 42)
ancova_xB = ancova_normalRandom(ancova_nPerGroup, ancova_confoundShift, 1, 99)
ancova_errA = ancova_normalRandom(ancova_nPerGroup, 0, 1, 123)
ancova_errB = ancova_normalRandom(ancova_nPerGroup, 0, 1, 456)

ancova_yA = ancova_xA.map((x, i) => ancova_trueEffect * 0 + ancova_betaX * x + ancova_errA[i])
ancova_yB = ancova_xB.map((x, i) => ancova_trueEffect * 1 + ancova_betaX * x + ancova_errB[i])

ancova_data = {
  const rows = [];
  for (let i = 0; i < ancova_nPerGroup; i++) {
    rows.push({ group: "Group A", x: ancova_xA[i], y: ancova_yA[i] });
  }
  for (let i = 0; i < ancova_nPerGroup; i++) {
    rows.push({ group: "Group B", x: ancova_xB[i], y: ancova_yB[i] });
  }
  return rows;
}
```

```{ojs}
//| echo: false
ancova_meanYA = d3.mean(ancova_yA)
ancova_meanYB = d3.mean(ancova_yB)
ancova_meanXA = d3.mean(ancova_xA)
ancova_meanXB = d3.mean(ancova_xB)
ancova_meanXGrand = d3.mean([...ancova_xA, ...ancova_xB])

ancova_bWithin = {
  let ssxy = 0, ssxx = 0;
  for (let i = 0; i < ancova_nPerGroup; i++) {
    ssxy += (ancova_xA[i] - ancova_meanXA) * (ancova_yA[i] - ancova_meanYA);
    ssxx += (ancova_xA[i] - ancova_meanXA) ** 2;
  }
  for (let i = 0; i < ancova_nPerGroup; i++) {
    ssxy += (ancova_xB[i] - ancova_meanXB) * (ancova_yB[i] - ancova_meanYB);
    ssxx += (ancova_xB[i] - ancova_meanXB) ** 2;
  }
  return ssxy / ssxx;
}

ancova_adjMeanA = ancova_meanYA - ancova_bWithin * (ancova_meanXA - ancova_meanXGrand)
ancova_adjMeanB = ancova_meanYB - ancova_bWithin * (ancova_meanXB - ancova_meanXGrand)
ancova_unadjDiff = ancova_meanYB - ancova_meanYA
ancova_adjDiff = ancova_adjMeanB - ancova_adjMeanA
ancova_bias = ancova_unadjDiff - ancova_adjDiff
```

```{ojs}
//| echo: false
ancova_tc = ancova_themeColors()

ancova_scatterPlot = {
  const xAll = [...ancova_xA, ...ancova_xB];
  const yAll = [...ancova_yA, ...ancova_yB];
  const xMin = d3.min(xAll) - 0.5;
  const xMax = d3.max(xAll) + 0.5;
  const xLine = ancova_seq(xMin, xMax, 50);

  const interceptA = ancova_meanYA - ancova_bWithin * ancova_meanXA;
  const interceptB = ancova_meanYB - ancova_bWithin * ancova_meanXB;

  const lineA = xLine.map(x => ({ x, y: interceptA + ancova_bWithin * x, group: "Group A" }));
  const lineB = xLine.map(x => ({ x, y: interceptB + ancova_bWithin * x, group: "Group B" }));

  return Plot.plot(ancova_plotDefaults({
    width: 420,
    height: 360,
    color: { domain: ["Group A", "Group B"], range: [ancova_ber640.accent, ancova_ber640.secondary] },
    x: { label: "Covariate (X)", domain: [xMin, xMax] },
    y: { label: "Outcome (Y)" },
    marks: [
      Plot.dot(ancova_data, {
        x: "x", y: "y", fill: "group",
        fillOpacity: 0.45, r: 3.5, stroke: "none"
      }),
      Plot.line(lineA, { x: "x", y: "y", stroke: ancova_ber640.accent, strokeWidth: 2.5 }),
      Plot.line(lineB, { x: "x", y: "y", stroke: ancova_ber640.secondary, strokeWidth: 2.5 }),
      Plot.ruleX([ancova_meanXGrand], {
        stroke: ancova_tc.muted, strokeDasharray: "2 2", strokeOpacity: 0.6
      }),
      Plot.text([{ x: ancova_meanXGrand, y: d3.max(yAll) + 0.3 }], {
        x: "x", y: "y", text: ["X̄ grand"], fontSize: 10,
        fill: ancova_tc.text, textAnchor: "middle"
      })
    ]
  }));
}

ancova_barData = [
  { type: "Unadjusted", group: "Group A", mean: ancova_meanYA },
  { type: "Unadjusted", group: "Group B", mean: ancova_meanYB },
  { type: "Adjusted", group: "Group A", mean: ancova_adjMeanA },
  { type: "Adjusted", group: "Group B", mean: ancova_adjMeanB }
]

ancova_barPlot = {
  return Plot.plot(ancova_plotDefaults({
    width: 380,
    height: 360,
    x: { label: null, padding: 0.2 },
    y: { label: "Mean Outcome (Y)", grid: true },
    fx: { label: null, padding: 0.15 },
    color: { domain: ["Group A", "Group B"], range: [ancova_ber640.accent, ancova_ber640.secondary], legend: true },
    marks: [
      Plot.barY(ancova_barData, {
        fx: "type", x: "group", y: "mean", fill: "group", fillOpacity: 0.8
      }),
      Plot.ruleY([0], { stroke: ancova_tc.muted }),
      Plot.text(ancova_barData, {
        fx: "type", x: "group",
        y: d => d.mean >= 0 ? d.mean + 0.12 : d.mean - 0.12,
        text: d => ancova_formatCoef(d.mean, 2),
        fontSize: 11, fill: ancova_tc.text, textAnchor: "middle"
      })
    ]
  }));
}
```

```{ojs}
//| echo: false
ancova_panels = html`<div style="display:flex; flex-wrap:wrap; gap:24px; align-items:flex-start; justify-content:center;">
  <div>
    <div style="text-align:center; font-weight:600; margin-bottom:4px; color:${ancova_tc.text};">Scatter Plot with Regression Lines</div>
    ${ancova_scatterPlot}
  </div>
  <div>
    <div style="text-align:center; font-weight:600; margin-bottom:4px; color:${ancova_tc.text};">Unadjusted vs. Adjusted Means</div>
    ${ancova_barPlot}
  </div>
</div>`
```

```{ojs}
//| echo: false
ancova_readout = {
  const biasColor = Math.abs(ancova_bias) > 0.3 ? ancova_ber640.secondary : ancova_ber640.success;
  const row = (label, value, color) =>
    `<tr>
       <td style="padding:4px 12px 4px 0; font-weight:500;">${label}</td>
       <td style="padding:4px 0; font-variant-numeric:tabular-nums; color:${color || ancova_tc.text};">
         ${ancova_formatCoef(value, 3)}
       </td>
     </tr>`;

  return html`<div style="max-width:480px; margin:8px auto 0; padding:12px 16px;
    border:1px solid ${ancova_tc.muted}; border-radius:8px;
    background:${ancova_withAlpha(ancova_tc.bg, 0.6)}; color:${ancova_tc.text};">
    <table style="border-collapse:collapse; font-size:0.92em;">
      <thead>
        <tr><th style="text-align:left; padding-bottom:6px;" colspan="2">Summary of Group Differences</th></tr>
      </thead>
      <tbody>
        ${row("True effect (τ)", ancova_trueEffect, ancova_ber640.success)}
        ${row("Unadjusted difference (Ȳ_B − Ȳ_A)", ancova_unadjDiff)}
        ${row("Adjusted difference (ANCOVA)", ancova_adjDiff, ancova_ber640.accent)}
        ${row("Bias (Unadjusted − Adjusted)", ancova_bias, biasColor)}
        ${row("Pooled within-group slope (b_w)", ancova_bWithin)}
      </tbody>
    </table>
  </div>`;
}
```

### Visualizing the ANCOVA Model

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-ancova-plot
#| fig-cap: "ANCOVA model: parallel regression lines for Battersea and Chelsea. The vertical distance between the lines represents the LRT-adjusted school difference."
#| fig-alt: "Scatter plot of GCSE versus LRT with two parallel regression lines, one for Battersea and one for Chelsea. The lines share a common slope but differ in intercept, and a vertical arrow between them indicates the LRT-adjusted school difference. Shaded prediction interval bands surround each line, showing the range of plausible individual GCSE scores at each LRT value."
#| code-fold: true
#| code-summary: "Show R Code"

plot_predictions(model_ancova, condition = c("lrt", "school_name")) +
  scale_color_manual(values = c(ber640_colors["primary"], ber640_colors["accent"])) +
  scale_fill_manual(values = c(ber640_colors["primary"], ber640_colors["accent"])) +
  labs(
    title = "ANCOVA: Parallel Regression Lines",
    subtitle = "Vertical gap = adjusted school difference, controlling for LRT",
    x = "LRT Score",
    y = "Predicted GCSE Score",
    color = "School",
    fill = "School"
  )
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-ancova-plot"
#| code-fold: true
#| code-summary: "Show R Code"
```
- The two **parallel lines** confirm the ANCOVA model structure: same slope, different intercepts
- The **shaded ribbon** around each line shows the 95% confidence interval for the predicted mean at each LRT value
- The **vertical gap** between the lines is constant across all LRT values and equals the `school_nameChelsea` coefficient ($\hat{\beta}_2$)
- This visualization makes clear that ANCOVA answers: "At any given LRT score, how much do the schools differ in expected GCSE?"

### Code Explanation
```{r}
#| ref.label: "fig-ch2-ancova-plot"
#| eval: false
```
- `plot_predictions()` from `marginaleffects` visualizes the model predictions
- `condition = c("lrt", "school_name")` shows how predicted GCSE varies across LRT values, separately for each school
- The parallel lines reflect the ANCOVA assumption: the LRT slope is the same for both schools


:::


## R-squared and Multiple Correlation {#sec-ch2-r2}

The coefficient of determination generalizes straightforwardly to multiple regression. With $k$ predictors:

$$
R^2 = \frac{\text{MSS}}{\text{TSS}} = 1 - \frac{\text{SSE}}{\text{TSS}}
$$

The **multiple correlation** $R$ is the Pearson correlation between observed $y_i$ and predicted $\hat{y}_i$. The squared multiple correlation equals $R^2$.

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-r2-compare
#| code-fold: true
#| code-summary: "Show R Code"

# Compare R-squared across nested models
m0 <- lm(gcse ~ 1, data = two_schools)            # Intercept only
m1 <- lm(gcse ~ school_name, data = two_schools)   # School only
m2 <- lm(gcse ~ lrt, data = two_schools)            # LRT only
m3 <- model_ancova                                   # Both

# Names appear in output: m0 = intercept, m1 = school, m2 = LRT, m3 = ANCOVA
compare_performance(m0, m1, m2, m3) |> print_md()
```

### Output Interpretation
```{r}
#| ref.label: "ch2-r2-compare"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **$R^2$ progression**: As we add predictors, $R^2$ increases (or stays the same). The intercept-only model explains 0% of variance; adding school or LRT alone explains some; the ANCOVA model with both explains the most
- **AIC/BIC**: Lower values indicate better model fit, balancing complexity against explanatory power. The ANCOVA model should have the lowest values if both predictors contribute meaningfully
- **RMSE**: The root mean squared error decreases as the model improves --- predictions become more accurate
- The improvement from the school-only model to the ANCOVA model shows the value of controlling for LRT

### Code Explanation
```{r}
#| ref.label: "ch2-r2-compare"
#| eval: false
```
- `compare_performance()` from `easystats` provides a side-by-side comparison of model fit statistics
- We can see how much each predictor (and their combination) contributes to explaining variance in GCSE
- $R^2$ for the ANCOVA model (both predictors) is greater than for either predictor alone


:::


## Sequential and Partial Tests {#sec-ch2-tests}

When we have multiple predictors, there are different ways to partition the Model Sum of Squares. The two main approaches are:

### Type I (Sequential) Sums of Squares

Predictors are added one at a time in the order specified. The SS for each predictor represents its **additional** contribution given the predictors already in the model.

### Type III (Unique/Partial) Sums of Squares

Each predictor's SS represents its **unique** contribution, controlling for all other predictors. This is typically what we want for hypothesis testing in observational studies.

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-ss-types
#| code-fold: true
#| code-summary: "Show R Code"

# Type I (Sequential) SS - order matters!
anova(model_ancova)
```

### Output Interpretation
```{r}
#| ref.label: "ch2-ss-types"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Type I SS**: The SS for `lrt` is tested first (its total contribution), then `school_name` is tested for its *additional* contribution beyond LRT. Reversing the order would change the individual SS values
- **Type III SS**: Each predictor's SS reflects its *unique* contribution after controlling for the other. The results are the same regardless of the order in which variables are listed in the model formula
- **$F$-statistics and $p$-values**: In both types, these test whether each predictor significantly improves the model. A significant $p$-value for `school_name` in the Type III test confirms that schools differ in GCSE even after controlling for LRT

### Code Explanation
```{r}
#| ref.label: "ch2-ss-types"
#| eval: false
```
- `anova()` produces **Type I** (sequential) SS: each predictor is tested after adding it to all preceding predictors
- `car::Anova(..., type = 3)` produces **Type III** (unique) SS: each predictor is tested controlling for all others
- In balanced designs, Types I and III give the same results. In unbalanced designs (common in observational research), they can differ


:::

::: {.callout-note}
## Type I vs Type III Sums of Squares
| Feature | Type I (Sequential) | Type III (Unique) |
|:--------|:-------------------|:-----------------|
| Depends on order? | Yes | No |
| Tests for... | Additional contribution | Unique contribution |
| Best for... | Hierarchical model building | Hypothesis testing |
| Default in R | `anova()` | `car::Anova(type = 3)` |

**Rule of thumb**: Use Type III for testing individual predictors in observational studies, where there is no natural ordering of variables.
:::

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-ss-venn
#| fig-cap: "Variance partitioning in ANCOVA. Type I SS assigns shared variance to the first predictor entered; Type III SS tests only the unique portions."
#| fig-alt: "Venn diagram with two overlapping circles representing the variance explained by school and LRT. The overlap represents shared variance. Annotations contrast Type I sums of squares, which assign the shared portion to whichever predictor is entered first, with Type III sums of squares, which test only each predictor's unique contribution. The diagram clarifies why predictor order matters for Type I but not Type III tests."
#| fig-width: 8
#| fig-height: 5
#| code-fold: true
#| code-summary: "Show R Code"

ss_lrt_only <- summary(lm(gcse ~ lrt, data = two_schools))$r.squared
ss_school_only <- summary(lm(gcse ~ school_name, data = two_schools))$r.squared
ss_both <- summary(model_ancova)$r.squared

unique_lrt <- ss_both - ss_school_only
unique_school <- ss_both - ss_lrt_only
shared_r2 <- ss_lrt_only + ss_school_only - ss_both

# Draw Venn with annotated circles
theta <- seq(0, 2 * pi, length.out = 100)
circle_lrt <- tibble(x = -0.55 + 1.3 * cos(theta), y = 1.3 * sin(theta), group = "LRT")
circle_sch <- tibble(x = 0.55 + 1.3 * cos(theta), y = 1.3 * sin(theta), group = "School")

ggplot() +
  geom_polygon(data = circle_lrt, aes(x, y), fill = ber640_colors["primary"], alpha = 0.2) +
  geom_polygon(data = circle_sch, aes(x, y), fill = ber640_colors["accent"], alpha = 0.2) +
  annotate("text", x = -1.3, y = 0, label = paste0("LRT unique\n", round(unique_lrt * 100, 1), "%"),
           color = ber640_colors["primary"], size = 4.5, fontface = "bold") +
  annotate("text", x = 0, y = 0, label = paste0("Shared\n", round(shared_r2 * 100, 1), "%"),
           color = ber640_colors["success"], size = 4.5, fontface = "bold") +
  annotate("text", x = 1.3, y = 0, label = paste0("School unique\n", round(unique_school * 100, 1), "%"),
           color = ber640_colors["accent"], size = 4.5, fontface = "bold") +
  annotate("text", x = 0, y = -2, hjust = 0.5, size = 3.8, color = "grey30", fontface = "italic",
           label = "Type I: shared → first predictor entered\nType III: only unique portions tested") +
  coord_equal(xlim = c(-2.5, 2.5), ylim = c(-2.5, 2)) +
  labs(title = "Type I vs Type III: Who Gets Credit for Shared Variance?",
       subtitle = paste0("Total R² = ", round(ss_both * 100, 1), "%")) +
  theme_void(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        plot.subtitle = element_text(color = "grey40", hjust = 0.5))
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-ss-venn"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **LRT unique**: Variance in GCSE explained only by LRT (not school). This is the same regardless of Type I or Type III
- **School unique**: Variance in GCSE explained only by school (not LRT). Also the same for both SS types
- **Shared**: Variance that both LRT and school could explain. Type I assigns this to whichever predictor is entered first; Type III ignores it entirely when testing each predictor
- The total $R^2$ is the union of all three regions. The unexplained variance ($1 - R^2$) lies outside both circles

### Code Explanation
```{r}
#| ref.label: "fig-ch2-ss-venn"
#| eval: false
```
- We compute $R^2$ from three models (LRT-only, school-only, both) to partition variance into unique and shared components
- `unique_lrt` = total $R^2$ minus school-only $R^2$: the variance explained by LRT that school cannot explain
- `unique_school` = total $R^2$ minus LRT-only $R^2$: the variance explained by school that LRT cannot explain
- `shared_r2` = the overlap where both predictors compete for credit
- The Venn diagram uses `geom_polygon()` with parametric circles to visualize these regions


:::


## Omitted Variable Bias {#sec-ch2-ovb}

One of the most important concepts in applied statistics is **omitted variable bias (OVB)**. When a relevant variable is left out of the model, the estimated coefficients for the included variables may be biased.

### The OVB Formula

Suppose the true model is:

$$
y_i = \alpha + \beta_1 x_{1i} + \beta_2 x_{2i} + \epsilon_i
$$

But we estimate the misspecified model that omits $x_2$:

$$
y_i = \alpha^* + \beta_1^* x_{1i} + \epsilon_i^*
$$

The expected value of $b_1^*$ is:

$$
\text{E}(b_1^*) = \beta_1 + r_{21} \frac{s_2}{s_1} \beta_2 \neq \beta_1
$$ {#eq-ch2-ovb}

where $r_{21}$ is the correlation between $x_1$ and $x_2$, and $s_1, s_2$ are their standard deviations.

::: {.callout-important}
## When Does OVB Occur?
Omitted variable bias requires **both**:

1. The omitted variable $x_2$ affects $y$ (i.e., $\beta_2 \neq 0$)
2. The omitted variable $x_2$ is correlated with the included variable $x_1$ (i.e., $r_{21} \neq 0$)

If either condition fails, the bias is zero. Randomization can eliminate condition (2), which is why randomized experiments are the gold standard for causal inference.
:::

::: {.callout-tip}
## Predicting the Direction of OVB Bias
The sign of the bias term $r_{21} \cdot (s_2/s_1) \cdot \beta_2$ tells you whether omitting $x_2$ will make $b_1^*$ too large or too small:

| Sign of $r_{21}$ | Sign of $\beta_2$ | Direction of bias |
|:------------------|:-------------------|:-----------------|
| Positive | Positive | Upward (overestimate) |
| Positive | Negative | Downward (underestimate) |
| Negative | Positive | Downward (underestimate) |
| Negative | Negative | Upward (overestimate) |
:::

### OVB Mechanism

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-ovb-mechanism
#| fig-cap: "The two conditions for OVB. (A) School correlates with LRT (condition 2). (B) LRT predicts GCSE (condition 1). Both conditions hold, so omitting LRT biases the school coefficient."
#| fig-alt: "Two-panel figure illustrating the conditions for omitted variable bias. Panel A shows a box plot or scatter demonstrating that school is correlated with LRT intake scores, satisfying the second OVB condition. Panel B shows that LRT is a significant predictor of GCSE, satisfying the first condition. Because both conditions hold simultaneously, omitting LRT from a regression of GCSE on school produces a biased estimate of the school effect."
#| fig-width: 12
#| fig-height: 4
#| code-fold: true
#| code-summary: "Show R Code"

p_a <- ggplot(two_schools, aes(x = school_name, y = lrt, fill = school_name)) +
  geom_violin(alpha = 0.35, color = NA) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = c(ber640_colors["primary"], ber640_colors["accent"])) +
  labs(title = "A: Condition 2", subtitle = "School correlated with LRT",
       x = NULL, y = "LRT Score") +
  theme(legend.position = "none")

p_b <- ggplot(two_schools, aes(x = lrt, y = gcse)) +
  geom_point(alpha = 0.25, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = ber640_colors["secondary"], fill = ber640_colors["secondary"], alpha = 0.15) +
  labs(title = "B: Condition 1", subtitle = "LRT predicts GCSE",
       x = "LRT Score", y = "GCSE Score")

school_means <- two_schools |> group_by(school_name) |>
  summarise(mean_gcse = mean(gcse, na.rm = TRUE), .groups = "drop")

p_c <- ggplot(two_schools, aes(x = school_name, y = gcse, fill = school_name)) +
  geom_violin(alpha = 0.3, color = NA) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.7) +
  geom_point(data = school_means, aes(y = mean_gcse), shape = 18, size = 5, color = "white") +
  scale_fill_manual(values = c(ber640_colors["primary"], ber640_colors["accent"])) +
  labs(title = "C: Biased Result", subtitle = "Raw school gap (confounded)",
       x = NULL, y = "GCSE Score") +
  theme(legend.position = "none")

p_a | p_b | p_c
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-ovb-mechanism"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Panel A**: The LRT distributions clearly differ between schools. Chelsea students enter with higher prior achievement, confirming that school and LRT are correlated ($r_{21} \neq 0$)
- **Panel B**: The strong positive slope confirms that LRT is a meaningful predictor of GCSE ($\beta_2 \neq 0$). Both OVB conditions are satisfied
- **Panel C**: The raw school gap (white diamonds) overestimates the true school effect because it includes the confounding influence of LRT. This is exactly what the OVB formula (@eq-ch2-ovb) predicts

### Code Explanation
```{r}
#| ref.label: "fig-ch2-ovb-mechanism"
#| eval: false
```
- **Panel A** (Condition 2): A violin + boxplot showing that Chelsea students have higher LRT scores at intake than Battersea students. This establishes that school assignment is correlated with the omitted variable (LRT)
- **Panel B** (Condition 1): A scatterplot confirming that LRT strongly predicts GCSE scores. This establishes that the omitted variable affects the outcome
- **Panel C** (Biased Result): The raw GCSE distributions by school, with diamond markers at the group means. The apparent school gap is confounded --- it mixes the true school effect with pre-existing ability differences
- Both conditions are met, so omitting LRT from the model biases the school coefficient upward


:::

### OVB Demonstration

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-ovb-demo
#| code-fold: true
#| code-summary: "Show R Code"

# Demonstrate OVB: compare bivariate vs ANCOVA estimate for school effect
model_bivariate <- lm(gcse ~ school_name, data = two_schools)
model_adjusted  <- lm(gcse ~ school_name + lrt, data = two_schools)

bind_rows(
  tidy_model(model_bivariate) |> mutate(model = "Bivariate (omits LRT)"),
  tidy_model(model_adjusted)  |> mutate(model = "Adjusted (includes LRT)")
) |>
  filter(term != "(Intercept)") |>
  select(model, term, estimate, std.error) |>
  kable(digits = 3)
```

### Output Interpretation
```{r}
#| ref.label: "ch2-ovb-demo"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Bivariate estimate**: The raw school difference ignoring LRT. This overestimates the school effect because it includes the confounding influence of pre-existing ability differences
- **Adjusted estimate**: The school difference controlling for LRT. This is smaller, reflecting the removal of the confounding bias
- **The reduction** between the two estimates is the **omitted variable bias** --- the amount by which ignoring LRT distorts the school effect estimate
- Use the OVB formula (@eq-ch2-ovb) to verify: the predicted bias should match the observed difference between the two estimates

### Code Explanation
```{r}
#| ref.label: "ch2-ovb-demo"
#| eval: false
```
- The bivariate model overestimates the school effect because it confounds the school difference with pre-existing LRT differences
- In the adjusted model, the school coefficient is smaller, reflecting the "true" school difference after accounting for intake ability
- This is OVB in action: LRT is a confounder (it affects GCSE and is correlated with school)


:::

### OVB Numerical Verification

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-ovb-verify
#| code-fold: true
#| code-summary: "Show R Code"

# Verify the OVB formula numerically
d_chelsea <- as.numeric(two_schools$school_name == "Chelsea")
r21 <- cor(d_chelsea, two_schools$lrt, use = "complete.obs")
s1  <- sd(d_chelsea)
s2  <- sd(two_schools$lrt, na.rm = TRUE)
b2  <- coef(model_adjusted)["lrt"]

predicted_bias <- r21 * (s2 / s1) * b2
observed_bias  <- coef(model_bivariate)["school_nameChelsea"] -
                  coef(model_adjusted)["school_nameChelsea"]

cat("Predicted bias (from OVB formula):", round(predicted_bias, 3), "\n")
cat("Observed bias (actual difference): ", round(observed_bias,  3), "\n")
```

### Output Interpretation
```{r}
#| ref.label: "ch2-ovb-verify"
#| code-fold: true
#| code-summary: "Show R Code"
```
- The predicted bias from the OVB formula exactly matches the observed difference between the bivariate and adjusted estimates. This confirms that the formula is not an approximation --- it is an algebraic identity in OLS regression
- The positive bias means the bivariate model **overestimates** the school effect, consistent with the OVB direction table: $r_{21} > 0$ (Chelsea students have higher LRT) and $\beta_2 > 0$ (higher LRT predicts higher GCSE), so the bias is upward
- This numerical verification makes the abstract OVB formula concrete and verifiable

### Code Explanation
```{r}
#| ref.label: "ch2-ovb-verify"
#| eval: false
```
- We compute each component of the OVB formula: the correlation between the school dummy and LRT ($r_{21}$), the standard deviations ($s_1, s_2$), and the LRT coefficient ($\beta_2$)
- `predicted_bias` applies the formula: $r_{21} \cdot (s_2/s_1) \cdot \beta_2$
- `observed_bias` is the actual difference between the bivariate and adjusted school coefficients
- If the formula is correct, these two values should be identical (up to rounding)


:::


## Confounding and Endogeneity {#sec-ch2-confounding}

The residual $\epsilon_i$ represents the combined effects of all omitted variables plus measurement error. When a confounder $x_2$ is omitted:

$$
\epsilon_i^* = \beta_2 x_{2i} + \epsilon_i
$$

Since $x_2$ is correlated with $x_1$ (and $x_2$ is now part of the error term), $\epsilon_i^*$ becomes correlated with $x_1$. In econometrics, a variable correlated with the error term is called **endogenous**, and its coefficient will be biased.

::: {.callout-warning}
## Confounders vs Mediators
Not every omitted variable is a confounder. A **mediator** is a variable on the causal pathway between $x$ and $y$:

$$
\text{Father's education} \rightarrow \text{Son's education} \rightarrow \text{Son's income}
$$

If we control for the mediator (son's education), we would remove the very effect we want to measure. **Never control for mediators** if the goal is to estimate the total effect of the predictor.

Key rule: Only control for variables that are **common causes** of both the predictor and the outcome (confounders), not variables that are **consequences** of the predictor (mediators).
:::

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-dag
#| fig-cap: "Confounder vs mediator. (A) LRT is a common cause (confounder) — control for it. (B) Teaching quality is on the causal pathway (mediator) — do NOT control for it."
#| fig-alt: "Two directed acyclic graph panels contrasting confounders and mediators. Panel A shows LRT as a common cause of both school assignment and GCSE scores, making it a confounder that should be controlled for. Panel B shows teaching quality as a variable on the causal pathway from school to GCSE, making it a mediator that should not be controlled for. The diagrams illustrate when statistical adjustment is appropriate versus when it blocks the causal effect of interest."
#| fig-width: 10
#| fig-height: 4
#| code-fold: true
#| code-summary: "Show R Code"

dag_theme <- theme_void(base_size = 13) +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        plot.subtitle = element_text(color = "grey40", hjust = 0.5, size = 11))

p_conf <- ggplot() +
  annotate("label", x = 0.5, y = 0.85, label = "LRT\n(Confounder)",
           fill = ber640_colors["warning"], color = "white", size = 4, fontface = "bold", label.size = 0) +
  annotate("label", x = 0.1, y = 0.2, label = "School",
           fill = ber640_colors["primary"], color = "white", size = 4, fontface = "bold", label.size = 0) +
  annotate("label", x = 0.9, y = 0.2, label = "GCSE",
           fill = ber640_colors["accent"], color = "white", size = 4, fontface = "bold", label.size = 0) +
  annotate("segment", x = 0.42, xend = 0.18, y = 0.75, yend = 0.33,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ber640_colors["warning"], linewidth = 1) +
  annotate("segment", x = 0.58, xend = 0.82, y = 0.75, yend = 0.33,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ber640_colors["warning"], linewidth = 1) +
  annotate("segment", x = 0.22, xend = 0.78, y = 0.2, yend = 0.2,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ber640_colors["primary"], linewidth = 1.2) +
  annotate("text", x = 0.5, y = 0.05, label = "Control for LRT \u2713",
           color = ber640_colors["success"], size = 3.5, fontface = "italic") +
  xlim(0, 1) + ylim(0, 1) +
  labs(title = "A: Confounder (control for it)") + dag_theme

p_med <- ggplot() +
  annotate("label", x = 0.1, y = 0.5, label = "School",
           fill = ber640_colors["primary"], color = "white", size = 4, fontface = "bold", label.size = 0) +
  annotate("label", x = 0.5, y = 0.5, label = "Teaching\nQuality",
           fill = ber640_colors["warning"], color = "white", size = 4, fontface = "bold", label.size = 0) +
  annotate("label", x = 0.9, y = 0.5, label = "GCSE",
           fill = ber640_colors["accent"], color = "white", size = 4, fontface = "bold", label.size = 0) +
  annotate("segment", x = 0.22, xend = 0.38, y = 0.5, yend = 0.5,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ber640_colors["primary"], linewidth = 1.2) +
  annotate("segment", x = 0.62, xend = 0.78, y = 0.5, yend = 0.5,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
           color = ber640_colors["warning"], linewidth = 1.2) +
  annotate("text", x = 0.5, y = 0.25, label = "Do NOT control for mediator \u2717",
           color = ber640_colors["secondary"], size = 3.5, fontface = "italic") +
  xlim(0, 1) + ylim(0, 1) +
  labs(title = "B: Mediator (do NOT control)") + dag_theme

p_conf | p_med
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-dag"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Panel A**: LRT is a confounder because it causes both school selection (higher-ability students attend Chelsea) and GCSE outcomes. Controlling for LRT removes the spurious association and reveals the direct school effect
- **Panel B**: Teaching quality is a mediator --- it is one of the *mechanisms* through which schools affect GCSE scores. If we controlled for teaching quality, we would remove part of the genuine school effect, underestimating the total impact of school choice
- The key distinction: confounders create **spurious** associations (control for them); mediators transmit **genuine** effects (do not control for them)

### Code Explanation
```{r}
#| ref.label: "fig-ch2-dag"
#| eval: false
```
- We use `annotate("label")` and `annotate("segment")` to draw DAG-style diagrams without external packages
- **Panel A** (Confounder): LRT sits above School and GCSE with arrows pointing down to both, forming a "fork" structure. This is the classic confounder pattern --- LRT is a common cause
- **Panel B** (Mediator): Teaching Quality sits between School and GCSE on the causal pathway. Controlling for it would block the very effect we want to measure
- `arrow()` creates directional edges; `label.size = 0` removes the border around node labels


:::

::: {.callout-warning}
## Statistical Control Is Not Causal Control
ANCOVA adjusts for **measured** confounders. But if Chelsea also enrolls students with higher parental SES, better peer effects, or more experienced teachers --- and these variables are not in the model --- the adjusted school difference still mixes school effects with unmeasured confounders.

The ANCOVA estimate is causal only under the **strong ignorability assumption**: that the model includes *all* common causes of school selection and GCSE outcomes. In observational education data, this is rarely guaranteed. This is why randomized experiments are preferred for causal claims.
:::


## Collinearity {#sec-ch2-collinearity}

When two predictors are highly correlated, they contain overlapping information about $y$. This creates **collinearity**, which inflates the standard errors of the regression coefficients.

### The Effect of Collinearity

::: {.panel-tabset}
### Output
```{r}
#| label: ch2-vif
#| code-fold: true
#| code-summary: "Show R Code"

# Check collinearity with VIF
check_collinearity(model_ancova)
```

### Output Interpretation
```{r}
#| ref.label: "ch2-vif"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **VIF**: A VIF of 1 means no collinearity; higher values indicate more overlap between predictors. For example, VIF = 2 means the standard error is $\sqrt{2} \approx 1.41$ times larger than it would be without collinearity
- **Tolerance**: The reciprocal of VIF ($1/\text{VIF}$). Values below 0.2 signal concern
- In our ANCOVA model, the VIF values should be low because LRT (continuous) and school (categorical) capture different types of information. High VIF would suggest that the two predictors are so correlated that their individual effects are hard to disentangle

### Code Explanation
```{r}
#| ref.label: "ch2-vif"
#| eval: false
```
- `check_collinearity()` from `performance` computes the **Variance Inflation Factor (VIF)** for each predictor
- VIF measures how much the variance of a coefficient is inflated due to correlation with other predictors
- **Rule of thumb**: VIF > 5 suggests moderate collinearity; VIF > 10 suggests serious collinearity


:::

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-collinearity
#| fig-cap: "Collinearity inflates standard errors. As predictor correlation increases, VIF grows and the standard error multiplier (\u221AVIF) shows how much less precise our estimates become."
#| fig-alt: "Line plot with predictor correlation on the horizontal axis and variance inflation factor on the vertical axis. As the correlation between predictors increases toward one, VIF rises sharply, and a second curve shows the square root of VIF representing the standard error multiplier. The plot demonstrates that even moderate collinearity noticeably inflates standard errors, and high collinearity makes coefficient estimates very imprecise."
#| fig-width: 10
#| fig-height: 4.5
#| code-fold: true
#| code-summary: "Show R Code"

r_seq <- seq(0, 0.99, by = 0.005)
vif_seq <- 1 / (1 - r_seq^2)
r_actual <- abs(cor(two_schools$lrt, as.numeric(two_schools$school_name) - 1, use = "complete.obs"))
vif_actual <- 1 / (1 - r_actual^2)

col_df <- tibble(r = r_seq, vif = vif_seq, se_mult = sqrt(vif_seq))

p1 <- ggplot(col_df, aes(x = r, y = vif)) +
  geom_line(color = ber640_colors["secondary"], linewidth = 1.2) +
  geom_vline(xintercept = r_actual, linetype = "dashed", color = ber640_colors["primary"]) +
  geom_point(aes(x = r_actual, y = vif_actual), color = ber640_colors["primary"], size = 4) +
  annotate("text", x = r_actual + 0.05, y = vif_actual + 1,
           label = paste0("Our model\nVIF = ", round(vif_actual, 2)),
           color = ber640_colors["primary"], size = 3.5, hjust = 0) +
  coord_cartesian(ylim = c(1, 15)) +
  labs(title = "A: VIF grows with predictor correlation",
       x = "|r| between predictors", y = "VIF")

p2 <- ggplot(col_df, aes(x = r, y = se_mult)) +
  geom_line(color = ber640_colors["accent"], linewidth = 1.2) +
  geom_hline(yintercept = 1, linetype = "dotted", color = "grey50") +
  geom_vline(xintercept = r_actual, linetype = "dashed", color = ber640_colors["primary"]) +
  geom_point(aes(x = r_actual, y = sqrt(vif_actual)), color = ber640_colors["primary"], size = 4) +
  coord_cartesian(ylim = c(1, 4)) +
  labs(title = "B: SE inflation (\u221AVIF)",
       x = "|r| between predictors", y = "SE multiplier")

p1 | p2
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-collinearity"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Panel A**: Our ANCOVA model (marked point) has low VIF, confirming that LRT and school are not highly collinear. The curve shows how dramatically VIF increases beyond $|r| = 0.8$
- **Panel B**: The SE multiplier for our model is close to 1, meaning collinearity has minimal impact on our standard errors. At $|r| = 0.9$, the SE would be more than doubled
- These plots illustrate why VIF > 10 (corresponding to $|r| > 0.95$) is considered problematic: standard errors become so inflated that coefficients are essentially uninterpretable

### Code Explanation
```{r}
#| ref.label: "fig-ch2-collinearity"
#| eval: false
```
- We compute VIF = $1/(1-r^2)$ across a range of correlations from 0 to 0.99
- **Panel A** shows VIF on the y-axis: it grows slowly at first, then explodes as $|r|$ approaches 1
- **Panel B** shows the SE multiplier ($\sqrt{\text{VIF}}$): this is how much wider the confidence interval becomes due to collinearity
- The dashed vertical line and point mark where our ANCOVA model falls on the curve


:::

::: {.callout-note}
## The Bias-Variance Tradeoff
Collinearity illustrates a fundamental tradeoff in regression:

- **Including more predictors**: Reduces bias (avoids OVB) but increases variance (larger standard errors)
- **Excluding correlated predictors**: Reduces variance (smaller standard errors) but may introduce bias (OVB)

The best approach depends on the research question. For causal inference, bias is usually a bigger concern than variance, so we prefer to include important confounders even if they create some collinearity.
:::

### Interactive Exploration: Collinearity Explorer {.unnumbered}

::: {.callout-tip collapse="false"}
## Try It: How Correlated Predictors Inflate Uncertainty

Drag the correlation slider to see how multicollinearity between two predictors inflates standard errors and widens confidence intervals. Notice that the coefficient *estimates* stay close to the true values --- collinearity does not create bias --- but the *precision* deteriorates rapidly as $r$ approaches 1.
:::

```{ojs}
//| echo: false

import { ber640 as collin_ber640, withAlpha as collin_withAlpha,
         themeColors as collin_themeColors, plotDefaults as collin_plotDefaults,
         normalRandom as collin_normalRandom, seq as collin_seq,
         formatCoef as collin_formatCoef } from "../ojs/shared.js"
```

```{ojs}
//| echo: false

viewof collin_r = Inputs.range([0, 0.99], {
  value: 0.3,
  step: 0.01,
  label: "Predictor correlation r(X₁, X₂)",
  width: 400
})
```

```{ojs}
//| echo: false

// --- Data generation (n = 100, all seeded) ---
collin_n = 100

collin_x1_raw = collin_normalRandom(collin_n, 0, 1, 42)
collin_eps = collin_normalRandom(collin_n, 0, 1, 43)
collin_ey = collin_normalRandom(collin_n, 0, 1, 44)

// X₂ = r * X₁ + sqrt(1 - r²) * ε
collin_x2 = {
  const r = collin_r;
  const s = Math.sqrt(1 - r * r);
  return collin_x1_raw.map((x1, i) => r * x1 + s * collin_eps[i]);
}

// Y = 1.0 * X₁ + 1.0 * X₂ + ε_y
collin_y = collin_x1_raw.map((x1, i) => 1.0 * x1 + 1.0 * collin_x2[i] + collin_ey[i])
```

```{ojs}
//| echo: false

// --- Multiple regression via normal equations (3×3 system) ---
collin_reg = {
  const n = collin_n;
  const x1 = collin_x1_raw;
  const x2 = collin_x2;
  const y = collin_y;

  // X'X elements (3×3: intercept, x1, x2)
  let s1 = 0, s2 = 0, s11 = 0, s22 = 0, s12 = 0;
  let sy = 0, s1y = 0, s2y = 0;

  for (let i = 0; i < n; i++) {
    s1 += x1[i]; s2 += x2[i];
    s11 += x1[i] * x1[i]; s22 += x2[i] * x2[i]; s12 += x1[i] * x2[i];
    sy += y[i]; s1y += x1[i] * y[i]; s2y += x2[i] * y[i];
  }

  // X'X matrix (row-major 3×3)
  const XtX = [
    n,   s1,  s2,
    s1,  s11, s12,
    s2,  s12, s22
  ];

  // X'Y vector
  const XtY = [sy, s1y, s2y];

  // Invert 3×3 via cofactors
  function inv3(m) {
    const [a,b,c, d,e,f, g,h,k] = m;
    const det = a*(e*k - f*h) - b*(d*k - f*g) + c*(d*h - e*g);
    if (Math.abs(det) < 1e-15) return null;
    const invDet = 1 / det;
    return [
      (e*k - f*h)*invDet, (c*h - b*k)*invDet, (b*f - c*e)*invDet,
      (f*g - d*k)*invDet, (a*k - c*g)*invDet, (c*d - a*f)*invDet,
      (d*h - e*g)*invDet, (b*g - a*h)*invDet, (a*e - b*d)*invDet
    ];
  }

  const XtXinv = inv3(XtX);
  if (!XtXinv) return { b0: NaN, b1: NaN, b2: NaN, se1: NaN, se2: NaN, vif: NaN };

  // Coefficients: (X'X)⁻¹ X'Y
  const b0 = XtXinv[0]*XtY[0] + XtXinv[1]*XtY[1] + XtXinv[2]*XtY[2];
  const b1 = XtXinv[3]*XtY[0] + XtXinv[4]*XtY[1] + XtXinv[5]*XtY[2];
  const b2 = XtXinv[6]*XtY[0] + XtXinv[7]*XtY[1] + XtXinv[8]*XtY[2];

  // Residual variance: σ² = SSE / (n - 3)
  let sse = 0;
  for (let i = 0; i < n; i++) {
    const resid = y[i] - (b0 + b1 * x1[i] + b2 * x2[i]);
    sse += resid * resid;
  }
  const sigma2 = sse / (n - 3);

  // SE from diagonal of (X'X)⁻¹ * σ²
  const se1 = Math.sqrt(XtXinv[4] * sigma2);
  const se2 = Math.sqrt(XtXinv[8] * sigma2);

  // VIF
  const r = collin_r;
  const vif = 1 / (1 - r * r);

  return { b0, b1, b2, se1, se2, vif, sigma2 };
}
```

```{ojs}
//| echo: false

// --- Coefficient plot: β̂₁ and β̂₂ with 95% CIs ---
collin_coefPlot = {
  const t = collin_themeColors();
  const reg = collin_reg;

  const coefData = [
    {
      label: "β̂₁",
      estimate: reg.b1,
      lo: reg.b1 - 1.96 * reg.se1,
      hi: reg.b1 + 1.96 * reg.se1,
      se: reg.se1,
      trueVal: 1.0,
      color: collin_ber640.accent,
      y: 1
    },
    {
      label: "β̂₂",
      estimate: reg.b2,
      lo: reg.b2 - 1.96 * reg.se2,
      hi: reg.b2 + 1.96 * reg.se2,
      se: reg.se2,
      trueVal: 1.0,
      color: collin_ber640.secondary,
      y: 0
    }
  ];

  // Compute dynamic x-axis range
  const allVals = coefData.flatMap(d => [d.lo, d.hi, d.trueVal]);
  const xMin = Math.min(...allVals) - 0.3;
  const xMax = Math.max(...allVals) + 0.3;

  return Plot.plot(collin_plotDefaults({
    width: 640,
    height: 220,
    marginLeft: 65,
    marginTop: 25,
    marginBottom: 40,
    x: {
      label: "Coefficient value",
      domain: [xMin, xMax],
      grid: true
    },
    y: {
      domain: [-0.5, 1.5],
      axis: null
    },
    marks: [
      // True value reference line
      Plot.ruleX([1.0], {
        stroke: t.muted,
        strokeWidth: 2,
        strokeDasharray: "6,4"
      }),

      // True value label
      Plot.text([{ x: 1.0, y: 1.5 }], {
        x: "x", y: "y",
        text: d => "True β = 1.0",
        fill: t.muted,
        fontSize: 11,
        dy: -5
      }),

      // CI bars
      Plot.link(coefData, {
        x1: "lo",
        x2: "hi",
        y1: "y",
        y2: "y",
        stroke: "color",
        strokeWidth: 4,
        strokeLinecap: "round"
      }),

      // Point estimates
      Plot.dot(coefData, {
        x: "estimate",
        y: "y",
        r: 7,
        fill: "color",
        stroke: t.bg,
        strokeWidth: 2
      }),

      // Labels on left
      Plot.text(coefData, {
        x: d => xMin + 0.05,
        y: "y",
        text: "label",
        fill: "color",
        fontWeight: 700,
        fontSize: 14,
        textAnchor: "start",
        dx: 5
      })
    ]
  }));
}
```

```{ojs}
//| echo: false

// --- VIF indicator display ---
collin_vifDisplay = {
  const t = collin_themeColors();
  const vif = collin_reg.vif;

  // Color coding
  let vifColor, vifLabel, vifClass;
  if (vif < 5) {
    vifColor = collin_ber640.success;
    vifLabel = "Low concern";
    vifClass = "ojs-readout-good";
  } else if (vif < 10) {
    vifColor = collin_ber640.warning;
    vifLabel = "Moderate concern";
    vifClass = "ojs-readout-warning";
  } else {
    vifColor = collin_ber640.secondary;
    vifLabel = "Severe collinearity";
    vifClass = "ojs-readout-bad";
  }

  return html`<div style="
    text-align: center;
    padding: 0.75rem 1rem;
    margin: 0.5rem auto;
    max-width: 320px;
    border-radius: 8px;
    border: 2px solid ${vifColor};
    background: ${collin_withAlpha(vifColor, 0.08)};
    transition: all 0.3s ease;
  ">
    <div style="font-size: 0.8rem; font-weight: 600; color: ${t.muted}; text-transform: uppercase; letter-spacing: 0.05em;">
      Variance Inflation Factor
    </div>
    <div style="font-size: 2.2rem; font-weight: 800; color: ${vifColor}; font-family: 'SFMono-Regular', Menlo, monospace; line-height: 1.2; margin: 0.25rem 0;">
      ${vif > 999 ? "∞" : collin_formatCoef(vif, vif < 10 ? 2 : 1)}
    </div>
    <div style="font-size: 0.85rem; font-weight: 600; color: ${vifColor};">
      ${vifLabel}
    </div>
  </div>`;
}
```

```{ojs}
//| echo: false

// --- Readout cards ---
collin_readouts = {
  const t = collin_themeColors();
  const reg = collin_reg;
  const ci1Width = 2 * 1.96 * reg.se1;
  const ci2Width = 2 * 1.96 * reg.se2;

  const vifClass = reg.vif < 5 ? "ojs-readout-good" : (reg.vif < 10 ? "ojs-readout-warning" : "ojs-readout-bad");

  return html`<div class="ojs-readouts">
    <span class="ojs-readout">
      <span class="ojs-readout-label">β̂₁ =</span>
      <span class="ojs-readout-value" style="color: ${collin_ber640.accent};">${collin_formatCoef(reg.b1, 3)} ± ${collin_formatCoef(reg.se1, 3)}</span>
      <span class="ojs-readout-label" style="font-size:0.75rem;">(CI width: ${collin_formatCoef(ci1Width, 2)})</span>
    </span>
    <span class="ojs-readout">
      <span class="ojs-readout-label">β̂₂ =</span>
      <span class="ojs-readout-value" style="color: ${collin_ber640.secondary};">${collin_formatCoef(reg.b2, 3)} ± ${collin_formatCoef(reg.se2, 3)}</span>
      <span class="ojs-readout-label" style="font-size:0.75rem;">(CI width: ${collin_formatCoef(ci2Width, 2)})</span>
    </span>
    <span class="ojs-readout ${vifClass}">
      <span class="ojs-readout-label">VIF =</span>
      <span class="ojs-readout-value">${reg.vif > 999 ? "∞" : collin_formatCoef(reg.vif, 2)}</span>
    </span>
    <span class="ojs-readout ojs-readout-info">
      <span class="ojs-readout-label">r =</span>
      <span class="ojs-readout-value">${collin_formatCoef(collin_r, 2)}</span>
    </span>
  </div>`;
}
```

```{ojs}
//| echo: false

// --- Dynamic narrative ---
collin_narrative = {
  const t = collin_themeColors();
  const r = collin_r;
  const reg = collin_reg;

  let msg;
  if (r < 0.3) {
    msg = `With a low correlation of ${collin_formatCoef(r, 2)}, the two predictors carry mostly independent information. Standard errors are near their minimum, and both confidence intervals tightly bracket the true value of 1.0.`;
  } else if (r < 0.6) {
    msg = `At r = ${collin_formatCoef(r, 2)}, some overlap between the predictors begins to appear. The VIF of ${collin_formatCoef(reg.vif, 2)} indicates that standard errors are inflated by a factor of ${collin_formatCoef(Math.sqrt(reg.vif), 2)} compared to the zero-correlation baseline. The estimates remain close to 1.0, but notice the confidence intervals starting to widen.`;
  } else if (r < 0.8) {
    msg = `At r = ${collin_formatCoef(r, 2)}, collinearity is becoming noticeable. VIF = ${collin_formatCoef(reg.vif, 1)} means standard errors are ${collin_formatCoef(Math.sqrt(reg.vif), 2)}× larger than they would be without correlation. The coefficient estimates are still unbiased, but the wider CIs reflect our growing uncertainty about each predictor's unique contribution.`;
  } else if (r < 0.95) {
    msg = `At r = ${collin_formatCoef(r, 2)}, collinearity is severe. VIF = ${collin_formatCoef(reg.vif, 1)} means standard errors are ${collin_formatCoef(Math.sqrt(reg.vif), 1)}× inflated. The confidence intervals are very wide — the model struggles to separate the individual effects of X₁ and X₂ because they are so similar. Yet the estimates remain centered near the true values: collinearity does not cause bias.`;
  } else {
    msg = `With r = ${collin_formatCoef(r, 2)}, the predictors are nearly identical. VIF = ${reg.vif > 999 ? "∞" : collin_formatCoef(reg.vif, 0)} pushes standard errors to extreme levels. The confidence intervals span an enormous range, making both coefficients practically uninterpretable — even though the estimates are still unbiased in expectation.`;
  }

  return html`<div style="
    font-size: 0.9rem;
    line-height: 1.55;
    color: ${t.text};
    padding: 0.6rem 0.8rem;
    margin-top: 0.5rem;
    background: ${t.isDark ? 'rgba(255,255,255,0.04)' : 'rgba(0,0,0,0.02)'};
    border-left: 3px solid ${collin_ber640.accent};
    border-radius: 0 6px 6px 0;
  ">${msg}</div>`;
}
```

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch2-bias-variance
#| fig-cap: "Bias-variance tradeoff via simulation. Left: the biased estimator (omitting LRT) consistently overestimates; the unbiased ANCOVA estimator is centered on truth but wider. Right: MSE = Bias² + Variance shows ANCOVA wins in total error."
#| fig-alt: "Two-panel simulation figure illustrating the bias-variance tradeoff. The left panel shows overlapping density curves for two estimators: the biased estimator that omits LRT is narrow but shifted away from the true value, while the unbiased ANCOVA estimator is centered on the truth but wider. The right panel displays a stacked bar chart decomposing mean squared error into bias squared and variance components for each estimator, showing that ANCOVA achieves lower total MSE despite its greater variance."
#| fig-width: 10
#| fig-height: 4.5
#| code-fold: true
#| code-summary: "Show R Code"

set.seed(42)
alpha_true <- coef(model_ancova)["(Intercept)"]
b_lrt <- coef(model_ancova)["lrt"]
b_school <- coef(model_ancova)["school_nameChelsea"]
sigma_true <- sigma(model_ancova)
n <- nrow(two_schools)

sim_results <- replicate(500, {
  y_sim <- alpha_true + b_lrt * two_schools$lrt +
    b_school * (as.numeric(two_schools$school_name) - 1) + rnorm(n, 0, sigma_true)
  df_sim <- two_schools |> mutate(gcse_sim = y_sim)
  c(biased = unname(coef(lm(gcse_sim ~ school_name, data = df_sim))["school_nameChelsea"]),
    unbiased = unname(coef(lm(gcse_sim ~ school_name + lrt, data = df_sim))["school_nameChelsea"]))
}, simplify = TRUE) |> t() |> as_tibble()

sim_long <- sim_results |>
  pivot_longer(everything(), names_to = "estimator", values_to = "estimate") |>
  mutate(estimator = factor(estimator,
    levels = c("biased", "unbiased"),
    labels = c("Biased (omits LRT)", "Unbiased (ANCOVA)")))

p_dist <- ggplot(sim_long, aes(x = estimate, fill = estimator, color = estimator)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  geom_vline(xintercept = b_school, linewidth = 1, color = "grey20") +
  annotate("text", x = b_school + 0.3, y = Inf, label = paste0("True \u03b2\u2082 = ", round(b_school, 2)),
           vjust = 1.5, color = "grey20", fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c(ber640_colors["secondary"], ber640_colors["accent"])) +
  scale_color_manual(values = c(ber640_colors["secondary"], ber640_colors["accent"])) +
  labs(title = "A: Sampling Distributions (500 sims)",
       x = "Estimated School Coefficient", y = "Density", fill = NULL, color = NULL)

sim_summary <- sim_long |>
  group_by(estimator) |>
  summarise(bias2 = (mean(estimate) - b_school)^2, variance = var(estimate), .groups = "drop") |>
  pivot_longer(c(bias2, variance), names_to = "component", values_to = "value") |>
  mutate(component = factor(component, levels = c("variance", "bias2"), labels = c("Variance", "Bias\u00b2")))

p_mse <- ggplot(sim_summary, aes(x = estimator, y = value, fill = component)) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("Bias\u00b2" = ber640_colors["secondary"], "Variance" = ber640_colors["warning"])) +
  labs(title = "B: MSE = Bias\u00b2 + Variance", x = NULL, y = "MSE", fill = NULL)

p_dist | p_mse
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch2-bias-variance"
#| code-fold: true
#| code-summary: "Show R Code"
```
- **Panel A**: The biased estimator (red) is shifted to the right of the true value --- it consistently overestimates the school effect because it absorbs the LRT confounding. The unbiased ANCOVA estimator (blue) is centered on the truth
- **Panel B**: Although the unbiased estimator has slightly more variance (it estimates an additional parameter), its total MSE is much smaller because it has near-zero bias. The biased estimator's large Bias² component dominates its MSE
- This demonstrates why controlling for confounders is usually worth the small increase in variance: bias reduction outweighs the variance cost

### Code Explanation
```{r}
#| ref.label: "fig-ch2-bias-variance"
#| eval: false
```
- We simulate 500 datasets from the true ANCOVA model (using estimated parameters as "truth") and fit both the biased (school-only) and unbiased (ANCOVA) models to each
- **Panel A**: Density plots of the 500 estimated school coefficients. The biased estimator is centered away from the true value; the unbiased estimator is centered on truth but has wider spread
- **Panel B**: Stacked bar chart decomposing MSE into Bias² and Variance for each estimator
- `as_tibble()` and `pivot_longer()` reshape the simulation results for plotting


:::


## Summary and Key Takeaways {#sec-ch2-summary}

::: {.callout-important}
## Key Takeaways

1. **ANCOVA** combines a categorical factor and a continuous covariate in a single regression model, allowing adjusted group comparisons
2. **Adjusted means** represent what each group's outcome would be if all groups had the same covariate distribution --- they enable "fair" comparisons
3. **Type III (unique) SS** tests each predictor controlling for all others; use this for hypothesis testing in observational studies
4. **Omitted variable bias** occurs when a relevant confounder is left out of the model. The bias depends on both the confounder's effect on $y$ and its correlation with the included predictor
5. **Confounders** should be controlled for; **mediators** should not
6. **Collinearity** inflates standard errors but does not bias coefficients --- it reflects the difficulty of separating the effects of correlated predictors
7. Every regression coefficient in multiple regression has a "controlling for" interpretation: $\beta_j$ is the effect of $x_j$ holding all other predictors constant
:::

::: {.callout-note}
## ANCOVA in Education Research Practice
ANCOVA appears throughout education research:

- **Randomized experiments with pre-tests**: Including a pre-test score as a covariate reduces residual variance and increases statistical power, even when randomization ensures balance
- **Value-added models (VAMs)**: School accountability systems use ANCOVA-style models to estimate school "effects" adjusted for student intake characteristics
- **Program evaluation**: Quasi-experimental studies use covariate adjustment to approximate the comparisons that randomization would provide

Understanding the assumptions and limitations of ANCOVA --- particularly around unmeasured confounders (@sec-ch2-confounding) --- is essential for critically evaluating these research designs.
:::


## Exercises {#sec-ch2-exercises}

::: {.callout-caution collapse="true"}
## Exercise 2.1: Full ANCOVA with All Schools
Using the full `gcse` dataset (all schools), fit an ANCOVA model with `lrt` and `school` as predictors. How many dummy variables does R create? What is $R^2$ compared to the LRT-only model?

**Hint**: `lm(gcse ~ lrt + factor(school), data = gcse)`. Check `nlevels(factor(gcse$school))` to see how many schools (and therefore dummies) there are.
:::

::: {.callout-caution collapse="true"}
## Exercise 2.2: Predicting the Direction of OVB
In the `gcse` data, suppose the true model includes both `lrt` and `gender`. If you omit `gender`, in which direction would you expect the coefficient of `lrt` to be biased? Use the OVB formula in @eq-ch2-ovb to reason through this.

**Hint**: Since `gender` is categorical, compute the point-biserial correlation: `cor(gcse$lrt, as.numeric(factor(gcse$gender)))`. Then fit `lm(gcse ~ lrt + gender, data = gcse)` to get $\hat{\beta}_{\text{gender}}$. Multiply the signs together using the OVB direction table.
:::

::: {.callout-caution collapse="true"}
## Exercise 2.3: Sequential vs Partial SS
Fit the ANCOVA model with `lrt` + `school_name` and then again with `school_name` + `lrt` (reversed order). Compare the Type I SS from `anova()`. Then use `car::Anova(type = 3)` for both. What do you notice?

**Hint**: The Type I SS will differ because they are sequential, but Type III SS will be identical.
:::

::: {.callout-caution collapse="true"}
## Exercise 2.4: When Does ANCOVA Fail?
A researcher evaluates a new reading program by comparing the post-test scores of students who volunteered for the program versus those who did not, controlling for pre-test scores using ANCOVA. The researcher concludes the program is effective because the adjusted mean for participants is significantly higher.

(a) What is the key assumption the researcher is making about confounders?
(b) List two plausible unmeasured confounders that could explain the result.
(c) Would a randomized experiment solve both problems? Why or why not?

**Hint**: Review @sec-ch2-confounding and think about what drives students to *volunteer* for programs.
:::

::: {.callout-caution collapse="true"}
## Exercise 2.5: Why Does Adding a Covariate Change the Group Coefficient?
In a simple regression comparing two schools, School B's coefficient is $\hat{\beta} = 5.2$ (i.e., School B students score 5.2 points higher on average). After adding prior achievement (LRT score) as a covariate in an ANCOVA model, School B's coefficient drops to $\hat{\beta} = 1.8$. Explain in plain language why controlling for prior achievement reduced the estimated school difference. What does this tell you about the relationship between school assignment and prior achievement in this dataset?

**Hint**: Think about omitted variable bias. If School B tends to enroll students with higher prior achievement, part of the raw school difference reflects pre-existing ability rather than a school effect. Adding the covariate removes this confounding.
:::

::: {.callout-caution collapse="true"}
## Exercise 2.6: Interpreting Parallel Slopes
The ANCOVA model assumes parallel regression lines (equal slopes across groups). A colleague argues this is unrealistic because "obviously different schools benefit differently from prior knowledge." Explain what the parallel slopes assumption actually means in statistical terms. Describe a specific pattern in the data that would indicate the assumption is violated, and how you would detect it both graphically and with a formal test.

**Hint**: The parallel slopes assumption means no interaction between the group factor and the covariate. You could detect a violation by plotting separate regression lines and looking for crossing or diverging lines, or by adding an interaction term to the model and testing whether it is significant.
:::

::: {.callout-caution collapse="true"}
## Exercise 2.7: Adjusted Means and Extrapolation
In an ANCOVA comparing three treatment groups, the adjusted (least-squares) means are computed at the grand mean of the covariate. Group C has very few observations near the grand mean of the covariate---most of its students scored much higher on the pre-test. Should you be concerned about the adjusted mean for Group C? Explain why extrapolation is a risk here and how you would assess whether the adjusted mean for Group C is trustworthy.

**Hint**: Adjusted means rely on the fitted regression line being accurate at the evaluation point. If Group C has little data near the grand mean of the covariate, the adjusted mean depends on extrapolating the within-group regression line into a region with sparse data. Check the overlap in covariate distributions across groups.
:::
