# Models for Ordinal Responses {#sec-ch13}

{{< include ../_common.qmd >}}


## Introduction {#sec-ch13-intro}

In @sec-ch9 through @sec-ch12, we modeled binary outcomes --- variables with exactly two categories. But many outcome variables in education and the social sciences have more than two categories that follow a natural order:

- **Essay grades**: Poor (1), Fair (2), Good (3), Excellent (4)
- **Course satisfaction**: Very Dissatisfied (1), Dissatisfied (2), Satisfied (3), Very Satisfied (4)
- **Likert scale items**: Strongly Disagree (1) through Strongly Agree (5)
- **Degree attainment**: Less than HS (1), HS diploma (2), Some college (3), Bachelor's (4), Graduate (5)

These **ordinal** variables share a critical feature: the categories have a meaningful rank order, but the distances between categories are not necessarily equal. The gap between "Poor" and "Fair" may not be the same as the gap between "Good" and "Excellent." This distinguishes ordinal data from truly continuous data, where equal intervals are assumed.

This chapter extends the latent variable framework from @sec-ch12 to ordinal outcomes with $S > 2$ categories. The key idea is elegant: instead of a single threshold dividing the latent continuum into two regions, we use **multiple thresholds** to carve it into $S$ ordered regions.

::: {.callout-note}
## Learning Objectives
After completing this chapter, you will be able to:

1. Explain why treating ordinal outcomes as continuous (in linear regression) or as unordered categories (in multinomial logistic regression) is inappropriate
2. Describe the latent variable and threshold model for ordinal responses
3. Write the cumulative logit model and explain the role of threshold parameters $\kappa_s$
4. State and interpret the **proportional odds assumption**: the same $\beta$ applies at every cut-point
5. Fit an ordinal logistic regression in R using `MASS::polr()` and interpret the output
6. Compute predicted probabilities and average marginal effects for ordinal models using `marginaleffects`
:::


## Why Special Models for Ordinal Data? {#sec-ch13-why}

You might wonder: why not just use linear regression or multinomial logistic regression for ordinal outcomes?

**Problem with linear regression**: Treating "Poor = 1, Fair = 2, Good = 3, Excellent = 4" as a continuous variable assumes equal spacing between categories and normally distributed errors. These assumptions are almost never warranted for ordinal data with few categories.

**Problem with multinomial logistic regression**: Treating the categories as unordered (like "Red, Blue, Green") discards the ordering information entirely. A model that treats "Poor vs. Excellent" the same as "Fair vs. Good" wastes valuable information.

**The ordinal model** respects the rank order without assuming equal intervals. It uses a single set of regression coefficients $\boldsymbol{\beta}$ that apply uniformly across all cut-points, producing a parsimonious and interpretable model.

::: {.callout-note}
## Connection to Previous Chapters
The ordinal model builds directly on concepts from @sec-ch9 through @sec-ch12:

- **From @sec-ch9--@sec-ch11 (Logistic Regression)**: The cumulative logit model uses the same logit link function, but applies it to *cumulative* probabilities rather than a single event probability.
- **From @sec-ch12 (Latent Variable & GLM Framework)**: The threshold model extends the single-threshold binary probit/logit to multiple thresholds. The GLM framework (linear predictor + link function + distributional family) remains the unifying thread.

If you understand binary logistic regression and the latent variable interpretation, the ordinal model is a natural next step: simply add more thresholds.
:::


## The Latent Variable and Threshold Model {#sec-ch13-latent}

### Extending the Binary Latent Variable

In @sec-ch12, we modeled a binary outcome through a latent continuous variable $y^*$ with a single threshold at zero. For ordinal outcomes with $S$ categories, we extend this to $S - 1$ thresholds:

$$
y_i^* = \beta_1 x_{1i} + \beta_2 x_{2i} + \cdots + \varepsilon_i
$$ {#eq-ch13-latent}

::: {.callout-note}
## No Intercept in the Latent Model
Notice that @eq-ch13-latent has **no intercept**. The intercept is absorbed into the threshold parameters $\kappa_1, \kappa_2, \ldots, \kappa_{S-1}$. Including both an intercept and free thresholds would create an identification problem (one redundant parameter).
:::

The observed ordinal response is determined by where $y^*$ falls relative to the thresholds:

$$
y_i = \begin{cases}
1 & \text{if } y_i^* < \kappa_1 \\
2 & \text{if } \kappa_1 \leq y_i^* < \kappa_2 \\
\vdots \\
S & \text{if } y_i^* \geq \kappa_{S-1}
\end{cases}
$$ {#eq-ch13-threshold}

where $\kappa_1 < \kappa_2 < \cdots < \kappa_{S-1}$ are the **threshold parameters** (also called **cut-points**).

### Visualizing the Threshold Model

Consider an essay grading example with $S = 4$ categories (Poor, Fair, Good, Excellent). The latent variable $y^*$ is a continuous "essay quality" score. The three thresholds $\kappa_1, \kappa_2, \kappa_3$ divide this continuum into four regions:

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch13-threshold-concept
#| fig-cap: "The threshold model for ordinal responses. A latent continuous variable y* is divided into four ordered categories by three threshold parameters. Shifting a predictor value moves the entire distribution along the latent scale, changing the probability of falling into each category."
#| fig-alt: "Density plot showing two overlapping logistic distributions along a latent variable axis, one for a low predictor value and one for a high predictor value. Three vertical dashed lines mark the threshold parameters kappa-1, kappa-2, and kappa-3, dividing the horizontal axis into four regions labeled Poor, Fair, Good, and Excellent. The distribution for the high predictor is shifted to the right, placing more of its area in the Good and Excellent regions, while the low predictor distribution concentrates more area in the Poor and Fair regions."
#| code-fold: true
#| code-summary: "Show R Code"

# Visualize the threshold model concept
kappa <- c(-1.5, 0, 1.5)
x_seq <- seq(-5, 5, length.out = 300)

# Two distributions: low predictor vs high predictor
df_dist <- bind_rows(
  tibble(x = x_seq, density = dlogis(x_seq, location = -1),
         group = "Low predictor (η = −1)"),
  tibble(x = x_seq, density = dlogis(x_seq, location = 1),
         group = "High predictor (η = 1)")
)

ggplot(df_dist, aes(x = x, y = density, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = kappa, linetype = "dashed", alpha = 0.6) +
  annotate("text", x = -3.2, y = 0.28, label = "Poor", fontface = "bold", size = 3.5) +
  annotate("text", x = -0.75, y = 0.28, label = "Fair", fontface = "bold", size = 3.5) +
  annotate("text", x = 0.75, y = 0.28, label = "Good", fontface = "bold", size = 3.5) +
  annotate("text", x = 3.2, y = 0.28, label = "Excellent", fontface = "bold", size = 3.5) +
  scale_color_manual(values = c(ber640_colors["accent"], ber640_colors["secondary"])) +
  labs(
    title = "Latent Variable and Threshold Model",
    subtitle = "Ordinal outcome with S = 4 categories",
    x = expression(y^"*" ~ "(latent variable)"),
    y = "Density",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch13-threshold-concept"
#| code-fold: true
#| code-summary: "Show R Code"
```
When the predictor value is low (blue), most of the distribution falls in the "Poor" and "Fair" regions. When the predictor value is high (red), the distribution shifts rightward, and more area falls under "Good" and "Excellent." The thresholds are fixed --- predictors shift the latent distribution, not the cut-points.

### Code Explanation
```{r}
#| ref.label: "fig-ch13-threshold-concept"
#| eval: false
```
- The dashed vertical lines represent the three thresholds $\kappa_1, \kappa_2, \kappa_3$
- The blue curve shows the latent distribution when the linear predictor is low ($\eta = -1$); the red curve when it is high ($\eta = 1$)
- A higher predictor value shifts the distribution to the right, increasing the probability of higher categories


:::


### App 13.1: Multiple Thresholds on the Latent Continuum {.unnumbered}

The static figure above fixes the thresholds and predictor values. Use the sliders below to explore how the **threshold parameters** ($\kappa_1, \kappa_2, \kappa_3$) and the **linear predictor** ($\beta$) jointly determine the probability of each ordinal category. Moving $\beta$ shifts the entire latent distribution---this is the proportional odds assumption in action.

```{ojs}
//| echo: false

import { ber640 as ord_ber640,
         categorical as ord_categorical,
         withAlpha as ord_withAlpha,
         themeColors as ord_themeColors,
         plotDefaults as ord_plotDefaults,
         sigmoid as ord_sigmoid,
         formatCoef as ord_formatCoef,
         formatPercent as ord_formatPercent,
         seq as ord_seq } from "../ojs/shared.js"
```

```{ojs}
//| echo: false

viewof ord_kappa1 = Inputs.range([-4, 3.8], {
  label: "κ₁ (threshold 1)",
  step: 0.1,
  value: -1.5
})
```

```{ojs}
//| echo: false

viewof ord_kappa2 = Inputs.range([-3.9, 3.9], {
  label: "κ₂ (threshold 2)",
  step: 0.1,
  value: 0
})
```

```{ojs}
//| echo: false

viewof ord_kappa3 = Inputs.range([-3.8, 4], {
  label: "κ₃ (threshold 3)",
  step: 0.1,
  value: 1.5
})
```

```{ojs}
//| echo: false

viewof ord_beta = Inputs.range([-2, 2], {
  label: "β (linear predictor)",
  step: 0.1,
  value: 1
})
```

```{ojs}
//| echo: false

// Enforce ordering: clamp thresholds so κ₁ < κ₂ < κ₃
ord_k1 = ord_kappa1
ord_k2 = Math.max(ord_kappa2, ord_k1 + 0.1)
ord_k3 = Math.max(ord_kappa3, ord_k2 + 0.1)
```

```{ojs}
//| echo: false

// --- Logistic PDF (inline, not in shared.js) ---
ord_dlogis = function(x, mu) {
  const z = x - mu;
  const ez = Math.exp(-z);
  return ez / Math.pow(1 + ez, 2);
}

// --- Category labels & colors ---
ord_catLabels = ["Poor", "Fair", "Good", "Excellent"]
ord_catColors = [ord_categorical[0], ord_categorical[1], ord_categorical[2], ord_categorical[3]]

// --- Category probabilities ---
ord_probs = {
  const p1 = ord_sigmoid(ord_k1 - ord_beta);
  const p12 = ord_sigmoid(ord_k2 - ord_beta);
  const p123 = ord_sigmoid(ord_k3 - ord_beta);
  return [
    Math.max(p1, 0),
    Math.max(p12 - p1, 0),
    Math.max(p123 - p12, 0),
    Math.max(1 - p123, 0)
  ];
}

// --- Density curve data ---
ord_densityData = {
  const xs = ord_seq(-6, 6, 300);
  return xs.map(x => ({ x: x, density: ord_dlogis(x, ord_beta) }));
}

// --- Shaded region data (for area marks between thresholds) ---
ord_regionData = {
  const xs = ord_seq(-6, 6, 300);
  const kappas = [-Infinity, ord_k1, ord_k2, ord_k3, Infinity];
  const rows = [];
  for (const x of xs) {
    const d = ord_dlogis(x, ord_beta);
    for (let cat = 0; cat < 4; cat++) {
      if (x >= kappas[cat] && x < kappas[cat + 1]) {
        rows.push({ x: x, density: d, category: ord_catLabels[cat] });
      }
    }
  }
  return rows;
}

// --- Stacked bar data ---
ord_barData = {
  let cumX = 0;
  return ord_catLabels.map((label, i) => {
    const x1 = cumX;
    cumX += ord_probs[i];
    return { category: label, x1: x1, x2: cumX, prob: ord_probs[i], color: ord_catColors[i] };
  });
}
```

```{ojs}
//| echo: false

// --- Equation display ---
html`<div style="
  font-family: 'Cambria Math', 'STIX Two Math', Georgia, serif;
  font-size: 15px;
  text-align: center;
  padding: 0.5rem 1rem;
  margin: 0.25rem 0 0.5rem 0;
  background: ${ord_themeColors().isDark ? 'rgba(255,255,255,0.05)' : 'rgba(0,0,0,0.03)'};
  border-radius: 6px;
  color: ${ord_themeColors().text};
">
  P(<em>Y</em> &le; <em>s</em>) = logistic(<span style="font-weight:700;">&#954;<sub>s</sub></span> &minus; <span style="color:${ord_ber640.secondary}; font-weight:700;">${ord_formatCoef(ord_beta, 1)}</span>)
  &nbsp;&nbsp;|&nbsp;&nbsp;
  &#954; = (${ord_formatCoef(ord_k1, 1)}, ${ord_formatCoef(ord_k2, 1)}, ${ord_formatCoef(ord_k3, 1)})
</div>`
```

```{ojs}
//| echo: false

// --- Top panel: Latent density with shaded regions & threshold lines ---
Plot.plot(ord_plotDefaults({
  width: 640,
  height: 340,
  marginBottom: 40,
  x: { label: "y* (latent variable)", domain: [-6, 6], grid: true },
  y: { label: "Density", domain: [0, 0.30], grid: true },
  color: { domain: ord_catLabels, range: ord_catColors, legend: true },
  marks: [
    // Shaded category regions
    Plot.areaY(ord_regionData, {
      x: "x",
      y: "density",
      fill: "category",
      fillOpacity: 0.30,
      curve: "linear"
    }),

    // Density curve
    Plot.line(ord_densityData, {
      x: "x",
      y: "density",
      stroke: ord_ber640.primary,
      strokeWidth: 2.5
    }),

    // Threshold lines
    Plot.ruleX([ord_k1, ord_k2, ord_k3], {
      stroke: ord_themeColors().text,
      strokeWidth: 1.5,
      strokeDasharray: "6,4"
    }),

    // Threshold labels
    Plot.text([
      { x: ord_k1, label: `κ₁ = ${ord_formatCoef(ord_k1, 1)}` },
      { x: ord_k2, label: `κ₂ = ${ord_formatCoef(ord_k2, 1)}` },
      { x: ord_k3, label: `κ₃ = ${ord_formatCoef(ord_k3, 1)}` }
    ], {
      x: "x",
      y: 0.285,
      text: "label",
      fill: ord_themeColors().text,
      fontSize: 11,
      fontWeight: 600,
      textAnchor: "middle"
    }),

    // Category labels at midpoints
    Plot.text([
      { x: ((-6) + ord_k1) / 2, label: "Poor" },
      { x: (ord_k1 + ord_k2) / 2, label: "Fair" },
      { x: (ord_k2 + ord_k3) / 2, label: "Good" },
      { x: (ord_k3 + 6) / 2, label: "Excellent" }
    ], {
      x: "x",
      y: 0.265,
      text: "label",
      fill: (d, i) => ord_catColors[i],
      fontSize: 12,
      fontWeight: 700,
      textAnchor: "middle"
    }),

    // Eta arrow showing location of distribution center
    Plot.ruleX([ord_beta], {
      stroke: ord_ber640.secondary,
      strokeWidth: 2,
      strokeDasharray: "3,3",
      y1: 0,
      y2: 0.05
    }),
    Plot.text([{ x: ord_beta }], {
      x: "x",
      y: 0.06,
      text: d => `η = ${ord_formatCoef(ord_beta, 1)}`,
      fill: ord_ber640.secondary,
      fontSize: 10,
      fontWeight: 700,
      textAnchor: "middle"
    })
  ]
}))
```

```{ojs}
//| echo: false

// --- Bottom panel: Stacked probability bar ---
Plot.plot(ord_plotDefaults({
  width: 640,
  height: 80,
  marginTop: 10,
  marginBottom: 30,
  marginLeft: 55,
  x: { label: "Cumulative Probability", domain: [0, 1], grid: false,
       tickFormat: d => (d * 100) + "%" },
  y: { axis: null, domain: [0, 1] },
  marks: [
    Plot.barX(ord_barData, {
      x1: "x1",
      x2: "x2",
      y1: 0.1,
      y2: 0.9,
      fill: "color",
      stroke: ord_themeColors().bg,
      strokeWidth: 1.5
    }),
    // Probability labels inside each bar
    Plot.text(ord_barData.filter(d => d.prob > 0.04), {
      x: d => (d.x1 + d.x2) / 2,
      y: 0.5,
      text: d => ord_formatPercent(d.prob, 1),
      fill: "white",
      fontSize: 12,
      fontWeight: 700,
      textAnchor: "middle"
    })
  ]
}))
```

```{ojs}
//| echo: false

// --- Readout cards ---
html`<div style="
  display: grid;
  grid-template-columns: repeat(4, 1fr);
  gap: 0.5rem;
  margin-top: 0.25rem;
  text-align: center;
  font-size: 0.85rem;
">
${ord_catLabels.map((label, i) => `
  <div style="
    padding: 0.5rem 0.3rem;
    border-radius: 8px;
    background: ${ord_withAlpha(ord_catColors[i], 0.10)};
    border: 1px solid ${ord_withAlpha(ord_catColors[i], 0.30)};
  ">
    <div style="font-weight: 700; color: ${ord_catColors[i]}; font-size: 0.9rem;">${label}</div>
    <div style="font-size: 1.25rem; font-weight: 700; font-variant-numeric: tabular-nums; color: ${ord_themeColors().text};">
      ${ord_formatPercent(ord_probs[i], 1)}
    </div>
  </div>
`).join("")}
</div>`
```


## The Cumulative Logit Model {#sec-ch13-cumulative}

### From Thresholds to Cumulative Probabilities

The cumulative logit model (also called the **proportional odds model**) works with cumulative probabilities $P(y_i > s \mid \mathbf{x}_i)$ --- the probability of being in a category **higher than** $s$:

$$
\text{logit}\big(P(y_i > s \mid \mathbf{x}_i)\big) = \beta_1 x_{1i} + \beta_2 x_{2i} + \cdots - \kappa_s
$$ {#eq-ch13-cumlogit}

for $s = 1, 2, \ldots, S-1$.

This gives us $S - 1$ equations, one for each possible dichotomization of the ordinal scale:

- $s = 1$: $\text{logit}(P(y > 1)) = \mathbf{x}'\boldsymbol{\beta} - \kappa_1$ (Fair or better vs. Poor)
- $s = 2$: $\text{logit}(P(y > 2)) = \mathbf{x}'\boldsymbol{\beta} - \kappa_2$ (Good or better vs. Fair or worse)
- $s = 3$: $\text{logit}(P(y > 3)) = \mathbf{x}'\boldsymbol{\beta} - \kappa_3$ (Excellent vs. Good or worse)

::: {.callout-important}
## The Proportional Odds Assumption
The key feature of @eq-ch13-cumlogit is that the **same regression coefficients** $\boldsymbol{\beta}$ appear in every equation. Only the thresholds $\kappa_s$ differ. This is the **proportional odds assumption**: the effect of $x$ on the odds of being in a higher category is the same regardless of where you split the scale.
:::

### Interpreting the Odds Ratio

Exponentiating a coefficient gives the **cumulative odds ratio**:

$$
\exp(\beta_j) = \text{OR for higher vs. lower categories per one-unit increase in } x_j
$$

This odds ratio is **constant across all cut-points** (by the proportional odds assumption). If $\exp(\beta) = 2.0$, then a one-unit increase in $x$ doubles the odds of being in a higher category, whether we define "higher" as $y > 1$, $y > 2$, or $y > 3$.

::: {.callout-tip}
## Intuition: Why "Proportional Odds"?
The name "proportional odds" comes from the fact that the **odds** of exceeding any category are always multiplied by the same factor $\exp(\beta)$ when $x$ increases by one unit. Imagine three separate coin-flip games (one for each cut-point). The proportional odds assumption says each game is "tilted" by the same amount when $x$ changes --- the predictor has a uniform effect across the entire ordinal scale.

When this assumption holds, a single odds ratio cleanly summarizes the predictor's effect. When it fails, the predictor might push hard at one end of the scale but not the other, and a single OR would be misleading.
:::

### Visualizing Cumulative Probabilities {#sec-ch13-cumprob-viz}

To build intuition, let us visualize how cumulative probabilities $P(y > s)$ change across a predictor range. These are the S-shaped curves that the cumulative logit model produces --- one curve per cut-point, shifted by the threshold parameters:

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch13-cumprob-curves
#| fig-cap: "Cumulative probability curves P(Y > s) as a function of a predictor x. Each curve corresponds to a different cut-point. Under proportional odds, the curves are parallel on the logit scale (equivalently, they have the same shape, only shifted horizontally)."
#| fig-alt: "Line plot with a continuous predictor x on the horizontal axis and cumulative probability P(Y > s) on the vertical axis, showing three S-shaped sigmoid curves. Each curve corresponds to a different cumulative cut-point: P(Y > 1) for Fair or above, P(Y > 2) for Good or above, and P(Y > 3) for Excellent. The three curves have identical shapes but are shifted horizontally, illustrating the proportional odds assumption where all cut-points share the same slope on the logit scale. Higher cut-points shift the curve to the right, requiring larger predictor values to achieve the same cumulative probability."
#| code-fold: true
#| code-summary: "Show R Code"

# Simulate cumulative probability curves
beta_demo <- 1.2
kappa_demo <- c(-1.5, 0.3, 2.0)
x_range <- seq(-3, 5, length.out = 200)

cum_prob_df <- expand.grid(x = x_range, s = 1:3) |>
  as_tibble() |>
  mutate(
    cutpoint = factor(s, labels = c("P(Y > 1): Fair+", "P(Y > 2): Good+", "P(Y > 3): Excellent")),
    logit_val = beta_demo * x - kappa_demo[s],
    cum_prob = plogis(logit_val)
  )

ggplot(cum_prob_df, aes(x = x, y = cum_prob, color = cutpoint)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c(ber640_colors["accent"],
                                ber640_colors["warning"],
                                ber640_colors["secondary"])) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Cumulative Probability Curves",
    subtitle = expression("Proportional odds: same " * beta * " shifts all curves equally"),
    x = "Predictor (x)",
    y = "Cumulative Probability P(Y > s)",
    color = "Cut-point"
  ) +
  theme(legend.position = "bottom")
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch13-cumprob-curves"
#| code-fold: true
#| code-summary: "Show R Code"
```
The three curves show the probability of exceeding each category as the predictor increases. Notice the curves never cross --- this is a consequence of the ordered thresholds ($\kappa_1 < \kappa_2 < \kappa_3$). The vertical distance between adjacent curves at any $x$ value gives the probability of falling in that specific category. All three curves have the same steepness because the same $\beta$ governs each, which is the proportional odds assumption in action.

### Code Explanation
```{r}
#| ref.label: "fig-ch13-cumprob-curves"
#| eval: false
```
- We compute $P(Y > s) = \text{logistic}(\beta x - \kappa_s)$ for three cut-points
- `beta_demo = 1.2` is the common slope (proportional odds); `kappa_demo` are the three thresholds
- Each curve is an S-shaped logistic function, shifted horizontally by the threshold value
- Under proportional odds, the curves are **parallel on the logit scale**, meaning they have the same shape but are shifted


:::


## Motivating Example: Essay Grades {#sec-ch13-example}

### The `grades` Dataset

The `grades` dataset from `regdatasets` contains 198 student essays, each assigned a grade from 1 (Poor) to 4 (Excellent). Several writing features were measured for each essay.

::: {.panel-tabset}
### Output
```{r}
#| label: ch13-data-load
#| code-fold: true
#| code-summary: "Show R Code"

# Load the grades dataset
data(grades)

# Create an ordered factor for grade
grades <- grades |>
  mutate(
    grade_ord = ordered(grade, levels = 1:4,
                        labels = c("Poor", "Fair", "Good", "Excellent"))
  )

# Overview
glimpse(grades)

# Distribution of grades
grades |>
  count(grade_ord) |>
  mutate(pct = n / sum(n)) |>
  kable(digits = 3, col.names = c("Grade", "N", "Proportion"))
```

### Output Interpretation
```{r}
#| ref.label: "ch13-data-load"
#| code-fold: true
#| code-summary: "Show R Code"
```
The grade distribution shows how many essays fall into each category. This is important for ordinal models --- very sparse categories can cause estimation problems.

### Code Explanation
```{r}
#| ref.label: "ch13-data-load"
#| eval: false
```
- `ordered()` creates an ordered factor, which R recognizes as ordinal
- `levels = 1:4` specifies the ordering; `labels` gives human-readable names
- The `count()` call shows how grades are distributed across the four categories


:::

### Exploring the Relationship

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch13-explore
#| fig-cap: "Average word length by essay grade. Essays with higher grades tend to use longer words on average, suggesting that word length may be a useful predictor of essay quality."
#| fig-alt: "Box plot with essay grade categories (Poor, Fair, Good, Excellent) on the horizontal axis and average word length on the vertical axis. Each grade has a colored box showing the interquartile range with a median line, plus whiskers and outlier points. The median word length increases monotonically from lower to higher grade categories, with the Poor category having the shortest median word length and the Excellent category having the longest. The pattern suggests a positive association between word length and essay quality."
#| code-fold: true
#| code-summary: "Show R Code"

ggplot(grades, aes(x = grade_ord, y = wordlength, fill = grade_ord)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = unname(ber640_colors[1:4])) +
  labs(
    title = "Word Length by Essay Grade",
    x = "Essay Grade",
    y = "Average Word Length"
  )
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch13-explore"
#| code-fold: true
#| code-summary: "Show R Code"
```
The box plots show a general upward trend: essays with higher grades tend to use longer words. This suggests that average word length may be a useful predictor in our ordinal regression model.

### Code Explanation
```{r}
#| ref.label: "fig-ch13-explore"
#| eval: false
```
- Box plots show the distribution of average word length within each grade category
- If word length is positively associated with grade, we expect higher medians in better grade categories
- The `fill` aesthetic colors each grade differently for visual distinction


:::


## Fitting the Ordinal Logistic Regression {#sec-ch13-fitting}

### Using `MASS::polr()`

The `polr()` function (**p**roportional **o**dds **l**ogistic **r**egression) from the `MASS` package fits the cumulative logit model:

::: {.panel-tabset}
### Output
```{r}
#| label: ch13-polr-fit
#| code-fold: true
#| code-summary: "Show R Code"

# Fit ordinal logistic regression
# Hess = TRUE is needed for standard errors
fit_ord <- MASS::polr(grade_ord ~ wordlength + errors,
                      data = grades, Hess = TRUE)

# Model summary using easystats
model_parameters(fit_ord) |>
  print_md()
```

### Output Interpretation
```{r}
#| ref.label: "ch13-polr-fit"
#| code-fold: true
#| code-summary: "Show R Code"
```
The output shows two types of parameters:

- **Regression coefficients** ($\beta$): The effect of each predictor on the log-odds of being in a higher grade category
- **Threshold parameters** ($\kappa$): The cut-points on the latent scale separating adjacent categories (labeled as intercepts by `polr`)

A positive $\beta$ means the predictor increases the probability of higher categories; a negative $\beta$ decreases it.

### Code Explanation
```{r}
#| ref.label: "ch13-polr-fit"
#| eval: false
```
- `polr(grade_ord ~ wordlength + errors)` fits a proportional odds model with two predictors
- `wordlength`: average word length (higher $\rightarrow$ more complex vocabulary)
- `errors`: percentage of words with spelling errors (higher $\rightarrow$ more errors)
- `Hess = TRUE` requests the Hessian matrix, which is needed to compute standard errors
- `model_parameters()` provides a clean summary with coefficients, SEs, CIs, and $p$-values


:::

### Interpreting the Coefficients

::: {.panel-tabset}
### Output
```{r}
#| label: ch13-odds-ratios
#| code-fold: true
#| code-summary: "Show R Code"

# Exponentiate to get cumulative odds ratios
tibble(
  term = names(coef(fit_ord)),
  log_odds = coef(fit_ord),
  OR = exp(coef(fit_ord))
) |>
  kable(digits = 4, col.names = c("Predictor", "Log-Odds (β)", "Odds Ratio (exp(β))"))
```

### Output Interpretation
```{r}
#| ref.label: "ch13-odds-ratios"
#| code-fold: true
#| code-summary: "Show R Code"
```
Interpreting the odds ratios:

- **wordlength**: For each one-unit increase in average word length, the odds of receiving a higher grade (vs. any lower grade) are multiplied by $\exp(\hat{\beta})$. If the OR is 1.5, a one-unit increase in word length increases the odds of a higher grade by 50%.
- **errors**: For each one-percentage-point increase in spelling errors, the odds of a higher grade are multiplied by $\exp(\hat{\beta})$. An OR below 1 means more errors decrease the odds of a higher grade.

The beauty of the proportional odds model is that these ORs are the same whether you compare "Fair or better vs. Poor," "Good or better vs. Fair or worse," or "Excellent vs. the rest."

### Code Explanation
```{r}
#| ref.label: "ch13-odds-ratios"
#| eval: false
```
- `coef(fit_ord)` extracts only the regression coefficients (not the thresholds)
- `exp(coef(fit_ord))` converts to cumulative odds ratios
- The OR applies uniformly across all cut-points (proportional odds assumption)


:::


## Predicted Probabilities {#sec-ch13-predictions}

### Category-Specific Predictions

One of the most useful outputs from an ordinal model is the predicted probability for each category:

::: {.panel-tabset}
### Output
```{r}
#| label: ch13-predictions
#| code-fold: true
#| code-summary: "Show R Code"

# Predicted probabilities at representative values
pred_grid <- datagrid(model = fit_ord,
                      wordlength = c(4.0, 4.5, 5.0),
                      errors = median(grades$errors))

preds <- predictions(fit_ord, newdata = pred_grid, type = "probs") |>
  as_tibble()

preds |>
  select(wordlength, group, estimate) |>
  pivot_wider(names_from = group, values_from = estimate) |>
  kable(digits = 3,
        col.names = c("Word Length", "P(Poor)", "P(Fair)", "P(Good)", "P(Excellent)"))
```

### Output Interpretation
```{r}
#| ref.label: "ch13-predictions"
#| code-fold: true
#| code-summary: "Show R Code"
```
Each row shows the predicted probability distribution across the four grade categories at a given word length (with spelling errors held at the median). As word length increases, probability should shift from lower to higher categories. The four probabilities in each row sum to 1.

### Code Explanation
```{r}
#| ref.label: "ch13-predictions"
#| eval: false
```
- `datagrid()` creates a grid of predictor values for prediction; `errors` is held at its median
- `predictions(type = "probs")` returns predicted probabilities for **each** ordinal category
- The `group` column identifies which category each probability refers to
- We pivot to wide format for a cleaner table


:::

### Visualizing Predicted Probabilities

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch13-pred-probs
#| fig-cap: "Predicted probability of each essay grade category as a function of average word length, with spelling errors held at the median. As word length increases, the probability mass shifts from lower to higher grade categories."
#| fig-alt: "Stacked area chart with average word length on the horizontal axis and predicted probability on the vertical axis, stacking to one. Four colored bands represent the predicted probabilities for each grade category: Poor, Fair, Good, and Excellent. At low word lengths, the Poor and Fair categories dominate the probability mass. As word length increases, the bands for Good and Excellent expand while Poor and Fair shrink, showing a smooth leftward-to-rightward shift in predicted grade distribution with increasing word length. Spelling errors are held constant at the median."
#| code-fold: true
#| code-summary: "Show R Code"

# Predictions across the full range of word length
pred_full <- predictions(
  fit_ord,
  newdata = datagrid(wordlength = seq(3.5, 5.5, by = 0.05),
                     errors = median(grades$errors)),
  type = "probs"
) |>
  as_tibble()

ggplot(pred_full, aes(x = wordlength, y = estimate, fill = group)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = unname(ber640_colors[1:4])) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Predicted Grade Probabilities by Word Length",
    subtitle = "Spelling errors held at median",
    x = "Average Word Length",
    y = "Predicted Probability",
    fill = "Grade"
  )
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch13-pred-probs"
#| code-fold: true
#| code-summary: "Show R Code"
```
The stacked area chart shows the smooth shift in predicted probabilities. At low word lengths, most probability is in the lower categories. At high word lengths, the probability mass shifts to the right. The proportional odds assumption means that this shift follows a regular pattern --- the curves do not cross in unexpected ways.

### Code Explanation
```{r}
#| ref.label: "fig-ch13-pred-probs"
#| eval: false
```
- `geom_area()` creates a stacked area chart showing the four category probabilities
- The total height is always 100% (probabilities sum to 1)
- As word length increases, the areas for higher grades expand and lower grades shrink
- This is a direct visualization of how the cumulative logit model distributes probability across categories


:::


### Predicted Probability Stacked Bar Chart {#sec-ch13-stacked-bar}

A stacked bar chart at selected predictor values provides a discrete, easy-to-read summary of how the probability distribution shifts:

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch13-stacked-bar
#| fig-cap: "Predicted probability distribution across grade categories at selected word length values. Each bar shows how probability is allocated among the four categories, illustrating the shift from lower to higher grades as word length increases."
#| fig-alt: "Stacked bar chart with selected word length values on the horizontal axis and predicted probability on the vertical axis. Each vertical bar is divided into four colored segments representing the Poor, Fair, Good, and Excellent grade categories, with all segments summing to one. At the shortest word length, Poor and Fair dominate. As word length increases across the bars, the Good and Excellent segments grow progressively larger while Poor and Fair shrink, providing a discrete summary of how the predicted grade distribution shifts upward with longer average word length."
#| code-fold: true
#| code-summary: "Show R Code"

# Predicted probabilities at selected word lengths
pred_bar <- predictions(
  fit_ord,
  newdata = datagrid(
    wordlength = seq(3.5, 5.5, by = 0.5),
    errors = median(grades$errors)
  ),
  type = "probs"
) |>
  as_tibble() |>
  mutate(wordlength_label = factor(round(wordlength, 1)))

ggplot(pred_bar, aes(x = wordlength_label, y = estimate, fill = group)) +
  geom_col(position = "stack", width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = unname(ber640_colors[1:4])) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Predicted Grade Distribution by Word Length",
    subtitle = "Spelling errors held at median",
    x = "Average Word Length",
    y = "Predicted Probability",
    fill = "Grade"
  )
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch13-stacked-bar"
#| code-fold: true
#| code-summary: "Show R Code"
```
Each bar represents the full probability distribution across grade categories at a given word length. Moving from left to right, the bars show a clear shift: the proportion allocated to "Poor" and "Fair" shrinks while "Good" and "Excellent" grow. This is the hallmark of a positive predictor effect in the proportional odds model --- the entire distribution shifts toward higher categories.

### Code Explanation
```{r}
#| ref.label: "fig-ch13-stacked-bar"
#| eval: false
```
- `predictions(type = "probs")` returns category-specific probabilities at each grid point
- `geom_col(position = "stack")` stacks the four category probabilities in each bar (summing to 100%)
- Selected word lengths (3.5 to 5.5 in steps of 0.5) provide discrete comparison points
- This complements the continuous stacked area chart above with a more accessible discrete view


:::


## Average Marginal Effects {#sec-ch13-marginal}

### Marginal Effects for Ordinal Models

In ordinal models, marginal effects are more nuanced than in binary models because each predictor affects the probability of **every** category. The `avg_slopes()` function computes these category-specific effects:

::: {.panel-tabset}
### Output
```{r}
#| label: ch13-ame
#| code-fold: true
#| code-summary: "Show R Code"

# Average marginal effects
avg_slopes(fit_ord, type = "probs") |>
  as_tibble() |>
  select(term, group, estimate, std.error, p.value) |>
  kable(digits = 4,
        col.names = c("Predictor", "Grade Category", "AME", "SE", "p-value"))
```

### Output Interpretation
```{r}
#| ref.label: "ch13-ame"
#| code-fold: true
#| code-summary: "Show R Code"
```
The AME table shows how a one-unit change in each predictor shifts the probability distribution:

- For **word length**: a one-unit increase decreases the probability of lower grades and increases the probability of higher grades
- For **errors**: a one-unit increase in spelling errors shifts probability toward lower grades and away from higher grades
- The AMEs for each predictor across all categories sum to approximately zero, confirming that probability is being redistributed, not created

### Code Explanation
```{r}
#| ref.label: "ch13-ame"
#| eval: false
```
- `avg_slopes(type = "probs")` computes average marginal effects on the probability scale for each category
- For each predictor, there is one AME per category (four AMEs for four categories)
- The AMEs within each predictor sum to zero: increasing the probability of some categories necessarily decreases others


:::


## Testing the Proportional Odds Assumption {#sec-ch13-testing}

### Why Test?

The proportional odds assumption --- that the same $\boldsymbol{\beta}$ applies at every cut-point --- is the central assumption of the cumulative logit model. If it is violated, the effect of a predictor on the odds of exceeding category $s$ depends on which $s$ you choose. This would mean a single odds ratio per predictor is misleading.

### Informal Test: Separate Binary Logistic Regressions

One practical approach is to fit separate binary logistic regressions at each cut-point and check whether the coefficients are similar:

::: {.panel-tabset}
### Output
```{r}
#| label: ch13-po-check
#| code-fold: true
#| code-summary: "Show R Code"

# Create binary variables for each cut-point
grades <- grades |>
  mutate(
    y_gt1 = as.integer(grade > 1),
    y_gt2 = as.integer(grade > 2),
    y_gt3 = as.integer(grade > 3)
  )

# Fit separate binary logistic regressions
fit_cut1 <- glm(y_gt1 ~ wordlength + errors, family = binomial, data = grades)
fit_cut2 <- glm(y_gt2 ~ wordlength + errors, family = binomial, data = grades)
fit_cut3 <- glm(y_gt3 ~ wordlength + errors, family = binomial, data = grades)

# Compare coefficients across cut-points
tibble(
  term = names(coef(fit_cut1)),
  `Cut 1 (>Poor)` = coef(fit_cut1),
  `Cut 2 (>Fair)` = coef(fit_cut2),
  `Cut 3 (>Good)` = coef(fit_cut3)
) |>
  filter(term != "(Intercept)") |>
  kable(digits = 4)
```

### Output Interpretation
```{r}
#| ref.label: "ch13-po-check"
#| code-fold: true
#| code-summary: "Show R Code"
```
If the proportional odds assumption holds, the coefficients for `wordlength` and `errors` should be roughly consistent across the three cut-points. Large discrepancies suggest the assumption may be violated --- the effect of a predictor depends on where you split the scale.

### Code Explanation
```{r}
#| ref.label: "ch13-po-check"
#| eval: false
```
- We create three binary variables, one for each cut-point (e.g., `y_gt1 = 1` if grade $> 1$)
- Three separate logistic regressions are fit, one at each dichotomization
- Under proportional odds, the slope coefficients should be approximately equal across the three models
- Only the intercepts should differ


:::

### Visualizing the Parallel Lines {#sec-ch13-parallel-viz}

A powerful way to assess the proportional odds assumption visually is to plot the estimated log-odds from the separate binary regressions. Under proportional odds, the lines for each predictor should be approximately parallel across cut-points:

::: {.panel-tabset}
### Output
```{r}
#| label: fig-ch13-parallel-lines
#| fig-cap: "Parallel lines check for the proportional odds assumption. Each line shows the predicted log-odds of exceeding a given category as a function of word length (errors held at median). Under proportional odds, lines should be approximately parallel (same slope, different intercepts)."
#| fig-alt: "Line plot with average word length on the horizontal axis and predicted log-odds on the vertical axis, showing three lines corresponding to the three cumulative cut-points: exceeding Poor, exceeding Fair, and exceeding Good. Each line is estimated from a separate binary logistic regression with spelling errors held at the median. The three lines have similar positive slopes but different intercepts, appearing approximately parallel. This visual parallelism supports the proportional odds assumption, indicating that the effect of word length is roughly constant across all cut-points of the ordinal outcome."
#| code-fold: true
#| code-summary: "Show R Code"

# Predicted log-odds from separate binary models
wl_range <- seq(
  min(grades$wordlength, na.rm = TRUE),
  max(grades$wordlength, na.rm = TRUE),
  length.out = 100
)
med_errors <- median(grades$errors, na.rm = TRUE)

parallel_df <- bind_rows(
  tibble(wordlength = wl_range,
         logit = predict(fit_cut1,
                         newdata = tibble(wordlength = wl_range, errors = med_errors),
                         type = "link"),
         cutpoint = "P(Y > 1): Fair+"),
  tibble(wordlength = wl_range,
         logit = predict(fit_cut2,
                         newdata = tibble(wordlength = wl_range, errors = med_errors),
                         type = "link"),
         cutpoint = "P(Y > 2): Good+"),
  tibble(wordlength = wl_range,
         logit = predict(fit_cut3,
                         newdata = tibble(wordlength = wl_range, errors = med_errors),
                         type = "link"),
         cutpoint = "P(Y > 3): Excellent")
)

ggplot(parallel_df, aes(x = wordlength, y = logit, color = cutpoint)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c(ber640_colors["accent"],
                                ber640_colors["warning"],
                                ber640_colors["secondary"])) +
  labs(
    title = "Parallel Lines Check for Proportional Odds",
    subtitle = "Errors held at median; lines should be parallel if assumption holds",
    x = "Average Word Length",
    y = "Predicted Log-Odds",
    color = "Cut-point"
  ) +
  theme(legend.position = "bottom")
```

### Output Interpretation
```{r}
#| ref.label: "fig-ch13-parallel-lines"
#| code-fold: true
#| code-summary: "Show R Code"
```
If the three lines are approximately parallel (same slope but shifted vertically), the proportional odds assumption is reasonable for `wordlength`. Non-parallel lines --- especially lines that cross or diverge strongly --- suggest the effect of word length differs across cut-points, which would undermine the single-OR interpretation.

### Code Explanation
```{r}
#| ref.label: "fig-ch13-parallel-lines"
#| eval: false
```
- For each of the three separate binary logistic regressions, we compute the predicted log-odds across a range of `wordlength` values (with `errors` fixed at the median)
- `predict(..., type = "link")` returns the log-odds (linear predictor) from each `glm` model
- Under proportional odds, the slopes should be equal --- only the intercepts differ, producing parallel lines
- Deviations from parallelism indicate potential violations of the proportional odds assumption


:::

### Interactive Exploration: Proportional Odds Assumption {.unnumbered}

::: {.callout-tip collapse="true"}
## Try It: Violate the Proportional Odds Assumption

Toggle between the proportional odds model (parallel cumulative curves) and a violated version (non-parallel curves that cross). When the assumption holds, one β applies to all cutpoints; when violated, each cutpoint has its own slope.
:::

```{ojs}
//| echo: false
import { ber640 as po_ber640, plotDefaults as po_plotDefaults, themeColors as po_themeColors, sigmoid as po_sigmoid, seq as po_seq, categorical as po_categorical, formatCoef as po_formatCoef, withAlpha as po_withAlpha } from "../ojs/shared.js"
```

```{ojs}
//| echo: false
po_tc = po_themeColors()

viewof po_assumption = Inputs.radio(
  ["Proportional Odds", "Violated"],
  { value: "Proportional Odds", label: "Assumption" }
)

viewof po_beta = Inputs.range([-2, 2], {
  value: 0.8, step: 0.1, label: "β (common slope)"
})

viewof po_severity = Inputs.range([0, 1.5], {
  value: 0.5, step: 0.05,
  label: "Violation severity",
  disabled: po_assumption === "Proportional Odds"
})
```

```{ojs}
//| echo: false
po_kappa = [-1.5, 0, 1.5]

po_violated = po_assumption === "Violated"

po_betas = po_violated
  ? [po_beta - po_severity, po_beta, po_beta + po_severity]
  : [po_beta, po_beta, po_beta]

po_xvals = po_seq(-4, 4, 200)

po_cumData = {
  const rows = [];
  for (const x of po_xvals) {
    for (let j = 0; j < 3; j++) {
      const p = po_sigmoid(po_kappa[j] - po_betas[j] * x);
      rows.push({x, cumProb: p, cutpoint: `P(Y ≤ ${j + 1})`, j});
    }
  }
  return rows;
}

po_catData = {
  const rows = [];
  for (const x of po_xvals) {
    const cum = [0, ...po_kappa.map((k, j) => po_sigmoid(k - po_betas[j] * x)), 1];
    for (let c = 0; c < 4; c++) {
      rows.push({x, prob: cum[c + 1] - cum[c], category: `P(Y = ${c + 1})`});
    }
  }
  return rows;
}

po_curvesOverlap = {
  if (!po_violated) return false;
  for (const x of po_xvals) {
    const c1 = po_sigmoid(po_kappa[0] - po_betas[0] * x);
    const c2 = po_sigmoid(po_kappa[1] - po_betas[1] * x);
    const c3 = po_sigmoid(po_kappa[2] - po_betas[2] * x);
    if (c1 > c2 || c2 > c3) return true;
  }
  return false;
}

po_cumLabels = ["P(Y ≤ 1)", "P(Y ≤ 2)", "P(Y ≤ 3)"]
po_catLabels = ["P(Y = 1)", "P(Y = 2)", "P(Y = 3)", "P(Y = 4)"]
po_catColors = [po_categorical[0], po_categorical[1], po_categorical[2], po_categorical[3]]

html`<div style="display:inline-block; padding:4px 12px; border-radius:4px; font-weight:600; font-size:0.9em; margin-bottom:8px; background:${po_violated ? po_withAlpha(po_ber640.secondary, 0.15) : po_withAlpha(po_ber640.success, 0.15)}; color:${po_violated ? po_ber640.secondary : po_ber640.success}; border:1px solid ${po_violated ? po_withAlpha(po_ber640.secondary, 0.4) : po_withAlpha(po_ber640.success, 0.4)};">${po_violated ? "⚠ Assumption Violated" : "✓ Proportional Odds Holds"}</div>
${po_curvesOverlap ? html`<div style="display:inline-block; padding:4px 12px; border-radius:4px; font-size:0.85em; margin-left:8px; margin-bottom:8px; background:${po_withAlpha(po_ber640.secondary, 0.1)}; color:${po_ber640.secondary}; border:1px solid ${po_withAlpha(po_ber640.secondary, 0.3)};">Curves crossing — serious violation!</div>` : html``}`

po_cumPlot = Plot.plot(po_plotDefaults({
  title: "Cumulative Probability Curves",
  width: 640,
  height: 340,
  x: { label: "Predictor X", domain: [-4, 4] },
  y: { label: "Cumulative Probability P(Y ≤ j)", domain: [0, 1] },
  color: {
    domain: po_cumLabels,
    range: [po_categorical[0], po_categorical[1], po_categorical[2]],
    legend: true
  },
  marks: [
    Plot.ruleY([0.5], { stroke: po_tc.muted, strokeOpacity: 0.3, strokeDasharray: "4,4" }),
    Plot.line(po_cumData, {
      x: "x", y: "cumProb", stroke: "cutpoint", strokeWidth: 2.5, curve: "natural"
    })
  ]
}))

html`<div style="font-size:0.85em; color:${po_tc.text}; padding:6px 10px; margin:4px 0 8px 0; background:${po_withAlpha(po_tc.text, 0.05)}; border-radius:4px; font-family:var(--bs-font-monospace, monospace);">${po_violated
  ? `β₁ = ${po_formatCoef(po_betas[0])}, β₂ = ${po_formatCoef(po_betas[1])}, β₃ = ${po_formatCoef(po_betas[2])}`
  : `β = ${po_formatCoef(po_beta)} (same for all cutpoints)`
}  |  κ₁ = −1.5, κ₂ = 0, κ₃ = 1.5</div>`

po_areaPlot = Plot.plot(po_plotDefaults({
  title: "Category Probabilities",
  width: 640,
  height: 280,
  x: { label: "Predictor X", domain: [-4, 4] },
  y: { label: "Probability", domain: [0, 1] },
  color: {
    domain: po_catLabels,
    range: po_catColors,
    legend: true
  },
  marks: [
    Plot.areaY(po_catData, Plot.stackY({
      x: "x", y: "prob", fill: "category",
      order: po_catLabels, curve: "natural", fillOpacity: 0.7
    })),
    Plot.ruleY([0])
  ]
}))
```

::: {.callout-tip}
## Practical Guidance
Minor variation in coefficients across cut-points is normal (due to sampling variability). The proportional odds assumption is a simplifying approximation, not an exact truth. Focus on whether the coefficients tell a qualitatively similar story across cut-points. If a coefficient is positive at one cut-point and negative at another, that is a serious violation.
:::


## Summary and Key Takeaways {#sec-ch13-summary}

This chapter completes the arc from continuous to ordinal outcomes, extending the latent variable and GLM frameworks introduced in @sec-ch12.

::: {.callout-important}
## Key Takeaways

1. **Ordinal outcomes require special models.** Linear regression assumes equal intervals; multinomial logistic regression ignores the ordering. The cumulative logit model respects the rank order without assuming equal spacing.

2. **The latent variable model extends naturally.** With $S$ categories, we use $S - 1$ thresholds on the latent continuum: $y = s$ if $\kappa_{s-1} \leq y^* < \kappa_s$.

3. **No intercept in the regression equation.** The intercept is absorbed into the threshold parameters, avoiding identification problems.

4. **The proportional odds assumption is the key simplification.** The same regression coefficients $\boldsymbol{\beta}$ apply at every cut-point, producing a single odds ratio per predictor.

5. **$\exp(\beta)$ is the cumulative odds ratio.** It represents the change in odds of being in a higher (vs. lower) category per one-unit increase in $x$, and this ratio is constant across all cut-points.

6. **Predicted probabilities are category-specific.** Each observation gets a full probability distribution across all $S$ categories, and these probabilities sum to 1.

7. **Average marginal effects sum to zero across categories.** Increasing the probability of some categories necessarily decreases others.
:::

::: {.callout-note}
## The Course Arc: From SLR to Ordinal Models

Looking back across the full course, we can trace a clear progression:

| Chapters | Outcome Type | Key Model |
|:---------|:------------|:----------|
| 1--8 | Continuous | Linear regression ($y = \alpha + \mathbf{x}'\boldsymbol{\beta} + \varepsilon$, identity link, normal family) |
| 9--11 | Binary | Logistic regression (logit link, binomial family) |
| 12 | Binary (latent view) | Probit regression + GLM framework |
| **13** | **Ordinal** | **Cumulative logit model (proportional odds)** |

Every model shares the same core structure: a linear predictor $\eta = \alpha + \mathbf{x}'\boldsymbol{\beta}$, a link function connecting $\eta$ to the expected outcome, and a distributional assumption for the response. The GLM framework from @sec-ch12 is the thread that binds them all together.
:::


## Exercises {#sec-ch13-exercises}

::: {.callout-caution collapse="true"}
## Exercise 13.1: Satisfaction and Income
Using the `satisfaction` dataset from `regdatasets`, fit an ordinal logistic regression predicting `satisfaction` (an ordinal variable) from `income` and `gender`.

a. Create an ordered factor for `satisfaction` and fit the model using `polr()`.
b. Interpret the odds ratio for `income`. What does a one-unit increase in income do to the odds of higher satisfaction?
c. Compute predicted probabilities at income levels 1, 2, 3, and 4 for each gender. How does the probability distribution shift?
d. Compute and interpret the average marginal effects.
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.1

**Part (a):** Use `ordered()` to create the factor, then `MASS::polr(satisfaction_ord ~ income + gender, data = satisfaction, Hess = TRUE)`.

**Part (b):** Extract coefficients with `coef()` and exponentiate with `exp()`. Remember: the OR applies uniformly across all cut-points.

**Part (c):** Use `marginaleffects::predictions()` with `datagrid()` to create the prediction grid. Set `type = "probs"` to get category-specific probabilities.

**Part (d):** `avg_slopes(model, type = "probs")` gives one AME per predictor per category. Check that they sum to approximately zero within each predictor.
:::

::: {.callout-caution collapse="true"}
## Exercise 13.2: Testing Proportional Odds
Using your model from Exercise 13.1:

a. Create binary variables for each cut-point of `satisfaction` and fit separate logistic regressions.
b. Compare the coefficient for `income` across the separate models. Does the proportional odds assumption appear reasonable?
c. What would it mean substantively if the assumption were violated --- for example, if income had a strong effect on the odds of being "Very Satisfied" vs. everything else but a weak effect on other comparisons?
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.2

**Part (a):** If `satisfaction` has $S$ categories, create $S - 1$ binary variables: `y_gt1 = as.integer(satisfaction > 1)`, etc. Fit `glm(..., family = binomial)` for each.

**Part (b):** Arrange the coefficients for `income` side by side. Look for whether they are roughly similar in magnitude and sign. Minor variation is expected; a sign change or order-of-magnitude difference is concerning.

**Part (c):** Think about what it means for the effect to be "non-uniform." If income matters more at the top of the scale, it suggests a ceiling/floor effect or that the underlying construct (satisfaction) relates to income differently at different levels.
:::

::: {.callout-caution collapse="true"}
## Exercise 13.3: The Full Modeling Arc
Consider a study of student performance with the following variables:

- `final_grade`: ordinal (F, D, C, B, A)
- `study_hours`: continuous
- `tutoring`: binary (1 = received tutoring, 0 = did not)

a. Write the cumulative logit model equation, specifying all threshold parameters and regression coefficients.
b. If $\hat{\beta}_{\text{tutoring}} = 0.85$, compute and interpret the odds ratio. Does this OR depend on which cut-point is used?
c. A colleague suggests using linear regression with the coding F=0, D=1, C=2, B=3, A=4. List two specific problems with this approach.
d. Explain how this ordinal model relates to the GLM framework from @sec-ch12. What are the three GLM components?
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.3

**Part (a):** With $S = 5$ categories (F through A), you need $S - 1 = 4$ threshold parameters $\kappa_1, \ldots, \kappa_4$ and two regression coefficients. The model is $\text{logit}(P(Y > s)) = \beta_1 \cdot \text{study\_hours} + \beta_2 \cdot \text{tutoring} - \kappa_s$.

**Part (b):** $\exp(0.85) \approx 2.34$. This OR does **not** depend on the cut-point --- that is precisely the proportional odds assumption.

**Part (c):** Think about (1) the equal-interval assumption and (2) the distributional assumption. Is the "distance" from F to D really the same as from B to A?

**Part (d):** The three components are: (1) linear predictor $\eta = \beta_1 x_1 + \beta_2 x_2$, (2) link function = cumulative logit, (3) distributional assumption = multinomial (ordinal) with cumulative logit structure.
:::

::: {.callout-caution collapse="true"}
## Exercise 13.4: Visualizing and Communicating Ordinal Results
Using the `grades` dataset and the model `fit_ord` from this chapter:

a. Create a plot of predicted probabilities for the **errors** predictor (holding `wordlength` at its median). Use either a stacked area chart or a stacked bar chart.
b. Compute the predicted probability of receiving an "Excellent" grade when `wordlength = 5.0` and `errors = 2` versus when `wordlength = 5.0` and `errors = 8`. How much does the probability change?
c. A non-technical stakeholder asks: "Does word length matter for essay grades?" Write a 2--3 sentence plain-language summary using the predicted probabilities and/or AMEs to answer this question (avoid mentioning odds ratios or log-odds).
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.4

**Part (a):** Adapt the code from @fig-ch13-pred-probs, replacing `wordlength` as the x-axis variable with `errors` and holding `wordlength = median(grades$wordlength)` instead.

**Part (b):** Use `predictions(fit_ord, newdata = datagrid(wordlength = 5.0, errors = c(2, 8)), type = "probs")` and filter for the "Excellent" group.

**Part (c):** Focus on concrete probability changes. For example: "An essay with average word length of X has a Y% chance of receiving an Excellent grade, compared to Z% for an essay with word length W..." Avoid jargon.
:::

::: {.callout-caution collapse="true"}
## Exercise 13.5: Why Not Just Use Linear Regression for Ordinal Outcomes?
A survey measures customer satisfaction on a 5-point scale (1 = Very Dissatisfied, ..., 5 = Very Satisfied). A researcher proposes treating the response as numeric and running ordinary linear regression.

a. What implicit assumption does this approach make about the "distance" between adjacent categories? Why might the gap between "Dissatisfied" and "Neutral" differ from the gap between "Satisfied" and "Very Satisfied"?
b. Linear regression can produce predicted values like 3.7. What does a predicted satisfaction of 3.7 mean? How does the proportional odds model handle this differently by predicting probabilities for each category?
c. Under what circumstances might treating an ordinal outcome as continuous be a reasonable approximation? When would you insist on using an ordinal model instead?
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.5

**Part (a):** Linear regression assumes equal spacing between all categories, but ordinal scales only guarantee ordering. Respondents may perceive "Very Satisfied" as much further from "Satisfied" than "Neutral" is from "Dissatisfied."

**Part (b):** A predicted value of 3.7 has no direct categorical interpretation. The ordinal model instead gives $P(Y \leq k)$ for each cutpoint, yielding a full probability distribution across categories.

**Part (c):** When there are many categories (e.g., 7+), the outcome is approximately symmetric, and the equal-spacing assumption is roughly defensible, linear regression may give similar substantive conclusions. With few categories or skewed distributions, ordinal models are preferred.
:::

::: {.callout-caution collapse="true"}
## Exercise 13.6: The Proportional Odds Assumption in Practice
The proportional odds (PO) model assumes that each predictor shifts the cumulative log-odds by the same amount regardless of where you split the ordinal outcome.

a. Explain the PO assumption in non-technical terms using the essay grading context (Poor, Fair, Good, Excellent). What would it mean substantively if this assumption were violated for the `errors` predictor?
b. If the PO assumption is violated for a predictor, what alternatives are available? Name at least two modeling strategies and describe the trade-off each involves.
c. A Brant test yields a non-significant result ($p = 0.15$). Does this prove that the PO assumption holds? Why should you also examine graphical diagnostics rather than relying solely on the hypothesis test?
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.6

**Part (a):** The PO assumption means that the effect of one additional error on the odds of falling below any given grade boundary is the same. Violation might mean errors hurt the chances of getting "Excellent" more than they hurt the chances of avoiding "Poor."

**Part (b):** Alternatives include (1) a partial proportional odds model that relaxes PO for specific predictors, and (2) a multinomial logistic regression that treats each category comparison separately. The trade-off is parsimony vs. flexibility.

**Part (c):** Failure to reject does not prove the assumption holds---especially with small samples where the test has low power. Graphical checks (e.g., comparing category-specific log-odds slopes) can reveal patterns the test misses.
:::

::: {.callout-caution collapse="true"}
## Exercise 13.7: Interpreting Odds Ratios in Ordinal Models
In the essay grading model, the coefficient for `errors` is $\hat{\beta} = -0.56$, corresponding to an odds ratio of $e^{-0.56} \approx 0.57$.

a. Write a one-sentence interpretation of this odds ratio in the context of the ordinal model. Be precise about what "higher" means and what odds are being compared.
b. A critic argues: "An OR of 0.57 means that each additional error reduces the probability of getting an Excellent grade by 43%." Why is this interpretation incorrect? What does the OR actually describe?
c. If you had to communicate the effect of errors to a teacher who has never taken a statistics course, how would you translate the model results into language they could act on? Why might predicted probabilities or average marginal effects be more useful than odds ratios for this audience?
:::

::: {.callout-tip collapse="true"}
## Hint for Exercise 13.7

**Part (a):** Under the PO assumption, the OR applies to every cumulative split: for each additional error, the odds of being in a higher (better) grade category versus the current or lower category are multiplied by 0.57.

**Part (b):** The OR is a ratio of odds, not a ratio of probabilities. A 43% reduction in odds does not translate to a 43% reduction in probability. The actual probability change depends on the baseline probability.

**Part (c):** A teacher would likely find it most useful to hear something like: "An essay with 2 errors has an X% chance of being rated Excellent, while an essay with 8 errors has only a Y% chance." Predicted probabilities are directly interpretable without statistical training.
:::
