---
title: "Homework 1: Simple Linear Regression Fundamentals"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Distinguish between response and explanatory variables
- Connect intercepts and slopes to the visual shape of a line
- Check whether a prediction equation gives realistic values
- Interpret slopes and intercepts in the units of the problem
- Compute predicted values from a simple linear equation
- **Related notes:** [Chapter 1: Simple Linear Regression](../../notes/chapters/01-simple-linear-regression.html)
- **Related lab:** [Lab 2: Simple Linear Regression](../../labs/labs/lab02-simple-regression.html)
:::

::: {.callout-tip}
## Stata-to-R Translation
| Stata command | R equivalent | Used in |
|:---|:---|:---|
| `twoway function y = 7 + .5*x, range(0 10)` | `ggplot(...) + geom_line(...)` | Problem 2 |
| `display 0.5 + 0.7 * 3` | `0.5 + 0.7 * 3` | Problem 3 |
:::

This homework is mostly conceptual. The R code is intentionally simple: we use small tables to check arithmetic and `ggplot2` to draw the lines that students would otherwise sketch by hand.


## Problem 1

### Problem Statement

> For each of the following pairs of variables, specify which is the response variable and which is the explanatory variable.
>
> (a) College GPA and High School GPA
> (b) Number of children and mother's education level
> (c) Annual income and number of years of education
> (d) Annual income and assessed value of home

*Adapted from @agresti2018, Chapter 9.*

### Solution

The response variable is the outcome we want to explain or predict. The explanatory variable is the variable used to make that prediction. These roles follow the research question, so the choices below use the most natural research question for each pair.

```{r}
#| label: hw01-p1-variable-roles

variable_roles <- tibble(
  part = c("(a)", "(b)", "(c)", "(d)"),
  response_y = c(
    "College GPA",
    "Number of children",
    "Annual income",
    "Assessed value of home"
  ),
  explanatory_x = c(
    "High School GPA",
    "Mother's education level",
    "Number of years of education",
    "Annual income"
  ),
  research_question = c(
    "Use high school performance to predict later college performance.",
    "Study whether mother's education is associated with family size.",
    "Use schooling to explain or predict income.",
    "Use income or purchasing power to explain home value."
  )
)

variable_roles |>
  knitr::kable(
    col.names = c("Part", "Response variable (y)", "Explanatory variable (x)", "Research question"),
    align = c("c", "l", "l", "l"),
    caption = "Variable roles for Problem 1."
  )
```

```{r}
#| label: hw01-p1-checks
#| include: false

stopifnot(
  identical(variable_roles$response_y, c(
    "College GPA",
    "Number of children",
    "Annual income",
    "Assessed value of home"
  )),
  identical(variable_roles$explanatory_x, c(
    "High School GPA",
    "Mother's education level",
    "Number of years of education",
    "Annual income"
  ))
)
```

The main habit is to ask: "What are we trying to predict?" In part (a), high school GPA comes before college GPA, so it is natural to use it as the predictor. In part (c), years of education is usually treated as the predictor and annual income as the outcome. In part (d), the solution treats annual income as the predictor because income helps determine what home value a person can afford.


## Problem 2

### Problem Statement

> For $x$ between 0 and 10, sketch the following lines:
>
> (a) $y = 7 + 0.5x$
> (b) $y = 7 + x$
> (c) $y = 7 - x$
> (d) $y = 7 - 0.5x$
> (e) $y = 7$
> (f) $y = x$

### Solution

The intercept tells us where the line crosses the $y$-axis. The slope tells us how much $y$ changes when $x$ increases by 1.

```{r}
#| label: hw01-p2-line-table

line_specs <- tibble(
  line = c("(a)", "(b)", "(c)", "(d)", "(e)", "(f)"),
  equation = c(
    "y = 7 + 0.5x",
    "y = 7 + x",
    "y = 7 - x",
    "y = 7 - 0.5x",
    "y = 7",
    "y = x"
  ),
  intercept = c(7, 7, 7, 7, 7, 0),
  slope = c(0.5, 1, -1, -0.5, 0, 1),
  y_at_0 = c(7, 7, 7, 7, 7, 0),
  y_at_10 = c(12, 17, -3, 2, 7, 10),
  direction = c("Rises", "Rises", "Falls", "Falls", "Flat", "Rises"),
  steepness = c("Moderate", "Steep", "Steep", "Moderate", "None", "Steep")
)

line_specs |>
  select(line, equation, intercept, slope, y_at_0, y_at_10, direction, steepness) |>
  knitr::kable(
    col.names = c("Line", "Equation", "Intercept", "Slope", "y at x = 0", "y at x = 10", "Direction", "Steepness"),
    align = c("c", "l", "r", "r", "r", "r", "l", "l"),
    caption = "Arithmetic scaffold for sketching the six lines."
  )
```

```{r}
#| label: hw01-p2-checks
#| include: false

stopifnot(
  isTRUE(all.equal(line_specs$intercept, c(7, 7, 7, 7, 7, 0))),
  isTRUE(all.equal(line_specs$slope, c(0.5, 1, -1, -0.5, 0, 1))),
  isTRUE(all.equal(line_specs$y_at_0, c(7, 7, 7, 7, 7, 0))),
  isTRUE(all.equal(line_specs$y_at_10, c(12, 17, -3, 2, 7, 10)))
)
```

Lines (a) through (e) all start at $(0, 7)$. Line (f) starts at the origin, $(0, 0)$. Positive slopes rise, negative slopes fall, and a slope of 0 gives a horizontal line.

```{r}
#| label: fig-hw01-six-lines
#| fig-cap: "Six lines for x between 0 and 10. Lines (a)-(e) share y-intercept 7, while line (f) passes through the origin."
#| fig-alt: "A line plot showing six straight lines: a is y equals 7 plus 0.5x and rises moderately; b is y equals 7 plus x and rises steeply; c is y equals 7 minus x and falls steeply; d is y equals 7 minus 0.5x and falls moderately; e is y equals 7 and is flat; f is y equals x and rises from the origin."
#| fig-width: 8
#| fig-height: 5
#| echo: false

x_grid <- seq(0, 10, length.out = 201)

line_plot <- tidyr::expand_grid(line_specs, x = x_grid) |>
  mutate(y = intercept + slope * x)

line_labels <- line_plot |>
  filter(x == max(x_grid)) |>
  mutate(label = paste(line, equation))

line_colors <- c(
  "(a)" = unname(ber640_colors["primary"]),
  "(b)" = unname(ber640_colors["secondary"]),
  "(c)" = unname(ber640_colors["accent"]),
  "(d)" = unname(ber640_colors["success"]),
  "(e)" = unname(ber640_colors["warning"]),
  "(f)" = unname(ber640_colors["info"])
)

ggplot(line_plot, aes(x = x, y = y, color = line, linetype = line)) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = tibble(x = c(0, 0), y = c(7, 0), label = c("a-e", "f")),
    aes(x = x, y = y),
    inherit.aes = FALSE,
    color = "grey20",
    size = 2.5
  ) +
  geom_text(
    data = line_labels,
    aes(label = label),
    hjust = 0,
    nudge_x = 0.25,
    size = 3.3,
    show.legend = FALSE
  ) +
  annotate("label", x = 1.5, y = 8.0, label = "a-e share (0, 7)", size = 3.2) +
  annotate("label", x = 1.3, y = 0.9, label = "f starts at (0, 0)", size = 3.2) +
  geom_hline(yintercept = 7, linewidth = 0.4, color = "grey70") +
  scale_color_manual(values = line_colors) +
  scale_linetype_manual(values = c(
    "(a)" = "solid",
    "(b)" = "dashed",
    "(c)" = "dotdash",
    "(d)" = "longdash",
    "(e)" = "dotted",
    "(f)" = "twodash"
  )) +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  coord_cartesian(xlim = c(0, 11.6), ylim = c(-3.5, 18)) +
  labs(
    x = "x",
    y = "y",
    title = "Sketching Six Linear Equations"
  ) +
  guides(color = "none", linetype = "none")
```

::: {.callout-tip}
## Reading the Sketch
The sign of the slope gives the direction of the line. The absolute value of the slope gives the steepness. For example, line (b) rises more steeply than line (a) because $1 > 0.5$, and line (c) falls more steeply than line (d) because $|-1| > |-0.5|$.
:::


## Problem 3

### Problem Statement

> A college admissions officer claims that for students at that college the prediction equation $\hat{y}_i = 0.5 + 7x_i$ approximates the relationship between college GPA ($y$) and high school GPA ($x$), both ranging from 0 to 4.
>
> (a) Is this equation realistic? Why or why not?
> (b) Suppose the prediction equation is actually $\hat{y}_i = 0.5 + 0.7x_i$. Interpret the slope.
> (c) Using the prediction equation in (b), find the predicted GPA for a student having a high school GPA of (i) 3.0 and (ii) 4.0.
> (d) Suppose the prediction equation is $\hat{y}_i = x_i$. Identify the intercept and slope and interpret their values.

::: {.callout-caution}
## Proposed Line vs. Estimated Line
These equations are proposed prediction equations. They are not estimated from data in this problem. Later homework solutions use `lm()` to estimate coefficients from actual data, but here the coefficients are fixed numbers we are asked to interpret.
:::

### Part (a): Is $\hat{y} = 0.5 + 7x$ realistic?

No. A college GPA must stay between 0 and 4, but this equation gives impossible predicted values for ordinary high school GPAs.

For example, at $x = 1$:

$$
\hat{y} = 0.5 + 7(1) = 7.5.
$$

A predicted college GPA of 7.5 is impossible on a 0-to-4 GPA scale.

The exact boundary is:

$$
\begin{aligned}
0.5 + 7x &= 4 \\
7x &= 3.5 \\
x &= 0.5.
\end{aligned}
$$

So the equation predicts a GPA above 4 whenever $x > 0.5$.

```{r}
#| label: hw01-p3a-boundary-check
#| include: false

gpa_boundary <- (4 - 0.5) / 7
impossible_example <- 0.5 + 7 * 1

stopifnot(
  isTRUE(all.equal(gpa_boundary, 0.5)),
  isTRUE(all.equal(impossible_example, 7.5))
)
```

### Part (b): Interpret the slope of $\hat{y} = 0.5 + 0.7x$

The slope is $\hat{\beta}_1 = 0.7$. For every 1-point increase in high school GPA, the predicted college GPA increases by 0.7 points.

### Part (c): Find the predicted college GPA

Using $\hat{y} = 0.5 + 0.7x$:

```{r}
#| label: hw01-p3c-prediction-table

gpa_predictions <- tibble(
  high_school_gpa = c(3.0, 4.0),
  calculation = c("0.5 + 0.7 * 3.0", "0.5 + 0.7 * 4.0"),
  predicted_college_gpa = 0.5 + 0.7 * high_school_gpa
)

gpa_predictions |>
  knitr::kable(
    col.names = c("High school GPA", "Calculation", "Predicted college GPA"),
    align = c("r", "l", "r"),
    caption = "Predicted college GPA from y-hat = 0.5 + 0.7x."
  )
```

```{r}
#| label: hw01-p3c-checks
#| include: false

stopifnot(
  isTRUE(all.equal(gpa_predictions$predicted_college_gpa, c(2.6, 3.3)))
)
```

Thus:

- If high school GPA is 3.0, the predicted college GPA is 2.6.
- If high school GPA is 4.0, the predicted college GPA is 3.3.

### Part (d): Interpret $\hat{y} = x$

The equation $\hat{y} = x$ is the same as:

$$
\hat{y} = 0 + 1x.
$$

So the intercept is $\hat{\beta}_0 = 0$ and the slope is $\hat{\beta}_1 = 1$.

- **Intercept:** a student with high school GPA 0 has predicted college GPA 0.
- **Slope:** for every 1-point increase in high school GPA, the predicted college GPA increases by 1 point.

In other words, this equation predicts that college GPA equals high school GPA exactly.

```{r}
#| label: fig-hw01-gpa-equations
#| fig-cap: "Three proposed GPA prediction equations. The steep equation leaves the valid 0-to-4 GPA range immediately after x = 0.5."
#| fig-alt: "A plot of predicted college GPA against high school GPA from zero to four. A shaded band marks valid college GPA values from zero to four. The line y-hat equals 0.5 plus 7x crosses y equals four at x equals 0.5 and reaches y equals 7.5 at x equals one. The lines y-hat equals 0.5 plus 0.7x and y-hat equals x remain in the valid GPA range."
#| fig-width: 7.5
#| fig-height: 5
#| echo: false

gpa_equations <- tibble(
  equation = c("y-hat = 0.5 + 7x", "y-hat = 0.5 + 0.7x", "y-hat = x"),
  intercept = c(0.5, 0.5, 0),
  slope = c(7, 0.7, 1)
)

gpa_plot_data <- tidyr::expand_grid(gpa_equations, x = seq(0, 4, length.out = 201)) |>
  mutate(y = intercept + slope * x)

gpa_points <- tibble(
  x = c(0.5, 1, 3, 4),
  y = c(4, 7.5, 2.6, 3.3),
  label = c("(0.5, 4)", "(1, 7.5)", "(3, 2.6)", "(4, 3.3)")
)

ggplot(gpa_plot_data, aes(x = x, y = y, color = equation, linetype = equation)) +
  annotate("rect", xmin = 0, xmax = 4, ymin = 0, ymax = 4, fill = "grey90", alpha = 0.8) +
  geom_hline(yintercept = 4, color = "grey35", linetype = "dashed", linewidth = 0.5) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = gpa_points,
    aes(x = x, y = y),
    inherit.aes = FALSE,
    color = "grey15",
    size = 2.5
  ) +
  geom_text(
    data = gpa_points,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = c(1.1, -0.1, 1.1, 1.1),
    vjust = c(-0.6, -0.6, 1.4, -0.6),
    size = 3.2
  ) +
  annotate("label", x = 3.0, y = 3.75, label = "valid GPA range", size = 3.2) +
  scale_color_manual(values = c(
    "y-hat = 0.5 + 7x" = unname(ber640_colors["secondary"]),
    "y-hat = 0.5 + 0.7x" = unname(ber640_colors["primary"]),
    "y-hat = x" = unname(ber640_colors["accent"])
  )) +
  scale_linetype_manual(values = c(
    "y-hat = 0.5 + 7x" = "solid",
    "y-hat = 0.5 + 0.7x" = "dashed",
    "y-hat = x" = "dotdash"
  )) +
  scale_x_continuous(breaks = 0:4) +
  scale_y_continuous(breaks = 0:8) +
  coord_cartesian(xlim = c(0, 4), ylim = c(0, 8)) +
  labs(
    x = "High School GPA (x)",
    y = "Predicted College GPA",
    color = "Equation",
    linetype = "Equation",
    title = "Checking Whether Prediction Equations Are Realistic"
  )
```

::: {.callout-tip}
## Regression Habit
Before interpreting a prediction equation, check whether it gives possible values over the relevant range of $x$. A correct arithmetic substitution can still produce an impossible prediction if the equation itself is unrealistic.
:::


## Summary

In this homework, you practiced three foundations of simple linear regression:

1. **Variable roles:** identify the response variable and the explanatory variable from the research question.
2. **Line sketches:** use the intercept and slope to draw and compare linear equations.
3. **Prediction equations:** check realism, interpret slope and intercept, and compute predicted values.

These skills prepare you for HW2, where the same ideas are applied to real data using `lm()` and simple R summaries.
