---
title: "Homework 2: Simple Regression with a Dummy Variable"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Estimate and interpret regression with a single dummy predictor
- Understand the equivalence of regression t-test and independent samples t-test
- Identify Root MSE as the pooled within-group standard deviation
- **Related:** Chapter 1: *Simple Linear Regression* and Chapter 2: *ANCOVA* (in Lecture Notes)
:::

::: {.callout-tip}
## Stata-to-R Translation
| Stata Command | R Equivalent | Used In |
|:---|:---|:---|
| `regress wleread female if usa==1` | `lm(wleread ~ female, data = pisa_us)` | Problem 1 |
| `twoway lfit wleread female` | `ggplot() + geom_smooth(method = "lm")` | Problem 3 |
| `ttest wleread, by(female)` | `t.test(wleread ~ female, var.equal = TRUE)` | Problem 5 |
:::

**Dataset:** `pisa2000` from the `regdatasets` package (US subset, N = 1,020)

## Setup

```{r}
#| label: hw02-setup
#| code-fold: true
#| code-summary: "Show R Code"

data(pisa2000)
pisa_us <- pisa2000 |> filter(usa == 1)
```

## Problem 1

### Problem Statement

> Run the regression of `wleread` on `female` for American students.

::: {.panel-tabset}
#### Solution

```{r}
#| label: hw02-q1-regression
#| code-fold: true
#| code-summary: "Show R Code"

# Fit simple regression: reading score on female dummy
model1 <- lm(wleread ~ female, data = pisa_us)

# Display full regression summary
summary(model1)

# Tidy coefficient table with confidence intervals
tidy_model(model1)
```

- **Sample size:** $N = 1{,}020$ US students from the PISA 2000 dataset.
- **Model fit:** $F(1, 1018) = 21.61$, $p < .001$, indicating that `female` is a statistically significant predictor of reading scores.
- **Variance explained:** $R^2 = 0.0208$ --- the `female` dummy variable accounts for approximately 2.1% of the variance in reading scores. While statistically significant, this is a small effect.
- **Root MSE:** $\hat{\sigma} = 94.633$, which estimates the standard deviation of reading scores within each gender group (assuming equal variances).
- **Regression equation:**

$$\widehat{\text{wleread}} = 507.86 + 27.70 \times \text{female}$$

#### Code Explanation

```{r}
#| label: hw02-q1-regression-explain
#| eval: false

# Fit simple regression: reading score on female dummy
model1 <- lm(wleread ~ female, data = pisa_us)

# Display full regression summary
summary(model1)

# Tidy coefficient table with confidence intervals
tidy_model(model1)
```

- `lm(wleread ~ female, data = pisa_us)`: Fits an OLS regression predicting reading scores (`wleread`) from the dummy variable `female` (coded 0 = male, 1 = female). With a single dummy predictor, this is equivalent to comparing two group means.
- `summary(model1)`: Produces the full regression output including coefficients, standard errors, $t$-statistics, $p$-values, $R^2$, adjusted $R^2$, $F$-statistic, and residual standard error (Root MSE).
- `tidy_model(model1)`: A helper function (defined in `_common.R`) that calls `broom::tidy()` with confidence intervals and rounds numeric values to 4 decimal places for clean display.

:::

::: {.callout-tip}
## Key Takeaway
When the predictor is a dummy variable (0/1), the regression intercept equals the mean of the reference group (males), and the slope equals the difference in means between the two groups. This makes dummy variable regression equivalent to a two-sample comparison of means.
:::

## Problem 2

### Problem Statement

> Report and interpret the estimated regression coefficients and their standard errors.

::: {.panel-tabset}
#### Solution

```{r}
#| label: hw02-q2-interpretation
#| code-fold: true
#| code-summary: "Show R Code"

# Extract coefficients and standard errors
coefs <- tidy_model(model1)
coefs

# Verify: intercept = male mean, slope = female - male difference
pisa_us |>
  group_by(female) |>
  summarise(
    n = n(),
    mean_read = mean(wleread),
    sd_read = sd(wleread),
    .groups = "drop"
  )
```

**Intercept** ($\hat{\beta}_0 = 507.86$, $\text{SE} = 4.43$):

- The intercept is the predicted reading score when `female` $= 0$, i.e., the **mean reading score for male students**.
- Male US students scored an average of $507.9$ points on the PISA reading assessment.
- The standard error of $4.43$ is the standard deviation of $\hat{\beta}_0$ across hypothetical repeated samples. It quantifies the precision of our estimate of the male mean.

**Slope** ($\hat{\beta}_1 = 27.70$, $\text{SE} = 5.96$):

- The slope is the **difference in mean reading scores** between female and male students: $\bar{Y}_{\text{female}} - \bar{Y}_{\text{male}} = 27.7$ points.
- Female US students scored, on average, $27.7$ points higher than male students on the PISA reading assessment.
- The standard error of $5.96$ is the standard deviation of $\hat{\beta}_1$ across hypothetical repeated samples. It reflects how precisely we have estimated the gender gap in reading scores.

**What standard errors represent:**

- Standard errors are the **standard deviations of the sampling distributions** of the estimated coefficients.
- If we could draw many independent random samples of $N = 1{,}020$ US students and fit this regression each time, the standard errors tell us how much $\hat{\beta}_0$ and $\hat{\beta}_1$ would vary from sample to sample.
- Smaller standard errors indicate more precise estimates; larger standard errors indicate greater uncertainty.

#### Code Explanation

```{r}
#| label: hw02-q2-interpretation-explain
#| eval: false

# Extract coefficients and standard errors
coefs <- tidy_model(model1)
coefs

# Verify: intercept = male mean, slope = female - male difference
pisa_us |>
  group_by(female) |>
  summarise(
    n = n(),
    mean_read = mean(wleread),
    sd_read = sd(wleread),
    .groups = "drop"
  )
```

- `tidy_model(model1)`: Extracts the coefficient table with estimates, standard errors, test statistics, $p$-values, and 95% confidence intervals in a tidy data frame.
- `group_by(female) |> summarise(...)`: Computes descriptive statistics by gender group to verify that the regression intercept matches the male mean and the slope matches the difference in means.
- `mean(wleread)`: Computes the group mean of reading scores, which should match $\hat{\beta}_0$ for males and $\hat{\beta}_0 + \hat{\beta}_1$ for females.
- `sd(wleread)`: Computes the within-group standard deviation, which is related to (but not identical to) the Root MSE from the regression.

:::

::: {.callout-tip}
## Key Takeaway
In a regression with a single dummy predictor, the standard errors have a direct interpretation: $\text{SE}(\hat{\beta}_0)$ measures uncertainty about the reference group mean, and $\text{SE}(\hat{\beta}_1)$ measures uncertainty about the difference between groups. These are the standard deviations of the estimates if we could repeat the study many times.
:::

## Problem 3

### Problem Statement

> Graph the regression relationship.

::: {.panel-tabset}
#### Simple Version

```{r}
#| label: hw02-q3-plot-simple
#| code-fold: true
#| code-summary: "Show R Code"
#| fig-cap: "Regression of reading score on gender with jittered data points."
#| fig-alt: "Scatter plot with jittered points showing reading scores for males (x equals 0) and females (x equals 1). A blue regression line connects the two group means, rising from about 508 for males to about 536 for females, indicating a 28-point gender gap."
#| fig-width: 8
#| fig-height: 5

# Simple version: fitted line plot
ggplot(pisa_us, aes(x = female, y = wleread)) +
  geom_jitter(alpha = 0.1, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_x_continuous(breaks = c(0, 1), labels = c("Male", "Female")) +
  labs(x = "Gender", y = "Reading Score",
       title = "Regression of Reading Score on Gender")
```

#### Detailed Version

```{r}
#| label: hw02-q3-plot
#| code-fold: true
#| code-summary: "Show R Code"
#| fig-cap: "Regression of reading score on gender with annotated group means."
#| fig-alt: "Two large points connected by a blue line showing male mean reading score of 507.9 at x equals 0 and female mean of 535.6 at x equals 1. A dashed red vertical segment at x equals 0.5 indicates the 27.7-point difference between groups."
#| fig-width: 8
#| fig-height: 6

# Extract fitted values for plotting
b0 <- coef(model1)[1]  # intercept (male mean)
b1 <- coef(model1)[2]  # slope (female - male difference)

# Create data frame for the two group means
fitted_points <- tibble(
  female = c(0, 1),
  fitted = c(b0, b0 + b1),
  label = c(
    paste0("Males: ", round(b0, 1)),
    paste0("Females: ", round(b0 + b1, 1))
  )
)

# Plot the fitted regression line with group means
ggplot(fitted_points, aes(x = female, y = fitted)) +
  # Regression line connecting the two means
  geom_line(
    color = ber640_colors["accent"],
    linewidth = 1.2
  ) +
  # Points at each group mean
  geom_point(
    size = 5,
    color = ber640_colors["primary"]
  ) +
  # Labels for each point
  geom_label(
    aes(label = label),
    vjust = -1.2,
    size = 4.5,
    fontface = "bold",
    color = ber640_colors["primary"],
    fill = "white",
    label.padding = unit(0.4, "lines")
  ) +
  # Annotate the slope (difference)
  annotate(
    "segment",
    x = 0.5, xend = 0.5,
    y = b0, yend = b0 + b1,
    color = ber640_colors["secondary"],
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 0.55, y = b0 + b1 / 2,
    label = paste0("Difference = ", round(b1, 1)),
    hjust = 0, size = 4,
    color = ber640_colors["secondary"],
    fontface = "italic"
  ) +
  # Axis formatting
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c("0\n(Male)", "1\n(Female)"),
    limits = c(-0.3, 1.5)
  ) +
  scale_y_continuous(
    limits = c(480, 560)
  ) +
  labs(
    title = "Regression of Reading Score on Gender",
    subtitle = "PISA 2000, US Students (N = 1,020)",
    x = "Female (Dummy Variable)",
    y = "Predicted Reading Score (wleread)"
  )
```

:::

- **The plot displays the fitted regression line** connecting the predicted mean reading score for males ($x = 0$, $\hat{Y} = 507.9$) to the predicted mean for females ($x = 1$, $\hat{Y} = 535.6$).
- **The slope of the line** represents the estimated gender gap: female students scored $27.7$ points higher on average.
- **With a dummy predictor, the "regression line" is simply two points** --- the means of the two groups connected by a straight line. This highlights that dummy variable regression is fundamentally a comparison of group means.

::: {.callout-tip}
## Key Takeaway
With a dummy variable predictor, the regression "line" is just two points connected by a segment. The plot makes the regression equation tangible: the intercept is the left point (reference group mean) and the slope is the vertical rise to the right point (comparison group mean). This visual reinforces why dummy variable regression is simply a comparison of two group means expressed as a linear model.
:::

## Problem 4

### Problem Statement

> Test the null hypothesis that `female` has no effect on `wleread` at the 1% significance level.

::: {.panel-tabset}
#### Solution

```{r}
#| label: hw02-q4-test
#| code-fold: true
#| code-summary: "Show R Code"

# Extract coefficient table from the regression model
coef_table <- summary(model1)$coefficients

# Pull out the t-statistic and p-value for the female coefficient
t_stat <- coef_table["female", "t value"]
p_value <- coef_table["female", "Pr(>|t|)"]
df_error <- model1$df.residual

# Display the test results
cat("Hypothesis Test for the Effect of Female on Reading Score\n")
cat("=========================================================\n")
cat(sprintf("  t-statistic : %.2f\n", t_stat))
cat(sprintf("  df          : %d\n", df_error))
cat(sprintf("  p-value     : %.6f\n", p_value))
cat(sprintf("  alpha       : 0.01\n"))
cat("=========================================================\n")

# Critical value for two-tailed test at 1% level
t_crit <- qt(0.995, df = df_error)
cat(sprintf("  Critical value (two-tailed): +/- %.3f\n", t_crit))
cat(sprintf("  |t| = %.2f > %.3f = t_crit => Reject H0\n", abs(t_stat), t_crit))
```

**Formal hypothesis test:**

1. **Hypotheses:**
   - $H_0: \beta_1 = 0$ (no difference in mean reading scores between female and male students)
   - $H_a: \beta_1 \neq 0$ (there is a difference)

2. **Test statistic:** $t = \dfrac{\hat{\beta}_1}{\text{SE}(\hat{\beta}_1)} = 4.65$ with $df = 1018$

3. **Decision rule:** At $\alpha = 0.01$ (two-tailed), reject $H_0$ if $|t| > t_{0.005,\, 1018}$

4. **Conclusion:** Since $p < 0.001 < 0.01 = \alpha$, we **reject $H_0$** at the 1% significance level. There is statistically significant evidence that mean reading scores differ between female and male US students in PISA 2000.

#### Code Explanation

```{r}
#| label: hw02-q4-test-explain
#| eval: false

# Extract coefficient table from the regression model
coef_table <- summary(model1)$coefficients
```

- `summary(model1)$coefficients` returns the matrix of estimates, standard errors, $t$-values, and $p$-values for all coefficients.

```{r}
#| label: hw02-q4-test-explain2
#| eval: false

# Pull out the t-statistic and p-value for the female coefficient
t_stat <- coef_table["female", "t value"]
p_value <- coef_table["female", "Pr(>|t|)"]
df_error <- model1$df.residual
```

- We index the coefficient matrix by row name `"female"` and column name to extract the specific values.
- `model1$df.residual` gives the error degrees of freedom: $n - k - 1 = 1020 - 1 - 1 = 1018$.

```{r}
#| label: hw02-q4-test-explain3
#| eval: false

# Critical value for two-tailed test at 1% level
t_crit <- qt(0.995, df = df_error)
```

- `qt(0.995, df)` returns the 99.5th percentile of the $t$-distribution, which is the upper critical value for a two-tailed test at $\alpha = 0.01$.

:::

::: {.callout-tip}
## Key Takeaway
The regression $t$-test for $\beta_1$ is equivalent to testing whether the dummy variable's group means differ. A significant $\beta_1$ means the two groups have different population means. The $p$-value from the regression output can be directly used for hypothesis testing at any significance level.
:::

::: {.callout-note}
## Association, Not Causation
Because PISA 2000 is an observational study, we describe the relationship between gender and reading scores as an **association** or **mean difference**, not a causal **effect**. The hypothesis test asks whether the observed difference is statistically distinguishable from zero --- it does not tell us whether gender *causes* the difference. The original homework uses "effect" as Stata/statistics shorthand, but in research writing, "association" is safer language for observational data.
:::


## Problem 5

### Problem Statement

> Test the same null hypothesis using an independent samples t-test. Compare with the regression result.

::: {.panel-tabset}
#### Solution

```{r}
#| label: hw02-q5-ttest
#| code-fold: true
#| code-summary: "Show R Code"

# Compute descriptive statistics by group
group_stats <- pisa_us |>
  summarize(
    .by = female,
    n    = n(),
    mean = mean(wleread),
    sd   = sd(wleread)
  ) |>
  arrange(female)

# Combined statistics
combined_stats <- pisa_us |>
  summarize(
    n    = n(),
    mean = mean(wleread),
    sd   = sd(wleread)
  )

cat("Descriptive Statistics\n")
cat("======================\n")
cat(sprintf("  Male   (female=0): N = %d, Mean = %.3f, SD = %.5f\n",
            group_stats$n[1], group_stats$mean[1], group_stats$sd[1]))
cat(sprintf("  Female (female=1): N = %d, Mean = %.4f, SD = %.5f\n",
            group_stats$n[2], group_stats$mean[2], group_stats$sd[2]))
cat(sprintf("  Combined         : N = %d, Mean = %.4f, SD = %.5f\n",
            combined_stats$n, combined_stats$mean, combined_stats$sd))
cat("\n")

# Independent samples t-test (equal variance assumed)
t_result <- t.test(wleread ~ female, data = pisa_us, var.equal = TRUE)
print(t_result)
```

**Comparing the t-test with regression:**

```{r}
#| label: hw02-q5-compare
#| code-fold: true
#| code-summary: "Show Comparison Code"

# Extract values from both approaches
reg_t <- summary(model1)$coefficients["female", "t value"]
reg_p <- summary(model1)$coefficients["female", "Pr(>|t|)"]

ttest_t <- t_result$statistic
ttest_p <- t_result$p.value
ttest_diff <- diff(t_result$estimate)  # female mean - male mean

cat("Side-by-Side Comparison\n")
cat("============================================\n")
cat(sprintf("  Regression t-statistic :  %.2f\n", reg_t))
cat(sprintf("  t-test t-statistic     : %.4f\n", ttest_t))
cat(sprintf("  Regression p-value     :  %.6f\n", reg_p))
cat(sprintf("  t-test p-value         :  %.6f\n", ttest_p))
cat("============================================\n")
cat(sprintf("  Mean difference (t-test): %.5f\n", ttest_diff))
cat(sprintf("  Regression slope (b1)  :  %.5f\n", coef(model1)["female"]))
cat("============================================\n")
cat("\nThe t-statistics have the SAME absolute value but OPPOSITE signs.\n")
cat("The p-values are identical.\n")
```

- The $t$-test yields $t = -4.6486$ while the regression yields $t = 4.65$.
- The **absolute values are identical**; the sign difference arises because:
  - The $t$-test computes: $\bar{Y}_{\text{male}} - \bar{Y}_{\text{female}}$ (group 0 minus group 1)
  - The regression computes: $\hat{\beta}_1 = \bar{Y}_{\text{female}} - \bar{Y}_{\text{male}}$ (group 1 minus group 0)
- The **$p$-values are identical** ($p < 0.001$), so both approaches lead to the same conclusion: reject $H_0$ at the 1% level.

#### Code Explanation

```{r}
#| label: hw02-q5-ttest-explain
#| eval: false

# Compute descriptive statistics by group
group_stats <- pisa_us |>
  summarize(
    .by = female,
    n    = n(),
    mean = mean(wleread),
    sd   = sd(wleread)
  ) |>
  arrange(female)
```

- `.by = female` groups the summary by the dummy variable without using `group_by()`.
- We compute the sample size, mean, and standard deviation for each group.
- `arrange(female)` ensures male (0) appears first.

```{r}
#| label: hw02-q5-ttest-explain2
#| eval: false

# Independent samples t-test (equal variance assumed)
t_result <- t.test(wleread ~ female, data = pisa_us, var.equal = TRUE)
```

- `wleread ~ female` specifies the formula: outcome by grouping variable.
- `var.equal = TRUE` requests the **pooled-variance** $t$-test, which assumes equal population variances. This is the version that is algebraically equivalent to the regression $t$-test.
- Without `var.equal = TRUE`, R defaults to Welch's $t$-test (unequal variances), which would give different $df$ and a slightly different $t$-statistic.

```{r}
#| label: hw02-q5-ttest-explain3
#| eval: false

# Extract values from both approaches
ttest_diff <- diff(t_result$estimate)  # female mean - male mean
```

- `t_result$estimate` contains the two group means; `diff()` computes female minus male.
- This equals the regression slope $\hat{\beta}_1$.

:::

::: {.callout-important}
## Regression = t-test (with Equal Variances)
When the predictor is a single dummy variable, the regression $t$-test for $\beta_1$ is **algebraically identical** to the pooled-variance independent samples $t$-test. The test statistics have the same absolute value, the degrees of freedom are the same ($n - 2$), and the $p$-values are identical. The only difference is the sign convention.
:::


## Problem 6

### Problem Statement

> Where is the pooled within-group standard deviation in the regression output? Explain the connection between Root MSE and the pooled standard deviation.

::: {.panel-tabset}
#### Solution

**Step 1: Extract Root MSE from the regression**

```{r}
#| label: hw02-q6-pooled
#| code-fold: true
#| code-summary: "Show R Code"

# Root MSE from the regression
root_mse <- sigma(model1)
cat(sprintf("Root MSE (from regression): %.3f\n\n", root_mse))

# Manually compute the pooled within-group SD
n1 <- group_stats$n[1]  # male
n2 <- group_stats$n[2]  # female
s1 <- group_stats$sd[1]  # male SD
s2 <- group_stats$sd[2]  # female SD

pooled_var <- ((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2)
pooled_sd  <- sqrt(pooled_var)

cat(sprintf("Pooled within-group SD (manual calculation): %.3f\n\n", pooled_sd))
cat(sprintf("Are they equal? %s\n", all.equal(root_mse, pooled_sd)))
```

**Step 2: Verify using ANOVA table**

```{r}
#| label: hw02-q6-anova
#| code-fold: true
#| code-summary: "Show ANOVA Verification"

# ANOVA table
anova_table <- anova(model1)
print(anova_table)
cat("\n")

mse <- anova_table["Residuals", "Mean Sq"]
cat(sprintf("MSE from ANOVA table: %.3f\n", mse))
cat(sprintf("sqrt(MSE) = %.3f\n", sqrt(mse)))
cat(sprintf("Root MSE   = %.3f\n", root_mse))
```

**Step 3: Mathematical proof that $s_p = \sqrt{\text{MSE}}$**

The pooled within-group variance is defined as:

$$
s_p^2 = \frac{(n_1 - 1)\, s_1^2 + (n_2 - 1)\, s_2^2}{n_1 + n_2 - 2}
$$

Recall that the sample variance for group $j$ is:

$$
s_j^2 = \frac{\sum_{i=1}^{n_j} (Y_{ij} - \bar{Y}_j)^2}{n_j - 1} = \frac{\text{SS}_j}{n_j - 1}
$$

Substituting into the pooled variance formula:

$$
s_p^2 = \frac{(n_1 - 1) \cdot \dfrac{\text{SS}_1}{n_1 - 1} + (n_2 - 1) \cdot \dfrac{\text{SS}_2}{n_2 - 1}}{n_1 + n_2 - 2}
$$

The $(n_j - 1)$ terms cancel:

$$
s_p^2 = \frac{\text{SS}_1 + \text{SS}_2}{n_1 + n_2 - 2} = \frac{\text{SSE}}{df_{\text{error}}} = \text{MSE}
$$

where $\text{SS}_1 + \text{SS}_2 = \text{SSE}$ (the sum of squared residuals) because in a dummy variable regression, the residual for each observation is $e_i = Y_i - \bar{Y}_{j(i)}$, i.e., the deviation from its own group mean.

Therefore:

$$
\boxed{s_p = \sqrt{s_p^2} = \sqrt{\text{MSE}} = \text{Root MSE} = 94.633}
$$

#### Code Explanation

```{r}
#| label: hw02-q6-pooled-explain
#| eval: false

# Root MSE from the regression
root_mse <- sigma(model1)
```

- `sigma()` extracts the residual standard error (Root MSE) from an `lm` object. This is $\sqrt{\text{MSE}} = \sqrt{\text{SSE}/df_{\text{error}}}$.

```{r}
#| label: hw02-q6-pooled-explain2
#| eval: false

# Manually compute the pooled within-group SD
n1 <- group_stats$n[1]  # male
n2 <- group_stats$n[2]  # female
s1 <- group_stats$sd[1]  # male SD
s2 <- group_stats$sd[2]  # female SD

pooled_var <- ((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2)
pooled_sd  <- sqrt(pooled_var)
```

- This implements the textbook formula for the pooled variance: a weighted average of the two group variances, where the weights are $(n_j - 1)$.
- The denominator $n_1 + n_2 - 2 = 1018$ equals the error degrees of freedom in the regression.

```{r}
#| label: hw02-q6-anova-explain
#| eval: false

# Verify using ANOVA table
anova_table <- anova(model1)
mse <- anova_table["Residuals", "Mean Sq"]
```

- `anova()` produces the ANOVA decomposition of the regression.
- The "Residuals" row's "Mean Sq" column is $\text{MSE} = \text{SSE}/df_{\text{error}}$.
- We verify that $\sqrt{\text{MSE}}$ equals both Root MSE and the pooled SD.

:::

::: {.callout-warning}
## Combined SD is NOT the Pooled SD
The "combined" standard deviation from descriptive statistics ($\text{SD} = 95.586$ for all 1,020 students) is **not** the pooled within-group standard deviation (94.633). The combined SD ignores group membership and includes between-group variation in its calculation. The pooled SD averages only the **within-group** variation, which is exactly what the regression's Root MSE captures.
:::

::: {.callout-tip}
## Key Takeaway
Root MSE in a dummy variable regression is the pooled within-group standard deviation. This is not a coincidence --- it follows directly from the algebra. In a regression with a single dummy predictor, the fitted values are the group means, so the residuals are deviations from group means, and MSE is the pooled within-group variance. This insight generalizes: in any ANOVA model, Root MSE equals the pooled within-group SD.
:::


::: {.callout-caution}
## Common Mistake: Dismissing a Small $R^2$ as "No Relationship"
With $R^2 = 0.0208$, it is tempting to conclude that gender "doesn't matter." But $R^2$ measures the proportion of variance explained, not the importance or significance of the predictor. A 27.7-point gender gap is highly significant ($p < .001$) and may be substantively meaningful in an educational context, even though it explains only 2% of the total variance in reading scores. Always evaluate both statistical significance and practical significance.
:::


## Summary

This homework demonstrated three fundamental equivalences that connect simple regression with a dummy variable to classical two-group comparisons:

1. **Regression coefficients = group means.** The intercept $\hat{\beta}_0$ equals the reference group mean, and the slope $\hat{\beta}_1$ equals the difference in group means. Regression with a dummy predictor is simply a way of comparing two groups.

2. **Regression $t$-test = independent samples $t$-test.** The $t$-statistic for $\hat{\beta}_1$ has the same absolute value, the same degrees of freedom ($n - 2$), and the same $p$-value as the pooled-variance $t$-test. The sign depends only on the coding convention (which group is subtracted from which).

3. **Root MSE = pooled within-group SD.** This follows algebraically from the fact that $s_p^2 = \text{SSE}/df_{\text{error}} = \text{MSE}$. The combined (overall) SD is a different quantity that conflates within- and between-group variation.

::: {.callout-note}
## Looking Ahead
These equivalences generalize beyond two groups. In **Homework 4**, we will see that regression with multiple dummy variables is equivalent to one-way ANOVA, and the $F$-test from the regression is identical to the ANOVA $F$-test. The Root MSE will again equal the pooled within-group SD, but now pooled across more than two groups.
:::
