---
title: "Homework 2: Simple Regression with a Dummy Variable"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Estimate and interpret regression with a single dummy predictor
- Explain why the intercept is the reference-group mean
- Explain why the slope is the difference between two group means
- Compare the regression $t$-test with the pooled two-sample $t$-test
- Identify Root MSE as the pooled within-group standard deviation
- **Related notes:** [Chapter 1: Simple Linear Regression](../../notes/chapters/01-simple-linear-regression.html) and [Chapter 2: ANCOVA](../../notes/chapters/02-ancova.html)
- **Related lab:** [Lab 2: Simple Linear Regression](../../labs/labs/lab02-simple-regression.html)
:::

::: {.callout-tip}
## Stata-to-R Translation
| Stata command | R equivalent | Used in |
|:---|:---|:---|
| `regress wleread female if usa==1` | `lm(wleread ~ female, data = pisa_us)` | Problems 1-4 |
| `twoway lfit wleread female if usa==1` | `ggplot(...) + geom_line(...)` | Problem 3 |
| `ttest wleread if usa==1, by(female)` | `t.test(wleread ~ female, data = pisa_us, var.equal = TRUE)` | Problem 5 |
:::

**Dataset:** `pisa2000` from the `regdatasets` package.

We use only the US sample. In this homework, `female` is coded `0 = male` and `1 = female`.

## Setup

```{r}
#| label: hw02-setup

data(pisa2000)

pisa_us <- pisa2000 |>
  filter(usa == 1)

pisa_us |>
  summarise(
    n = n(),
    male = sum(female == 0),
    female = sum(female == 1),
    missing_reading = sum(is.na(wleread)),
    missing_female = sum(is.na(female))
  )
```

The analysis sample contains 1,020 US students: 457 males and 563 females.


## Problem 1

### Problem Statement

> Run the regression of `wleread` on `female` for American students.

### Solution

```{r}
#| label: hw02-p1-fit

model1 <- lm(wleread ~ female, data = pisa_us)

model_params(model1) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  select(Parameter, Coefficient, SE, CI_low, CI_high, t, df_error, p) |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw02-p1-fit-summary

hw02_fit <- model_glance(model1)

hw02_fit |>
  round_numeric(digits = 4)
```

The fitted regression equation is:

$$
\widehat{\text{wleread}} = 507.86 + 27.70 \times \text{female}.
$$

Key results:

- $N = 1{,}020$.
- The intercept is $507.86$.
- The coefficient for `female` is $27.70$.
- The overall regression test is $F(1, 1018) = 21.61$, $p < .001$.
- $R^2 = 0.0208$ and adjusted $R^2 = 0.0198$, so the gender dummy explains about 2.1% of the variation in reading scores; the adjusted value is almost identical because this model has only one predictor.
- Root MSE is $94.633$, which will matter again in Problem 6.

```{r}
#| label: hw02-p1-check
#| include: false

stopifnot(nrow(pisa_us) == 1020)
stopifnot(abs(coef(model1)[["(Intercept)"]] - 507.859) < 1e-3)
stopifnot(abs(coef(model1)[["female"]] - 27.69859) < 1e-5)
stopifnot(abs(summary(model1)$coefficients["female", "Std. Error"] - 5.958441) < 1e-6)
stopifnot(abs(summary(model1)$coefficients["female", "t value"] - 4.64863) < 1e-5)
stopifnot(abs(sigma(model1) - 94.6335) < 1e-4)
```

### Code Explanation

- `lm(wleread ~ female, data = pisa_us)` fits an ordinary least squares regression.
- The outcome is `wleread`, the PISA reading score.
- The predictor is `female`, a dummy variable coded `0 = male`, `1 = female`.
- `model_params(model1)` prints the coefficient table in a compact format.
- `format_p()` displays very small p-values as `< .001` rather than rounding them to 0.
- `model_glance(model1)` prints the model-level fit table: sample size, model and error degrees of freedom, the overall $F$-test, $R^2$, adjusted $R^2$, and Root MSE.
- In that fit table, `root_mse` is the regression Root MSE, the same quantity Stata labels Root MSE.

::: {.callout-tip}
## Key Takeaway
With a single dummy predictor, regression is a comparison of two means. The intercept is the mean for the reference group (`female = 0`, males), and the slope is the comparison-group mean minus the reference-group mean.
:::


## Problem 2

### Problem Statement

> Report and interpret the estimated regression coefficients and their standard errors.

### Solution

```{r}
#| label: hw02-p2-coefficients

model_params(model1) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  select(Parameter, Coefficient, SE, t, df_error, p, CI_low, CI_high) |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw02-p2-group-means

group_stats <- pisa_us |>
  summarise(
    .by = female,
    n = n(),
    mean_read = mean(wleread),
    sd_read = sd(wleread)
  ) |>
  arrange(female) |>
  mutate(group = if_else(female == 0, "Male", "Female")) |>
  select(group, female, n, mean_read, sd_read)

group_stats |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw02-p2-check
#| include: false

stopifnot(identical(group_stats$n, c(457L, 563L)))
stopifnot(abs(group_stats$mean_read[group_stats$group == "Male"] - 507.859) < 1e-3)
stopifnot(abs(group_stats$mean_read[group_stats$group == "Female"] - 535.5576) < 1e-4)
stopifnot(abs(diff(group_stats$mean_read) - coef(model1)[["female"]]) < 1e-10)
```

**Intercept** ($\hat{\beta}_0 = 507.86$, $\text{SE} = 4.43$):

- This is the predicted reading score when `female = 0`.
- Because `female = 0` means male, the intercept is the mean reading score for male US students.
- The standard error of $4.43$ describes the sampling uncertainty in the estimated male mean.

**Slope** ($\hat{\beta}_1 = 27.70$, $\text{SE} = 5.96$):

- This is the mean reading-score difference between female and male students.
- Female US students scored about $27.7$ points higher than male US students, on average.
- The standard error of $5.96$ describes the sampling uncertainty in that estimated difference.

Standard errors are standard deviations of estimates over hypothetical repeated samples. If we repeatedly drew new samples of 1,020 US students and refit this model, the estimated coefficients would vary from sample to sample. The standard errors summarize that typical variation.

### Code Explanation

- `model_params(model1)` gives a compact coefficient table using the same helper used in the R labs.
- `group_stats` verifies the dummy-regression interpretation directly: the male mean is the intercept, and the female mean is the intercept plus the slope.

::: {.callout-tip}
## Key Takeaway
The coefficient table and the group means tell the same story: $535.56 - 507.86 = 27.70$. For a dummy predictor, the slope is a mean difference.
:::


## Problem 3

### Problem Statement

> Graph the regression relationship.

### Solution

```{r}
#| label: fig-hw02-p3-regression-line
#| fig-cap: "Dummy-variable regression of reading score on gender. The slope is the female-minus-male mean difference, about 27.7 points."
#| fig-alt: "Jittered points show reading scores for male and female students. Large points connected by a line show the male mean around 508 and the female mean around 536, so the line rises by about 28 points."
#| fig-width: 8
#| fig-height: 5
#| echo: false

mean_points <- group_stats |>
  mutate(fitted = mean_read)

ggplot(pisa_us, aes(x = female, y = wleread)) +
  geom_jitter(width = 0.08, alpha = 0.12, color = ber640_colors["primary"]) +
  geom_line(
    data = mean_points,
    aes(x = female, y = fitted),
    color = ber640_colors["secondary"],
    linewidth = 1.2
  ) +
  geom_point(
    data = mean_points,
    aes(x = female, y = fitted),
    color = ber640_colors["secondary"],
    size = 4
  ) +
  geom_text(
    data = mean_points,
    aes(x = female, y = fitted, label = round(fitted, 1)),
    vjust = -1,
    fontface = "bold",
    color = ber640_colors["secondary"]
  ) +
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c("Male\n(female = 0)", "Female\n(female = 1)")
  ) +
  labs(
    x = "Gender dummy",
    y = "Reading score",
    title = "Dummy-variable regression connects the two group means"
  )
```

The plot has only two fitted values:

- for males (`female = 0`), $\widehat{Y} = 507.86$;
- for females (`female = 1`), $\widehat{Y} = 535.56$.

The vertical change between those two fitted values is the slope:

$$
535.56 - 507.86 = 27.70.
$$

::: {.callout-tip}
## Key Takeaway
With a dummy predictor, the "regression line" is just the two group means connected by a line segment. The intercept is the left point, and the slope is the rise from the left point to the right point.
:::


## Problem 4

### Problem Statement

> Test the null hypothesis that `female` has no effect on `wleread` at the 1% significance level.

### Solution

```{r}
#| label: hw02-p4-test

female_test <- model_params(model1) |>
  as_tibble() |>
  filter(Parameter == "female") |>
  transmute(
    estimate = Coefficient,
    SE,
    t,
    df = df_error,
    p = format_p(p)
  )

female_test |>
  round_numeric(digits = 4)
```

The hypotheses are:

$$
H_0: \beta_1 = 0
\qquad \text{versus} \qquad
H_a: \beta_1 \neq 0.
$$

The test statistic is:

$$
t = \frac{27.70}{5.96} = 4.65,
\qquad df = 1018.
$$

At the 1% significance level, we reject $H_0$ because $p < .001 < .01$. There is statistically significant evidence that mean reading scores differ between female and male US students in PISA 2000.

::: {.callout-note}
## Association, Not Causation
The original homework uses "effect" as statistical shorthand, but PISA is observational. A safer interpretation is that gender is **associated with** mean reading-score differences in this sample. The test does not prove that gender causes the difference.
:::


## Problem 5

### Problem Statement

> Test the same null hypothesis using an independent samples t-test. Compare with the regression result.

### Solution

```{r}
#| label: hw02-p5-ttest

t_result <- t.test(wleread ~ female, data = pisa_us, var.equal = TRUE)
t_result
```

```{r}
#| label: hw02-p5-compare

reg_result <- model_params(model1) |>
  as_tibble() |>
  filter(Parameter == "female")

comparison <- tibble(
  method = c("Regression", "Pooled t-test"),
  difference_reported = c("Female - Male", "Male - Female"),
  estimate = c(
    coef(model1)[["female"]],
    unname(t_result$estimate[1] - t_result$estimate[2])
  ),
  t = c(reg_result$t, unname(t_result$statistic)),
  df = c(df.residual(model1), unname(t_result$parameter)),
  p = format_p(c(reg_result$p, t_result$p.value))
)

comparison |>
  round_numeric(digits = 4)
```

The regression and pooled $t$-test give the same test:

- Regression reports the difference as female minus male, so the estimate and $t$ are positive.
- The R $t$-test reports the difference as group 0 minus group 1, or male minus female, so the estimate and $t$ are negative.
- The absolute $t$-statistics are the same: $|t| = 4.6486$.
- The degrees of freedom are the same: $df = 1018$.
- The two-sided $p$-values are identical.

::: {.callout-important}
## Regression = Pooled Two-Sample t-test
When the predictor is a single dummy variable, the regression $t$-test for the slope is algebraically identical to the equal-variance independent samples $t$-test. The only sign difference comes from which group is subtracted from which.
:::


## Problem 6

### Problem Statement

> Where is the pooled within-group standard deviation in the regression output? Explain the connection between Root MSE and the pooled standard deviation.

### Solution

The code below verifies the Root MSE identity numerically. This formula code is included because the question asks where the pooled within-group SD appears in the regression output.

```{r}
#| label: hw02-p6-root-mse

root_mse <- sigma(model1)

pooled_sd <- with(
  group_stats,
  sqrt(((n[1] - 1) * sd_read[1]^2 + (n[2] - 1) * sd_read[2]^2) / (sum(n) - 2))
)

anova_table <- anova(model1)
anova_root_mse <- sqrt(anova_table["Residuals", "Mean Sq"])
combined_sd <- sd(pisa_us$wleread)

tibble(
  quantity = c(
    "Root MSE from regression",
    "Pooled within-group SD",
    "sqrt(residual Mean Sq)",
    "Combined SD ignoring gender"
  ),
  value = c(root_mse, pooled_sd, anova_root_mse, combined_sd)
) |>
  round_numeric(digits = 3)
```

The pooled within-group standard deviation is the **Root MSE** in the regression output:

$$
\text{Root MSE} = 94.633.
$$

It is not the combined SD of all 1,020 students. The combined SD ignores gender and mixes within-group variation with the between-group difference.

The pooled within-group variance is:

$$
s_p^2 =
\frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}.
$$

In a dummy-variable regression, the fitted values are group means. Therefore, the residuals are deviations from group-specific means. The regression residual sum of squares is the same within-group sum of squares used in the pooled variance:

$$
s_p^2
= \frac{\text{SSE}}{df_{\text{error}}}
= \text{MSE}.
$$

Taking square roots gives:

$$
s_p = \sqrt{\text{MSE}} = \text{Root MSE}.
$$

```{r}
#| label: hw02-p6-check
#| include: false

stopifnot(abs(root_mse - pooled_sd) < 1e-10)
stopifnot(abs(root_mse - anova_root_mse) < 1e-10)
stopifnot(abs(root_mse - 94.6335) < 1e-4)
```

::: {.callout-warning}
## Combined SD Is Not the Pooled SD
The combined standard deviation is $95.586$, not $94.633$. The combined SD is computed after ignoring group membership. The pooled SD is computed from within-group variation and is the same quantity as Root MSE in this dummy-variable regression.
:::


## Summary

This homework connects regression with a dummy predictor to classical two-group comparisons:

1. **Regression coefficients equal group means and mean differences.** The intercept is the male mean, and the slope is the female-minus-male mean difference.

2. **The regression $t$-test equals the pooled two-sample $t$-test.** The signs differ only because regression and the t-test subtract the groups in opposite directions.

3. **Root MSE equals the pooled within-group SD.** In this model, residuals are deviations from group means, so the regression MSE is the pooled within-group variance.

::: {.callout-note}
## Looking Ahead
In Homework 4, this same logic expands from two groups to three groups. Regression with two dummy variables becomes equivalent to one-way ANOVA.
:::
