---
title: "Homework 3: Multiple Regression and Confounding"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Interpret a regression model with one continuous predictor and one dummy predictor
- Distinguish unadjusted group differences from adjusted group differences
- Explain why parallel-line models compare groups at the same value of a covariate
- Use "controlling for $x$" language accurately
- Reason about confounding and Simpson-style reversals
- **Related notes:** [Chapter 2: ANCOVA](../../notes/chapters/02-ancova.html)
- **Related lab:** [Lab 3: ANCOVA and Statistical Control](../../labs/labs/lab03-anova-ancova.html)
:::

This homework is equation-based. We do not analyze a dataset here. The short R chunks below define the equations once, compute the predicted values, and draw the same lines that would be sketched by hand.

*Adapted from @agresti2018, Chapter 13.*

## Setup

```{r}
#| label: hw03-setup
#| include: false

groups <- tibble(
  group = factor(c("Non-white", "White"), levels = c("Non-white", "White")),
  z = c(0, 1)
)

x_grid <- seq(4, 16, by = 0.25)

pred_grid <- tidyr::expand_grid(
  father_education = x_grid,
  groups
) |>
  mutate(
    unadjusted = 11 + 2 * z,
    adjusted = 3 + 0.8 * father_education - 0.6 * z
  )

contrast_table <- tibble(
  model = c("Regression 1: race only", "Regression 2: controlling for father's education"),
  white = c(11 + 2 * 1, 3 + 0.8 * 11 - 0.6 * 1),
  nonwhite = c(11 + 2 * 0, 3 + 0.8 * 11 - 0.6 * 0),
  white_minus_nonwhite = white - nonwhite
)
```


## Problem 1

### Problem Statement

> Consider two regression models for years of education ($y$) with father's education ($x$) and a white/non-white dummy ($z$, where $z = 1$ for white and $z = 0$ for non-white):
>
> - Regression 1: $\hat{y} = 11 + 2z$
> - Regression 2: $\hat{y} = 3 + 0.8x - 0.6z$
>
> (a) From the first regression, find the estimated mean education for whites and non-whites. What is the difference?
> (b) Plot the relationship between $\hat{y}$ and $x$ for the first regression.
> (c) From the second regression, find the estimated difference in mean education between whites and non-whites after controlling for father's education.
> (d) Using the second regression, find the predicted means when father's education = 11 years.

### Solution

#### Part (a): Regression 1 Means

Regression 1 is:

$$
\hat{y} = 11 + 2z.
$$

Substitute the two dummy-variable values:

```{r}
#| label: hw03-p1a-unadjusted-means

regression1_means <- tibble(
  group = factor(c("Non-white", "White"), levels = c("Non-white", "White")),
  z = c(0, 1),
  predicted_education = c(11, 13)
)

regression1_means |>
  select(group, z, predicted_education) |>
  knitr::kable(
    col.names = c("Group", "z", "Predicted education"),
    align = c("l", "r", "r"),
    caption = "Predicted group means from Regression 1."
  )
```

```{r}
#| label: hw03-p1a-check
#| include: false

stopifnot(
  isTRUE(all.equal(regression1_means$predicted_education, c(11, 13))),
  isTRUE(all.equal(
    contrast_table$white_minus_nonwhite[contrast_table$model == "Regression 1: race only"],
    2
  ))
)
```

For whites, the estimated mean is $11 + 2(1) = 13$. For non-whites, the estimated mean is $11 + 2(0) = 11$. The unadjusted difference is:

$$
13 - 11 = 2.
$$

Without controlling for father's education, whites are estimated to have 2 more years of education than non-whites.

#### Part (b): Plot Regression 1

Regression 1 does not include $x$, so the predicted value does not change as father's education changes. The two lines are horizontal.

#### Parts (c) and (d): Regression 2 Adjusted Difference

Regression 2 is:

$$
\hat{y} = 3 + 0.8x - 0.6z.
$$

For non-whites ($z = 0$):

$$
\hat{y}_{z=0} = 3 + 0.8x.
$$

For whites ($z = 1$):

$$
\hat{y}_{z=1} = 3 + 0.8x - 0.6 = 2.4 + 0.8x.
$$

At any fixed value of father's education, the white minus non-white adjusted difference is:

$$
(2.4 + 0.8x) - (3 + 0.8x) = -0.6.
$$

After controlling for father's education, whites are estimated to have 0.6 fewer years of education than non-whites. The sign matters: the adjusted difference is negative even though the unadjusted difference was positive.

At father's education $x = 11$:

```{r}
#| label: hw03-p1d-adjusted-predictions

predictions_at_11 <- tibble(
  group = factor(c("Non-white", "White"), levels = c("Non-white", "White")),
  z = c(0, 1),
  father_education = 11,
  predicted_education = c(11.8, 11.2)
)

predictions_at_11 |>
  select(group, z, father_education, predicted_education) |>
  knitr::kable(
    col.names = c("Group", "z", "Father's education", "Predicted education"),
    align = c("l", "r", "r", "r"),
    caption = "Predicted education from Regression 2 when father's education is 11."
  )
```

```{r}
#| label: hw03-p1-adjusted-checks
#| include: false

gap_by_x <- pred_grid |>
  select(father_education, group, adjusted) |>
  tidyr::pivot_wider(names_from = group, values_from = adjusted) |>
  mutate(gap_white_minus_nonwhite = White - `Non-white`)

stopifnot(
  isTRUE(all.equal(predictions_at_11$predicted_education, c(11.8, 11.2))),
  isTRUE(all.equal(
    contrast_table$white_minus_nonwhite[
      contrast_table$model == "Regression 2: controlling for father's education"
    ],
    -0.6
  )),
  all(abs(gap_by_x$gap_white_minus_nonwhite + 0.6) < 1e-12)
)
```

For whites:

$$
\hat{y} = 3 + 0.8(11) - 0.6(1) = 11.2.
$$

For non-whites:

$$
\hat{y} = 3 + 0.8(11) - 0.6(0) = 11.8.
$$

The table below summarizes the main contrast.

```{r}
#| label: hw03-p1-contrast-table

contrast_table |>
  knitr::kable(
    col.names = c("Model", "White", "Non-white", "White minus non-white"),
    align = c("l", "r", "r", "r"),
    digits = 1,
    caption = "Unadjusted and adjusted white-minus-non-white differences."
  )
```

```{r}
#| label: fig-hw03-unadjusted-adjusted
#| fig-cap: "Unadjusted and adjusted predicted education by white/non-white group. Regression 1 gives horizontal race-only means; Regression 2 gives parallel lines controlling for father's education."
#| fig-alt: "Two-panel plot of predicted education against father's education. The first panel shows two horizontal lines: white at thirteen and non-white at eleven, so the raw white minus non-white gap is positive two. The second panel shows two parallel upward-sloping lines with slope zero point eight. At father's education eleven, non-white is eleven point eight and white is eleven point two, so the adjusted white minus non-white gap is negative zero point six."
#| fig-width: 9
#| fig-height: 5.5
#| echo: false

plot_grid <- pred_grid |>
  tidyr::pivot_longer(
    cols = c(unadjusted, adjusted),
    names_to = "model",
    values_to = "predicted_education"
  ) |>
  mutate(
    model = factor(
      model,
      levels = c("unadjusted", "adjusted"),
      labels = c("Regression 1: race only", "Regression 2: controlling for father's education")
    )
  )

gap_segments <- tibble(
  model = factor(
    c("Regression 1: race only", "Regression 2: controlling for father's education"),
    levels = levels(plot_grid$model)
  ),
  father_education = c(11, 11),
  y_start = c(11, 11.2),
  y_end = c(13, 11.8),
  label = c("white - non-white = +2", "white - non-white = -0.6")
) |>
  mutate(
    y_mid = (y_start + y_end) / 2,
    label_x = father_education + 0.6
  )

points_at_11 <- plot_grid |>
  filter(father_education == 11)

ggplot(plot_grid, aes(father_education, predicted_education, color = group, linetype = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(data = points_at_11, size = 2.5, show.legend = FALSE) +
  geom_segment(
    data = gap_segments,
    aes(x = father_education, xend = father_education, y = y_start, yend = y_end),
    inherit.aes = FALSE,
    linewidth = 0.8,
    linetype = "dotted",
    color = "grey20"
  ) +
  geom_label(
    data = gap_segments,
    aes(x = label_x, y = y_mid, label = label),
    inherit.aes = FALSE,
    size = 3.1,
    label.size = 0.25
  ) +
  facet_wrap(~ model) +
  scale_color_manual(values = c(
    "Non-white" = unname(ber640_colors["primary"]),
    "White" = unname(ber640_colors["secondary"])
  )) +
  scale_linetype_manual(values = c(
    "Non-white" = "solid",
    "White" = "dashed"
  )) +
  scale_x_continuous(breaks = seq(4, 16, by = 4)) +
  scale_y_continuous(breaks = seq(5, 16, by = 1)) +
  coord_cartesian(xlim = c(4, 16), ylim = c(5, 16.3)) +
  labs(
    x = "Father's education (years)",
    y = "Predicted years of education",
    color = "Group",
    linetype = "Group",
    title = "Unadjusted vs. Father's-Education-Adjusted Differences"
  )
```

::: {.callout-tip}
## Key Takeaway
The unadjusted model compares group means directly: white minus non-white is $+2$. The adjusted model compares the two groups at the same value of father's education: white minus non-white is $-0.6$. This is the vertical gap between the parallel lines.
:::


## Problem 2

### Problem Statement

> True or false for the model $y = \beta_0 + \beta_1 x + \beta_2 z + \varepsilon$, where $x$ is a continuous predictor and $z$ is a qualitative dummy predictor:
>
> (a) The qualitative predictor has two categories.
> (b) One line has slope $\beta_1$ and the other has slope $\beta_2$.
> (c) $\beta_2$ is the difference between the mean of $y$ for the second and first categories of the qualitative variable.
> (d) $\beta_2$ is the difference between the adjusted mean of $y$ (controlling for $x$) for the second and first categories of the qualitative variable.

### Solution

```{r}
#| label: hw03-p2-key-check
#| include: false

tf_key_p2 <- c(a = TRUE, b = FALSE, c = FALSE, d = TRUE)
stopifnot(identical(tf_key_p2, c(a = TRUE, b = FALSE, c = FALSE, d = TRUE)))
```

**(a) True.** A single dummy variable $z$ has two values, usually 0 and 1. Those two values represent two categories.

**(b) False.** Both lines have slope $\beta_1$. When $z = 0$:

$$
E(y \mid x, z = 0) = \beta_0 + \beta_1 x.
$$

When $z = 1$:

$$
E(y \mid x, z = 1) = (\beta_0 + \beta_2) + \beta_1 x.
$$

The two lines have different intercepts, but the same slope.

**(c) False.** This is almost right, but it is missing the phrase "controlling for $x$." In a multiple regression, $\beta_2$ is not the raw group mean difference. It is the difference between the two predicted means at the same value of $x$.

**(d) True.** At any fixed value of $x$:

$$
\begin{aligned}
E(y \mid x, z = 1) - E(y \mid x, z = 0)
&= [(\beta_0 + \beta_2) + \beta_1 x] - [\beta_0 + \beta_1 x] \\
&= \beta_2.
\end{aligned}
$$

So $\beta_2$ is the adjusted mean difference for the second category minus the first category, controlling for $x$.

::: {.callout-warning}
## Common Mistake
In a dummy-only regression, the dummy coefficient is a raw group mean difference. In a multiple regression with $x$ included, the dummy coefficient is an adjusted group difference. The words "controlling for $x$" are not decoration; they change the estimand.
:::


## Problem 3

### Problem Statement

> In the United States, mean annual income for Black people ($\mu_1$) is lower than for white people ($\mu_2$). Education level is lower on average for Black people, and income is positively related to education.
>
> The difference in mean income between white and Black people, controlling for education, is:
>
> (a) Less than $\mu_2 - \mu_1$
> (b) Greater than $\mu_2 - \mu_1$
> (c) Possibly equal to $\mu_2 - \mu_1$

### Solution

```{r}
#| label: hw03-p3-key-check
#| include: false

tf_key_p3 <- c(a = TRUE, b = FALSE, c = FALSE)
stopifnot(identical(tf_key_p3, c(a = TRUE, b = FALSE, c = FALSE)))
```

**(a) True.** The education-adjusted difference is less than $\mu_2 - \mu_1$.

The raw gap $\mu_2 - \mu_1$ compares white and Black people without holding education constant. But the problem states that education is lower on average for Black people and that income increases with education. Therefore, part of the raw income gap reflects the education gap.

When we compare white and Black people at the same education level, we remove that education-related part of the raw difference. Under the intended confounding logic of the problem, the education-adjusted white-Black income difference is smaller than the raw difference.

**(b) False.** Controlling for education would not make the gap greater in this setup. Because education is positively related to income and is higher on average for the group with higher raw income, adjustment removes a positive contribution to the raw gap.

**(c) False.** The problem states that education is associated with both group membership and income. If a variable is related to both the predictor and the outcome, controlling for it changes the group coefficient, even if the change is small. That is exactly why it is a confounder here.

::: {.callout-important}
## Confounding Logic
Education is a confounder in this problem because it is associated with both:

- the group comparison, since average education differs by group;
- the outcome, since income is positively related to education.

The adjusted comparison is a model-adjusted within-education comparison. It is not automatically causal, but it answers a clearer regression question: among people with the same education level, how large is the remaining income difference?
:::


## Summary

This homework connects dummy-variable regression to the ANCOVA idea:

1. A dummy-only model gives an unadjusted group mean difference.
2. A model with a dummy predictor and a continuous covariate gives an adjusted group difference.
3. In a no-interaction ANCOVA model, the group lines are parallel and the adjusted difference is the constant vertical gap between them.
4. Confounding can shrink or reverse a raw group comparison after adjustment.

In HW4, we move from one dummy variable to several group indicators. The same coefficient-interpretation discipline carries over: always identify the reference group, the comparison group, and whether the coefficient is adjusted for other variables.
