---
title: "Homework 4: Multiple Dummies and One-Way ANOVA"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Use multiple dummy variables to represent a categorical predictor with three groups
- Interpret dummy-regression coefficients relative to a reference group
- Explain why regression with country dummies and one-way ANOVA give the same omnibus $F$-test
- Use Sidak-adjusted pairwise comparisons to identify which country means differ
- **Related notes:** [Chapter 3: *One-Way ANOVA*](../../notes/chapters/03-one-way-anova.html)
- **Related lab:** [Lab 4: *One-Way ANOVA and Categorical Predictors*](../../labs/labs/lab04-categorical-contrasts.html)
:::

::: {.callout-tip}
## Stata-to-R Translation
| Stata command | R equivalent | Used in |
|:---|:---|:---|
| `regress wleread uk germany` | `lm(wleread ~ uk + germany, data = pisa_3)` | Problem 1 |
| `anova wleread country` | `lm(wleread ~ country_f, data = pisa_3)` followed by `anova()` | Problem 2(a) |
| `oneway wleread country, sidak` | `emmeans()` followed by `pairs(adjust = "sidak")` | Problem 2(c) |
:::

**Dataset:** `pisa2000` from the `regdatasets` package.

We use students from the USA, UK, and Germany ($N = 4{,}528$). In Problem 1, USA is the reference group.


## Setup

```{r}
#| label: hw04-setup

data(pisa2000)

pisa_3 <- pisa2000 |>
  filter(usa == 1 | uk == 1 | germany == 1) |>
  mutate(
    country_f = factor(
      case_when(
        usa == 1 ~ "USA",
        uk == 1 ~ "UK",
        germany == 1 ~ "Germany"
      ),
      levels = c("USA", "UK", "Germany")
    )
  )

pisa_3 |>
  count(country_f, name = "n") |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r}
#| label: hw04-setup-check
#| include: false

stopifnot(nrow(pisa_3) == 4528)
stopifnot(identical(levels(pisa_3$country_f), c("USA", "UK", "Germany")))
stopifnot(all(rowSums(pisa_3[c("usa", "uk", "germany")] == 1) == 1L))
```


## Problem 1

### Problem Statement

> Choose one country as the reference group and regress `wleread` on dummy variables for the other two countries. Interpret the regression coefficients.

### Solution

```{r}
#| label: hw04-p1-fit

model1 <- lm(wleread ~ uk + germany, data = pisa_3)

model_params(model1) |>
  as_tibble() |>
  mutate(p = format_p(p)) |>
  select(Parameter, Coefficient, SE, CI_low, CI_high, t, df_error, p) |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw04-p1-fit-summary

hw04_fit <- model_glance(model1)

hw04_fit |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw04-p1-group-means

group_means <- pisa_3 |>
  summarise(
    .by = country_f,
    n = n(),
    mean_read = mean(wleread),
    sd_read = sd(wleread)
  ) |>
  arrange(country_f)

group_means |>
  round_numeric(digits = 4)
```

The fitted regression equation is:

$$
\widehat{\text{wleread}}
= 523.15 + 15.04 \times \text{uk} - 5.40 \times \text{germany}.
$$

Because USA is the reference group:

- **Intercept** ($\hat{\beta}_0 = 523.15$, $p < .001$): the estimated mean reading score for US students.
- **`uk`** ($\hat{\beta}_1 = 15.04$, $p < .001$): UK students scored about 15.0 points higher than US students, on average.
- **`germany`** ($\hat{\beta}_2 = -5.40$, $p = .150$): German students scored about 5.4 points lower than US students, on average. This difference is not statistically significant.

::: {.callout-note}
## Calculation Note
If you see a source key marking the Germany-vs-USA coefficient as $p < .001$, treat that as a typo. The coefficient table has $t \approx -1.44$, which corresponds to the unadjusted regression p-value $p = .150$. In the Sidak-adjusted pairwise table below, the USA-Germany comparison remains non-significant with $p = .386$.
:::

The group means make the dummy coding transparent:

$$
\bar{Y}_{\text{USA}} \approx 523.15,
\qquad
\bar{Y}_{\text{UK}} \approx 538.18,
\qquad
\bar{Y}_{\text{Germany}} \approx 517.75.
$$

Using the unrounded coefficients, the UK mean is the USA mean plus the UK coefficient, and the Germany mean is the USA mean plus the Germany coefficient.

```{r}
#| label: hw04-p1-check
#| include: false

stopifnot(abs(coef(model1)[["(Intercept)"]] - 523.1475) < 1e-4)
stopifnot(abs(coef(model1)[["uk"]] - 15.03654) < 1e-5)
stopifnot(abs(coef(model1)[["germany"]] + 5.395565) < 1e-6)
stopifnot(abs(summary(model1)$coefficients["germany", "Pr(>|t|)"] - 0.1501) < 1e-4)
stopifnot(abs(model_glance(model1)$F - 21.75714) < 1e-5)
stopifnot(abs(model_glance(model1)$r_squared - 0.009524819) < 1e-9)
stopifnot(abs(sigma(model1) - 94.36021) < 1e-5)
stopifnot(abs(group_means$mean_read[group_means$country_f == "USA"] - 523.1475) < 1e-4)
stopifnot(abs(group_means$mean_read[group_means$country_f == "UK"] - 538.1841) < 1e-4)
stopifnot(abs(group_means$mean_read[group_means$country_f == "Germany"] - 517.7520) < 1e-4)
```

### Code Explanation

- `lm(wleread ~ uk + germany, data = pisa_3)` fits a regression with two dummy variables.
- With three countries, we need $3 - 1 = 2$ dummy variables.
- USA is omitted, so USA is the reference group.
- `model_params(model1)` prints the coefficient table.
- `model_glance(model1)` gives model-level quantities such as the overall $F$-test, $R^2$, and Root MSE.

::: {.callout-tip}
## Key Takeaway
With $k$ groups, regression uses $k - 1$ dummy variables. The intercept is the reference-group mean, and each dummy coefficient is the difference between that group's mean and the reference-group mean.
:::


## Problem 2

### Problem Statement

> (a) Using ANOVA, is the effect of country significant at the 5% level?
>
> (b) Could you answer (a) from the regression output in Problem 1?
>
> (c) Perform pairwise comparisons of country means using Sidak's method. Which countries differ significantly?


### Part (a): ANOVA F-Test

```{r}
#| label: hw04-p2a-anova

anova_model <- lm(wleread ~ country_f, data = pisa_3)

anova_table <- anova(anova_model)

anova_table |>
  as.data.frame() |>
  tibble::rownames_to_column("Source") |>
  mutate(p = format_p(`Pr(>F)`)) |>
  select(Source, Df, `Sum Sq`, `Mean Sq`, F = `F value`, p) |>
  round_numeric(digits = 4)
```

Yes. The ANOVA table gives:

$$
F(2, 4525) = 21.76,
\qquad
p < .001.
$$

At the 5% significance level, we reject:

$$
H_0: \mu_{\text{USA}} = \mu_{\text{UK}} = \mu_{\text{Germany}}.
$$

Country is statistically associated with mean reading score. The effect is statistically significant, but small: $R^2 = 0.0095$, so country explains about 1% of the variation in reading scores.

Stata's `oneway` output also reports Bartlett's test for equal variances. Here it is not significant, $\chi^2(2) = 0.45$, $p = .799$, so it does not flag unequal country variances. This is an assumption diagnostic, not the omnibus test of equal country means.

```{r}
#| label: hw04-p2a-effect-size

anova_effect <- tibble(
  ss_country = anova_table["country_f", "Sum Sq"],
  ss_residual = anova_table["Residuals", "Sum Sq"],
  ss_total = ss_country + ss_residual,
  r_squared = ss_country / ss_total,
  root_mse = sqrt(anova_table["Residuals", "Mean Sq"])
)

anova_effect |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw04-p2a-bartlett-check
#| include: false

bartlett_hw04 <- bartlett.test(wleread ~ country_f, data = pisa_3)

stopifnot(abs(unname(bartlett_hw04$statistic) - 0.44791) < 1e-5)
stopifnot(abs(bartlett_hw04$p.value - 0.7994) < 1e-4)
```

::: {.callout-note}
## What the Omnibus F-Test Says
The overall ANOVA $F$-test tells us whether at least one country mean differs from the others. It does not tell us which countries differ. That is why we need pairwise comparisons in Part (c).
:::


### Part (b): Regression and ANOVA Give the Same F-Test

```{r}
#| label: hw04-p2b-compare

regression_f <- model_glance(model1) |>
  transmute(
    method = "Regression with dummies",
    df_model,
    df_error,
    F,
    p,
    root_mse
  )

anova_f <- tibble(
  method = "One-way ANOVA",
  df_model = anova_table["country_f", "Df"],
  df_error = anova_table["Residuals", "Df"],
  F = anova_table["country_f", "F value"],
  p = format_p(anova_table["country_f", "Pr(>F)"]),
  root_mse = sqrt(anova_table["Residuals", "Mean Sq"])
)

bind_rows(regression_f, anova_f) |>
  round_numeric(digits = 4)
```

Yes. The regression output in Problem 1 already contains the same omnibus test:

$$
H_0: \beta_{\text{UK}} = \beta_{\text{Germany}} = 0.
$$

Because USA is the reference group, this null hypothesis says that the UK mean and Germany mean are both equal to the USA mean. That is the same as saying all three country means are equal.

Regression with two country dummies and one-way ANOVA with a three-level country factor are the same linear model written two ways. They have the same overall $F$-test, $p$-value, residual sum of squares, and Root MSE.

```{r}
#| label: hw04-p2b-check
#| include: false

stopifnot(abs(anova_table["country_f", "F value"] - 21.75714) < 1e-5)
stopifnot(abs(anova_table["country_f", "Sum Sq"] - 387444.5) < 0.1)
stopifnot(abs(anova_table["Residuals", "Mean Sq"] - sigma(model1)^2) < 1e-8)
stopifnot(abs(anova_table["country_f", "F value"] - model_glance(model1)$F) < 1e-10)
```

::: {.callout-important}
## ANOVA Is Regression
One-way ANOVA is regression with a categorical predictor. In R, `lm(wleread ~ country_f, data = pisa_3)` lets R create the dummies from the factor. In Problem 1, `lm(wleread ~ uk + germany, data = pisa_3)` used the same dummy coding manually.
:::


### Part (c): Sidak-Adjusted Pairwise Comparisons

```{r}
#| label: hw04-p2c-pairwise

country_means <- emmeans::emmeans(anova_model, "country_f")
country_means |>
  as.data.frame() |>
  select(country_f, emmean, SE, df, lower.CL, upper.CL) |>
  round_numeric(digits = 4)

pairwise_sidak <- pairs(country_means, adjust = "sidak")
```

```{r}
#| label: tbl-hw04-p2c-pairwise
#| tbl-cap: "Sidak-adjusted pairwise comparisons of country reading means"

pairwise_table <- as.data.frame(pairwise_sidak) |>
  transmute(
    Comparison = contrast,
    Difference = estimate,
    SE = SE,
    df = df,
    t = t.ratio,
    p = format_p(p.value),
    Significant = if_else(p.value < .05, "Yes", "No")
  )

pairwise_table |>
  gt::gt() |>
  gt::cols_label(
    Comparison = "Comparison",
    Difference = "Mean diff.",
    SE = "Std. error",
    df = "df",
    t = "t ratio",
    p = "Sidak p",
    Significant = "Significant"
  ) |>
  gt::fmt_number(columns = c(Difference, SE, t), decimals = 4) |>
  gt::fmt_number(columns = df, decimals = 0, use_seps = FALSE) |>
  gt::cols_align("left", columns = Comparison) |>
  gt::cols_align("right", columns = c(Difference, SE, df, t, p)) |>
  gt::cols_align("center", columns = Significant) |>
  gt::cols_width(
    Comparison ~ gt::pct(28),
    Difference ~ gt::pct(14),
    SE ~ gt::pct(14),
    df ~ gt::pct(8),
    t ~ gt::pct(11),
    p ~ gt::pct(13),
    Significant ~ gt::pct(8)
  ) |>
  gt::tab_options(table.width = gt::pct(100))
```

The Sidak-adjusted comparisons show:

1. **USA vs. UK:** USA is about 15.0 points lower than the UK. This difference is statistically significant ($p < .001$).

2. **USA vs. Germany:** USA is about 5.4 points higher than Germany. This difference is not statistically significant ($p = .386$).

3. **UK vs. Germany:** UK is about 20.4 points higher than Germany. This difference is statistically significant ($p < .001$).

So the UK mean is significantly higher than both the USA and Germany means. The USA-Germany difference is not statistically significant after Sidak correction.

```{r}
#| label: hw04-p2c-check
#| include: false

pairwise_df <- as.data.frame(pairwise_sidak)

stopifnot(abs(pairwise_df$estimate[pairwise_df$contrast == "USA - UK"] + 15.03654) < 1e-5)
stopifnot(abs(pairwise_df$p.value[pairwise_df$contrast == "USA - UK"] - 0.0001377) < 1e-6)
stopifnot(abs(pairwise_df$estimate[pairwise_df$contrast == "USA - Germany"] - 5.395565) < 1e-6)
stopifnot(abs(pairwise_df$p.value[pairwise_df$contrast == "USA - Germany"] - 0.3860129) < 1e-7)
stopifnot(abs(pairwise_df$estimate[pairwise_df$contrast == "UK - Germany"] - 20.43211) < 1e-5)
stopifnot(abs(pairwise_df$SE[pairwise_df$contrast == "UK - Germany"] - 3.189637) < 1e-6)
stopifnot(abs(pairwise_df$t.ratio[pairwise_df$contrast == "UK - Germany"] - 6.405777) < 1e-6)
stopifnot(pairwise_df$p.value[pairwise_df$contrast == "UK - Germany"] < .001)
```

::: {.callout-warning}
## Common Mistake: Ignoring Multiple Comparisons
After a significant overall $F$-test, it is tempting to run several ordinary two-sample tests. That inflates the chance of at least one false positive. Sidak adjustment corrects the pairwise p-values for the fact that we are making three comparisons.
:::

### Optional Mean Plot

The table above is the main answer. The plot below shows the same estimated country means with confidence intervals.

```{r}
#| label: fig-hw04-country-means
#| fig-cap: "Estimated mean reading scores for USA, UK, and Germany with 95% confidence intervals."
#| fig-alt: "Dot-and-whisker plot with confidence intervals for mean reading scores. The UK mean is highest around 538, the USA mean is around 523, and the Germany mean is around 518."
#| fig-width: 8
#| fig-height: 5
#| echo: false

country_means_df <- as.data.frame(country_means)

ggplot(country_means_df, aes(x = country_f, y = emmean, color = country_f)) +
  geom_pointrange(
    aes(ymin = lower.CL, ymax = upper.CL),
    size = 1.1,
    linewidth = 0.8
  ) +
  geom_text(
    aes(label = round(emmean, 1)),
    vjust = -1.1,
    fontface = "bold"
  ) +
  scale_color_manual(values = unname(ber640_colors[c("accent", "primary", "warning")])) +
  labs(
    x = "Country",
    y = "Estimated mean reading score",
    color = "Country"
  ) +
  theme(legend.position = "none")
```


## Summary

This homework extends the dummy-variable logic from Homework 2 to a three-group predictor:

1. **Multiple dummy regression:** With three groups, we use two dummy variables and one reference group.

2. **Regression and ANOVA equivalence:** The regression overall $F$-test and the one-way ANOVA $F$-test are the same test written in two languages.

3. **Pairwise comparisons:** A significant omnibus $F$-test tells us that at least one mean differs, but Sidak-adjusted pairwise comparisons tell us which pairs differ.

::: {.callout-note}
## Looking Ahead
In **Homework 6**, we extend this framework to two-way ANOVA with interactions. Instead of asking whether one factor matters, we ask whether the effect of one factor depends on the level of another.
:::
