---
title: "Homework 5: Multiple Regression and Interactions"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Interpret coefficients in a multiple regression with many predictors
- Read an interaction term from a published regression table
- Distinguish main effects from conditional effects in interaction models
- Translate one interaction into group-specific slopes and a group-difference formula
- **Related notes:** [Chapter 4: *Continuous Predictors*](../../notes/chapters/04-continuous-predictors.html) and [Chapter 5: *Interactions*](../../notes/chapters/05-interactions.html)
- **Related labs:** [Lab 5: *Multiple Regression with Continuous Predictors*](../../labs/labs/lab05-multiple-regression.html) and [Lab 6: *Interactions in Regression*](../../labs/labs/lab06-interactions.html)
:::

This homework is entirely **conceptual**: there are no raw data to analyze, only a published regression table to interpret. The table comes from @langton1994, who studied salary determinants among tenured and tenure-track faculty at 308 U.S. colleges and universities in 1969 ($N = 35{,}491$).

As in the interaction lab, the key idea is simple: an interaction means **it depends**. Here, the salary-variation slope depends on sex, and the male-female salary gap depends on salary variation.


## The Regression Table

The table below reproduces Table 3 from @langton1994. Model 1 is a main-effects model. Model 2 adds salary variation within the professor's field and an interaction between sex and salary variation.

The raw data are not used here. The table is reproduced from the published coefficients, so the chapter interprets the table rather than refitting the model.

```{r}
#| label: tbl-hw05-regression
#| tbl-cap: "Unstandardized OLS Coefficients for Faculty Salary [@langton1994, Table 3]"
#| echo: false

reg_table_df <- data.frame(
  Variable = c(
    "Years of experience",
    "Terminal degree (Ph.D. = 1)",
    "Degree quality",
    "Research productivity",
    "Marital status (married = 1)",
    "Race (nonwhite = 1)",
    "Sex (female = 1)",
    "Salary variation",
    "Sex × salary variation",
    "Intercept",
    "R² (adjusted)"
  ),
  Model_1 = c(
    "162.68*** (2.05)",
    "264.050*** (40.09)",
    "616.77*** (34.48)",
    "616.52*** (7.68)",
    "850.26*** (51.19)",
    "-310.37** (106.96)",
    "-1863.70*** (62.07)",
    "—",
    "—",
    "10033.24***",
    ".40"
  ),
  Model_2 = c(
    "162.59*** (2.05)",
    "270.15*** (40.12)",
    "657.08*** (34.64)",
    "614.96*** (7.68)",
    "816.00*** (53.23)",
    "-342.13*** (106.83)",
    "-698.46 (566.87)",
    "-5571.50*** (585.62)",
    "-3932.71* (1872.05)",
    "11712.80***",
    ".40"
  )
)

knitr::kable(
  reg_table_df,
  col.names = c("Variable", "Model 1", "Model 2"),
  align = c("l", "r", "r"),
  caption = NULL
)
```

**Note:** \*$p < .05$, \*\*$p < .01$, \*\*\*$p < .001$ (one-tailed). Numbers in parentheses are standard errors. The nonwhite (race) coefficient reads \*\* in Model 1 ($t \approx 2.90$) but \*\*\* in Model 2 ($t \approx 3.20$) despite an almost unchanged estimate, because only the Model 2 $t$-ratio clears the one-tailed $p<.001$ cutoff ($z \approx 3.09$); under two-tailed tests ($z_{.001} \approx 3.29$) both cells would read \*\*.

```{r}
#| label: hw05-reference-check
#| include: false

# Race stars follow the source's one-tailed tests: Model 1 t ~ 2.90 -> ** (p<.01);
# Model 2 t ~ 3.20 clears the one-tailed p<.001 cutoff (z ~ 3.09) -> *** (two-tailed would be **).
# Do not "correct" the Model 2 *** to ** without also reverting the legend above to two-tailed.
stopifnot(reg_table_df$Model_1[reg_table_df$Variable == "Race (nonwhite = 1)"] == "-310.37** (106.96)")
stopifnot(reg_table_df$Model_2[reg_table_df$Variable == "Race (nonwhite = 1)"] == "-342.13*** (106.83)")
stopifnot(reg_table_df$Model_2[reg_table_df$Variable == "Sex × salary variation"] == "-3932.71* (1872.05)")
stopifnot(reg_table_df$Model_2[reg_table_df$Variable == "Salary variation"] == "-5571.50*** (585.62)")
```

::: {.callout-note}
## Variable Definitions
- **Degree quality:** Rating of institution from which degree was received (Carnegie Commission), scale 0--4.
- **Research productivity:** Scale based on number of articles/books published.
- **Salary variation:** Variability in salary within the professor's field controlling for the other variables; a measure of uncertainty and dispersion in wages paid in the labor market.
- **Sex:** $1 =$ female, $0 =$ male.
:::

```{r}
#| label: hw05-interaction-constants
#| include: false

model2_sex <- -698.46
model2_salary_variation <- -5571.50
model2_interaction <- -3932.71

men_sv_slope <- model2_salary_variation
women_sv_slope <- model2_salary_variation + model2_interaction
female_minus_male_gap_at_0 <- model2_sex
male_minus_female_gap_at_0 <- -model2_sex
male_minus_female_gap_slope <- -model2_interaction

stopifnot(
  any(reg_table_df$Variable == "Sex (female = 1)"),
  isTRUE(all.equal(women_sv_slope, -9504.21)),
  isTRUE(all.equal(male_minus_female_gap_at_0, 698.46)),
  isTRUE(all.equal(male_minus_female_gap_slope, 3932.71))
)
```


## Problem 1

### Problem Statement

> Interpret each of the regression coefficients in Model 1 (main effects only) from the @langton1994 table on faculty salaries.

### Direct Answer

Model 1 contains seven predictors: three continuous/scale predictors and four dummy variables. Terminal degree, marital status, race, and sex are dummy variables. Each coefficient is interpreted after controlling for all other variables in the model.

After controlling for each of the other variables, the estimated mean 9-month salary:

1. **Years of experience** ($\hat{\beta}_j = 162.68$, $p < .001$): increases by **\$162.68** for each additional year of experience.

2. **Terminal degree** ($\hat{\beta}_j = 264.050$, $p < .001$): is **\$264.05 higher** for Ph.D. holders than for those without a Ph.D.

3. **Degree quality** ($\hat{\beta}_j = 616.77$, $p < .001$): increases by **\$616.77** for each one-unit increase in degree quality on the 0--4 Carnegie scale.

4. **Research productivity** ($\hat{\beta}_j = 616.52$, $p < .001$): increases by **\$616.52** for each one-unit increase in research productivity.

5. **Marital status** ($\hat{\beta}_j = 850.26$, $p < .001$): is **\$850.26 higher** for married faculty than for non-married faculty.

6. **Race** ($\hat{\beta}_j = -310.37$, $p < .01$): is **\$310.37 lower** for nonwhite faculty than for white faculty.

7. **Sex** ($\hat{\beta}_j = -1863.70$, $p < .001$): is **\$1,863.70 lower** for female faculty than for male faculty.

::: {.callout-tip}
## Interpretation Template
- **Continuous predictor:** "After controlling for the other variables, each one-unit increase in $X$ is associated with a \$___ change in estimated mean salary."
- **Dummy predictor:** "After controlling for the other variables, the estimated mean salary for the group coded 1 is \$___ higher/lower than for the reference group coded 0."
:::

::: {.callout-important}
## Reference Groups Matter
The race and sex variables are dummy variables. Because `Race = 1` means nonwhite and `Sex = 1` means female, negative coefficients mean the coded group has a lower estimated mean salary than the reference group.
:::


## Problem 2

### Problem Statement

> In Model 2, interpret the coefficients of Sex, Salary variation, and Sex $\times$ Salary variation.

### Direct Answer

Model 2 includes an interaction between Sex and Salary variation. The three key coefficients are:

- **Sex:** $\hat{\beta}_1 = -698.46$
- **Salary variation:** $\hat{\beta}_2 = -5571.50$
- **Sex $\times$ salary variation:** $\hat{\beta}_3 = -3932.71$

```{r}
#| label: hw05-p2-coefficient-interpretation-table
#| echo: true
#| code-fold: true

model2_interpretation <- tibble(
  term = c("Sex (female = 1)", "Salary variation", "Sex × salary variation"),
  coefficient = c(model2_sex, model2_salary_variation, model2_interaction),
  reference_condition = c(
    "Only when salary variation = 0",
    "Only for men, because Female = 0",
    "Difference between the women's and men's salary-variation slopes"
  ),
  correct_reading = c(
    "At salary variation 0, women are estimated 698.46 dollars lower than men.",
    "For men, a 1-unit increase in salary variation is associated with 5571.50 dollars lower salary.",
    "The women's salary-variation slope is 3932.71 dollars more negative than the men's slope."
  ),
  common_misread = c(
    "Treating this as the overall gender gap.",
    "Treating this as the salary-variation slope for everyone.",
    "Forgetting that it changes both the women's slope and the gender gap."
  )
)

model2_interpretation |>
  gt::gt() |>
  gt::cols_label(
    term = "Term",
    coefficient = "Coefficient",
    reference_condition = "Reference condition",
    correct_reading = "Correct reading",
    common_misread = "Common misread"
  ) |>
  gt::fmt_number(columns = coefficient, decimals = 2) |>
  gt::cols_align("left", columns = c(term, reference_condition, correct_reading, common_misread)) |>
  gt::cols_align("right", columns = coefficient) |>
  gt::cols_width(
    term ~ gt::pct(16),
    coefficient ~ gt::pct(12),
    reference_condition ~ gt::pct(22),
    correct_reading ~ gt::pct(30),
    common_misread ~ gt::pct(20)
  ) |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::tab_caption("How to read the three Model 2 interaction coefficients.")
```

Holding the other predictors fixed:

- **Sex:** When salary variation equals 0, women's estimated mean 9-month salary is **\$698.46 lower** than men's. This is a conditional baseline difference, not the overall gender gap. The answer key notes that this value is probably not substantively meaningful if salary variation is never actually 0.
- **Salary variation:** For men, the reference group where Female $= 0$, a one-unit increase in salary variation is associated with a **\$5,571.50 lower** estimated mean salary.
- **Sex $\times$ salary variation:** The salary-variation slope for women is **\$3,932.71 more negative** than the salary-variation slope for men.

Therefore, women's conditional salary-variation slope is:

$$
-5571.50 - 3932.71 = -9504.21
$$

So for women, a one-unit increase in salary variation is associated with about a **\$9,504 lower** estimated mean salary, holding the other predictors fixed.

```{r}
#| label: hw05-p2-derived-quantities
#| echo: true
#| code-fold: true

interaction_quantities <- tibble(
  quantity = c(
    "Men's salary-variation slope",
    "Women's salary-variation slope",
    "Female-minus-male gap",
    "Male-minus-female gap"
  ),
  expression = c(
    "-5571.50",
    "-5571.50 - 3932.71",
    "-698.46 - 3932.71 * SV",
    "698.46 + 3932.71 * SV"
  ),
  interpretation = c(
    "Men's estimated salary decreases by 5571.50 dollars for a 1-unit increase in salary variation.",
    "Women's estimated salary decreases by 9504.21 dollars for a 1-unit increase in salary variation.",
    "Women are estimated lower than men, and the female-minus-male gap becomes more negative as salary variation increases.",
    "Men are estimated higher than women, and the male-minus-female gap increases as salary variation increases."
  )
)

interaction_quantities |>
  knitr::kable(
    col.names = c("Derived quantity", "Expression", "Interpretation"),
    align = c("l", "l", "l"),
    caption = "Derived slopes and gender-gap formulas implied by Model 2."
  )
```

### Algebra

Focus on the part of the regression equation involving Sex, Salary variation, and their interaction:

$$
\hat{y} =
\hat{\beta}_0
+ \hat{\beta}_1 \text{Female}
+ \hat{\beta}_2 \text{SV}
+ \hat{\beta}_3 (\text{Female} \times \text{SV})
+ \cdots
$$

Substituting the published Model 2 coefficients gives:

$$
\hat{y} =
\hat{\beta}_0
- 698.46 \text{Female}
- 5571.50 \text{SV}
- 3932.71(\text{Female} \times \text{SV})
+ \cdots
$$

For **men** (Female $= 0$):

$$
\hat{y}_{\text{men}} =
\hat{\beta}_0
- 5571.50 \text{SV}
+ \cdots
$$

For **women** (Female $= 1$):

$$
\hat{y}_{\text{women}} =
(\hat{\beta}_0 - 698.46)
+ (-5571.50 - 3932.71)\text{SV}
+ \cdots
$$

$$
\hat{y}_{\text{women}} =
(\hat{\beta}_0 - 698.46)
- 9504.21\text{SV}
+ \cdots
$$

### Two Equivalent Readings

You can read the interaction in two equivalent ways.

**Reading 1: The salary-variation slope depends on sex.**

- For men, the salary-variation slope is $-5571.50$.
- For women, the salary-variation slope is $-9504.21$.
- The interaction term $-3932.71$ is the difference between those two slopes.

**Reading 2: The gender gap depends on salary variation.**

Because Female $= 1$, the sex coefficient is the female-minus-male difference. If we report the male-minus-female gap, we reverse the sign:

$$
\hat{y}_{\text{men}} - \hat{y}_{\text{women}}
= 698.46 + 3932.71 \times \text{SV}
$$

At salary variation $\text{SV}$, the estimated male-female gap is $698.46 + 3932.71 \times \text{SV}$. As salary variation increases, this estimated gap widens.

::: {.callout-warning}
## Common Mistake: Treating Main Effects as Average Effects
When an interaction is in the model, the main effects are conditional. In Model 2, the Sex coefficient is the female-male difference only when salary variation equals 0. The Salary variation coefficient is the slope only for men, the reference group where Female $= 0$. Do not interpret either coefficient as an unconditional average effect.
:::

::: {.callout-warning}
## Common Mistake: Reversing the Gender Gap
Because Female $= 1$, the sex coefficient is female minus male. A negative coefficient means women are estimated to earn less than men at that salary-variation value. If you report the male-minus-female gap, reverse the sign: $698.46 + 3932.71 \times \text{SV}$.
:::

### Optional Illustration

The figure below is an illustration based only on the published Model 2 coefficients. Its purpose is to show the non-parallel lines implied by the interaction; the raw data are not being plotted or refit. Because the published table does not give the observed range of salary variation or covariate means, the vertical placement and x-range are schematic. The slope difference and widening gap are the important features.

```{r}
#| label: fig-hw05-interaction-lines
#| fig-cap: "Schematic illustration of the Model 2 interaction component. The salary-variation axis is drawn on a 0-to-1 schematic scale, and the vertical axis shows only the sex and salary-variation component, not full predicted salary."
#| fig-alt: "Two downward-sloping schematic lines show the model component from sex and salary variation terms. Men start at zero on the component scale and fall with salary variation. Women start lower and fall more steeply, so the vertical gap widens as salary variation increases."
#| fig-width: 8
#| fig-height: 5
#| echo: false

salary_variation <- seq(0, 1, length.out = 100)

plot_data <- data.frame(
  salary_variation = rep(salary_variation, 2),
  salary_component = c(
    men_sv_slope * salary_variation,
    model2_sex + women_sv_slope * salary_variation
  ),
  sex = rep(c("Men", "Women"), each = length(salary_variation))
)

ggplot(plot_data, aes(x = salary_variation, y = salary_component, color = sex, linetype = sex)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c(
    "Men" = unname(ber640_colors["primary"]),
    "Women" = unname(ber640_colors["secondary"])
  )) +
  scale_linetype_manual(values = c("Men" = "solid", "Women" = "dashed")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Salary variation (schematic 0-1 scale)",
    y = "Model component",
    color = "Sex",
    linetype = "Sex"
  )
```

The two lines are not parallel. That is the visual signature of this continuous $\times$ categorical interaction: the salary-variation slope differs by sex, and the male-female gap changes as salary variation changes.


## Problem 3

### Problem Statement

> By looking at the estimates, can you answer the question of whether the male-female wage differential is exacerbated by more uncertainty and dispersion in wages?

### Direct Answer

**Yes, based on these estimates.** Greater salary variation is associated with a larger male-female wage differential.

The interaction coefficient is negative:

$$
\hat{\beta}_{\text{Female} \times \text{SV}} = -3932.71
$$

This means that the salary-variation slope is more negative for women than for men. Equivalently, the estimated male-female gap increases by **\$3,932.71** for each one-unit increase in salary variation:

$$
\text{Male-female gap}
= 698.46 + 3932.71 \times \text{SV}
$$

So, as salary variation increases, the estimated male-female gap widens.

::: {.callout-important}
## What Interaction Terms Tell Us
An interaction term tells us that the association or slope for one variable **depends on** the level of another variable. Here, both statements describe the same statistical pattern:

- The salary-variation slope differs for men and women.
- The male-female salary gap differs across fields with different salary variation.

The published estimates support the conclusion that greater wage uncertainty and dispersion are associated with a larger male-female salary gap. This is an interpretation of the regression table, not a claim that the homework refit the model from raw data.
:::

::: {.callout-caution}
## Common Mistake: Reading the Sex Main Effect Alone
Do not conclude that there is no gender gap just because the Model 2 Sex coefficient is not statistically significant. With an interaction in the model, the Sex coefficient refers only to salary variation $= 0$. The gender gap depends on salary variation, so the interaction term and the gap formula are the more relevant pieces of Model 2.
:::


## Summary

This homework practiced three key skills for interpreting multiple regression with interaction terms:

1. **Coefficient interpretation in main-effects models.** Each coefficient in Model 1 represents the estimated change in salary associated with a one-unit change in the predictor, controlling for all other variables.

2. **Interaction interpretation.** When a model includes $X_1$, $X_2$, and $X_1 \times X_2$, the coefficient of $X_1$ is its conditional slope or difference when $X_2 = 0$, and the interaction coefficient tells us how that slope or difference changes as $X_2$ changes.

3. **Conditional effects and non-parallel lines.** The interaction between sex and salary variation produces different slopes for men and women. The male-female salary gap is not a fixed number in Model 2; it depends on salary variation.

::: {.callout-note}
## Looking Ahead
In **Homework 6** (Two-Way ANOVA), we extend interaction concepts to models with two categorical predictors. Instead of diverging slopes for a continuous $\times$ dummy interaction, you will see crossing or non-parallel cell means. The core logic is the same: the difference associated with one variable depends on the level of another.
:::
