---
title: "Homework 6: Two-Way ANOVA and Interactions"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Conduct and interpret a two-way ANOVA with interaction between two categorical predictors
- Create and interpret interaction plots using predicted cell means
- Demonstrate the equivalence between two-way ANOVA and regression with dummy variables and interaction terms
- **Related:** Chapter 3: *One-Way ANOVA* and Chapter 5: *Interactions* (in Lecture Notes)
:::

::: {.callout-tip}
## Stata-to-R Translation
| Stata Command | R Equivalent | Used In |
|:---|:---|:---|
| `anova wleread i.test_lan i.country i.test_lan#i.country` | `lm()` + `car::Anova(type = "III")` with `contr.sum` | Problem 1 |
| `margins test_lan#country` + `marginsplot` | `marginaleffects::avg_predictions()` + `ggplot()` | Problem 2 |
| `regress wleread uk germany test_lan lang_germ lang_uk` | `lm(wleread ~ uk + germany + test_lan + lang_germ + lang_uk)` | Problem 3 |
:::

**Dataset:** `pisa2000` from the `regdatasets` package --- reading literacy scores for 15-year-olds in three countries (USA, UK, Germany). We examine how reading scores vary by **country** and whether the student speaks the **test language** at home ($N = 4{,}528$).


## Setup

```{r}
#| label: hw06-setup
#| code-fold: true
#| code-summary: "Show R Code"

# --- Load data and filter to three countries ---
data(pisa2000)

pisa_3 <- pisa2000 |>
  filter(usa == 1 | uk == 1 | germany == 1) |>
  mutate(
    # Factor for country (USA as reference to match Stata)
    country_f = factor(
      case_when(usa == 1 ~ "USA", uk == 1 ~ "UK", germany == 1 ~ "Germany"),
      levels = c("USA", "UK", "Germany")
    ),
    # Factor for test language
    test_lan_f = factor(test_lan, levels = c(0, 1), labels = c("No", "Yes")),
    # Interaction dummy variables for Problem 3
    lang_germ = test_lan * germany,
    lang_uk   = test_lan * uk
  )

cat("N =", nrow(pisa_3), "\n")
cat("Countries:", levels(pisa_3$country_f), "\n")
cat("test_lan distribution:\n")
table(pisa_3$test_lan_f, pisa_3$country_f)
```


## Problem 1 {#sec-hw06-p1}

### Problem Statement

> Using a two-way ANOVA, test whether there is a **main effect of test language**, a **main effect of country**, and an **interaction** between the two on reading literacy scores (`wleread`).

### Solution

::: {.panel-tabset}

#### Conceptual

A **two-way ANOVA with interaction** partitions the variance in reading scores into four components:

1. **Main effect of `test_lan`** --- Does speaking the test language at home matter (averaging across countries)?
2. **Main effect of `country`** --- Do countries differ in reading scores (averaging across test-language groups)?
3. **Interaction (`test_lan` $\times$ `country`)** --- Does the test-language effect **differ across countries**?
4. **Residual** --- Within-cell variation not explained by the model.

The statistical model is:

$$
Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}
$$

where $\alpha_i$ is the effect of test language ($i = 0, 1$), $\beta_j$ is the effect of country ($j = \text{USA, UK, Germany}$), $(\alpha\beta)_{ij}$ is the interaction effect, and $\varepsilon_{ijk} \sim N(0, \sigma^2)$.

The interaction term is the key focus: if significant, the effect of one factor **depends on the level** of the other factor.

#### R Code

```{r}
#| label: hw06-p1-anova
#| code-fold: true
#| code-summary: "Show R Code"

# --- Fit the two-way ANOVA model with interaction ---
# IMPORTANT: For Type III SS to match Stata, we need sum-to-zero contrasts.
# R's default (contr.treatment) gives different Type III SS for main effects.
old_contrasts <- options(contrasts = c("contr.sum", "contr.poly"))

model_anova <- lm(wleread ~ factor(test_lan) * country_f, data = pisa_3)

# Type III SS (now matches Stata's anova with i. notation)
anova_table <- car::Anova(model_anova, type = "III")
print(anova_table)

# Restore default contrasts
options(old_contrasts)

# Overall model fit
cat("\n--- Overall Model Fit ---\n")
cat("N =", nobs(model_anova), "\n")
cat("R² =", round(summary(model_anova)$r.squared, 4), "\n")
cat("Root MSE =", round(summary(model_anova)$sigma, 4), "\n")

# Model-level F-test
model_summary <- summary(model_anova)
cat("F(", model_summary$fstatistic[2], ",",
    model_summary$fstatistic[3], ") =",
    round(model_summary$fstatistic[1], 2), "\n")
```

```{r}
#| label: hw06-p1-anova-detail
#| code-fold: true
#| code-summary: "Show detailed ANOVA table"

# Detailed ANOVA decomposition
# Model SS and Residual SS
anova_seq <- anova(model_anova)
model_ss <- sum(anova_seq$`Sum Sq`[1:3])
resid_ss <- anova_seq$`Sum Sq`[4]
resid_df <- anova_seq$Df[4]
resid_ms <- resid_ss / resid_df

cat("Model SS =", format(round(model_ss, 2), big.mark = ","), "\n")
cat("Model df = 5\n")
cat("Model MS =", format(round(model_ss / 5, 3), big.mark = ","), "\n")
cat("Residual SS =", format(round(resid_ss, 0), big.mark = ","), "\n")
cat("Residual df =", resid_df, "\n")
cat("Residual MS =", format(round(resid_ms, 5), big.mark = ","), "\n")
```

:::

**Key Results:**

| Source | SS | df | MS | F | p |
|:-------|---:|---:|---:|---:|:---|
| Model | 1,368,555.21 | 5 | 273,711.04 | 31.49 | $< .001$ |
| `test_lan` | 517,794.27 | 1 | 517,794.27 | 59.57 | $< .001$ |
| `country` | 452,092.82 | 2 | 226,046.41 | 26.00 | $< .001$ |
| `test_lan` $\times$ `country` | 255,837.10 | 2 | 127,918.55 | 14.72 | $< .001$ |
| Residual | 39,308,808 | 4,522 | 8,692.79 | | |
| **Total** | **40,677,363** | **4,527** | | | |

$N = 4{,}528$, $R^2 = 0.0336$, Root MSE $= 93.2351$

All three effects are **statistically significant**:

- **Test language** ($F(1, 4522) = 59.57$, $p < .001$): Students who speak the test language at home score higher on average.
- **Country** ($F(2, 4522) = 26.00$, $p < .001$): Significant differences in reading scores across countries.
- **Interaction** ($F(2, 4522) = 14.72$, $p < .001$): The test-language effect is **not the same** across countries.

::: {.callout-important}
## What the Significant Interaction Means
A significant interaction means that you **cannot simply talk about "the effect of test language"** without specifying which country you are referring to. The main effects alone tell an incomplete story. We need to examine the predicted cell means and interaction plots (Problem 2) to understand the nature of the interaction.
:::

::: {.callout-warning}
## R Contrast Settings for Type III Sums of Squares
Stata's `anova` command with `i.` notation produces **Type III** sums of squares by default. To replicate this in R using `car::Anova(type = "III")`, you must set **sum-to-zero contrasts** (`contr.sum`) before fitting the model. R's default `contr.treatment` (dummy coding) produces different Type III SS for main effects, because each main effect is tested conditional on the other being at its reference level rather than being averaged across all levels. The interaction term and residuals are unaffected by the contrast choice.
:::

::: {.callout-warning}
## Main Effects After a Significant Interaction
When the interaction is significant, interpret the main effect F-tests with caution. A significant main effect of country does not mean "country matters equally for all test-language groups." The interaction tells us the country effect *depends on* test language. Focus on the cell means and interaction plot (Problem 2) rather than the main effects alone.
:::


## Problem 2 {#sec-hw06-p2}

### Problem Statement

> Produce interaction plots to visualize the nature of the interaction between country and test language. Interpret the pattern.

### Solution

::: {.panel-tabset}

#### Conceptual

An **interaction plot** displays the predicted (or observed) cell means for each combination of the two factors. If the lines connecting the means are **parallel**, there is no interaction --- the effect of one factor is the same at every level of the other. When the lines are **not parallel** (they converge, diverge, or cross), that is visual evidence of an interaction.

We compute predicted means from the ANOVA model for each of the $2 \times 3 = 6$ cells:

| `test_lan` | Country | Predicted Mean |
|:-----------|:--------|---------------:|
| 0 (No) | Germany | 423.40 |
| 0 (No) | UK | 538.19 |
| 0 (No) | USA | 465.77 |
| 1 (Yes) | Germany | 521.93 |
| 1 (Yes) | UK | 538.18 |
| 1 (Yes) | USA | 528.63 |

The **test-language effect** (difference between Yes and No) within each country:

- **Germany:** $521.93 - 423.40 = 98.53$ points
- **USA:** $528.63 - 465.77 = 62.87$ points
- **UK:** $538.18 - 538.19 \approx 0$ points

The effect of speaking the test language at home is **huge in Germany**, **moderate in the USA**, and **essentially zero in the UK**. This is the interaction: the same variable has dramatically different effects depending on the country.

#### R Code

```{r}
#| label: hw06-p2-means
#| code-fold: true
#| code-summary: "Show R Code"

# --- Compute predicted cell means ---
pred_means <- marginaleffects::avg_predictions(
  model_anova,
  by = c("test_lan", "country_f")
) |>
  as.data.frame()

# Display the table
pred_means |>
  select(test_lan, country_f, estimate, std.error, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(
    col.names = c("test_lan", "Country", "Predicted Mean", "SE",
                   "95% CI Low", "95% CI High"),
    align = c("c", "l", "r", "r", "r", "r"),
    caption = "Predicted cell means from the two-way ANOVA model"
  )
```

::: {.panel-tabset group="fig-style"}

##### Simple

```{r}
#| label: fig-hw06-interaction-simple
#| fig-cap: "Interaction plot: Predicted reading scores by test language and country. Non-parallel lines indicate an interaction --- the test-language effect is largest in Germany, moderate in the USA, and negligible in the UK."
#| fig-alt: "Interaction plot with three lines connecting predicted reading scores from No to Yes for test language spoken at home. Germany shows the steepest increase (about 98 points), USA moderate increase (about 63 points), and UK nearly flat (close to zero increase), demonstrating the country-by-test-language interaction."
#| fig-width: 9
#| fig-height: 6.5
#| code-fold: true
#| code-summary: "Show R Code"

# Simple: Interaction plot using predicted means
pred_means_simple <- pisa_3 |>
  group_by(country_f, test_lan_f) |>
  summarize(mean_read = mean(wleread), .groups = "drop")

ggplot(pred_means_simple, aes(x = test_lan_f, y = mean_read,
                       color = country_f, group = country_f)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(x = "Test Language Spoken at Home?", y = "Mean Reading Score",
       color = "Country",
       title = "Interaction: Country x Test Language")
```

##### Detailed

```{r}
#| label: fig-hw06-interaction
#| fig-cap: "Interaction plot: Predicted reading scores by test language and country. Non-parallel lines indicate an interaction --- the test-language effect is largest in Germany, moderate in the USA, and negligible in the UK."
#| fig-alt: "Interaction plot with error bars showing predicted reading scores by test language and country. Germany has the steepest slope from 423 to 522, USA from 466 to 529, and UK is nearly flat around 538. Non-parallel lines confirm the significant interaction between country and test language."
#| fig-width: 9
#| fig-height: 6.5
#| code-fold: true
#| code-summary: "Show R Code"

# --- Interaction plot ---
plot_data <- pred_means |>
  mutate(
    test_lan_label = factor(test_lan, levels = c(0, 1),
                            labels = c("No", "Yes")),
    country_f = factor(country_f, levels = c("Germany", "USA", "UK"))
  )

ggplot(plot_data, aes(x = test_lan_label, y = estimate,
                      color = country_f, group = country_f)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.08, linewidth = 0.7) +
  # Annotate the test-language effect for each country

  annotate("text", x = 1.5, y = 460,
           label = "Germany: +98.5 pts",
           color = ber640_colors["primary"],
           fontface = "bold", size = 4.2) +
  annotate("text", x = 1.5, y = 505,
           label = "USA: +62.9 pts",
           color = ber640_colors["secondary"],
           fontface = "bold", size = 4.2) +
  annotate("text", x = 1.5, y = 545,
           label = "UK: ~0 pts",
           color = ber640_colors["accent"],
           fontface = "bold", size = 4.2) +
  scale_color_manual(
    name = "Country",
    values = c(
      "Germany" = unname(ber640_colors["primary"]),
      "USA"     = unname(ber640_colors["secondary"]),
      "UK"      = unname(ber640_colors["accent"])
    )
  ) +
  labs(
    title = "Interaction: Test Language x Country",
    subtitle = "Non-parallel lines reveal the interaction effect",
    x = "Speaks Test Language at Home",
    y = "Predicted Reading Score (wleread)"
  )
```

:::

:::

**Interpretation:**

The interaction plot reveals two key patterns:

1. **For students who do NOT speak the test language at home** (left side), the country differences are **very large**: UK students score highest (538), followed by the USA (466), with Germany far behind (423). The spread across countries is roughly 115 points.

2. **For students who DO speak the test language at home** (right side), the three countries **converge**: UK (538), USA (529), and Germany (522) are all within about 16 points of each other.

The converging lines from left to right are the hallmark of this interaction. In Germany, students who do not speak the test language at home are at a much greater disadvantage than in the other two countries. In the UK, whether or not the student speaks the test language at home makes virtually no difference to their reading score.

::: {.callout-important}
## Why Are the Lines Not Parallel?
If there were **no interaction**, the test-language effect would be the same in every country, and the three lines would be parallel. The fact that Germany's line rises steeply while the UK's line is nearly flat tells us that the test-language effect **depends on the country** --- the defining feature of an interaction.
:::


## Problem 3 {#sec-hw06-p3}

### Problem Statement

> By forming appropriate dummy variables for two of the countries and their interactions with `test_lan`, estimate the two-way ANOVA model using **multiple regression**. Use the USA as the reference group. Verify that the regression results match the ANOVA output.

### Solution

::: {.panel-tabset}

#### Conceptual

The two-way ANOVA model with interaction can be written as an equivalent regression model. With USA as the reference category for country, we need:

- **Main effect dummies:** `uk` (1 if UK, 0 otherwise) and `germany` (1 if Germany, 0 otherwise)
- **Main effect of test language:** `test_lan` (1 if speaks test language at home)
- **Interaction dummies:** `lang_germ` $=$ `test_lan` $\times$ `germany` and `lang_uk` $=$ `test_lan` $\times$ `uk`

The regression equation is:

$$
\widehat{\text{wleread}} = \hat{\beta}_0 + \hat{\beta}_1 \cdot \text{uk} + \hat{\beta}_2 \cdot \text{germany} + \hat{\beta}_3 \cdot \text{test\_lan} + \hat{\beta}_4 \cdot \text{lang\_germ} + \hat{\beta}_5 \cdot \text{lang\_uk}
$$

Each coefficient has a precise meaning tied to specific **cell mean differences**:

- $\hat{\beta}_0$ = Mean for the reference cell (USA, No test language)
- $\hat{\beta}_1$ = UK vs. USA difference **among non-speakers** ($\bar{Y}_{\text{UK,No}} - \bar{Y}_{\text{USA,No}}$)
- $\hat{\beta}_2$ = Germany vs. USA difference **among non-speakers** ($\bar{Y}_{\text{Ger,No}} - \bar{Y}_{\text{USA,No}}$)
- $\hat{\beta}_3$ = Test-language effect **in the USA** ($\bar{Y}_{\text{USA,Yes}} - \bar{Y}_{\text{USA,No}}$)
- $\hat{\beta}_4$ = How much **more/less** the test-language effect is in Germany vs. USA
- $\hat{\beta}_5$ = How much **more/less** the test-language effect is in the UK vs. USA

The total test-language effect for each country:

- **USA:** $\hat{\beta}_3 = 62.87$
- **Germany:** $\hat{\beta}_3 + \hat{\beta}_4 = 62.87 + 35.66 = 98.53$
- **UK:** $\hat{\beta}_3 + \hat{\beta}_5 = 62.87 + (-62.87) = 0.00$

#### R Code

```{r}
#| label: hw06-p3-regression
#| code-fold: true
#| code-summary: "Show R Code"

# --- Regression with dummy variables and interactions ---
# uk and germany are pre-existing dummies in pisa2000
# lang_germ and lang_uk were created in Setup

model_reg <- lm(wleread ~ uk + germany + test_lan + lang_germ + lang_uk,
                data = pisa_3)

# Full summary
summary(model_reg)
```

```{r}
#| label: tbl-hw06-p3-coefs
#| tbl-cap: "Regression coefficients for the dummy-coded two-way ANOVA model (USA as reference)"
#| code-fold: true
#| code-summary: "Show R Code"

# --- Tidy coefficient table ---
tidy_model(model_reg) |>
  knitr::kable(
    digits = c(0, 4, 4, 4, 4, 4, 4),
    col.names = c("Term", "Estimate", "SE", "Statistic", "p-value",
                   "95% CI Low", "95% CI High"),
    align = c("l", rep("r", 6))
  )
```

```{r}
#| label: hw06-p3-equivalence
#| code-fold: true
#| code-summary: "Show ANOVA-Regression equivalence"

# --- Verify equivalence with ANOVA ---
cat("=== ANOVA vs. Regression Equivalence ===\n\n")

# R-squared
cat("R² (ANOVA model):", round(summary(model_anova)$r.squared, 4), "\n")
cat("R² (Regression):  ", round(summary(model_reg)$r.squared, 4), "\n\n")

# Root MSE
cat("Root MSE (ANOVA):    ", round(summary(model_anova)$sigma, 4), "\n")
cat("Root MSE (Regression):", round(summary(model_reg)$sigma, 4), "\n\n")

# F-statistic
cat("F (ANOVA):     ", round(summary(model_anova)$fstatistic[1], 2), "\n")
cat("F (Regression):", round(summary(model_reg)$fstatistic[1], 2), "\n\n")

# Model SS
anova_reg <- anova(model_reg)
model_ss_reg <- sum(anova_reg$`Sum Sq`[1:5])
cat("Model SS (ANOVA):     ", format(round(model_ss, 2), big.mark = ","), "\n")
cat("Model SS (Regression):", format(round(model_ss_reg, 2), big.mark = ","), "\n\n")

# Predicted cell means from regression
cat("--- Predicted Cell Means from Regression ---\n")
cat("USA, No test_lan  (intercept):        ",
    round(coef(model_reg)["(Intercept)"], 4), "\n")
cat("UK, No test_lan   (intercept + uk):   ",
    round(coef(model_reg)["(Intercept)"] + coef(model_reg)["uk"], 4), "\n")
cat("Germany, No test_lan (intercept + ger):",
    round(coef(model_reg)["(Intercept)"] + coef(model_reg)["germany"], 4), "\n")
cat("USA, Yes test_lan (intercept + tl):   ",
    round(coef(model_reg)["(Intercept)"] + coef(model_reg)["test_lan"], 4), "\n")
cat("UK, Yes test_lan  (int + uk + tl + lu):",
    round(coef(model_reg)["(Intercept)"] + coef(model_reg)["uk"] +
          coef(model_reg)["test_lan"] + coef(model_reg)["lang_uk"], 4), "\n")
cat("Germany, Yes test_lan (int + g + tl + lg):",
    round(coef(model_reg)["(Intercept)"] + coef(model_reg)["germany"] +
          coef(model_reg)["test_lan"] + coef(model_reg)["lang_germ"], 4), "\n")
```

:::

**Regression Results** ($N = 4{,}528$, $F(5, 4522) = 31.49$, $R^2 = 0.0336$, Root MSE $= 93.235$):

| Predictor | $\hat{\beta}$ | SE | $t$ | $p$ |
|:----------|---:|---:|---:|:---|
| `uk` | 72.424 | 17.182 | 4.22 | $< .001$ |
| `germany` | $-42.365$ | 14.836 | $-2.86$ | $.004$ |
| `test_lan` | 62.867 | 10.345 | 6.08 | $< .001$ |
| `lang_germ` | 35.663 | 15.325 | 2.33 | $.020$ |
| `lang_uk` | $-62.873$ | 17.591 | $-3.57$ | $< .001$ |
| Intercept | 465.766 | 9.883 | 47.13 | $< .001$ |

**Coefficient interpretation:**

1. **Intercept** ($\hat{\beta}_0 = 465.77$): The predicted mean reading score for **USA students who do NOT speak the test language at home** (the reference cell).

2. **`uk`** ($\hat{\beta}_1 = 72.42$, $p < .001$): Among students who do **not** speak the test language at home, **UK students score 72.42 points higher** than USA students.

3. **`germany`** ($\hat{\beta}_2 = -42.37$, $p = .004$): Among non-test-language speakers, **German students score 42.37 points lower** than USA students.

4. **`test_lan`** ($\hat{\beta}_3 = 62.87$, $p < .001$): In the **USA**, students who speak the test language at home score **62.87 points higher** than those who do not.

5. **`lang_germ`** ($\hat{\beta}_4 = 35.66$, $p = .020$): The test-language effect in **Germany** is **35.66 points larger** than in the USA. The total test-language effect in Germany is:

$$
\hat{\beta}_3 + \hat{\beta}_4 = 62.87 + 35.66 = 98.53 \text{ points}
$$

6. **`lang_uk`** ($\hat{\beta}_5 = -62.87$, $p < .001$): The test-language effect in the **UK** is **62.87 points smaller** than in the USA. The total test-language effect in the UK is:

$$
\hat{\beta}_3 + \hat{\beta}_5 = 62.87 + (-62.87) \approx 0 \text{ points}
$$

This is a striking result --- in the UK, whether or not the student speaks the test language at home is **completely unrelated** to reading performance.

::: {.callout-important}
## ANOVA = Regression: A Fundamental Equivalence
The regression model produces **identical** results to the two-way ANOVA:

- Same $R^2 = 0.0336$
- Same $F(5, 4522) = 31.49$
- Same Model $\text{SS} = 1{,}368{,}555.21$ with 5 df
- Same Root MSE $= 93.235$
- Same predicted cell means

This is not a coincidence --- **ANOVA is a special case of regression** where all predictors are dummy-coded categorical variables. The ANOVA table partitions variance by factor, while the regression table shows individual coefficient tests, but the underlying model is the same.
:::


::: {.callout-caution}
## Common Mistake: Interpreting Main Effects After a Significant Interaction
When the country $\times$ test-language interaction is significant, the main effects of country and test language are difficult to interpret on their own. A significant main effect of country does not mean "country matters equally for speakers and non-speakers." The interaction tells us the answer depends on which test-language group we examine. Focus on the cell means and interaction plot rather than the main effect $F$-tests.
:::


## Summary

This homework demonstrated three perspectives on the same analysis:

1. **Two-way ANOVA** (Problem 1): Tests the overall significance of each factor and the interaction using $F$-tests. The significant interaction ($F(2, 4522) = 14.72$, $p < .001$) tells us the test-language effect varies across countries, but does not tell us *how*.

2. **Interaction plots** (Problem 2): Visualize the cell means to reveal the *nature* of the interaction. The lines converge from left to right: country differences are large among non-speakers but nearly disappear among speakers. Germany has the steepest slope (largest test-language effect), and the UK's line is essentially flat.

3. **Dummy-coded regression** (Problem 3): Decomposes the interaction into **specific, interpretable coefficients**. Each interaction term ($\hat{\beta}_4$, $\hat{\beta}_5$) quantifies exactly how much the test-language effect in one country differs from the reference country (USA). The regression and ANOVA are mathematically equivalent --- same $R^2$, same $F$, same SS.

The core lesson: **two-way ANOVA with interaction is simply regression with dummy variables and their products**. The interaction means the effect of one variable depends on the level of the other. Always follow up a significant interaction with cell means or interaction plots to understand the pattern.

::: {.callout-note}
## Looking Ahead
In **Homework 7**, we transition from linear models to **logistic regression**, learning how to interpret odds ratios and confidence intervals from published tables. Then in **Homework 8**, we apply logistic regression to the classic UC Berkeley admissions dataset, encountering **Simpson's Paradox** --- where a marginal association reverses direction after controlling for a confounding variable.
:::
