---
title: "Homework 7: Logistic Regression Interpretation"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Convert logistic regression coefficients to odds ratios using $\exp(\hat{\beta}_j)$
- Interpret adjusted odds ratios for continuous and dummy predictors
- Construct confidence intervals on the log-odds scale and then exponentiate
- Use reciprocal odds ratios when the original odds ratio is less than 1
- Convert odds ratios to percentage changes in odds
- **Related notes:** [Chapter 9: Simple Logistic Regression](../../notes/chapters/09-simple-logistic-regression.html) and [Chapter 10: Multiple Logistic Regression](../../notes/chapters/10-multiple-logistic-regression.html)
- **Related labs:** [Lab 9: Simple Logistic Regression](../../labs/labs/lab09-logistic-intro.html) and [Lab 10: Multiple Logistic Regression](../../labs/labs/lab10-logistic-extended.html)
:::

This homework is conceptual. We do not fit a model from raw data. Instead, we interpret a published logistic regression table from @kalmijn1994.

**Data:** National Survey of Families and Households (NSFH), 1987-1988. Non-Hispanic white respondents born 1900-1963, raised by two parents ($N = 6{,}092$).

**Response variable:** whether the respondent completed high school ($1 =$ yes, $0 =$ no).


## The Logistic Regression Table

The table below reproduces Model B from Table 2 of @kalmijn1994. The coefficients are maximum-likelihood estimates from a logistic regression predicting high school completion.

```{r}
#| label: tbl-hw07-kalmijn
#| tbl-cap: "Logistic regression coefficients for obtaining a high school degree [@kalmijn1994, Table 2, Model B]."
#| echo: false

kalmijn_table <- tibble(
  Variable = c(
    "Respondent's year of birth (-63 to 0)",
    "Father's education (0 to 17)",
    "Mother's education (0 to 17)",
    "Father's occupation (1 to 9)",
    "Mother's occupation (0, 1 to 9)",
    "Mother employed (0 to 1)",
    "Number of siblings",
    "Intercept",
    "N"
  ),
  Model_B = c(
    ".350** (.033)",
    ".126** (.016)",
    ".093** (.018)",
    ".295** (.047)",
    ".212** (.068)",
    "-.921** (.189)",
    "-.162** (.016)",
    ".747** (.247)",
    "6,092"
  )
)

kalmijn_table |>
  knitr::kable(
    col.names = c("Independent variable", "Model B"),
    align = c("l", "r"),
    caption = NULL
  )
```

**Note:** \*$p < .05$, \*\*$p < .01$ (two-tailed). Numbers in parentheses are estimated standard errors. Coefficients are maximum-likelihood estimates. Coefficients for year of birth are multiplied by 10.

::: {.callout-note}
## Variable Definitions
- **Respondent's year of birth:** year of birth minus 1963.
- **Father's/Mother's education:** completed years of schooling, 0 to 17.
- **Father's occupation:** socioeconomic index of father's occupation when respondent was 16, scale 1 to 9.
- **Mother's occupation:** socioeconomic index of mother's last occupation before respondent turned 18, scale 0 and 1 to 9.
- **Mother employed:** $1 =$ mother was employed between respondent's birth and age 18; $0 =$ not employed.
- **Number of siblings:** count of siblings.
:::

```{r}
#| label: hw07-reference-check
#| include: false

stopifnot(
  kalmijn_table$Model_B[kalmijn_table$Variable == "Mother's education (0 to 17)"] == ".093** (.018)",
  kalmijn_table$Model_B[kalmijn_table$Variable == "Mother employed (0 to 1)"] == "-.921** (.189)",
  kalmijn_table$Model_B[kalmijn_table$Variable == "Mother's occupation (0, 1 to 9)"] == ".212** (.068)"
)

z_star <- 1.96
```

::: {.callout-warning}
## Odds Are Not Probabilities
An odds ratio of 1.10 does not mean the probability increases by 10%. It means the **odds** are multiplied by 1.10. For common outcomes, odds ratios and probability ratios can differ substantially, so always use odds language unless you explicitly convert to predicted probabilities.
:::


## Problem 1

### Problem Statement

> For Model B, calculate and interpret the estimated odds ratio for mother's education.

### Solution

For mother's education, the published coefficient is:

$$
\hat{\beta}_j = 0.093.
$$

Exponentiating gives the odds ratio:

```{r}
#| label: hw07-p1-mother-education-or

beta_mother_education <- 0.093
or_mother_education <- exp(beta_mother_education)

mother_education <- tibble(
  predictor = "Mother's education",
  beta = beta_mother_education,
  odds_ratio = or_mother_education,
  percent_change_in_odds = 100 * (or_mother_education - 1)
)

mother_education |>
  mutate(
    odds_ratio = round(odds_ratio, 2),
    percent_change_in_odds = round(percent_change_in_odds, 0)
  ) |>
  knitr::kable(
    col.names = c("Predictor", "Coefficient", "Odds ratio", "Percent change in odds"),
    align = c("l", "r", "r", "r"),
    caption = "Odds ratio for mother's education."
  )
```

```{r}
#| label: hw07-p1-check
#| include: false

stopifnot(
  isTRUE(all.equal(round(mother_education$odds_ratio, 2), 1.10)),
  isTRUE(all.equal(round(mother_education$percent_change_in_odds, 0), 10))
)
```

Thus:

$$
\exp(0.093) = 1.10.
$$

The odds of obtaining a high school degree are estimated to increase by about **10%** for each additional year of mother's schooling, holding constant the other explanatory variables in Model B.

In odds-ratio language: comparing two respondents who are the same on the other variables but differ by one year in mother's education, the respondent whose mother has one additional year of schooling has **1.10 times the odds** of completing high school.


## Problem 2

### Problem Statement

> For Model B, calculate the estimated odds ratio for "Mother employed" and form an approximate 95% confidence interval for the odds ratio. Interpret these results.

### Solution

For `Mother employed`, the published coefficient and standard error are:

$$
\hat{\beta}_j = -0.921, \qquad SE = 0.189.
$$

We first compute the odds ratio:

```{r}
#| label: hw07-p2-mother-employed-or

beta_mother_employed <- -0.921
se_mother_employed <- 0.189
or_mother_employed <- exp(beta_mother_employed)

mother_employed <- tibble(
  predictor = "Mother employed",
  beta = beta_mother_employed,
  se = se_mother_employed,
  odds_ratio = or_mother_employed,
  reciprocal_odds_ratio = 1 / or_mother_employed
)

mother_employed |>
  transmute(
    predictor,
    beta,
    odds_ratio = round(odds_ratio, 2),
    reciprocal_odds_ratio = round(reciprocal_odds_ratio, 1)
  ) |>
  knitr::kable(
    col.names = c("Predictor", "Coefficient", "Odds ratio", "Reciprocal odds ratio"),
    align = c("l", "r", "r", "r"),
    caption = "Odds ratio and reciprocal odds ratio for mother employed."
  )
```

The odds ratio is:

$$
\exp(-0.921) = 0.40.
$$

Because this odds ratio is less than 1, it is often clearer to reverse the comparison:

$$
\frac{1}{0.40} = 2.5.
$$

Respondents whose mothers were **not employed** are estimated to have about **2.5 times the odds** of graduating from high school as respondents whose mothers were employed, holding the other explanatory variables constant.

### Confidence Interval

Construct the confidence interval on the log-odds scale first:

$$
\hat{\beta}_j \pm 1.96 \times SE
= -0.921 \pm 1.96(0.189).
$$

The answer key reports the rounded coefficient interval and then exponentiates those endpoints.

```{r}
#| label: hw07-p2-mother-employed-ci

mother_employed_beta_lower <- round(beta_mother_employed - z_star * se_mother_employed, 3)
mother_employed_beta_upper <- round(beta_mother_employed + z_star * se_mother_employed, 3)

mother_employed_ci <- tibble(
  predictor = "Mother employed",
  beta_lower = mother_employed_beta_lower,
  beta_upper = mother_employed_beta_upper,
  or_lower = exp(mother_employed_beta_lower),
  or_upper = exp(mother_employed_beta_upper)
) |>
  mutate(
    reciprocal_lower = 1 / or_upper,
    reciprocal_upper = 1 / or_lower
  )

mother_employed_ci |>
  transmute(
    predictor,
    beta_lower,
    beta_upper,
    or_lower = round(or_lower, 3),
    or_upper = round(or_upper, 3),
    reciprocal_lower = round(reciprocal_lower, 1),
    reciprocal_upper = round(reciprocal_upper, 1)
  ) |>
  knitr::kable(
    col.names = c(
      "Predictor",
      "Coefficient CI lower",
      "Coefficient CI upper",
      "OR CI lower",
      "OR CI upper",
      "Reciprocal CI lower",
      "Reciprocal CI upper"
    ),
    align = c("l", "r", "r", "r", "r", "r", "r"),
    caption = "Answer-key rounded log-odds, odds-ratio, and reciprocal confidence intervals for mother employed."
  )
```

```{r}
#| label: hw07-p2-check
#| include: false

stopifnot(
  isTRUE(all.equal(round(mother_employed$odds_ratio, 2), 0.40)),
  isTRUE(all.equal(round(mother_employed$reciprocal_odds_ratio, 1), 2.5)),
  isTRUE(all.equal(mother_employed_ci$beta_lower, -1.291)),
  isTRUE(all.equal(mother_employed_ci$beta_upper, -0.551)),
  isTRUE(all.equal(round(mother_employed_ci$or_lower, 3), 0.275)),
  isTRUE(all.equal(round(mother_employed_ci$or_upper, 3), 0.576)),
  isTRUE(all.equal(round(mother_employed_ci$reciprocal_lower, 1), 1.7)),
  isTRUE(all.equal(round(mother_employed_ci$reciprocal_upper, 1), 3.6))
)
```

The 95% confidence interval for the log-odds coefficient is:

$$
[-1.291,\ -0.551].
$$

Exponentiating gives the odds-ratio interval:

$$
[\exp(-1.291),\ \exp(-0.551)] = [0.275,\ 0.576].
$$

For the reciprocal interpretation, reverse the endpoints:

$$
\left[\frac{1}{0.576},\ \frac{1}{0.275}\right] = [1.7,\ 3.6].
$$

So the reciprocal interpretation is: respondents whose mothers were not employed are estimated to have between **1.7 and 3.6 times the odds** of graduating from high school as respondents whose mothers were employed. As with any 95% confidence interval, the true population odds ratio could fall outside this interval in repeated sampling.

::: {.callout-tip}
## Confidence Interval Workflow
For odds ratios, use this order:

1. Build the confidence interval for the coefficient on the log-odds scale.
2. Exponentiate both endpoints.
3. If using a reciprocal interpretation, take reciprocals and reverse the endpoint order.
:::


## Problem 3

### Problem Statement

> (a) The author reported that a one-point increase in mother's occupational level is associated with a 24% increase in the odds of a high school diploma. Explain how he made this interpretation.
>
> (b) Form a 95% confidence interval for the percentage increase in the odds.

### Solution

For mother's occupation, the coefficient and standard error are:

$$
\hat{\beta}_j = 0.212, \qquad SE = 0.068.
$$

The odds ratio is:

```{r}
#| label: hw07-p3-mother-occupation-or

beta_mother_occupation <- 0.212
se_mother_occupation <- 0.068
or_mother_occupation <- exp(beta_mother_occupation)

mother_occupation <- tibble(
  predictor = "Mother's occupation",
  beta = beta_mother_occupation,
  se = se_mother_occupation,
  odds_ratio = or_mother_occupation,
  percent_change = 100 * (or_mother_occupation - 1)
)

mother_occupation |>
  transmute(
    predictor,
    beta,
    odds_ratio = round(odds_ratio, 2),
    percent_change = round(percent_change, 0)
  ) |>
  knitr::kable(
    col.names = c("Predictor", "Coefficient", "Odds ratio", "Percent change in odds"),
    align = c("l", "r", "r", "r"),
    caption = "Odds ratio and percentage change for mother's occupation."
  )
```

Thus:

$$
\exp(0.212) = 1.24.
$$

The percentage increase in odds is:

$$
100\% \times (1.24 - 1) = 24\%.
$$

Each one-point increase in mother's occupational level is associated with a **24% increase in the odds** of completing high school, holding constant the other explanatory variables.

### Confidence Interval for the Percentage Increase

Again, start on the log-odds scale:

```{r}
#| label: hw07-p3-mother-occupation-ci

mother_occupation_beta_lower <- round(beta_mother_occupation - z_star * se_mother_occupation, 3)
mother_occupation_beta_upper <- round(beta_mother_occupation + z_star * se_mother_occupation, 3)

mother_occupation_ci <- tibble(
  predictor = "Mother's occupation",
  beta_lower = mother_occupation_beta_lower,
  beta_upper = mother_occupation_beta_upper,
  or_lower = exp(mother_occupation_beta_lower),
  or_upper = exp(mother_occupation_beta_upper)
) |>
  mutate(
    percent_lower = 100 * (or_lower - 1),
    percent_upper = 100 * (or_upper - 1)
  )

mother_occupation_ci |>
  transmute(
    predictor,
    beta_lower,
    beta_upper,
    or_lower = round(or_lower, 2),
    or_upper = round(or_upper, 2),
    percent_lower = round(percent_lower, 0),
    percent_upper = round(percent_upper, 0)
  ) |>
  knitr::kable(
    col.names = c(
      "Predictor",
      "Coefficient CI lower",
      "Coefficient CI upper",
      "OR CI lower",
      "OR CI upper",
      "Percent lower",
      "Percent upper"
    ),
    align = c("l", "r", "r", "r", "r", "r", "r"),
    caption = "Answer-key rounded log-odds, odds-ratio, and percent-change confidence intervals for mother's occupation."
  )
```

```{r}
#| label: hw07-p3-check
#| include: false

stopifnot(
  isTRUE(all.equal(round(mother_occupation$odds_ratio, 2), 1.24)),
  isTRUE(all.equal(round(mother_occupation$percent_change, 0), 24)),
  isTRUE(all.equal(mother_occupation_ci$beta_lower, 0.079)),
  isTRUE(all.equal(mother_occupation_ci$beta_upper, 0.345)),
  isTRUE(all.equal(round(mother_occupation_ci$or_lower, 2), 1.08)),
  isTRUE(all.equal(round(mother_occupation_ci$or_upper, 2), 1.41)),
  isTRUE(all.equal(round(mother_occupation_ci$percent_lower, 0), 8)),
  isTRUE(all.equal(round(mother_occupation_ci$percent_upper, 0), 41))
)
```

The coefficient interval is:

$$
[0.079,\ 0.345].
$$

Exponentiating gives:

$$
[\exp(0.079),\ \exp(0.345)] = [1.08,\ 1.41].
$$

Converting the odds-ratio interval to a percentage-change interval gives:

$$
100\% \times (1.08 - 1) = 8\%,
\qquad
100\% \times (1.41 - 1) = 41\%.
$$

So the 95% confidence interval for the percentage increase in the odds is approximately **8% to 41%**.


## Summary

This homework used one published logistic regression table to practice three transformations:

1. **Coefficient to odds ratio:** $\exp(\hat{\beta}_j)$.
2. **Coefficient CI to odds-ratio CI:** build $\hat{\beta}_j \pm 1.96SE$ first, then exponentiate.
3. **Odds ratio to percentage change in odds:** $100\% \times (\text{OR} - 1)$.

For odds ratios less than 1, a reciprocal interpretation can be easier to read, but the comparison group must be reversed and the confidence-interval endpoints must be reversed.

::: {.callout-note}
## Looking Ahead
In **Homework 8**, you will fit logistic regression models to real data and use these same odds-ratio interpretation skills in the presence of confounding and Simpson's paradox.
:::
