---
title: "Homework 8: Logistic Regression and Simpson's Paradox"
---

{{< include ../_common.qmd >}}

## Overview

::: {.callout-note}
## Learning Objectives
- Build three-way and collapsed frequency tables from binary data
- Compute and interpret a crude odds ratio
- Fit a multiple logistic regression with categorical predictors
- Compare crude and adjusted odds ratios to explain Simpson's paradox
- Conduct a Wald test for one logistic-regression coefficient
- Compute a grouped Pearson goodness-of-fit test for categorical predictor patterns
- **Related notes:** [Chapter 10: *Multiple Logistic Regression*](../../notes/chapters/10-multiple-logistic-regression.html) and [Chapter 11: *Logistic Regression Diagnostics*](../../notes/chapters/11-logistic-diagnostics.html)
- **Related labs:** [Lab 10: *Multiple Logistic Regression*](../../labs/labs/lab10-logistic-extended.html) and [Lab 11: *Logistic Diagnostics and Classification*](../../labs/labs/lab11-logistic-diagnostics.html)
:::

::: {.callout-tip}
## Stata-to-R Translation
| Stata command | R equivalent | Used in |
|:---|:---|:---|
| `table admitted female, contents(freq) by(departme)` | `count(department, admitted, female)` | Problem 1 |
| `cs admitted female, or` | `glm(admitted ~ female, family = binomial)` or a manual odds ratio | Problem 2 |
| `logit admitted female i.departme, or` | `glm(admitted ~ female + factor(department), family = binomial)` | Problem 3 |
| `testparm female` | the Wald $z$ test in `model_params()` | Problem 5 |
| `estat gof, table` | grouped observed/expected counts by covariate pattern | Problem 6 |
:::

**Dataset:** `berkeley` from the `regdatasets` package, based on the UC Berkeley graduate-admissions data discussed by @bickel1975.

The Stata reference uses the variable name `departme`; the R dataset uses `department`. They refer to the same five Berkeley graduate departments.

**Variables:**

- `department`: department code, 1--5
- `female`: applicant gender, where $0 =$ male and $1 =$ female
- `admitted`: admission outcome, where $0 =$ not admitted and $1 =$ admitted

The Berkeley admissions data are a classic example of **Simpson's paradox**: an aggregate association changes after conditioning on a confounding variable.


## Setup

```{r}
#| label: hw08-setup

data(berkeley)

berk <- berkeley |>
  mutate(
    department_f = factor(department, levels = 1:5, labels = paste("Dept", 1:5)),
    female_f = factor(female, levels = c(0, 1), labels = c("Male", "Female")),
    admitted_f = factor(
      admitted,
      levels = c(0, 1),
      labels = c("Not admitted", "Admitted")
    )
  )

berk |>
  summarise(
    n = n(),
    departments = n_distinct(department),
    female_applicants = sum(female),
    admitted_applicants = sum(admitted),
    admission_rate = mean(admitted)
  ) |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw08-setup-check
#| include: false

stopifnot(nrow(berk) == 3593)
stopifnot(identical(sort(unique(berk$department)), 1:5))
stopifnot(all(berk$female %in% c(0, 1)))
stopifnot(all(berk$admitted %in% c(0, 1)))
stopifnot(sum(is.na(berk)) == 0)
stopifnot(sum(berk$female) == 1727)
stopifnot(sum(berk$admitted) == 1154)
```


## Problem 1

### Problem Statement

> Read the data and form a three-way table of frequencies: admitted by gender, separately for each department.

### Solution

```{r}
#| label: hw08-p1-three-way-table

dept_counts <- berk |>
  count(department, admitted_f, female_f, name = "n") |>
  tidyr::pivot_wider(
    names_from = female_f,
    values_from = n
  ) |>
  arrange(department, admitted_f)

dept_counts
```

The same table can be summarized as admission rates within each department-by-gender cell:

```{r}
#| label: hw08-p1-department-rates

dept_rates <- berk |>
  group_by(department_f, female_f) |>
  summarise(
    n = n(),
    admitted = sum(admitted),
    not_admitted = n - admitted,
    admission_rate = admitted / n,
    odds = admission_rate / (1 - admission_rate),
    .groups = "drop"
  )

dept_rates |>
  gt::gt() |>
  gt::cols_label(
    department_f = "Department",
    female_f = "Gender",
    n = "N",
    admitted = "Admitted",
    not_admitted = "Not admitted",
    admission_rate = "Admission rate",
    odds = "Odds"
  ) |>
  gt::fmt_number(columns = c(n, admitted, not_admitted), decimals = 0, use_seps = FALSE) |>
  gt::fmt_number(columns = c(admission_rate, odds), decimals = 4) |>
  gt::cols_align("left", columns = c(department_f, female_f)) |>
  gt::cols_align("right", columns = c(n, admitted, not_admitted, admission_rate, odds)) |>
  gt::cols_width(
    department_f ~ gt::pct(17),
    female_f ~ gt::pct(14),
    n ~ gt::pct(8),
    admitted ~ gt::pct(14),
    not_admitted ~ gt::pct(16),
    admission_rate ~ gt::pct(18),
    odds ~ gt::pct(13)
  ) |>
  gt::tab_options(table.width = gt::pct(100))
```

Department matters for two reasons. First, departments differ sharply in admission rates. Second, the gender composition of applicant pools differs across departments. That is exactly the setup in which a collapsed gender comparison can be misleading.

```{r}
#| label: hw08-p1-check
#| include: false

count_key <- berk |>
  mutate(key = paste(department, admitted, female, sep = "_")) |>
  count(key) |>
  tibble::deframe()

expected_counts <- c(
  "1_0_0" = 207, "1_0_1" = 8,   "1_1_0" = 353, "1_1_1" = 17,
  "2_0_0" = 205, "2_0_1" = 391, "2_1_0" = 120, "2_1_1" = 202,
  "3_0_0" = 279, "3_0_1" = 244, "3_1_0" = 138, "3_1_1" = 131,
  "4_0_0" = 138, "4_0_1" = 299, "4_1_0" = 53,  "4_1_1" = 94,
  "5_0_0" = 351, "5_0_1" = 317, "5_1_0" = 22,  "5_1_1" = 24
)

stopifnot(identical(
  as.integer(count_key[names(expected_counts)]),
  as.integer(expected_counts)
))
stopifnot(abs(dept_rates$admission_rate[dept_rates$department_f == "Dept 1" & dept_rates$female_f == "Male"] - 353 / 560) < 1e-12)
stopifnot(abs(dept_rates$admission_rate[dept_rates$department_f == "Dept 1" & dept_rates$female_f == "Female"] - 17 / 25) < 1e-12)
```


## Problem 2

### Problem Statement

> Form a two-way table collapsing over departments. Calculate the odds ratio and interpret it.

### Solution

```{r}
#| label: hw08-p2-collapsed-table

aggregate_by_gender <- berk |>
  group_by(female_f) |>
  summarise(
    n = n(),
    admitted = sum(admitted),
    not_admitted = n - admitted,
    admission_rate = admitted / n,
    odds = admitted / not_admitted,
    .groups = "drop"
  )

aggregate_by_gender |>
  round_numeric(digits = 4)
```

```{r}
#| label: hw08-p2-crude-or

odds_female <- aggregate_by_gender$odds[aggregate_by_gender$female_f == "Female"]
odds_male <- aggregate_by_gender$odds[aggregate_by_gender$female_f == "Male"]
crude_or <- odds_female / odds_male

tibble(
  comparison = "Female vs male, ignoring department",
  odds_female = odds_female,
  odds_male = odds_male,
  odds_ratio = crude_or,
  percent_change_in_odds = 100 * (crude_or - 1)
) |>
  round_numeric(digits = 4)
```

The crude odds ratio is:

$$
\widehat{\text{OR}}_{\text{crude}}
= \frac{468 / 1259}{686 / 1180}
= 0.64.
$$

Ignoring department, female applicants have estimated admission odds about 36% lower than male applicants. This is the aggregate comparison; it has not adjusted for the department to which applicants applied.

::: {.callout-warning}
## Why This Is Only a First Look
The crude odds ratio compares all female applicants to all male applicants after pooling departments together. If department is associated with both gender and admission, this pooled comparison mixes the gender comparison with department composition.
:::

```{r}
#| label: hw08-p2-check
#| include: false

stopifnot(aggregate_by_gender$not_admitted[aggregate_by_gender$female_f == "Male"] == 1180)
stopifnot(aggregate_by_gender$not_admitted[aggregate_by_gender$female_f == "Female"] == 1259)
stopifnot(aggregate_by_gender$admitted[aggregate_by_gender$female_f == "Male"] == 686)
stopifnot(aggregate_by_gender$admitted[aggregate_by_gender$female_f == "Female"] == 468)
stopifnot(abs(crude_or - 0.6394079247) < 1e-10)
```


## Problem 3

### Problem Statement

> Fit a logistic regression of `admitted` on `female` and `department` as a categorical variable. Report the odds ratios and interpret the coefficient of `female`.

### Solution

Department is a set of labels, not a numeric scale. In R, use `factor(department)` so the model creates dummy variables for departments 2--5, with department 1 as the reference group.

```{r}
#| label: hw08-p3-adjusted-model

model_adjusted <- glm(
  admitted ~ female + factor(department),
  data = berk,
  family = binomial
)

adjusted_or_table <- model_params(
  model_adjusted,
  exponentiate = TRUE,
  ci_method = "wald"
) |>
  as_tibble() |>
  mutate(
    Parameter = dplyr::recode(
      Parameter,
      `(Intercept)` = "Baseline odds: Dept 1 male",
      female = "Female vs male",
      `factor(department)2` = "Dept 2 vs Dept 1",
      `factor(department)3` = "Dept 3 vs Dept 1",
      `factor(department)4` = "Dept 4 vs Dept 1",
      `factor(department)5` = "Dept 5 vs Dept 1"
    ),
    p = format_p(p)
  ) |>
  rename(Odds_ratio = Coefficient) |>
  select(Parameter, Odds_ratio, SE, CI_low, CI_high, z, p)

adjusted_or_table |>
  round_numeric(digits = 4) |>
  rename(`Baseline odds or odds ratio` = Odds_ratio) |>
  knitr::kable(
    col.names = c(
      "Term",
      "Baseline odds or odds ratio",
      "Log-odds SE",
      "CI lower",
      "CI upper",
      "z",
      "p"
    ),
    align = c("l", "r", "r", "r", "r", "r", "r"),
    caption = "Adjusted logistic regression odds ratios by gender and department."
  )
```

The table above is built from `model_params()` so we can relabel terms and match the Stata-style answer key. For a report-ready regression table, `gtsummary::tbl_regression()` can format the same adjusted model in one short call:

```{r}
#| label: hw08-p3-gtsummary-table
#| message: false
#| warning: false

gtsummary::tbl_regression(
  model_adjusted,
  exponentiate = TRUE,
  label = list(female ~ "Female vs male")
) |>
  gtsummary::bold_labels()
```

```{r}
#| label: hw08-p3-model-fit

model_null <- glm(admitted ~ 1, data = berk, family = binomial)
lr_chisq <- as.numeric(deviance(model_null) - deviance(model_adjusted))
lr_df <- attr(logLik(model_adjusted), "df") - attr(logLik(model_null), "df")

adjusted_fit <- tibble(
  n = nobs(model_adjusted),
  log_likelihood = as.numeric(logLik(model_adjusted)),
  lr_chisq = lr_chisq,
  df = lr_df,
  p = format_p(pchisq(lr_chisq, df = lr_df, lower.tail = FALSE)),
  pseudo_r2 = 1 - as.numeric(logLik(model_adjusted) / logLik(model_null))
)

adjusted_fit |>
  round_numeric(digits = 4)
```

The adjusted odds ratio for `female` is:

$$
\widehat{\text{OR}}_{\text{female} \mid \text{department}} = 0.97,
\qquad
95\% \ \text{CI} = [0.82, 1.15],
\qquad
p = 0.724.
$$

Holding department constant, the estimated odds of admission for female applicants are about 0.97 times the odds for male applicants. This is very close to 1, and the confidence interval includes 1. The adjusted model therefore does not provide evidence that female and male applicants had different admission odds after accounting for department.

::: {.callout-warning}
## Do Not Treat Department as Continuous
The values 1, 2, 3, 4, and 5 are department codes. A model with `+ department` would assume a constant log-odds change from department 1 to 2, 2 to 3, and so on. That is not meaningful for nominal department labels.
:::

```{r}
#| label: hw08-p3-check
#| include: false

or_key <- adjusted_or_table |>
  select(Parameter, Odds_ratio, SE, CI_low, CI_high, z) |>
  tibble::column_to_rownames("Parameter")

stopifnot(abs(or_key["Female vs male", "Odds_ratio"] - 0.9697744) < 1e-7)
stopifnot(abs(or_key["Female vs male", "SE"] - 0.0841372) < 1e-7)
stopifnot(abs(or_key["Female vs male", "CI_low"] - 0.8181272) < 1e-7)
stopifnot(abs(or_key["Female vs male", "CI_high"] - 1.149531) < 1e-6)
stopifnot(abs(or_key["Female vs male", "z"] + 0.3537573) < 1e-7)
stopifnot(abs(or_key["Dept 2 vs Dept 1", "Odds_ratio"] - 0.3197945) < 1e-7)
stopifnot(abs(or_key["Dept 5 vs Dept 1", "Odds_ratio"] - 0.0405479) < 1e-7)
stopifnot(nobs(model_adjusted) == 3593)
stopifnot(abs(as.numeric(logLik(model_adjusted)) + 1987.0756) < 1e-4)
stopifnot(abs(lr_chisq - 536.9016) < 1e-4)
stopifnot(abs(adjusted_fit$pseudo_r2 - 0.1190191) < 1e-7)
```


## Problem 4

### Problem Statement

> Contrast the odds ratio from Problem 3 with the odds ratio from Problem 2 and explain why they differ so much.

### Solution

```{r}
#| label: hw08-p4-or-comparison

model_unadjusted <- glm(admitted ~ female, data = berk, family = binomial)

crude_model_or <- model_params(
  model_unadjusted,
  exponentiate = TRUE,
  ci_method = "wald"
) |>
  as_tibble() |>
  filter(Parameter == "female") |>
  transmute(
    model = "Crude: female only",
    odds_ratio = Coefficient,
    CI_low,
    CI_high,
    p = format_p(p)
  )

adjusted_female_or <- adjusted_or_table |>
  filter(Parameter == "Female vs male") |>
  transmute(
    model = "Adjusted: female + department",
    odds_ratio = Odds_ratio,
    CI_low,
    CI_high,
    p
  )

or_comparison <- bind_rows(crude_model_or, adjusted_female_or)

or_comparison |>
  round_numeric(digits = 4)
```

For a compact side-by-side report table, `modelsummary::modelsummary()` can show the crude and adjusted female odds ratios with one model list:

```{r}
#| label: hw08-p4-modelsummary-comparison
#| message: false
#| warning: false
#| results: asis

modelsummary::modelsummary(
  list("Crude" = model_unadjusted, "Adjusted" = model_adjusted),
  exponentiate = TRUE,
  coef_map = c("female" = "Female vs male"),
  statistic = "conf.int",
  fmt = 3,
  gof_omit = "AIC|BIC|Log.Lik|RMSE|F|R2",
  output = "markdown"
)
```

```{r}
#| label: fig-hw08-crude-adjusted-or
#| fig-cap: "Crude and department-adjusted odds ratios for female applicants. The crude odds ratio is below 1, while the adjusted odds ratio is close to 1."
#| fig-alt: "Forest plot comparing two odds ratios for female applicants. The crude female odds ratio is about 0.64, below the null line at 1. The department-adjusted odds ratio is about 0.97 and its confidence interval crosses 1."
#| fig-width: 7
#| fig-height: 3.8
#| echo: false

ggplot(or_comparison, aes(x = odds_ratio, y = model)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey45") +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), height = 0.15) +
  geom_point(size = 3, color = ber640_colors["primary"]) +
  geom_text(
    aes(label = sprintf("OR = %.2f", odds_ratio)),
    nudge_y = 0.18,
    size = 4
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.64, 0.8, 1, 1.15),
    labels = scales::number_format(accuracy = 0.01),
    limits = c(0.5, 1.25)
  ) +
  labs(
    x = "Odds Ratio for Female Applicants (log scale)",
    y = NULL
  )
```

The crude OR is about 0.64, suggesting lower admission odds for female applicants when departments are ignored. The adjusted OR is about 0.97, suggesting nearly equal admission odds after conditioning on department.

This difference is Simpson's paradox. Department is a confounder because it is associated with both:

- **gender composition:** department 1 has a very small female share, while departments 2 and 4 have much larger female shares;
- **admission probability:** department 1 has a high admission rate, while departments 4 and 5 have low admission rates.

When departments are collapsed, applicant gender is partly standing in for department choice. The adjusted model separates the gender comparison from department differences.

```{r}
#| label: hw08-p4-department-summary

department_summary <- berk |>
  group_by(department_f) |>
  summarise(
    n = n(),
    percent_female = 100 * mean(female),
    admit_rate_male = 100 * mean(admitted[female == 0]),
    admit_rate_female = 100 * mean(admitted[female == 1]),
    admit_rate_overall = 100 * mean(admitted),
    .groups = "drop"
  )

department_summary |>
  round_numeric(digits = 1)
```

```{r}
#| label: hw08-p4-check
#| include: false

stopifnot(abs(crude_model_or$odds_ratio - crude_or) < 1e-10)
stopifnot(abs(adjusted_female_or$odds_ratio - 0.9697744) < 1e-7)
stopifnot(abs(department_summary$percent_female[department_summary$department_f == "Dept 1"] - 100 * 25 / 585) < 1e-10)
stopifnot(abs(department_summary$admit_rate_overall[department_summary$department_f == "Dept 5"] - 100 * 46 / 714) < 1e-10)
```


## Problem 5

### Problem Statement

> Conduct a Wald test for the coefficient of `female`. Interpret.

### Solution

The Wald test for a single coefficient tests:

$$
H_0: \beta_{\text{female}} = 0.
$$

For one coefficient, the Wald test can be reported as a $z$ test or as a one-degree-of-freedom chi-square test:

$$
X^2(1) = z^2.
$$

```{r}
#| label: hw08-p5-wald

female_logit <- model_params(
  model_adjusted,
  exponentiate = FALSE,
  ci_method = "wald"
) |>
  as_tibble() |>
  filter(Parameter == "female")

wald_female <- female_logit |>
  transmute(
    term = "female",
    log_odds_coefficient = Coefficient,
    SE,
    z,
    chi_square = z^2,
    df = 1,
    p = format_p(p)
  )

wald_female |>
  round_numeric(digits = 4)
```

The Wald test gives:

$$
z = -0.35,
\qquad
X^2(1) = 0.12,
\qquad
p = 0.724.
$$

We do not reject the null hypothesis that the adjusted log-odds coefficient for `female` is zero. This is the hypothesis-test version of the odds-ratio interpretation from Problem 3.

```{r}
#| label: hw08-p5-check
#| include: false

stopifnot(abs(wald_female$z + 0.3537573) < 1e-7)
stopifnot(abs(wald_female$chi_square - 0.1251442) < 1e-7)
stopifnot(female_logit$p > 0.72 & female_logit$p < 0.73)
```


## Problem 6

### Problem Statement

> Test the goodness of fit of the model. Interpret.

### Solution

Because both predictors are categorical, the data have 10 covariate patterns: 5 departments crossed with 2 gender groups. Stata's `estat gof, table` computes observed and expected admissions within those 10 patterns.

Within each gender-department pattern, every applicant has the same fitted probability because the adjusted model includes only `female` and `factor(department)`. The code below therefore uses `first(predicted_probability)` to keep one fitted probability per pattern.

```{r}
#| label: hw08-p6-gof-table

gof_table <- berk |>
  mutate(predicted_probability = fitted(model_adjusted)) |>
  group_by(female, department) |>
  summarise(
    predicted_probability = first(predicted_probability),
    observed_admitted = sum(admitted),
    observed_not_admitted = sum(1 - admitted),
    total = n(),
    .groups = "drop"
  ) |>
  mutate(
    expected_admitted = total * predicted_probability,
    expected_not_admitted = total * (1 - predicted_probability),
    pearson_component =
      (observed_admitted - expected_admitted)^2 / expected_admitted +
      (observed_not_admitted - expected_not_admitted)^2 / expected_not_admitted
  ) |>
  arrange(predicted_probability) |>
  mutate(group = row_number(), .before = 1)

gof_table |>
  mutate(
    gender = if_else(female == 1, "Female", "Male"),
    department_label = paste("Dept", department)
  ) |>
  select(
    group,
    department_label,
    gender,
    predicted_probability,
    observed_admitted,
    expected_admitted,
    observed_not_admitted,
    expected_not_admitted,
    total
  ) |>
  round_numeric(digits = 4) |>
  gt::gt() |>
  gt::cols_label(
    group = "Group",
    department_label = "Dept.",
    gender = "Gender",
    predicted_probability = gt::md("Predicted<br>prob."),
    observed_admitted = "Observed",
    expected_admitted = "Expected",
    observed_not_admitted = "Observed",
    expected_not_admitted = "Expected",
    total = "Total"
  ) |>
  gt::tab_spanner("Admitted", columns = c(observed_admitted, expected_admitted)) |>
  gt::tab_spanner("Not admitted", columns = c(observed_not_admitted, expected_not_admitted)) |>
  gt::fmt_number(columns = c(group, observed_admitted, observed_not_admitted, total), decimals = 0, use_seps = FALSE) |>
  gt::fmt_number(columns = c(predicted_probability, expected_admitted, expected_not_admitted), decimals = 4) |>
  gt::cols_align("left", columns = c(department_label, gender)) |>
  gt::cols_align(
    "right",
    columns = c(
      group,
      predicted_probability,
      observed_admitted,
      expected_admitted,
      observed_not_admitted,
      expected_not_admitted,
      total
    )
  ) |>
  gt::cols_width(
    group ~ gt::pct(7),
    department_label ~ gt::pct(10),
    gender ~ gt::pct(10),
    predicted_probability ~ gt::pct(15),
    observed_admitted ~ gt::pct(11),
    expected_admitted ~ gt::pct(12),
    observed_not_admitted ~ gt::pct(12),
    expected_not_admitted ~ gt::pct(13),
    total ~ gt::pct(8)
  ) |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::tab_caption("Observed and expected admissions by gender-department pattern.")
```

```{r}
#| label: hw08-p6-gof-summary

pearson_chisq <- sum(gof_table$pearson_component)
df_gof <- nrow(gof_table) - length(coef(model_adjusted))
p_gof <- pchisq(pearson_chisq, df = df_gof, lower.tail = FALSE)

gof_summary <- tibble(
  statistic = "Grouped Pearson X2",
  value = pearson_chisq,
  df = df_gof,
  p = format_p(p_gof)
)

gof_summary |>
  round_numeric(digits = 4)
```

The goodness-of-fit result is:

$$
X^2(4) = 2.56,
\qquad
p = 0.634.
$$

There is no evidence of lack of fit. In this categorical model, lack of fit would correspond to a gender-by-department interaction. The additive model with `female` plus department is adequate for these data.

::: {.callout-warning}
## Grouped, Not Individual, Pearson Residuals
For this Stata-style goodness-of-fit test, compute observed and expected counts by covariate pattern. Summing individual-level Pearson residuals is not the same test and does not reproduce `estat gof, table`.
:::

```{r}
#| label: hw08-p6-check
#| include: false

stopifnot(nrow(gof_table) == 10)
stopifnot(df_gof == 4)
stopifnot(abs(pearson_chisq - 2.5581815) < 1e-7)
stopifnot(abs(p_gof - 0.6342487) < 1e-7)
stopifnot(abs(gof_table$predicted_probability[gof_table$group == 1] - 0.0635) < 1e-4)
stopifnot(gof_table$observed_admitted[gof_table$group == 1] == 24)
stopifnot(abs(gof_table$expected_admitted[gof_table$group == 10] - 354.4) < 0.1)
```


## Summary

This homework shows Simpson's paradox in a logistic-regression setting:

1. The collapsed table gives a crude female-vs-male odds ratio of about 0.64.

2. After adjusting for department, the female odds ratio is about 0.97, with $p = 0.724$.

3. The difference occurs because department is associated with both applicant gender and admission probability.

4. The Wald test does not provide evidence against equal adjusted admission odds for female and male applicants.

5. The grouped Pearson goodness-of-fit test gives $X^2(4) = 2.56$, $p = 0.634$, so there is no evidence that the additive gender-plus-department model lacks fit.

The core lesson is practical: always compare crude and adjusted estimates when a confounder is plausible. In this example, the aggregate gender comparison is misleading because it ignores department.
