# Linear Regression Example: Proposal {#sec-linear-proposal}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 1--3** of the linear regression project using
the PIAAC Korea dataset. Think of it as a worked example: we show you *exactly*
how we wrote each section so that you can follow the same pattern for your own
proposal. Each section begins with a callout reminding you of the guideline
requirement, followed by the example text.
:::


## Part 1: Background and Research Questions {#sec-linear-ex-part1}

::: {.callout-important}
## Guideline Reminder --- Part 1
Provide a brief paragraph motivating **why** the topic matters, then state
**2--3 numbered research questions** that you will address with linear
regression. Do **not** describe the data or variables here --- save that for
Part 2. (See @sec-linear-part1 for the full expectations.)
:::

### Background

Understanding the determinants of earnings is a central question in labor
economics and education research. The **Mincer earnings function**
[@mincer1974schooling] established that years of education and work experience
are the primary predictors of individual wages. This simple framework has guided
decades of empirical work and remains a starting point for most earnings
analyses.

However, the traditional Mincer model relies on *quantity* of schooling (years
of education) and does not directly measure what individuals actually know or
can do. Measures of adult skill proficiency allow researchers to ask whether
cognitive skills add explanatory power beyond what years of schooling already
captures.

South Korea is a particularly interesting context for this question. Korea has
one of the highest rates of tertiary education attainment among OECD countries,
which means that years of education may have limited variation and reduced
predictive power compared with other settings. At the same time, the gender
wage gap in Korea is among the largest in the OECD, raising the question of
whether education and cognitive skills are associated with earnings in the same
way for men and women.

In this project, we examine how traditional human capital variables
(education, experience), cognitive skills (numeracy), and occupational
classification are associated with monthly earnings among Korean full-time
workers, and whether these associations vary by gender.

### Research Questions

1. **RQ1:** How are years of education and work experience associated with
   monthly earnings for Korean full-time workers, controlling for gender?
2. **RQ2:** Do numeracy and occupational classification improve the explanation
   of monthly earnings beyond the baseline human-capital model?
3. **RQ3:** Do the associations of education and numeracy with monthly earnings
   differ by gender?

::: {.callout-tip}
## Writing Tip
Notice how each research question maps to a specific model extension. RQ1 asks
about the baseline human-capital variables, RQ2 asks about adding occupation
and numeracy, and RQ3 asks about interaction effects. This structure makes it
easy to link each question to a specific analysis later in Part 3.
:::


## Part 2: Methods --- Subjects and Variables {#sec-linear-ex-part2}

::: {.callout-important}
## Guideline Reminder --- Part 2
Describe the **data source** and study design, the **subjects** and final
analytic sample size, the **missing-data rule**, and provide a table listing
**every variable** with its name, type, coding, and definition. For categorical
predictors, list all categories and the reference group. For transformed
variables, explain what the transformation means. (See @sec-linear-part2 for
the full expectations.)
:::

### Study Design and Data Source

We use data from the **Programme for the International Assessment of Adult
Competencies** [PIAAC; @oecd2013piaac], a large-scale cross-sectional survey
administered by the OECD in 2011--2012. PIAAC assessed adult skills in
literacy, numeracy, and problem solving in technology-rich environments across
24 countries. We analyze the Korean module of PIAAC Round 1.

### Subjects

The original PIAAC Korea sample included 6,667 Korean adults aged 16--65. The
provided analysis file is an already-filtered complete-case subset of the
Korean PIAAC Round 1 data. It contains adults from the original Korean PIAAC
sample who met the project inclusion criteria:

- **Age:** 35--54 years (prime working-age adults)
- **Employment status:** Full-time employed (excluding part-time workers and the
  self-employed)
- **Complete cases:** Complete data on all variables used in the planned linear
  regression models

The supplied analysis file has **$N$ = 1,357** rows and no missing values on
the planned analysis variables, so all three linear regression models use the
same analytic sample.

Because this course dataset is an unweighted teaching subset of PIAAC Korea,
the analysis treats the 1,357 complete cases as the study sample rather than as
a survey-weighted national estimate. The standard errors, confidence intervals,
and p-values from the planned linear regression models are model-based and do
not incorporate PIAAC's complex survey design, sampling weights, replicate
weights, or plausible values. Results should therefore be framed as sample
associations in this teaching file, not as population-level estimates for all
Korean adults or workers.

::: {.callout-tip}
## Writing Tip
Be explicit about your inclusion and exclusion criteria, even when the course
dataset has already been filtered for you. This allows the reader to judge
how far your findings may extend beyond the teaching subset and to understand any
selection issues.
:::

### Variables

Let us load the dataset, define the exact variables used in the planned models,
and create one complete-case analysis dataset.

```{r}
#| label: load-piaac
#| message: false

analysis_vars <- c(
  "ln_inc", "female", "schyrs_c", "exper_c",
  "exper2_c", "occupation", "numeracy"
)

analysis_df <- readRDS(here::here("data", "piaac_kor.rds")) |>
  mutate(occupation = fct_relevel(occupation, "Skilled"))

analysis_df <- analysis_df |>
  drop_na(all_of(analysis_vars))

tibble(
  final_n = nrow(analysis_df),
  missing_analysis_values = sum(is.na(analysis_df[analysis_vars])),
  occupation_reference = levels(analysis_df$occupation)[1]
)
```

::: {.callout-important}
## Setting the Reference Category
The supplied `piaac_kor.rds` file already stores `occupation` with `Skilled` as
the first factor level. We still relevel it explicitly so the regression
reference category is unambiguous. `Skilled` is used as the reference because it
is an interpretable comparison group for the earnings models; in this sample,
`Semi-white` is slightly more common than `Skilled`. If you create a factor from
character values, R may sort levels alphabetically, so always verify the
reference category with `levels(your_factor)` before fitting models.
:::

@tbl-linear-variables presents all variables used in the analysis, including
their roles, types, and definitions.

```{r}
#| label: tbl-linear-variables
#| tbl-cap: "Variables used in the linear regression analysis of monthly earnings in the PIAAC Korea teaching subset."

var_table <- tibble::tribble(
  ~Variable,      ~Role,       ~Type,                    ~Coding, ~Definition,
  "`ln_inc`",     "Outcome",   "Continuous",             "Natural log scale", "Natural log of monthly income as supplied in the analysis file",
  "`female`",     "Predictor", "Binary",                 "1 = female; 0 = male (ref)", "Worker gender",
  "`schyrs_c`",   "Predictor", "Continuous",             "Centered at 13.65 years", "Years of formal education, centered at the sample mean",
  "`exper_c`",    "Predictor", "Continuous",             "Centered experience in decades", "Work experience divided by 10, then centered at the sample mean",
  "`exper2_c`",   "Predictor", "Continuous",             "Squared centered experience term", "Quadratic term used to allow a curved experience-earnings association",
  "`occupation`", "Predictor", "Categorical (4 levels)", "Skilled (ref), Semi-white, Semi-blue, Elementary", "Broad occupational classification",
  "`numeracy`",   "Predictor", "Continuous",             "Approximately z-scored", "PIAAC numeracy proficiency; about 1 unit = 1 SD"
)

kable(var_table, align = "lllll")
```

Below we describe the measurement and coding decisions for each variable.

**Outcome variable.** The outcome is the natural logarithm of monthly income
(`ln_inc`) as supplied in the analysis file. Log-income is useful for earnings
models because income distributions are often right-skewed, and coefficients in
a log-income model have a convenient percentage-change interpretation: the
exact percent difference is $100 \times [\exp(\beta) - 1]$, holding other
variables constant. For small coefficients, $100 \times \beta$ is a close
approximation.

**Gender.** The variable `female` is a binary indicator coded 1 for female
workers and 0 for male workers. Male is the reference category.

**Years of education (mean-centered).** The variable `schyrs_c` is years of
formal education, centered at the sample mean of 13.65 years. Mean-centering
serves two purposes: (a) it makes the intercept interpretable as the predicted
log-income for a person with *average* education (and zero on all other centered
predictors), and (b) it reduces collinearity between the main effect and any
interaction terms involving this variable.

**Work experience (mean-centered, in decades).** The variable `exper_c` is
years of work experience divided by 10 and then centered at the sample mean of
16.46 years.
Rescaling by 10 means that a one-unit change in `exper_c` corresponds to a
change of 10 years of experience, producing coefficients that are easier to
interpret than per-year coefficients (which tend to be very small). The squared
term `exper2_c` captures the well-known concavity in the experience--earnings
profile: earnings increase with experience but at a diminishing rate.
When making adjusted predictions, `exper_c` and `exper2_c` should be set
coherently. Because `exper2_c` is the squared centered experience term, setting
both `exper_c = 0` and `exper2_c = 0` represents average experience.

**Occupation.** The variable `occupation` is a four-level categorical predictor
representing broad occupational classifications:

- **Skilled** (reference category): Managers, professionals, and technicians
- **Semi-white**: Clerical support and service/sales workers
- **Semi-blue**: Skilled agricultural, craft, and machine operators
- **Elementary**: Elementary occupations

When `occupation` is included as a factor in `lm()`, R estimates three contrasts:
Semi-white versus Skilled, Semi-blue versus Skilled, and Elementary versus
Skilled. The supplied data also include equivalent 0/1 dummy columns
(`semiwhite`, `semiblue`, `elementary`), but the main worked example uses the
factor directly so R creates the contrasts automatically.

**Numeracy (z-scored).** The variable `numeracy` is a course-supplied PIAAC
numeracy proficiency measure, approximately standardized to have a mean of 0
and a standard deviation of 1 in the analysis sample. A one-unit increase in
`numeracy` thus represents about a one-standard-deviation increase in numeracy
proficiency. The worked example does not use the full set of PIAAC plausible
values, so proficiency-measurement uncertainty is not incorporated.

```{r}
#| label: check-variable-summary
#| message: false

analysis_df |>
  summarise(
    final_n = n(),
    missing_analysis_values = sum(is.na(analysis_df[analysis_vars])),
    ln_income_mean = mean(ln_inc),
    ln_income_sd = sd(ln_inc),
    numeracy_mean = mean(numeracy),
    numeracy_sd = sd(numeracy)
  )

analysis_df |>
  count(occupation, name = "n")
```

::: {.callout-tip}
## Writing Tip
Mean-centering and z-scoring are not required for obtaining correct regression
estimates --- the slope coefficients for uncentered and centered predictors are
identical. The advantage is purely about **interpretability**: centered
predictors give the intercept a meaningful interpretation, and z-scored
predictors allow you to compare effect sizes across variables measured on
different scales.
:::


## Part 3: Methods --- Statistical Analysis {#sec-linear-ex-part3}

::: {.callout-important}
## Guideline Reminder --- Part 3
Write out your regression model(s) using proper notation. If you use a model
comparison strategy, describe each model, explain which research question each
model addresses, and state exactly which models will be compared. State your
significance level and your planned diagnostic checks. (See @sec-linear-part3
for the full expectations.)
:::

### Analytic Strategy

We estimate a series of **three nested linear regression models**, each building
on the previous one by adding predictors. This nested approach allows us to
quantify how much additional variance each block of predictors explains and to
address each research question in turn. All three models are fit to the same
complete-case analysis dataset (`analysis_df`, $N$ = 1,357), so differences in
model fit are due to model specification rather than changes in the analytic
sample.

All analyses are conducted in R (version 4.4+). We set the significance level
at $\alpha = 0.05$ for all statistical tests. These tests use ordinary
unweighted regression for instructional purposes; they are not
survey-design-based tests for national population inference.

### Model 1: Traditional Mincer Model (RQ1) {#sec-linear-ex-model1}

The first model estimates a baseline human-capital earnings model, regressing
log-income on gender, years of education, and work experience. We include
`exper2_c` because the experience-earnings association is expected to be curved:
earnings may rise with experience but at a diminishing rate.

$$
\begin{aligned}
\text{ln\_inc}_i ={}& \beta_0 + \beta_1 \, \text{female}_i +
\beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i \\
&+ \beta_4 \, \text{exper2\_c}_i + \varepsilon_i
\end{aligned}
$$ {#eq-model1}

where $\varepsilon_i \sim N(0, \sigma^2)$.

This model addresses **RQ1** by estimating the associations of education
($\beta_2$) and experience ($\beta_3$, $\beta_4$) with log-income, while
controlling for gender ($\beta_1$). The intercept $\beta_0$ represents the predicted
log-income for a *male* worker with *average* education and *average*
experience (because both predictors are mean-centered).

### Model 2: Adding Cognitive Skills and Occupation (RQ2) {#sec-linear-ex-model2}

The second model extends Model 1 by adding occupational classification and
numeracy proficiency. In the equation below, `semiwhite`, `semiblue`, and
`elementary` represent the factor contrasts for `occupation`, with `Skilled` as
the reference group.

$$
\begin{aligned}
\text{ln\_inc}_i ={}& \beta_0 + \beta_1 \, \text{female}_i +
\beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i +
\beta_4 \, \text{exper2\_c}_i \\
&+ \beta_5 \, \text{semiwhite}_i +
\beta_6 \, \text{semiblue}_i +
\beta_7 \, \text{elementary}_i +
\beta_8 \, \text{numeracy}_i + \varepsilon_i
\end{aligned}
$$ {#eq-model2}

This model addresses **RQ2**. Comparing the $R^2$ values of Model 1 and Model 2
tells us how much additional variance in log-income is explained by occupation
and numeracy, above and beyond the baseline human-capital variables. We also
test whether the block of new predictors is statistically significant using an
incremental $F$-test (i.e., `anova(model1, model2)`).

Key coefficients of interest:

- $\beta_8$ (numeracy): the expected difference in log-income associated with
  about a one-SD increase in numeracy, holding all other variables constant.
- $\beta_5$, $\beta_6$, $\beta_7$: the expected difference in log-income
  between each occupation category and the skilled reference group, holding all
  other variables constant.

### Model 3: Gender Interactions (RQ3) {#sec-linear-ex-model3}

The third model extends Model 2 by adding interaction terms between gender and
two key predictors: education and numeracy. These interaction terms are included
because **RQ3** asks whether the education-income and numeracy-income
associations differ between men and women.

$$
\begin{aligned}
\text{ln\_inc}_i ={}& \beta_0 + \beta_1 \, \text{female}_i +
\beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i +
\beta_4 \, \text{exper2\_c}_i \\
&+ \beta_5 \, \text{semiwhite}_i +
\beta_6 \, \text{semiblue}_i +
\beta_7 \, \text{elementary}_i +
\beta_8 \, \text{numeracy}_i \\
&+ \beta_9 \, (\text{female}_i \times \text{schyrs\_c}_i) +
\beta_{10} \, (\text{female}_i \times \text{numeracy}_i) +
\varepsilon_i
\end{aligned}
$$ {#eq-model3}

This model addresses **RQ3**. The interaction coefficients tell us whether the
associations of education ($\beta_9$) and numeracy ($\beta_{10}$) with
log-income differ between men and women:

- $\beta_9 > 0$ would indicate that the education-income association is more
  positive for women than for men.
- $\beta_{10} > 0$ would indicate that the numeracy-income association is more
  positive for women than for men.

We test the significance of the interaction block using an incremental
$F$-test comparing Model 2 and Model 3.

### Summary of Model--Research Question Mapping

@tbl-model-summary summarizes how the nested model sequence maps onto the
three research questions.

| Model | Predictors Added | Research Question | Planned comparison |
|-------|------------------|-------------------|--------------------|
| Model 1 | `female`, `schyrs_c`, `exper_c`, `exper2_c` | RQ1: Education and experience associations | Interpret focal coefficients |
| Model 2 | + `occupation`, `numeracy` | RQ2: Added value of occupation and numeracy | Compare Model 1 vs. Model 2 |
| Model 3 | + `female` $\times$ `schyrs_c`, `female` $\times$ `numeracy` | RQ3: Gender differences in associations | Compare Model 2 vs. Model 3 |

: Summary of the three nested regression models. {#tbl-model-summary}

### Assumption Checks

For the planned model sequence, we will assess the standard linear regression
assumptions:

1. **Linearity and constant spread:** Residual-vs-fitted plots and the
   `check_model()` dashboard.
2. **Residual distribution:** Q--Q plots and residual histograms.
3. **Multicollinearity:** VIFs from `check_collinearity()`, using VIF > 5 as a
   conservative flag and VIF > 10 as a stronger warning.
4. **Influential observations:** Cook's distance and related influence checks,
   treating $4/n$ as a sensitive screening rule rather than a deletion rule.
5. **Independence/design:** We will discuss the cross-sectional individual-level
   design and avoid causal language.

If diagnostic problems are detected, we will describe them clearly and discuss
how they affect interpretation. Diagnostics will be reported in the Results
section, not in the proposal.

::: {.callout-tip}
## Writing Tip
Stating your assumption-checking plan in the proposal signals to the reader
that you understand regression assumptions and plan to verify them. You do not
need to report the results of assumption checks until the Results section
(Part 5).
:::


### R Setup for the Analysis

The code below shows the R commands we will use to fit the three models. The
same `analysis_df` created in Part 2 is used for every model. The results will
be reported and interpreted in @sec-linear-results.

```{r}
#| label: linear-model-setup
#| eval: false

# Model 1: baseline human-capital model (RQ1)
model1 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c,
             data = analysis_df)

# Model 2: add occupation and numeracy (RQ2)
model2 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               occupation + numeracy,
             data = analysis_df)

# Model 3: add gender interactions (RQ3)
model3 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               occupation + numeracy +
               female:schyrs_c + female:numeracy,
             data = analysis_df)

# Simple easystats-style summaries
model_params(model3)
model_fit(model3)
compare_models(model1, model2, model3, rank = FALSE)

# Incremental F-tests for the planned model comparisons
anova(model1, model2)
anova(model2, model3)
```

::: {.callout-note}
## What Comes Next
In @sec-linear-results, we will fit these three models, report the coefficient
estimates in combined tables, interpret the results in the context of each
research question, compare the nested models, and verify the regression
diagnostics.
:::
