# Linear Regression Example: Proposal {#sec-linear-proposal}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 1--3** of the linear regression project using
the PIAAC Korea dataset. Think of it as a worked example: we show you *exactly*
how we wrote each section so that you can follow the same pattern for your own
proposal. Each section begins with a callout reminding you of the guideline
requirement, followed by the example text.
:::


## Part 1: Background and Research Questions {#sec-linear-ex-part1}

::: {.callout-important}
## Guideline Reminder --- Part 1
Provide a brief paragraph motivating **why** the topic matters, then state
**2--3 numbered research questions** that you will address with linear
regression. Do **not** describe the data or variables here --- save that for
Part 2. (See @sec-linear-part1 for the full expectations.)
:::

### Background

Understanding the determinants of earnings is a central question in labor
economics and education research. The **Mincer earnings function**
[@mincer1974schooling] established that years of education and work experience
are the primary predictors of individual wages. This simple framework has guided
decades of empirical work and remains a starting point for most earnings
analyses.

However, the traditional Mincer model relies on *quantity* of schooling (years
of education) and does not directly measure what individuals actually know or
can do. The Programme for the International Assessment of Adult Competencies
(**PIAAC**), administered by the OECD, provides direct measures of cognitive
skills --- including literacy, numeracy, and problem solving --- for nationally
representative samples of adults in 24 participating countries and sub-national regions. These data allow
researchers to ask whether *skill proficiency* adds explanatory power beyond
what years of schooling already captures.

South Korea is a particularly interesting context for this question. Korea has
one of the highest rates of tertiary education attainment among OECD countries,
which means that years of education may have limited variation and reduced
predictive power compared with other settings. At the same time, the gender
wage gap in Korea is among the largest in the OECD, raising the question of
whether the returns to education and cognitive skills differ between men and
women.

In this project, we analyze a sample of Korean full-time workers to examine
how traditional human capital variables (education, experience), cognitive
skills (numeracy), and occupational classification relate to monthly earnings,
and whether these relationships vary by gender.

### Research Questions

1. **RQ1:** What are the returns to education and work experience on monthly
   earnings for Korean workers, controlling for gender?
2. **RQ2:** Do cognitive skills (numeracy) and occupational classification add
   explanatory power beyond the traditional Mincer model?
3. **RQ3:** Do the returns to education and cognitive skills differ by gender?

::: {.callout-tip}
## Writing Tip
Notice how each research question maps to a specific model extension. RQ1 asks
about the baseline Mincer variables, RQ2 asks about adding new predictors, and
RQ3 asks about interaction effects. This structure makes it easy to link each
question to a specific analysis later in Part 3.
:::


## Part 2: Methods --- Subjects and Variables {#sec-linear-ex-part2}

::: {.callout-important}
## Guideline Reminder --- Part 2
Describe the **data source** and study design, the **subjects** (who, how
sampled, sample size), and provide a table listing **every variable** with its
name, type, and definition. For categorical predictors, list all categories and
the reference group. (See @sec-linear-part2 for the full expectations.)
:::

### Study Design and Data Source

We use data from the **Programme for the International Assessment of Adult
Competencies** [PIAAC; @oecd2013piaac], a large-scale cross-sectional survey
administered by the OECD in 2011--2012. PIAAC assessed adult skills in
literacy, numeracy, and problem solving in technology-rich environments across
24 countries. We analyze the Korean module of PIAAC Round 1.

### Subjects

The original PIAAC Korea sample included 6,667 Korean adults aged 16--65. We
restrict our analysis sample using the following inclusion criteria:

- **Age:** 35--54 years (prime working-age adults)
- **Employment status:** Full-time employed (excluding part-time workers and the
  self-employed)
- **Complete cases:** No missing data on any analysis variable

After applying these restrictions, the final analytic sample includes
**$N$ = 1,357** adults.

::: {.callout-tip}
## Writing Tip
Be explicit about your inclusion and exclusion criteria. This allows the reader
to judge whether your findings generalize to a broader population and to
understand any selection issues.
:::

### Variables

Let us load the dataset and examine its structure before defining the
variables.

```{r}
#| label: load-piaac
#| message: false

# Load the PIAAC Korea analysis dataset
piaac <- readRDS(here::here("data", "piaac_kor.rds"))

# Explicitly set Skilled as the reference category for occupation
piaac <- piaac |>
  mutate(occupation = fct_relevel(occupation, "Skilled"))

# Quick overview of the data
glimpse(piaac)
```

::: {.callout-important}
## Setting the Reference Category
R orders factor levels alphabetically by default, which would make "Elementary" the reference. We explicitly set "Skilled" as the reference because it is the largest and most interpretable comparison group. The original Stata analysis used "Elementary" as the reference — both choices are valid, but they produce different coefficient signs. Always verify your reference category with `levels(your_factor)` before fitting models.
:::

@tbl-linear-variables presents all variables used in the analysis, including
their roles, types, and definitions.

```{r}
#| label: tbl-linear-variables
#| tbl-cap: "Variables used in the linear regression analysis of Korean adults' monthly earnings."

var_table <- tibble::tribble(
  ~Variable,      ~Role,        ~Type,                    ~Definition,
  "`ln_inc`",     "Outcome",    "Continuous",             "Natural log of monthly income (PPP-adjusted)",
  "`female`",     "Predictor",  "Binary",                 "1 = female, 0 = male",
  "`schyrs_c`",   "Predictor",  "Continuous",             "Mean-centered years of education",
  "`exper_c`",    "Predictor",  "Continuous",             "Mean-centered work experience (in decades)",
  "`exper2_c`",   "Predictor",  "Continuous",             "Squared term of mean-centered experience",
  "`occupation`", "Predictor",  "Categorical (4 levels)", "Skilled (ref), Semi-white, Semi-blue, Elementary",
  "`numeracy`",   "Predictor",  "Continuous",             "Z-scored numeracy proficiency (1 unit = 1 SD)"
)

kable(var_table, align = "lllll")
```

Below we describe the measurement and coding decisions for each variable.

**Outcome variable.** The outcome is the natural logarithm of monthly income
(`ln_inc`). We use the log transformation for two reasons: (a) income
distributions are typically right-skewed, and the log transformation yields a
more symmetric distribution that better satisfies the normality assumption of
linear regression; and (b) coefficients in a log-income model have a convenient
percentage-change interpretation (i.e., a one-unit change in a predictor is
associated with approximately a $100 \times \beta$% change in income).

**Gender.** The variable `female` is a binary indicator coded 1 for female
workers and 0 for male workers. Male is the reference category.

**Years of education (mean-centered).** The variable `schyrs_c` is years of
formal education, centered at the sample mean. Mean-centering serves two
purposes: (a) it makes the intercept interpretable as the predicted log-income
for a person with *average* education (and zero on all other centered
predictors), and (b) it reduces collinearity between the main effect and any
interaction terms involving this variable.

**Work experience (mean-centered, in decades).** The variable `exper_c` is
years of work experience divided by 10 and then centered at the sample mean.
Rescaling by 10 means that a one-unit change in `exper_c` corresponds to a
change of 10 years of experience, producing coefficients that are easier to
interpret than per-year coefficients (which tend to be very small). The squared
term `exper2_c` captures the well-known concavity in the experience--earnings
profile: earnings increase with experience but at a diminishing rate.

**Occupation.** The variable `occupation` is a four-level categorical predictor
representing broad occupational classifications:

- **Skilled** (reference category): Managers, professionals, and technicians
- **Semi-white**: Clerical support and service/sales workers
- **Semi-blue**: Skilled agricultural, craft, and machine operators
- **Elementary**: Elementary occupations

In the regression models, three dummy variables are created automatically
(semi-white, semi-blue, elementary), with skilled occupations as the
reference group.

**Numeracy (z-scored).** The variable `numeracy` is the PIAAC numeracy
proficiency score, standardized to have a mean of 0 and a standard deviation of
1 in the analysis sample. A one-unit increase in `numeracy` thus represents a
one-standard-deviation increase in numeracy proficiency.

```{r}
#| label: check-variable-summary
#| message: false

# Verify sample size and key variable summaries
cat("Sample size:", nrow(piaac), "\n\n")

# Outcome variable
summary(piaac$ln_inc)

# Check the distribution of the categorical predictor
if ("occupation" %in% names(piaac)) {
  cat("\nOccupation distribution:\n")
  table(piaac$occupation)
}
```

::: {.callout-tip}
## Writing Tip
Mean-centering and z-scoring are not required for obtaining correct regression
estimates --- the slope coefficients for uncentered and centered predictors are
identical. The advantage is purely about **interpretability**: centered
predictors give the intercept a meaningful interpretation, and z-scored
predictors allow you to compare effect sizes across variables measured on
different scales.
:::


## Part 3: Methods --- Statistical Analysis {#sec-linear-ex-part3}

::: {.callout-important}
## Guideline Reminder --- Part 3
Write out your regression model(s) using proper notation. If you use a model
comparison strategy (e.g., nested models), describe each model and explain which
research question each model addresses. State your significance level.
(See @sec-linear-part3 for the full expectations.)
:::

### Analytic Strategy

We estimate a series of **three nested linear regression models**, each building
on the previous one by adding predictors. This nested approach allows us to
quantify how much additional variance each block of predictors explains and to
address each research question in turn.

All analyses are conducted in R (version 4.4+). We set the significance level
at $\alpha = 0.05$ for all statistical tests.

### Model 1: Traditional Mincer Model (RQ1) {#sec-linear-ex-model1}

The first model estimates the traditional Mincer earnings function, regressing
log-income on gender, years of education, and work experience (including the
quadratic term):

$$
\text{ln\_inc}_i = \beta_0 + \beta_1 \, \text{female}_i + \beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i + \beta_4 \, \text{exper2\_c}_i + \varepsilon_i
$$ {#eq-model1}

where $\varepsilon_i \sim N(0, \sigma^2)$.

This model addresses **RQ1** by estimating the returns to education
($\beta_2$) and experience ($\beta_3$, $\beta_4$), while controlling for
gender ($\beta_1$). The intercept $\beta_0$ represents the predicted
log-income for a *male* worker with *average* education and *average*
experience (because both predictors are mean-centered).

### Model 2: Adding Cognitive Skills and Occupation (RQ2) {#sec-linear-ex-model2}

The second model extends Model 1 by adding occupational classification and
numeracy proficiency:

$$
\text{ln\_inc}_i = \beta_0 + \beta_1 \, \text{female}_i + \beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i + \beta_4 \, \text{exper2\_c}_i + \beta_5 \, \text{semiwhite}_i + \beta_6 \, \text{semiblue}_i + \beta_7 \, \text{elementary}_i + \beta_8 \, \text{numeracy}_i + \varepsilon_i
$$ {#eq-model2}

This model addresses **RQ2**. Comparing the $R^2$ values of Model 1 and
Model 2 tells us how much additional variance in log-income is explained by
occupation and numeracy, above and beyond the traditional Mincer variables. We
also test whether the block of new predictors is statistically significant
using an incremental $F$-test (i.e., `anova(model1, model2)`).

Key coefficients of interest:

- $\beta_8$ (numeracy): the expected change in log-income for a one-SD increase
  in numeracy, holding all other variables constant.
- $\beta_5$, $\beta_6$, $\beta_7$: the expected difference in log-income
  between each occupation category and the skilled reference group, holding all
  other variables constant.

### Model 3: Gender Interactions (RQ3) {#sec-linear-ex-model3}

The third model extends Model 2 by adding interaction terms between gender and
two key predictors --- education and numeracy:

$$
\text{ln\_inc}_i = \beta_0 + \cdots + \beta_9 \, (\text{female} \times \text{schyrs\_c})_i + \beta_{10} \, (\text{female} \times \text{numeracy})_i + \varepsilon_i
$$ {#eq-model3}

where "$\cdots$" represents all terms from Model 2.

This model addresses **RQ3**. The interaction coefficients tell us whether the
returns to education ($\beta_9$) and numeracy ($\beta_{10}$) differ between men
and women:

- $\beta_9 > 0$ would indicate that the returns to education are *larger* for
  women than for men.
- $\beta_{10} > 0$ would indicate that the returns to numeracy are *larger* for
  women than for men.

We test the significance of the interaction block using an incremental
$F$-test comparing Model 2 and Model 3.

### Summary of Model--Research Question Mapping

| Model | Predictors Added | Research Question |
|-------|------------------|-------------------|
| Model 1 | `female`, `schyrs_c`, `exper_c`, `exper2_c` | RQ1: Returns to education and experience |
| Model 2 | + `occupation` dummies, `numeracy` | RQ2: Added value of skills and occupation |
| Model 3 | + `female` $\times$ `schyrs_c`, `female` $\times$ `numeracy` | RQ3: Gender differences in returns |

: Summary of the three nested regression models. {#tbl-model-summary}

### Assumption Checks

For each model, we will assess the standard linear regression assumptions:

1. **Linearity:** Residual-vs-fitted plots to check for nonlinear patterns.
2. **Normality:** Q--Q plots of standardized residuals.
3. **Homoscedasticity:** Scale-location plots and the Breusch--Pagan test.
4. **Multicollinearity:** Variance inflation factors (VIF $< 5$ as a
   guideline).

If assumption violations are detected, we will note them and, where
appropriate, apply robust standard errors (HC3).

::: {.callout-tip}
## Writing Tip
Stating your assumption-checking plan in the proposal signals to the reader
that you understand regression assumptions and plan to verify them. You do not
need to report the results of assumption checks until the Results section
(Part 5).
:::


### R Setup for the Analysis

The code below shows the R commands we will use to fit the three models. We
present the code here to make the analysis fully transparent; the results will
be reported and interpreted in @sec-linear-results.

```{r}
#| label: linear-model-setup
#| eval: false

# --- Model 1: Traditional Mincer model (RQ1) ---
model1 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c,
             data = piaac)

# --- Model 2: + Occupation and numeracy (RQ2) ---
model2 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               occupation + numeracy,
             data = piaac)

# --- Model 3: + Gender interactions (RQ3) ---
model3 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               occupation + numeracy +
               female:schyrs_c + female:numeracy,
             data = piaac)

# --- Incremental F-tests ---
anova(model1, model2)   # Tests whether occupation + numeracy add explanatory power
anova(model2, model3)   # Tests whether gender interactions are significant
```

::: {.callout-note}
## What Comes Next
In @sec-linear-results, we will fit these three models, report the coefficient
estimates in a combined table, interpret the results in the context of each
research question, and verify the regression assumptions.
:::
