# Linear Regression Example: Results & Discussion {#sec-linear-results}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 4--6** of the linear regression project
(Descriptive Statistics, Results, and Discussion). This is the main analysis
chapter --- the heart of the project. Every code block runs on the PIAAC Korea
dataset and produces real output, so you can follow along step by step.
If you copy code into your own project, start from the complete script in
@sec-r-script or the appendix templates; isolated chunks may depend on earlier
objects such as `analysis_df`, fitted models, helper functions, and
`ber640_colors`.
:::


## Part 4: Descriptive Analysis {#sec-linear-ex-part4}

::: {.callout-important}
## Guideline Reminder --- Part 4
Before running any regressions, describe your data. Report the sample size,
present summary statistics for all continuous variables (mean, SD, min, max)
and frequency tables for categorical variables, and include exploratory
visualizations that preview the relationships you will model.
:::

### Sample Overview {#sec-linear-ex-sample-overview}

We begin by creating the same complete-case analysis dataset used for the
descriptives, figures, and regression models.

```{r}
#| label: load-data-results
#| message: false

analysis_vars <- c(
  "ln_inc", "female", "schyrs", "schyrs_c", "exper", "exper_c",
  "exper2_c", "occupation", "numeracy"
)

analysis_df <- readRDS(here::here("data", "piaac_kor.rds")) |>
  mutate(occupation = fct_relevel(occupation, "Skilled")) |>
  drop_na(all_of(analysis_vars))

tibble(
  final_n = nrow(analysis_df),
  analysis_variables = length(analysis_vars),
  missing_analysis_values = sum(is.na(analysis_df[analysis_vars])),
  occupation_reference = levels(analysis_df$occupation)[1]
)
```

The analytic dataset contains **N = 1,357** Korean workers and has complete
data on all variables retained for this analysis. The age and employment-status
restrictions were applied during data preparation. We use this same
`analysis_df` for every descriptive table, figure, and regression model in the
chapter.

### Continuous Variables {#sec-linear-ex-continuous}

@tbl-desc-continuous presents summary statistics for the four key continuous
variables: log-income, years of education, work experience, and numeracy.

```{r}
#| label: tbl-desc-continuous
#| tbl-cap: "Descriptive Statistics for Continuous Variables"

continuous_summary <- analysis_df |>
  summarise(
    across(
      c(ln_inc, schyrs, exper, numeracy),
      list(Mean = mean, SD = sd, Min = min, Max = max),
      .names = "{.col}__{.fn}"
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = c("Variable", ".value"),
    names_sep = "__"
  ) |>
  mutate(
    Variable = dplyr::recode(
      Variable,
      ln_inc = "Log monthly income",
      schyrs = "Years of education",
      exper = "Work experience (years)",
      numeracy = "Numeracy (z-score)"
    )
  ) |>
  round_numeric(digits = 2)

kable(continuous_summary, align = "lrrrr")
```

Several features of the continuous variable distributions are worth noting:

- **Log-income** has a mean of about 8.00 and ranges from 5.62 to 10.44. The
  distribution is approximately symmetric after the log transformation, which
  is desirable for linear regression.
- **Years of education** averages approximately 13.7 years, reflecting Korea's
  high educational attainment. The range (6--22 years) shows meaningful
  variation.
- **Work experience** averages about 16.5 years. The range (0--40 years) is
  wide enough to make the curved experience pattern worth checking.
- **Numeracy** is z-scored (mean $\approx$ 0, SD $\approx$ 1), ranging from
  about $-3.8$ to $+3.0$ standard deviations.

### Categorical Variables {#sec-linear-ex-categorical}

@tbl-desc-categorical presents the frequency distributions of gender and
occupational classification.

```{r}
#| label: tbl-desc-categorical
#| tbl-cap: "Frequency Distributions of Categorical Variables"

gender_summary <- analysis_df |>
  mutate(
    Category = factor(
      if_else(female == 1, "Female", "Male"),
      levels = c("Male", "Female")
    )
  ) |>
  count(Category, name = "n") |>
  mutate(
    Variable = "Gender",
    Percent = 100 * n / sum(n)
  )

occupation_summary <- analysis_df |>
  count(occupation, name = "n") |>
  mutate(
    Variable = "Occupation",
    Category = as.character(occupation),
    Percent = 100 * n / sum(n)
  ) |>
  select(Variable, Category, n, Percent)

bind_rows(gender_summary, occupation_summary) |>
  select(Variable, Category, n, Percent) |>
  round_numeric(digits = 1) |>
  kable(align = "llrr")
```

The sample is about 60% male and 40% female. The occupational distribution
shows that skilled occupations (managers, professionals, technicians) and
semi-white-collar workers (clerical, service/sales) are the most common
categories, together accounting for about 68% of the sample. Semi-blue-collar
workers make up about 21%, and elementary occupations represent about 10% of
the sample.


### Visualizations {#sec-linear-ex-visualizations}

#### Distribution of the Outcome Variable

@fig-hist-lninc shows the distribution of log monthly income, our outcome
variable. The log transformation produces a roughly symmetric outcome
distribution, which is appropriate for linear modeling. Residual normality is
assessed later using model diagnostics.

```{r}
#| label: fig-hist-lninc
#| fig-cap: "Distribution of Log Monthly Income"
#| fig-alt: "Histogram of log monthly income centered around 8.0 with a dashed vertical line at the sample mean."
#| fig-height: 4

ggplot(analysis_df, aes(x = ln_inc)) +
  geom_histogram(
    bins = 30, fill = ber640_colors["accent"], color = "white", alpha = 0.8
  ) +
  geom_vline(
    xintercept = mean(analysis_df$ln_inc),
    linetype = "dashed", color = ber640_colors["secondary"], linewidth = 0.8
  ) +
  labs(
    x = "Log monthly income",
    y = "Count",
    title = "Distribution of Log Monthly Income",
    subtitle = "Dashed line = sample mean"
  )
```

#### Scatterplots of Continuous Predictors

@fig-scatter-predictors displays the bivariate relationships between
log-income and each continuous predictor. Each panel includes both a linear fit
and a LOESS smooth to help preview whether a straight-line term is a reasonable
starting point.

```{r}
#| label: fig-scatter-predictors
#| fig-cap: "Scatterplots of Log Income vs. Continuous Predictors"
#| fig-alt: "Three-panel scatterplot showing log income versus education, experience, and numeracy with linear and LOESS smooths."
#| fig-height: 5

scatter_df <- analysis_df |>
  select(ln_inc, schyrs, exper, numeracy) |>
  pivot_longer(
    cols = c(schyrs, exper, numeracy),
    names_to = "predictor",
    values_to = "value"
  ) |>
  mutate(
    predictor = dplyr::recode(
      predictor,
      schyrs = "Years of education",
      exper = "Work experience (years)",
      numeracy = "Numeracy (z-score)"
    ),
    predictor = factor(
      predictor,
      levels = c("Years of education", "Work experience (years)", "Numeracy (z-score)")
    )
  )

ggplot(scatter_df, aes(x = value, y = ln_inc)) +
  geom_point(alpha = 0.16, color = ber640_colors["primary"]) +
  geom_smooth(
    method = "lm", formula = y ~ x, se = FALSE,
    color = ber640_colors["secondary"], linewidth = 0.9
  ) +
  geom_smooth(
    method = "loess", formula = y ~ x, se = FALSE,
    color = ber640_colors["accent"], linewidth = 0.9, linetype = "dashed"
  ) +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(
    x = NULL,
    y = "Log monthly income",
    caption = "Solid line = linear fit; dashed line = LOESS smooth."
  ) +
  theme(plot.caption = element_text(size = 10, hjust = 0))
```

Education and numeracy both show positive associations with log-income. The
relationships are broadly linear in the middle of the distributions, though the
LOESS smooths show mild curvature at the tails. Experience shows a positive but
concave pattern, consistent with diminishing returns at higher experience
levels.

#### Boxplots by Categorical Predictors

@fig-boxplot-categorical displays the distribution of log-income across gender
and occupational categories.

```{r}
#| label: fig-boxplot-categorical
#| fig-cap: "Log Income by Gender and Occupation"
#| fig-alt: "Two-panel boxplot showing log income by gender and occupation."
#| fig-height: 5

boxplot_df <- analysis_df |>
  mutate(
    Gender = if_else(female == 1, "Female", "Male"),
    Occupation = as.character(occupation)
  ) |>
  select(ln_inc, Gender, Occupation) |>
  pivot_longer(
    cols = c(Gender, Occupation),
    names_to = "predictor",
    values_to = "group"
  ) |>
  mutate(
    group = factor(
      group,
      levels = c("Male", "Female", "Skilled", "Semi-white", "Semi-blue", "Elementary")
    )
  )

ggplot(boxplot_df, aes(x = group, y = ln_inc)) +
  geom_boxplot(fill = ber640_colors["accent"], alpha = 0.75) +
  facet_wrap(~ predictor, scales = "free_x") +
  scale_x_discrete(labels = \(x) stringr::str_wrap(x, width = 10)) +
  labs(x = NULL, y = "Log monthly income") +
  theme(axis.text.x = element_text(size = 10))
```

Two patterns stand out:

- **Gender gap:** Male workers have noticeably higher log-income than female
  workers. The median for males is above the 75th percentile for females.
- **Occupation differences:** Skilled workers have the highest log-income and
  elementary workers have the lowest. Semi-white-collar and semi-blue-collar
  workers occupy the middle of the distribution and overlap substantially.

::: {.callout-tip}
## Writing Tip
Exploratory visualizations serve two purposes: (1) they preview the
relationships you will model, and (2) they help you check whether the
linearity assumption is reasonable *before* fitting the regression. Always
describe what you see in each plot --- do not just drop in a figure without
interpretation.
:::


## Part 5: Regression Analysis {#sec-linear-ex-part5}

::: {.callout-important}
## Guideline Reminder --- Part 5
Fit your planned models, present the results in a publication-style table,
interpret the focal results substantively, and check regression assumptions.
Link each model back to the research question it addresses. Interpret every
non-intercept coefficient in the final model, and summarize key coefficient
changes across earlier nested models.
:::

### Fit the Planned Model Sequence {#sec-linear-ex-fit-sequence}

We fit the same three nested models described in the proposal. All three models
use the same complete-case `analysis_df`, so model comparisons are based on the
same $N = 1{,}357$ rows.

```{r}
#| label: fit-linear-models

model1 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c,
             data = analysis_df)

model2 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               occupation + numeracy,
             data = analysis_df)

model3 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               occupation + numeracy +
               female:schyrs_c + female:numeracy,
             data = analysis_df)
```


### Model 1: Traditional Mincer Model (RQ1) {#sec-linear-ex-results-m1}

Model 1 regresses log-income on gender, years of education, work experience,
and the squared experience term. This model answers **RQ1**.

```{r}
#| label: params-model1

model_params(model1)
```

**Interpretation of Model 1:**

- **Gender:** The coefficient for `female` is $b = -0.400$ ($p < .001$). Among
  workers with the same education and experience values, the expected monthly
  income for women is about 33% lower than for men because
  $100 \times [\exp(-0.400) - 1] \approx -33\%$.
- **Education:** The coefficient for `schyrs_c` is $b = 0.084$ ($p < .001$).
  Each additional year of education is associated with 0.084 higher log monthly
  income, holding gender and experience constant. On the original income scale,
  the exact percent difference is $100 \times [\exp(0.084) - 1] \approx 8.8\%$;
  the shortcut $100b$ gives the close approximation of 8.4%.
- **Experience:** The coefficient for `exper_c` is positive ($b = 0.169$,
  $p < .001$), while the squared term is negative ($b = -0.073$, $p < .001$).
  This pattern is consistent with a curved association: income increases with
  experience, but the increase becomes smaller at higher experience levels.
- **Model fit:** Model 1 explains about 39.5% of the variance in log-income
  ($R^2 = .395$).


### Model 2: Adding Occupation and Numeracy (RQ2) {#sec-linear-ex-results-m2}

Model 2 extends Model 1 by adding occupation and numeracy proficiency. Because
`occupation` is a factor, R automatically compares `Semi-white`, `Semi-blue`,
and `Elementary` workers with the `Skilled` reference group. This model answers
**RQ2**.

```{r}
#| label: params-model2

model_params(model2)
```

**Interpretation of Model 2:**

- **Occupation:** Relative to skilled workers, expected monthly income is about
  13% lower for semi-white-collar workers ($b = -0.143$), about 24% lower for
  semi-blue-collar workers ($b = -0.275$), and about 31% lower for elementary
  workers ($b = -0.377$), holding the other predictors constant. All three
  differences are statistically significant at $\alpha = .05$.
- **Numeracy:** The numeracy coefficient is $b = 0.061$ ($p < .001$). A
  one-unit increase in numeracy, roughly a one-SD increase, is associated with
  0.061 higher log monthly income after controlling for education, experience,
  gender, and occupation. On the original income scale, this is
  $100 \times [\exp(0.061) - 1] \approx 6.3\%$ higher monthly income; the
  $100b$ approximation is 6.1%.
- **Education:** The education coefficient decreases from $0.084$ in Model 1 to
  $0.048$ in Model 2. This suggests that part of the baseline education-income
  association is shared with occupation and numeracy.
- **Model fit:** Model 2 explains about 43.5% of the variance in log-income
  ($R^2 = .435$), about 4 percentage points more than Model 1.

#### Incremental F-Test: Model 1 vs. Model 2

We formally test whether the four additional predictors (three occupation
factor contrasts plus numeracy) significantly improve model fit using an
incremental $F$-test.

```{r}
#| label: ftest-m1-m2

anova(model1, model2)
```

The $F$-test is highly significant ($F(4, 1348) = 23.99$, $p < .001$),
confirming that occupation and numeracy jointly add significant explanatory
power beyond the baseline human-capital variables. This answers **RQ2** in the
affirmative: occupation and numeracy improve the explanation of monthly
earnings.


### Model 3: Gender Interactions (RQ3) {#sec-linear-ex-results-m3}

Model 3 extends Model 2 by adding interaction terms between gender and two key
predictors: education and numeracy. These interaction terms test whether the
education-income and numeracy-income associations differ between men and women.
This model answers **RQ3**.

```{r}
#| label: params-model3

model_params(model3)
```

Because Model 3 includes interactions, the lower-order coefficients are
conditional. The `female` coefficient compares women with men when education
and numeracy are at their centered/zero values, experience is held constant,
and occupation is at the skilled reference category. The `schyrs_c` and
`numeracy` coefficients describe the education-income and numeracy-income
associations for men, because men are the gender reference group.

**Interpretation of the interaction terms:**

- **Female $\times$ Education ($b = -0.015$, $p = .160$):** The interaction
  is not statistically significant at $\alpha = .05$. We do not have clear
  evidence that the education-income association differs by gender.
- **Female $\times$ Numeracy ($b = -0.055$, $p = .068$):** This interaction
  is close to, but does not reach, the conventional .05 significance level.
  The estimate suggests the numeracy-income association may be smaller for
  women than for men, but the individual coefficient is not strong enough for a
  firm conclusion.
- **$R^2 = .439$:** The addition of two interaction terms produces only a
  small increase in $R^2$ (from .435 to .439).

#### Adjusted Predictions for the Gender-by-Numeracy Pattern

Interaction coefficients are easier to understand when we convert them into
predicted values. `ggeffects::predict_response()` lets us do this with one
short command. Here we hold education and experience at their centered values,
set the occupation reference category to skilled workers, and compare men and
women across selected numeracy values.

Because the model includes a squared experience term, the prediction profile
must keep `exper_c` and `exper2_c` coherent. Here, setting both to 0 represents
average experience because `exper2_c` is the squared centered experience term.

```{r}
#| label: tbl-numeracy-gender-predictions
#| tbl-cap: "Adjusted predicted log-income by gender and numeracy."

numeracy_gender_predictions <- ggeffects::predict_response(
  model3,
  terms = c("numeracy [-2,-1,0,1,2]", "female"),
  condition = list(
    schyrs_c = 0,
    exper_c = 0,
    exper2_c = 0,
    occupation = "Skilled"
  )
) |>
  as_tibble() |>
  transmute(
    numeracy = x,
    Gender = factor(
      group,
      levels = c("0", "1"),
      labels = c("Male", "Female")
    ),
    predicted_log_income = predicted,
    conf.low,
    conf.high
  )

numeracy_gender_predictions |>
  round_numeric(digits = 3) |>
  kable(align = "llrrr")
```

```{r}
#| label: fig-numeracy-gender-predictions
#| fig-cap: "Adjusted predicted log-income by numeracy and gender."
#| fig-alt: "Line plot of adjusted predicted log-income across five numeracy values for men and women. Both lines increase with numeracy, and the male line is higher than the female line across the displayed range."
#| fig-width: 8
#| fig-height: 5

ggplot(
  numeracy_gender_predictions,
  aes(x = numeracy, y = predicted_log_income, color = Gender, fill = Gender)
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c(
    "Male" = unname(ber640_colors["primary"]),
    "Female" = unname(ber640_colors["secondary"])
  )) +
  scale_fill_manual(values = c(
    "Male" = unname(ber640_colors["primary"]),
    "Female" = unname(ber640_colors["secondary"])
  )) +
  labs(
    x = "Numeracy (z-score)",
    y = "Adjusted predicted log monthly income",
    color = "Gender",
    fill = "Gender"
  )
```

@tbl-numeracy-gender-predictions and
@fig-numeracy-gender-predictions show that predicted log-income is higher for
men than for women across the displayed numeracy range. The male line rises
more steeply: among skilled workers with average education and experience,
predicted log-income increases from about 8.19 to 8.51 as numeracy moves from
$-2$ to $+2$, while the female prediction increases from about 7.88 to 7.98.
This pattern is consistent with the negative `female:numeracy` interaction.
Because the individual interaction term is only marginal by conventional
standards ($p = .068$), we describe this as suggestive rather than definitive
evidence of gender moderation.

#### Incremental F-Test: Model 2 vs. Model 3

```{r}
#| label: ftest-m2-m3

anova(model2, model3)
```

The incremental $F$-test for the two interaction terms yields
$F(2, 1346) = 5.17$, $p = .006$. Although the block is statistically
significant at $\alpha = .05$, neither individual interaction term reaches
significance on its own. This is a case where the *joint* test detects a small
combined improvement in model fit, while the individual terms do not identify a
clear single source of gender moderation. For **RQ3**, we conclude that there
is limited evidence of gender moderation, but the substantive pattern is still
dominated by the large gender level difference in earnings.


### Model Fit Summary {#sec-linear-ex-fit-summary}

`compare_models()` gives a quick easystats comparison of the planned model
sequence.

```{r}
#| label: compare-linear-models

compare_models(model1, model2, model3, rank = FALSE)
```

@tbl-linear-fit-summary summarizes the same fit statistics in a compact table.

```{r}
#| label: tbl-linear-fit-summary
#| tbl-cap: "Model fit statistics for the nested linear regression models."

fit_summary <- bind_rows(
  model_fit(model1) |>
    as_tibble() |>
    mutate(Model = "Model 1"),
  model_fit(model2) |>
    as_tibble() |>
    mutate(Model = "Model 2"),
  model_fit(model3) |>
    as_tibble() |>
    mutate(Model = "Model 3")
) |>
  select(Model, R2, R2_adjusted, RMSE, AIC, BIC) |>
  round_numeric(digits = 3)

kable(fit_summary, align = "lrrrrr")
```

Model 2 is a clear improvement over Model 1. Model 3 adds a small amount of fit
relative to Model 2, but the practical gain is modest.


### Publication-Style Regression Table {#sec-linear-ex-reg-table}

@tbl-regression-models presents all three models side by side in a
publication-ready format. For this side-by-side comparison, `modelsummary()`
is a compact way to control row labels and keep the nested models aligned.

```{r}
#| label: tbl-regression-models
#| tbl-cap: "Nested linear regression models predicting log monthly income."

model_list <- list(
  "Model 1" = model1,
  "Model 2" = model2,
  "Model 3" = model3
)

coef_map <- c(
  "(Intercept)" = "Intercept",
  "female" = "Female",
  "schyrs_c" = "Education (years, centered)",
  "exper_c" = "Experience (decades, centered)",
  "exper2_c" = "Experience squared",
  "occupationSemi-white" = "Occupation: Semi-white vs Skilled",
  "occupationSemi-blue" = "Occupation: Semi-blue vs Skilled",
  "occupationElementary" = "Occupation: Elementary vs Skilled",
  "numeracy" = "Numeracy (z-score)",
  "female:schyrs_c" = "Female x education",
  "female:numeracy" = "Female x numeracy"
)

modelsummary::modelsummary(
  model_list,
  coef_map = coef_map,
  gof_map = c("nobs", "r.squared", "adj.r.squared", "rmse"),
  statistic = c("SE = {std.error}", "95% CI [{conf.low}, {conf.high}]", "p = {p.value}"),
  fmt = 3,
  output = "markdown"
)
```

::: {.callout-tip}
## Writing Tip
A combined regression table is the standard way to present nested models. It
allows the reader to see at a glance how coefficients change as new predictors
are added. Always include confidence intervals and standard errors, and include
p-values or another clear indication of statistical evidence.
:::


### Collinearity Diagnostics {#sec-linear-ex-vif}

We check for multicollinearity using `performance::check_collinearity()` for
the full model (Model 3).

```{r}
#| label: tbl-vif
#| tbl-cap: "Variance Inflation Factors for Model 3"

collinearity_table <- performance::check_collinearity(model3, ci = FALSE) |>
  as_tibble() |>
  select(
    Term,
    VIF,
    `SE inflation factor` = SE_factor,
    Tolerance
  ) |>
  round_numeric(digits = 2)

kable(collinearity_table, align = "lrrr")
```

@tbl-vif shows that all VIF values are well below 5 (the highest is about 3.1
for `schyrs_c`), indicating that multicollinearity is not a concern in this
analysis. Even the interaction terms (`female:schyrs_c`, `female:numeracy`)
have modest VIFs, which is expected because we used mean-centered predictors.

::: {.callout-note}
## VIF Guidelines
- VIF < 5: Usually not concerning
- VIF 5--10: Conservative flag for discussion; inspect whether coefficients are
  unstable or redundant
- VIF > 10: Stronger warning; affected coefficients may be difficult to
  interpret
:::


### Residual Diagnostics {#sec-linear-ex-residuals}

We assess the regression assumptions by examining residual plots for Model 3
(the most complex model).

#### Histogram and Q--Q Plot of Residuals

```{r}
#| label: fig-residual-hist-qq
#| fig-cap: "Normality Assessment of Residuals (Model 3)"
#| fig-alt: "Two-panel diagnostic display. The histogram shows the distribution of model residuals, and the Q-Q plot compares standardized residuals with theoretical normal quantiles."
#| fig-height: 4

resid_df <- broom::augment(model3, data = analysis_df) |>
  mutate(
    case_id = row_number(),
    cook_threshold = 4 / n(),
    cook_flag = .cooksd > cook_threshold,
    large_std_resid = abs(.std.resid) > 3
  )

p_hist <- ggplot(resid_df, aes(x = .resid)) +
  geom_histogram(bins = 30, fill = ber640_colors["accent"],
                 color = "white", alpha = 0.8) +
  labs(x = "Residuals", y = "Count",
       title = "Histogram of Residuals")

p_qq <- ggplot(resid_df, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.3, color = ber640_colors["primary"]) +
  stat_qq_line(color = ber640_colors["secondary"], linewidth = 0.8) +
  labs(x = "Theoretical quantiles", y = "Standardized residuals",
       title = "Q-Q Plot")

p_hist + p_qq
```

@fig-residual-hist-qq shows that the residual histogram is approximately
symmetric and bell-shaped, with slight heavy tails. The Q--Q plot shows that
most points follow the diagonal line closely, with some deviation in the tails.
This is typical for earnings data and does not pose a serious threat to
inference given the large sample size ($N = 1{,}357$).

#### Residuals vs. Fitted Values

```{r}
#| label: fig-residual-fitted
#| fig-cap: "Residuals vs. Fitted Values (Model 3)"
#| fig-alt: "Scatterplot of residuals versus fitted values with a horizontal zero line and a LOESS smooth."
#| fig-height: 4

ggplot(resid_df, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.15, color = ber640_colors["primary"]) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = ber640_colors["secondary"]) +
  geom_smooth(method = "loess", se = FALSE,
              color = ber640_colors["accent"], linewidth = 0.8) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted Values",
    subtitle = "Checking linearity and homoscedasticity"
  )
```

@fig-residual-fitted shows no obvious systematic pattern in the residuals
against fitted values. The LOESS smooth is approximately flat around zero,
suggesting that the linearity assumption is reasonable. The spread of residuals
is roughly constant across fitted values, supporting the homoscedasticity
assumption.

#### Influence and Outlier Screening

`broom::augment()` also gives us leverage, standardized residuals, and Cook's
distance. These values help us decide whether a small number of cases might be
driving the fitted model.

```{r}
#| label: tbl-influence-summary
#| tbl-cap: "Influence and residual screening summary for Model 3."

influence_summary <- resid_df |>
  summarise(
    `Cook threshold (4/n)` = first(cook_threshold),
    `Cases above 4/n` = sum(cook_flag),
    `|Standardized residual| > 3` = sum(large_std_resid),
    `Maximum Cook's D` = max(.cooksd)
  ) |>
  round_numeric(digits = 3)

kable(influence_summary, align = "rrrr")
```

```{r}
#| label: tbl-influential-cases
#| tbl-cap: "Five cases with the largest Cook's distance values."

resid_df |>
  arrange(desc(.cooksd)) |>
  slice_head(n = 5) |>
  transmute(
    Case = case_id,
    `Fitted value` = .fitted,
    Residual = .resid,
    `Std. residual` = .std.resid,
    `Cook's D` = .cooksd
  ) |>
  round_numeric(digits = 3) |>
  kable(align = "rrrrr")
```

@tbl-influence-summary summarizes the influence screening rules, and
@tbl-influential-cases lists the five largest Cook's distance values. The
common $4/n$ rule is a sensitive screening rule, so flagged cases should be
inspected rather than automatically deleted. Here, the maximum Cook's distance
is only about 0.08, so no single case appears to dominate the regression
results. In a real final project, this is
where you would inspect the flagged cases, check for data-entry problems, and
report whether a sensitivity analysis changed the conclusions. You should not
delete observations only because they are inconvenient.

#### Heteroscedasticity and Robust-SE Sensitivity

The residual plot is the main diagnostic, but `performance` also provides a
simple heteroscedasticity check.

```{r}
#| label: test-heteroscedasticity

performance::check_heteroscedasticity(model3)
```

For this model, the check does not indicate a heteroscedasticity problem. If
you are concerned about unequal residual variance, a simple sensitivity check
is to compare the main results with HC3 robust standard errors. The table below
focuses on the interaction terms because those are the most delicate part of
RQ3.

```{r}
#| label: tbl-hc3-sensitivity
#| tbl-cap: "Conventional and HC3 robust standard errors for Model 3 interaction terms."

robust_interaction_check <- bind_rows(
  model_params(model3) |>
    as_tibble() |>
    mutate(SE_type = "Conventional"),
  model_params(model3, vcov = "HC3") |>
    as_tibble() |>
    mutate(SE_type = "HC3 robust")
) |>
  filter(Parameter %in% c("female:schyrs_c", "female:numeracy")) |>
  select(
    `SE type` = SE_type,
    Parameter,
    Coefficient,
    `Std. Error` = SE,
    `p-value` = p
  ) |>
  round_numeric(digits = 3)

kable(robust_interaction_check, align = "llrrr")
```

@tbl-hc3-sensitivity shows that the HC3 sensitivity check does not change the
substantive conclusion: the education interaction remains weak, and the
numeracy interaction remains suggestive rather than definitive.
HC3 robust SEs are used here only as a sensitivity check for possible unequal
residual variance. They do not address other threats such as omitted variables,
nonlinearity, influential observations, or survey-design features.

#### Comprehensive Diagnostics with `performance::check_model()`

For a comprehensive view, @fig-check-model shows the `performance` package's
diagnostic suite. This plot is useful as a dashboard, but the report text
should still explain the specific residual, collinearity, influence, and
heteroscedasticity checks above.

```{r}
#| label: fig-check-model
#| fig-cap: "Comprehensive Residual Diagnostics (Model 3)"
#| fig-alt: "Four-panel diagnostic suite from the performance package showing residual patterns, residual spread, Q-Q departures, and influence screening."
#| fig-height: 8
#| fig-width: 9
#| fig-dpi: 200

performance::check_model(
  model3,
  check = c("linearity", "homogeneity", "qq", "outliers")
)
```

**Overall assessment:** The regression assumptions are broadly adequate for
this worked example. The Q--Q plot shows visible tail departures, and several
observations are worth screening, but the residual-vs-fitted pattern is
reasonably stable, the heteroscedasticity check is not concerning, and no
single case appears to dominate the model. These diagnostics support using
Model 3 while reporting the limitations honestly.


## Part 6: Discussion {#sec-linear-ex-part6}

::: {.callout-important}
## Guideline Reminder --- Part 6
Summarize your key findings by directly answering each research question,
discuss practical significance, note at least 2--3 limitations, and use
association language unless the study design supports causal inference. This
is where you show that you understand what the numbers mean.
:::

### Summary of Key Findings {#sec-linear-ex-findings}

This study examined how education, experience, occupation, numeracy, and
gender are associated with monthly earnings among Korean full-time workers
using a series of nested linear regression models applied to PIAAC Korea data
($N = 1{,}357$). The results yield three main findings.

**RQ1: Education, experience, and earnings.** RQ1 is answered in the
affirmative: education is positively associated with earnings, and experience
shows a concave association, with earnings rising at a slowing rate. The
education association remains practically meaningful after adding occupation
and numeracy, but its smaller size suggests that part of the baseline
education-income association overlaps with observed skills and occupation.
Together, the traditional Mincer variables explain about 39.5% of the
variance in log-income.

**RQ2: Added value of cognitive skills and occupation.** RQ2 is also answered
in the affirmative. Occupational classification and numeracy proficiency add
statistically clear explanatory power beyond the traditional Mincer variables
($\Delta R^2 = .040$, $F(4, 1348) = 23.99$, $p < .001$). The practical gain is
moderate overall: occupation differences are large, while numeracy adds a
smaller adjusted association beyond schooling. In substantive terms, measured
numeracy contains earnings-relevant information not captured by years of
schooling alone.

**RQ3: Gender differences in education and numeracy associations.** RQ3 has a
mixed answer. The joint interaction block improves model fit statistically,
and the adjusted predictions suggest that the numeracy-income association may
be steeper for men than for women. However, the practical gain is small, the
individual interaction estimates are weak or suggestive, and the HC3
sensitivity check does not change that conclusion. There is little evidence
that the education-income association differs by gender. The clearest
gender-related finding is a persistent conditional level difference in
earnings, not definitive evidence of different education or numeracy returns.

### The Persistent Gender Gap {#sec-linear-ex-gender-gap}

One of the most striking findings is the large and persistent gender gap.
The `female` coefficient is around $-0.40$ to $-0.43$ across all three
models, corresponding to roughly a 33--35% earnings disadvantage for women
compared with men who have the same education, experience, occupation, and
cognitive skills. This pattern is consistent with the well-documented gender
wage gap in South Korea, which is among the largest in the OECD. The gap is
not accounted for by the measured human-capital variables, broad occupation
categories, or numeracy scores in this model. Possible unmeasured contributors
include discrimination, career interruptions, job tenure, firm or industry
sorting, hours, and within-occupation segregation, but this analysis cannot
distinguish among those mechanisms.

### Limitations {#sec-linear-ex-limitations}

Several limitations should be noted when interpreting these results.

1. **Cross-sectional design.** Because PIAAC is a cross-sectional survey, we
   cannot establish causal relationships. The associations between education
   and earnings, for example, may partly reflect selection effects (e.g.,
   more able individuals obtain more education *and* earn more) rather than a
   true causal return to schooling.

2. **Single-country sample.** Our analysis focuses on South Korea, which has
   distinctive features (very high educational attainment, large gender wage
   gap, specific labor market institutions). The findings may not generalize
   to other countries or cultural contexts.

3. **Teaching subset, survey design, and numeracy simplification.** This
   worked example analyzes an unweighted PIAAC Korea teaching subset as a
   complete-case analytic sample. The standard errors, confidence intervals,
   $F$-tests, and p-values are ordinary model-based quantities; they do not
   incorporate PIAAC sampling weights, replicate weights, complex-survey
   variance estimation, or the full set of plausible values for numeracy.
   Results should therefore be framed as sample associations in the teaching
   file, not as nationally representative estimates.

4. **Selection bias.** By restricting the sample to full-time workers aged
   35--54, we exclude part-time workers, the unemployed, and those outside the
   labor force. This sample selection may bias our estimates, particularly for
   the gender gap, because women are disproportionately represented among
   nonworkers and part-time workers.

5. **Self-reported earnings.** Monthly income in PIAAC is self-reported and
   may contain measurement error. The log transformation reduces the influence
   of extremely high earnings values, but it does not eliminate measurement
   error.

6. **Model diagnostics and sensitivity.** Residual diagnostics showed some
   tail departures and several cases worth screening. No single case appeared
   to dominate the model by Cook's distance, and HC3 robust standard errors
   did not change the RQ3 conclusion. However, these checks do not eliminate
   the possibility of model misspecification or unmeasured confounding.

### Model Adequacy and Confidence {#sec-linear-ex-confidence}

The diagnostics in Part 5 support using Model 3 as a reasonable summary model.
VIF values were low, residual plots did not show severe nonlinearity or
unequal spread, and no case dominated the model by Cook's distance. At the same
time, the Q--Q plot showed visible tail departures and several observations
warranted screening. This means the strongest conclusions are the broad,
stable patterns: the positive education and numeracy associations, the
occupation hierarchy, and the persistent gender level difference. The
borderline interaction pattern should be described more cautiously.

### Implications {#sec-linear-ex-implications}

Despite these limitations, the findings have practical implications for how we
think about education, skills, occupation, and gender inequality:

- The numeracy-income association, net of schooling and occupation, suggests
  that measured skills should be considered alongside years of schooling.
  However, the cross-sectional model does not show that a numeracy
  intervention would cause earnings gains.
- The persistent adjusted gender earnings gap warrants attention to structural
  barriers such as workplace practices, caregiving penalties, and
  discrimination, although this model cannot identify which mechanism is
  responsible.
- The occupational earnings hierarchy shows that occupational location is
  strongly associated with earnings. Implications for counseling, training, or
  mobility programs should be framed as hypotheses for further study rather
  than direct policy effects from this model.


## Submission Checklist {#sec-linear-ex-checklist}

::: {.callout-important}
## Parts 4--6 Submission Checklist
Before submitting your Results & Discussion write-up, confirm that you have
completed all of the following:

- [ ] **Descriptive statistics** for all continuous variables (mean, SD, min,
  max) and frequency tables for categorical variables
- [ ] **Exploratory visualizations** (histograms, scatterplots, boxplots)
  with written interpretations
- [ ] **Regression models** fitted and summarized (show `model_params()` output
  while learning, and include a publication-style table in the report)
- [ ] **Coefficient interpretation** for every predictor --- direction,
  magnitude, significance, and substantive meaning
- [ ] **Model comparison** using $F$-tests or $R^2$ changes to evaluate
  incremental contributions
- [ ] **Adjusted predictions or conditional-effect plots/tables** for
  interaction terms, interpreted as conditional associations; copy-and-adapt
  templates are available in @sec-template-linear-pred-table and
  @sec-template-linear-interaction-plot
- [ ] **Assumption diagnostics** including residual plots, VIF,
  influence/Cook's distance, heteroscedasticity/HC3 sensitivity when relevant,
  and a brief conclusion about model adequacy; see
  @sec-template-linear-diagnostics for templates
- [ ] **Discussion** that answers each research question directly, discusses
  practical significance, uses non-causal language for observational results,
  frames results as sample associations when appropriate, notes limitations,
  and explains how diagnostics/sensitivity checks affect confidence in the
  findings
- [ ] All code runs without errors (`eval: true` for all code chunks)
- [ ] Major report figures and tables are labeled, captioned, and
  cross-referenced in prose
:::
