# Linear Regression Example: Results & Discussion {#sec-linear-results}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 4--6** of the linear regression project
(Descriptive Statistics, Results, and Discussion). This is the main analysis
chapter --- the heart of the project. Every code block runs on the PIAAC Korea
dataset and produces real output, so you can follow along step by step.
:::


## Part 4: Descriptive Analysis {#sec-linear-ex-part4}

::: {.callout-important}
## Guideline Reminder --- Part 4
Before running any regressions, describe your data. Report the sample size,
present summary statistics for all continuous variables (mean, SD, min, max)
and frequency tables for categorical variables, and include exploratory
visualizations that preview the relationships you will model.
:::

### Sample Overview {#sec-linear-ex-sample-overview}

We begin by loading the analysis dataset and confirming the sample size.

```{r}
#| label: load-data-results
#| message: false

# Load the PIAAC Korea analysis dataset
piaac <- readRDS(here::here("data", "piaac_kor.rds"))

# Confirm sample size and check for missing data
cat("Sample size:", nrow(piaac), "\n")
cat("Variables:  ", ncol(piaac), "\n")
cat("Missing values:", sum(is.na(piaac)), "\n")
```

The analytic sample contains **N = 1,357** Korean full-time workers aged
35--54 with complete data on all analysis variables. Because we constructed a
complete-case dataset during data preparation, there are no missing values.

### Continuous Variables {#sec-linear-ex-continuous}

@tbl-desc-continuous presents summary statistics for the four key continuous
variables: log-income, years of education, work experience, and numeracy.

```{r}
#| label: tbl-desc-continuous
#| tbl-cap: "Descriptive Statistics for Continuous Variables"

piaac |>
  select(ln_inc, schyrs, exper, numeracy) |>
  tbl_summary(
    type = everything() ~ "continuous",
    statistic = all_continuous() ~ "{mean} ({sd})",
    label = list(
      ln_inc   ~ "Log monthly income",
      schyrs   ~ "Years of education",
      exper    ~ "Work experience (years)",
      numeracy ~ "Numeracy (z-score)"
    ),
    digits = everything() ~ 2
  ) |>
  bold_labels()
```

Several features of the continuous variable distributions are worth noting:

- **Log-income** has a mean of about 8.00 and ranges from 5.63 to 10.44. The
  distribution is approximately symmetric after the log transformation, which
  is desirable for linear regression.
- **Years of education** averages approximately 13.7 years, reflecting Korea's
  high educational attainment. The range (6--22 years) shows meaningful
  variation.
- **Work experience** averages about 16.5 years, as expected for a sample of
  prime working-age adults (35--54). The range (0--40 years) is wide.
- **Numeracy** is z-scored (mean $\approx$ 0, SD $\approx$ 1), ranging from
  about $-3.8$ to $+3.0$ standard deviations.

### Categorical Variables {#sec-linear-ex-categorical}

@tbl-desc-categorical presents the frequency distributions of gender and
occupational classification.

```{r}
#| label: tbl-desc-categorical
#| tbl-cap: "Frequency Distributions of Categorical Variables"

piaac |>
  select(female, occupation) |>
  mutate(
    female = factor(female, levels = c(0, 1),
                    labels = c("Male", "Female"))
  ) |>
  tbl_summary(
    label = list(
      female     ~ "Gender",
      occupation ~ "Occupation"
    )
  ) |>
  bold_labels()
```

The sample is about 60% male and 40% female. The occupational distribution
shows that skilled occupations (managers, professionals, technicians) and
semi-white-collar workers (clerical, service/sales) are the most common
categories, together accounting for about 68% of the sample. Semi-blue-collar
workers make up about 21%, and elementary occupations represent about 10% of
the sample.


### Visualizations {#sec-linear-ex-visualizations}

#### Distribution of the Outcome Variable

@fig-hist-lninc shows the distribution of log monthly income, our outcome
variable. The log transformation produces a distribution that is approximately
normal, which supports the normality assumption for the regression residuals.

```{r}
#| label: fig-hist-lninc
#| fig-cap: "Distribution of Log Monthly Income"
#| fig-alt: "Histogram of log monthly income showing an approximately normal, symmetric distribution centered around 8.0 with a dashed vertical line at the sample mean"
#| fig-height: 4

ggplot(piaac, aes(x = ln_inc)) +
  geom_histogram(
    bins = 30, fill = ber640_colors["accent"], color = "white", alpha = 0.8
  ) +
  geom_vline(
    xintercept = mean(piaac$ln_inc),
    linetype = "dashed", color = ber640_colors["secondary"], linewidth = 0.8
  ) +
  labs(
    x = "Log monthly income",
    y = "Count",
    title = "Distribution of Log Monthly Income",
    subtitle = "Dashed line = sample mean"
  )
```

#### Scatterplots of Continuous Predictors

@fig-scatter-predictors displays the bivariate relationships between
log-income and each continuous predictor. Each panel includes both a linear
fit (red) and a LOESS smooth (blue) to help detect nonlinearity.

```{r}
#| label: fig-scatter-predictors
#| fig-cap: "Scatterplots of Log Income vs. Continuous Predictors"
#| fig-alt: "Three-panel scatterplot showing log income versus education, experience, and numeracy. Education and numeracy show positive linear trends with LOESS curves confirming linearity. Experience shows a positive but slightly concave relationship where the LOESS curve bends downward at higher values, indicating diminishing returns."
#| fig-height: 5

p_schyrs <- ggplot(piaac, aes(x = as.numeric(schyrs), y = ln_inc)) +
  geom_point(alpha = 0.15, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = FALSE,
              color = ber640_colors["secondary"], linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE,
              color = ber640_colors["accent"], linewidth = 1,
              linetype = "dashed") +
  labs(x = "Years of education", y = "Log income",
       title = "Education")

p_exper <- ggplot(piaac, aes(x = as.numeric(exper), y = ln_inc)) +
  geom_point(alpha = 0.15, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = FALSE,
              color = ber640_colors["secondary"], linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE,
              color = ber640_colors["accent"], linewidth = 1,
              linetype = "dashed") +
  labs(x = "Work experience (years)", y = "Log income",
       title = "Experience")

p_numeracy <- ggplot(piaac, aes(x = numeracy, y = ln_inc)) +
  geom_point(alpha = 0.15, color = ber640_colors["primary"]) +
  geom_smooth(method = "lm", se = FALSE,
              color = ber640_colors["secondary"], linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE,
              color = ber640_colors["accent"], linewidth = 1,
              linetype = "dashed") +
  labs(x = "Numeracy (z-score)", y = "Log income",
       title = "Numeracy")

p_schyrs + p_exper + p_numeracy +
  plot_annotation(
    subtitle = "Red = linear fit; Blue dashed = LOESS smooth"
  )
```

Key observations from the scatterplots:

- **Education** shows a clear positive linear relationship with log-income.
  The LOESS curve closely tracks the linear fit, suggesting linearity is
  reasonable.
- **Experience** shows a positive but slightly curved relationship. The LOESS
  smooth bends downward at higher experience levels, consistent with the
  diminishing-returns pattern that justifies including the squared term
  (`exper2_c`) in our models.
- **Numeracy** shows a moderate positive association with log-income. The
  relationship appears approximately linear.

#### Boxplots by Categorical Predictors

@fig-boxplot-categorical displays the distribution of log-income across gender
and occupational categories.

```{r}
#| label: fig-boxplot-categorical
#| fig-cap: "Log Income by Gender and Occupation"
#| fig-alt: "Side-by-side boxplots in two panels. The gender panel shows males have a noticeably higher median log income than females, with the male median above the female 75th percentile. The occupation panel shows a clear earnings gradient: skilled workers earn the most, followed by semi-white-collar, semi-blue-collar, and elementary workers."
#| fig-height: 5

p_gender <- piaac |>
  mutate(Gender = factor(female, levels = c(0, 1),
                         labels = c("Male", "Female"))) |>
  ggplot(aes(x = Gender, y = ln_inc, fill = Gender)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(ber640_colors["primary"],
                               ber640_colors["accent"])) +
  labs(x = NULL, y = "Log income", title = "Gender")

p_occupation <- ggplot(piaac, aes(x = occupation, y = ln_inc,
                                   fill = occupation)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = unname(ber640_colors[1:4])) +
  labs(x = NULL, y = "Log income", title = "Occupation") +
  theme(axis.text.x = element_text(size = 10))

p_gender + p_occupation
```

Two patterns stand out:

- **Gender gap:** Male workers have noticeably higher log-income than female
  workers. The median for males is above the 75th percentile for females.
- **Occupational hierarchy:** There is a clear earnings gradient across
  occupational categories. Skilled workers earn the most, followed by
  semi-white-collar, semi-blue-collar, and elementary workers.

::: {.callout-tip}
## Writing Tip
Exploratory visualizations serve two purposes: (1) they preview the
relationships you will model, and (2) they help you check whether the
linearity assumption is reasonable *before* fitting the regression. Always
describe what you see in each plot --- do not just drop in a figure without
interpretation.
:::


## Part 5: Regression Analysis {#sec-linear-ex-part5}

::: {.callout-important}
## Guideline Reminder --- Part 5
Fit your planned models, present the results in a publication-style table,
interpret every coefficient substantively, and check regression assumptions.
Link each model back to the research question it addresses.
:::

### Model 1: Traditional Mincer Model (RQ1) {#sec-linear-ex-results-m1}

We begin by fitting Model 1, which regresses log-income on gender, years of
education (mean-centered), work experience (mean-centered, in decades), and
the squared experience term.

```{r}
#| label: fit-model1

# --- Model 1: Traditional Mincer model ---
model1 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c,
             data = piaac)

summary(model1)
```

**Interpretation of Model 1:**

- **Intercept (8.20):** The predicted log-income for a *male* worker with
  *average* education and *average* experience. In original units, this
  corresponds to approximately $e^{8.20} \approx 3{,}641$ (PPP-adjusted
  monthly income units).
- **Female ($b = -0.40$, $p < .001$):** Women earn approximately
  $100 \times (e^{-0.40} - 1) \approx -33\%$ less than men with the same
  education and experience. This is a substantial gender gap.
- **Years of education ($b = 0.084$, $p < .001$):** Each additional year of
  education is associated with approximately an 8.4% increase in monthly
  income, holding other variables constant. This is consistent with typical
  Mincer estimates.
- **Experience ($b = 0.169$, $p < .001$):** An additional decade of work
  experience is associated with approximately a 16.9% increase in income.
- **Experience squared ($b = -0.073$, $p < .001$):** The negative coefficient
  confirms diminishing returns to experience --- the income gains from
  additional experience decrease at higher experience levels.
- **$R^2 = .395$:** The traditional Mincer variables explain about 39.5% of
  the variance in log-income.


### Model 2: Adding Occupation and Numeracy (RQ2) {#sec-linear-ex-results-m2}

Model 2 extends Model 1 by adding three occupation dummy variables (with
skilled occupations as the reference category) and numeracy proficiency.

```{r}
#| label: fit-model2

# --- Model 2: + Occupation and numeracy ---
model2 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               semiwhite + semiblue + elementary + numeracy,
             data = piaac)

summary(model2)
```

**Interpretation of Model 2:**

- **Occupation effects** (all relative to skilled workers):
  - *Semi-white* ($b = -0.14$, $p < .001$): Semi-white-collar workers earn
    about 13% less than skilled workers.
  - *Semi-blue* ($b = -0.28$, $p < .001$): Semi-blue-collar workers earn
    about 24% less than skilled workers.
  - *Elementary* ($b = -0.38$, $p < .001$): Elementary workers earn about 31%
    less than skilled workers.
  - The clear gradient across occupation categories confirms that occupational
    classification captures meaningful earnings differences.
- **Numeracy ($b = 0.061$, $p < .001$):** A one-standard-deviation increase
  in numeracy proficiency is associated with approximately a 6.1% increase in
  income, even after controlling for education, experience, and occupation.
  This means that *what workers know* (cognitive skills) matters for earnings
  above and beyond *how long they went to school*.
- **Education ($b = 0.048$):** Note that the education coefficient decreased
  from 0.084 in Model 1 to 0.048 in Model 2. This reduction occurs because
  some of the education effect in Model 1 was actually operating through
  occupational sorting and cognitive skills. With those pathways controlled,
  the direct return to an additional year of education is about 4.8%.
- **$R^2 = .435$:** Model 2 explains about 43.5% of the variance in
  log-income, a meaningful increase of 4 percentage points over Model 1.

#### Incremental F-Test: Model 1 vs. Model 2

We formally test whether the four additional predictors (three occupation
dummies + numeracy) significantly improve model fit using an incremental
$F$-test.

```{r}
#| label: ftest-m1-m2

anova(model1, model2)
```

The $F$-test is highly significant ($F(4, 1348) = 23.99$, $p < .001$),
confirming that occupation and numeracy jointly add significant explanatory
power beyond the traditional Mincer variables. This answers **RQ2** in the
affirmative.


### Model 3: Gender Interactions (RQ3) {#sec-linear-ex-results-m3}

Model 3 extends Model 2 by adding interaction terms between gender and two key
predictors: education and numeracy. These interaction terms test whether the
*returns* to education and cognitive skills differ between men and women.

```{r}
#| label: fit-model3

# --- Model 3: + Gender interactions ---
model3 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               semiwhite + semiblue + elementary + numeracy +
               fem_schyrs_c + fem_numeracy,
             data = piaac)

summary(model3)
```

**Interpretation of the interaction terms:**

- **Female $\times$ Education ($b = -0.015$, $p = .160$):** The interaction
  is not statistically significant at $\alpha = .05$. There is no evidence
  that the returns to education differ by gender.
- **Female $\times$ Numeracy ($b = -0.055$, $p = .068$):** This interaction
  is marginally nonsignificant. There is a suggestive pattern that the
  returns to numeracy may be somewhat smaller for women, but the evidence is
  not strong enough to conclude that the difference is statistically reliable.
- **$R^2 = .439$:** The addition of two interaction terms produces only a
  small increase in $R^2$ (from .435 to .439).

#### Incremental F-Test: Model 2 vs. Model 3

```{r}
#| label: ftest-m2-m3

anova(model2, model3)
```

The incremental $F$-test for the two interaction terms yields
$F(2, 1346) = 5.17$, $p = .006$. Although the block is statistically
significant at $\alpha = .05$, neither individual interaction term reaches
significance on its own. This is a case where the *joint* test detects a
small combined effect that the individual $t$-tests lack power to isolate.
Given the marginal nature of these results, the most conservative
interpretation --- and the one we adopt for **RQ3** --- is that the returns to
education and numeracy are broadly **similar for men and women**. The gender
gap in earnings appears to operate primarily as a level difference (the main
effect of `female`) rather than through differential returns to human capital.


### Publication-Style Regression Table {#sec-linear-ex-reg-table}

@tbl-regression-models presents all three models side by side in a
publication-ready format.

```{r}
#| label: tbl-regression-models
#| tbl-cap: "Nested Linear Regression Models Predicting Log Monthly Income"

# Create individual tbl_regression objects
t1 <- tbl_regression(
  model1,
  intercept = TRUE,
  label = list(
    female   ~ "Female",
    schyrs_c ~ "Education (years, centered)",
    exper_c  ~ "Experience (decades, centered)",
    exper2_c ~ "Experience squared"
  )
) |> bold_p(t = 0.05)

t2 <- tbl_regression(
  model2,
  intercept = TRUE,
  label = list(
    female    ~ "Female",
    schyrs_c  ~ "Education (years, centered)",
    exper_c   ~ "Experience (decades, centered)",
    exper2_c  ~ "Experience squared",
    semiwhite ~ "Semi-white-collar",
    semiblue  ~ "Semi-blue-collar",
    elementary ~ "Elementary",
    numeracy  ~ "Numeracy (z-score)"
  )
) |> bold_p(t = 0.05)

t3 <- tbl_regression(
  model3,
  intercept = TRUE,
  label = list(
    female       ~ "Female",
    schyrs_c     ~ "Education (years, centered)",
    exper_c      ~ "Experience (decades, centered)",
    exper2_c     ~ "Experience squared",
    semiwhite    ~ "Semi-white-collar",
    semiblue     ~ "Semi-blue-collar",
    elementary   ~ "Elementary",
    numeracy     ~ "Numeracy (z-score)",
    fem_schyrs_c ~ "Female x Education",
    fem_numeracy ~ "Female x Numeracy"
  )
) |> bold_p(t = 0.05)

# Merge into one table
tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
)
```

::: {.callout-tip}
## Writing Tip
A combined regression table is the standard way to present nested models. It
allows the reader to see at a glance how coefficients change as new predictors
are added. Always include confidence intervals (or standard errors) and
significance indicators.
:::


### Collinearity Diagnostics {#sec-linear-ex-vif}

We check for multicollinearity by computing variance inflation factors (VIFs)
for the full model (Model 3).

```{r}
#| label: tbl-vif
#| tbl-cap: "Variance Inflation Factors for Model 3"

car::vif(model3) |>
  as.data.frame() |>
  setNames("VIF") |>
  knitr::kable(digits = 2)
```

All VIF values are well below 5 (the highest is about 3.1 for `schyrs_c`),
indicating that multicollinearity is not a concern in this analysis. Even the
interaction terms (`fem_schyrs_c`, `fem_numeracy`) have modest VIFs, which is
expected because we used mean-centered predictors to construct the
interactions.

::: {.callout-note}
## VIF Guidelines
- VIF < 5: No concern
- VIF 5--10: Moderate concern; consider whether centering or combining
  predictors might help
- VIF > 10: Serious multicollinearity problem; the affected coefficients may
  be unstable
:::


### Residual Diagnostics {#sec-linear-ex-residuals}

We assess the regression assumptions by examining residual plots for Model 3
(the most complex model).

#### Histogram and Q--Q Plot of Residuals

```{r}
#| label: fig-residual-hist-qq
#| fig-cap: "Normality Assessment of Residuals (Model 3)"
#| fig-alt: "Two-panel diagnostic display. The histogram of residuals shows an approximately symmetric, bell-shaped distribution with slight heavy tails. The Q-Q plot shows points closely following the diagonal reference line with minor deviations in the tails, indicating approximate normality."
#| fig-height: 4

# Extract residuals
resid_df <- data.frame(
  residuals = residuals(model3),
  fitted    = fitted(model3),
  std_resid = rstandard(model3)
)

p_hist <- ggplot(resid_df, aes(x = residuals)) +
  geom_histogram(bins = 30, fill = ber640_colors["accent"],
                 color = "white", alpha = 0.8) +
  labs(x = "Residuals", y = "Count",
       title = "Histogram of Residuals")

p_qq <- ggplot(resid_df, aes(sample = std_resid)) +
  stat_qq(alpha = 0.3, color = ber640_colors["primary"]) +
  stat_qq_line(color = ber640_colors["secondary"], linewidth = 0.8) +
  labs(x = "Theoretical quantiles", y = "Standardized residuals",
       title = "Q-Q Plot")

p_hist + p_qq
```

The residual histogram is approximately symmetric and bell-shaped, with
slight heavy tails. The Q--Q plot shows that most points follow the diagonal
line closely, with some deviation in the tails. This is typical for earnings
data and does not pose a serious threat to inference given the large sample
size ($N = 1{,}357$).

#### Residuals vs. Fitted Values

```{r}
#| label: fig-residual-fitted
#| fig-cap: "Residuals vs. Fitted Values (Model 3)"
#| fig-alt: "Scatterplot of residuals versus fitted values showing a random scatter of points around the horizontal zero line with no systematic pattern. The LOESS smooth is approximately flat near zero, supporting the linearity and homoscedasticity assumptions."
#| fig-height: 4

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.15, color = ber640_colors["primary"]) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = ber640_colors["secondary"]) +
  geom_smooth(method = "loess", se = FALSE,
              color = ber640_colors["accent"], linewidth = 0.8) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs. Fitted Values",
    subtitle = "Checking linearity and homoscedasticity"
  )
```

The residuals-vs-fitted plot shows no obvious systematic pattern. The LOESS
smooth is approximately flat around zero, suggesting that the linearity
assumption is reasonable. The spread of residuals is roughly constant across
fitted values, supporting the homoscedasticity assumption.

#### Comprehensive Diagnostics with `performance::check_model()`

For a comprehensive view, we can also use the `performance` package's
diagnostic suite.

```{r}
#| label: fig-check-model
#| fig-cap: "Comprehensive Residual Diagnostics (Model 3)"
#| fig-alt: "Four-panel diagnostic suite from the performance package showing linearity check, homogeneity of variance, Q-Q plot of residuals, and outlier detection. All panels indicate that regression assumptions are reasonably well met with no evidence of serious violations."
#| fig-height: 8
#| fig-width: 9

performance::check_model(
  model3,
  check = c("linearity", "homogeneity", "qq", "outliers")
)
```

**Overall assessment:** The regression assumptions are reasonably well met.
There is no evidence of serious nonlinearity, heteroscedasticity, or
problematic outliers. The slight heavy-tailedness in the residual distribution
is minor and unlikely to affect our conclusions given the sample size.


## Part 6: Discussion {#sec-linear-ex-part6}

::: {.callout-important}
## Guideline Reminder --- Part 6
Summarize your key findings, answer each research question directly, discuss
at least 2--3 limitations, and note implications. This is where you show
that you understand what the numbers mean.
:::

### Summary of Key Findings {#sec-linear-ex-findings}

This study examined the determinants of monthly earnings among Korean
full-time workers using a series of nested linear regression models applied
to PIAAC Korea data ($N = 1{,}357$). The results yield three main findings.

**RQ1: Returns to education and experience.** Consistent with the Mincer
earnings function, both years of education and work experience are
significant predictors of log-income. Each additional year of education is
associated with approximately an 8.4% increase in monthly income in the
baseline model (Model 1), which decreases to about 4.8% after controlling
for occupation and numeracy (Model 2). Work experience shows the expected
concave (diminishing-returns) pattern: income rises with experience but at a
decreasing rate. Together, the traditional Mincer variables explain about
39.5% of the variance in log-income.

**RQ2: Added value of cognitive skills and occupation.** Occupational
classification and numeracy proficiency add significant explanatory power
beyond the traditional Mincer variables ($\Delta R^2 = .040$,
$F(4, 1348) = 23.99$, $p < .001$). The occupational dummies reveal a clear
earnings hierarchy: relative to skilled workers, semi-white-collar workers
earn about 13% less, semi-blue-collar workers earn about 24% less, and
elementary workers earn about 31% less. Importantly, numeracy proficiency is
a significant predictor even after controlling for years of education and
occupation: a one-SD increase in numeracy is associated with approximately a
6.1% increase in income. This suggests that what workers *know* matters for
earnings above and beyond how long they attended school.

**RQ3: Gender differences in returns.** The interaction terms between gender
and education ($p = .160$) and gender and numeracy ($p = .068$) are not
individually significant at $\alpha = .05$. The returns to education and
cognitive skills appear to be broadly similar for men and women. The gender
gap in earnings operates primarily as a *level* difference: women earn
approximately 33--35% less than men across all three models, and this gap
persists regardless of education, experience, occupation, or cognitive
skills.

### The Persistent Gender Gap {#sec-linear-ex-gender-gap}

One of the most striking findings is the large and persistent gender gap.
The `female` coefficient is around $-0.40$ to $-0.43$ across all three
models, corresponding to roughly a 33--35% earnings disadvantage for women
compared with men who have the same education, experience, occupation, and
cognitive skills. This finding is consistent with the well-documented gender
wage gap in South Korea, which is among the largest in the OECD. The fact
that the gap barely changes as we add controls suggests that it is not
explained by differences in human capital or occupational sorting --- it
likely reflects structural factors such as discrimination, career
interruptions, or within-occupation segregation that our model cannot capture.

### Limitations {#sec-linear-ex-limitations}

Several limitations should be noted when interpreting these results.

1. **Cross-sectional design.** Because PIAAC is a cross-sectional survey, we
   cannot establish causal relationships. The associations between education
   and earnings, for example, may partly reflect selection effects (e.g.,
   more able individuals obtain more education *and* earn more) rather than a
   true causal return to schooling.

2. **Single-country sample.** Our analysis focuses on South Korea, which has
   distinctive features (very high educational attainment, large gender wage
   gap, specific labor market institutions). The findings may not generalize
   to other countries or cultural contexts.

3. **Selection bias.** By restricting the sample to full-time workers aged
   35--54, we exclude part-time workers, the unemployed, and those outside the
   labor force. This sample selection may bias our estimates, particularly for
   the gender gap, because women are disproportionately represented among
   nonworkers and part-time workers.

4. **Self-reported earnings.** Monthly income in PIAAC is self-reported and
   may contain measurement error. However, the log transformation helps
   reduce the influence of extreme values.

### Implications {#sec-linear-ex-implications}

Despite these limitations, the findings have practical implications for
education policy and workforce development:

- The significant association between numeracy and earnings, above and beyond years of education,
  suggests that policies focused solely on increasing educational attainment
  (e.g., expanding college access) may be insufficient. Programs that develop
  *actual cognitive skills* could have meaningful returns in the labor market.
- The persistent gender earnings gap, even after extensive controls, points
  to the need for structural interventions such as anti-discrimination
  policies, family leave reforms, and workplace culture changes.
- The clear occupational earnings hierarchy highlights the importance of
  occupational counseling and training programs for workers in lower-tier
  occupations.


## Submission Checklist {#sec-linear-ex-checklist}

::: {.callout-important}
## Part 5 Submission Checklist
Before submitting your Results & Discussion write-up, confirm that you have
completed all of the following:

- [ ] **Descriptive statistics** for all continuous variables (mean, SD, min,
  max) and frequency tables for categorical variables
- [ ] **Exploratory visualizations** (histograms, scatterplots, boxplots)
  with written interpretations
- [ ] **Regression models** fitted and summarized (show `summary()` output
  or a publication-style table)
- [ ] **Coefficient interpretation** for every predictor --- direction,
  magnitude, significance, and substantive meaning
- [ ] **Model comparison** using $F$-tests or $R^2$ changes to evaluate
  incremental contributions
- [ ] **Assumption diagnostics** (residual plots, VIF) with a brief
  assessment of whether assumptions are met
- [ ] **Discussion** that answers each research question, notes limitations,
  and discusses implications
- [ ] All code runs without errors (`eval: true` for all code chunks)
- [ ] All figures and tables are labeled, captioned, and cross-referenced
:::
