# Logistic Regression Example: Proposal {#sec-logistic-proposal}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 1--3** of the logistic regression project
using the PIAAC Korea dataset. The same data and predictor variables from
Project 1 are reused, but the outcome changes to a **binary** variable:
whether a worker holds a supervisory (managerial) position. Each section
begins with a guideline callout, followed by the example text.
:::


## Part 1: Background and Research Question {#sec-logistic-ex-part1}

::: {.callout-important}
## Guideline Reminder --- Part 1
Provide a brief paragraph motivating **why** the topic matters, then state
**one focused research question** involving a binary outcome. Name the modeled
event explicitly: which outcome category is $Y = 1$? Do **not** describe the
data or variables here --- save that for Part 2. (See @sec-logistic-part1 for
the full expectations.)
:::

### Background

Beyond earnings, career advancement to supervisory positions is another
important labor market outcome. Supervisory roles are often associated with
higher wages, greater autonomy, and greater influence over organizational
decisions. Understanding which individual characteristics are associated with
supervisory status can help us think more carefully about labor-market
advancement and inequality.

In South Korea, where educational attainment is nearly universal at the
tertiary level, the traditional credential signal may have limited power to
differentiate workers for promotion into supervisory roles. Cognitive skills
measured directly --- such as numeracy --- may provide additional predictive
value beyond years of schooling. Furthermore, a persistent gender gap in
leadership positions raises the question of whether education and skill
associations with supervisory status differ between men and women.

This project examines how human capital, cognitive skills, occupation, and
gender relate to supervisory status within the supplied PIAAC Korea teaching
subset.

### Research Question

**Research question:** Among Korean full-time workers, how are gender,
education, work experience, occupational classification, and numeracy
associated with the probability of holding a supervisory position
(`manager = 1`), including whether the education or numeracy associations
differ by gender?

::: {.callout-tip}
## Writing Tip
Project 2 uses **one focused logistic research question**. The planned models
below still build in stages --- baseline model, extended model, interaction
model --- but those stages support the same overarching question rather than
creating separate numbered research questions.
:::


## Part 2: Methods --- Subjects and Variables {#sec-logistic-ex-part2}

::: {.callout-important}
## Guideline Reminder --- Part 2
Describe the **data source** and study design, the **subjects** and final
analytic sample size, the **missing-data rule**, and provide a table listing
**every variable** with its name, type, coding, and definition. For the binary
outcome, identify the event and non-event. For categorical predictors, list all
categories and the reference group. (See @sec-logistic-part2 for the full
expectations.)
:::

### Study Design and Data Source

We use the same data from the **Programme for the International Assessment of
Adult Competencies** [PIAAC; @oecd2013piaac] described in @sec-linear-ex-part2.
The PIAAC provides detailed information on adult skills and labor market
outcomes, including supervisory responsibilities.

### Subjects

The supplied analysis file is the same already-filtered Korean PIAAC subset
used in Project 1. It contains prime working-age adults (35--54) who were
full-time employees. After applying the complete-case rule to the planned
logistic outcome and predictors, the final analytic sample includes
**$N$ = 1,357** respondents in the teaching file.

As in Project 1, this is an unweighted teaching subset. The logistic models do
not use PIAAC survey weights, replicate weights, plausible values, or complex
sample design information, so inference is limited to this teaching analysis
file. Interpret odds ratios and predicted probabilities as sample associations
with supervisory status in this subset, not as nationally representative
estimates.

### Variables

Because this project reuses the same predictor variables from Project 1, we
focus on describing the **new outcome variable** and provide a complete
variable table for reference.

```{r}
#| label: logistic-load-data
#| message: false

analysis_vars <- c(
  "manager", "female", "schyrs_c", "exper_c",
  "exper2_c", "occupation", "numeracy"
)

analysis_df <- readRDS(here::here("data", "piaac_kor.rds")) |>
  mutate(occupation = fct_relevel(occupation, "Skilled")) |>
  drop_na(all_of(analysis_vars))

tibble(
  final_n = nrow(analysis_df),
  modeled_event = sum(analysis_df$manager == 1),
  non_event = sum(analysis_df$manager == 0),
  occupation_reference = levels(analysis_df$occupation)[1]
)
```

::: {.callout-note}
As in the linear regression example (@sec-linear-proposal), we set **Skilled** as the reference category for occupation. See Chapter 4 for the rationale.
:::

**Outcome variable.** The outcome is `manager`, a binary indicator of
supervisory status. The modeled event is `manager = 1`: the respondent
supervises other employees as part of their job. The non-event is
`manager = 0`: the respondent does not supervise other employees. This
variable is derived from the PIAAC item asking whether the respondent manages
or supervises other workers at their current job.

@tbl-manager-distribution shows the number and percentage of workers in the
modeled event and non-event categories.

```{r}
#| label: tbl-manager-distribution
#| tbl-cap: "Distribution of the outcome variable (manager)."

# Frequency table of the modeled event and non-event
manager_freq <- table(analysis_df$manager)
manager_pct  <- round(prop.table(manager_freq) * 100, 1)

manager_tbl <- tibble::tibble(
  `Supervisory Status` = c(
    "Non-manager (0, non-event)",
    "Manager (1, event)",
    "Total"
  ),
  n   = c(manager_freq[1], manager_freq[2], sum(manager_freq)),
  `%` = c(manager_pct[1], manager_pct[2], 100.0)
)

kable(manager_tbl, align = "lrr")
```

The sample contains **580 managers (42.7%)** and **777 non-managers (57.3%)**.
The outcome is not extremely rare, so logistic regression is an appropriate
modeling choice with no need for specialized rare-event methods.

**Predictor variables.** All predictor variables are identical to those used in
Project 1. @tbl-logistic-variables provides the complete variable table.

```{r}
#| label: tbl-logistic-variables
#| tbl-cap: "Variables used in the logistic regression analysis of supervisory status in the PIAAC Korea teaching subset."

var_table <- tibble::tribble(
  ~Variable,      ~Role,        ~Type,                    ~Definition,
  "`manager`",    "Outcome",    "Binary",                 "1 = supervises other employees (event), 0 = does not (non-event)",
  "`female`",     "Predictor",  "Binary",                 "1 = female, 0 = male (ref)",
  "`schyrs_c`",   "Predictor",  "Continuous",             "Mean-centered years of education",
  "`exper_c`",    "Predictor",  "Continuous",             "Mean-centered work experience (in decades)",
  "`exper2_c`",   "Predictor",  "Continuous",             "Squared term of mean-centered experience",
  "`occupation`", "Predictor",  "Categorical (4 levels)", "Skilled (ref), Semi-white, Semi-blue, Elementary",
  "`numeracy`",   "Predictor",  "Continuous",             "Z-scored numeracy proficiency (1 unit = 1 SD)"
)

kable(var_table, align = "llll")
```

::: {.callout-tip}
## Writing Tip
When the same predictors are used across projects, you do not need to repeat
lengthy descriptions of every variable. Instead, reference the earlier project
and focus your writing on what has changed --- in this case, the outcome
variable and the rationale for using logistic regression.
:::


## Part 3: Methods --- Statistical Analysis {#sec-logistic-ex-part3}

::: {.callout-important}
## Guideline Reminder --- Part 3
Write out your logistic regression model(s) using proper notation. If you use a
model comparison strategy (e.g., nested models), describe each model stage and
explain how it supports the single research question. State your significance
level. (See @sec-logistic-part3 for the full expectations.)
:::

### Analytic Strategy

We estimate a planned sequence of **three nested logistic regression models**
to build evidence for the single research question. Model 1 estimates a
baseline human-capital specification, Model 2 adds cognitive skill and
occupation, and Model 3 tests whether the education and numeracy associations
differ by gender. All three models use the same complete-case `analysis_df`,
so model comparisons are based on the same $N = 1{,}357$ rows.

All analyses are conducted in R (version 4.4+). We set the significance level
at $\alpha = 0.05$ for all statistical tests. Results will be presented as
**odds ratios** with 95% confidence intervals, and we will compute **predicted
probabilities** to illustrate practical significance.

Let $Y_i = 1$ if worker $i$ supervises other employees and $Y_i = 0$
otherwise. Let $p_i = P(Y_i = 1)$.

### Model 1: Baseline Human Capital Model {#sec-logistic-ex-model1}

The first model regresses supervisory status on gender, education, and work
experience. The baseline model is shown in @eq-logistic-model1.

$$
\begin{aligned}
\text{logit}(p_i) ={} & \beta_0 + \beta_1 \, \text{female}_i
  + \beta_2 \, \text{schyrs\_c}_i
  + \beta_3 \, \text{exper\_c}_i \\
& + \beta_4 \, \text{exper2\_c}_i
\end{aligned}
$$ {#eq-logistic-model1}

where $\text{logit}(p) = \ln\!\big(\frac{p}{1-p}\big)$.

This baseline stage estimates the association of gender, education, and
experience with the odds of the modeled event, `manager = 1`. The
exponentiated coefficient $e^{\beta_2}$ gives the **odds ratio** for a
one-year increase in education, holding all other variables constant. The
intercept $\beta_0$ represents the log-odds of being a manager for a male
worker with average education and average experience because both predictors
are mean-centered.

### Model 2: Adding Cognitive Skills and Occupation {#sec-logistic-ex-model2}

The second model extends Model 1 by adding occupational classification and
numeracy proficiency. The extended model is shown in @eq-logistic-model2.

For compact notation, define $D_{SW,i}$, $D_{SB,i}$, and $D_{EL,i}$ as
indicators for the semi-white, semi-blue, and elementary occupation categories,
respectively. Skilled workers are the reference group. In R, we use
`occupation` as a factor and let R create these contrasts automatically; the
course data also include equivalent 0/1 columns (`semiwhite`, `semiblue`,
`elementary`) if you want to inspect the coding.

$$
\begin{aligned}
\text{logit}(p_i) ={} & \beta_0 + \beta_1 \, \text{female}_i
  + \beta_2 \, \text{schyrs\_c}_i
  + \beta_3 \, \text{exper\_c}_i
  + \beta_4 \, \text{exper2\_c}_i \\
& + \beta_5 \, D_{SW,i}
  + \beta_6 \, D_{SB,i}
  + \beta_7 \, D_{EL,i}
  + \beta_8 \, \text{numeracy}_i
\end{aligned}
$$ {#eq-logistic-model2}

This stage tests whether occupation and numeracy add evidence about
supervisory status beyond the baseline human-capital variables. We compare
Model 1 and Model 2 using a **likelihood ratio test** (the logistic regression
analogue of the incremental $F$-test) to determine whether the additional
predictors significantly improve model fit.

Key coefficients of interest:

- $e^{\beta_8}$ (numeracy): the multiplicative change in the odds of being a
  manager for a one-SD increase in numeracy, holding all other variables
  constant.
- $e^{\beta_5}$, $e^{\beta_6}$, $e^{\beta_7}$: the odds of being a manager in
  each listed occupation category relative to the skilled reference group.

### Model 3: Gender Interactions {#sec-logistic-ex-model3}

The third model extends Model 2 by adding interaction terms between gender and
two key predictors --- education and numeracy. The interaction model is shown
in @eq-logistic-model3.

$$
\begin{aligned}
\text{logit}(p_i) ={} & \beta_0 + \beta_1 \, \text{female}_i
  + \beta_2 \, \text{schyrs\_c}_i
  + \beta_3 \, \text{exper\_c}_i
  + \beta_4 \, \text{exper2\_c}_i \\
& + \beta_5 \, D_{SW,i}
  + \beta_6 \, D_{SB,i}
  + \beta_7 \, D_{EL,i}
  + \beta_8 \, \text{numeracy}_i \\
& + \beta_9 \, (\text{female}_i \times \text{schyrs\_c}_i)
  + \beta_{10} \, (\text{female}_i \times \text{numeracy}_i)
\end{aligned}
$$ {#eq-logistic-model3}

This stage tests whether the education and numeracy associations with the
log-odds of the modeled event differ between men and women:

- $\beta_9 > 0$ would indicate that the education log-odds association is
  larger for women than for men.
- $\beta_{10} > 0$ would indicate that the numeracy log-odds association is
  larger for women than for men.

Because interaction terms in logistic regression are on the log-odds scale,
the probability-scale implications will be interpreted using predicted
probabilities in the results chapter.

We test the significance of the interaction block using a likelihood ratio
test comparing Model 2 and Model 3.

### Summary of Model Stages

@tbl-logistic-model-summary summarizes how the planned nested model stages
support the single logistic research question.

| Model | Predictors Added | Purpose for the Single Research Question |
|-------|------------------|------------------------------------------|
| Model 1 | `female`, `schyrs_c`, `exper_c`, `exper2_c` | Baseline association of gender, education, and experience with the odds of `manager = 1` |
| Model 2 | + `occupation` factor contrasts, `numeracy` | Tests whether cognitive skill and occupation add evidence about supervisory status |
| Model 3 | + `female` $\times$ `schyrs_c`, `female` $\times$ `numeracy` | Tests whether education and numeracy associations differ by gender |

: Summary of the planned logistic model stages. {#tbl-logistic-model-summary}

### Model Evaluation

For logistic regression, we will assess model quality using the following:

1. **Overall model fit:** Model $N$ and log-likelihood from `nobs()` and
   `logLik()`, plus AIC, BIC, Tjur's $R^2$, and RMSE from `model_fit()`.
2. **Model comparison:** Likelihood ratio tests comparing the planned nested
   model stages
   (analogous to the incremental $F$-test in linear regression).
3. **Sparse cells and separation:** Outcome-by-category tables, convergence
   warnings, unusually large standard errors, and fitted probabilities close to
   0 or 1.
4. **Categorical predictor block:** An overall likelihood ratio test for the
   multi-category occupation predictor.
5. **Linearity in the logit:** A simple check for the continuous predictors
   before relying on a linear logit specification.
6. **Multicollinearity:** Variance inflation factors for the full logistic
   model, using `performance::check_collinearity()`; values above 5 will be
   flagged for discussion.
7. **Influence:** Cook's distance or related influence screening, followed by
   a sensitivity check if influential observations appear to drive conclusions.

::: {.callout-tip}
## Writing Tip
Logistic regression does not have the same residual-based assumption checks as
linear regression (no normality or homoscedasticity of residuals). Instead,
focus on overall model fit statistics, the likelihood ratio test, and
predicted probabilities. Mentioning your evaluation plan in the proposal
shows methodological awareness.
:::


### R Setup for the Analysis

The code below shows the R commands we will use to fit the three models. The
results will be reported and interpreted in @sec-logistic-results.

```{r}
#| label: logistic-model-setup
#| eval: false

# --- Model 1: Baseline human capital model ---
logit1 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c,
              data = analysis_df, family = binomial)

# --- Model 2: Add occupation and numeracy ---
logit2 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                occupation + numeracy,
              data = analysis_df, family = binomial)

# --- Model 3: Add gender interactions ---
logit3 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                occupation + numeracy +
                female:schyrs_c + female:numeracy,
              data = analysis_df, family = binomial)

# --- Odds ratios, fit, and planned model comparisons ---
model_params(logit3, exponentiate = TRUE)
model_fit(logit3)
logLik(logit3)
performance::check_collinearity(logit3)
anova(logit1, logit2, test = "Chisq")   # Occupation + numeracy block
anova(logit2, logit3, test = "Chisq")   # Gender interaction block
```

::: {.callout-note}
## What Comes Next
In @sec-logistic-results, we will fit these three models, report the odds ratios
in a combined table, compute predicted probabilities for key comparisons,
interpret the results in the context of the single research question, and
evaluate model fit.
:::
