# Logistic Regression Example: Proposal {#sec-logistic-proposal}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 1--3** of the logistic regression project
using the PIAAC Korea dataset. The same data and predictor variables from
Project 1 are reused, but the outcome changes to a **binary** variable:
whether a worker holds a supervisory (managerial) position. Each section
begins with a guideline callout, followed by the example text.
:::


## Part 1: Background and Research Questions {#sec-logistic-ex-part1}

::: {.callout-important}
## Guideline Reminder --- Part 1
Provide a brief paragraph motivating **why** the topic matters, then state
**2--3 numbered research questions** that you will address with logistic
regression. Do **not** describe the data or variables here --- save that for
Part 2. (See @sec-logistic-part1 for the full expectations.)
:::

### Background

Beyond earnings, career advancement to supervisory positions is another
important labor market outcome. Managers typically earn higher wages, exercise
greater autonomy, and play a critical role in organizational decision-making.
Understanding which individual characteristics predict who becomes a manager
has implications for both human resource policy and educational investment.

In South Korea, where educational attainment is nearly universal at the
tertiary level, the traditional credential signal may have limited power to
differentiate workers for promotion into supervisory roles. Cognitive skills
measured directly --- such as numeracy --- may provide additional predictive
value beyond years of schooling. Furthermore, a persistent gender gap in
leadership positions raises the question of whether the pathways to
managerial roles differ between men and women.

In this project, we analyze the same sample of Korean full-time workers used
in Project 1 to examine how education, cognitive skills, work experience, and
occupational classification relate to the likelihood of holding a supervisory
position, and whether these relationships vary by gender.

### Research Questions

1. **RQ1:** After controlling for gender, do education and work experience
   predict the likelihood of holding a supervisory position among Korean
   workers?
2. **RQ2:** Do cognitive skills (numeracy) and occupational classification
   add predictive power beyond the baseline human capital model?
3. **RQ3:** Do the effects of education and cognitive skills on the
   probability of being a manager differ by gender?

::: {.callout-tip}
## Writing Tip
Notice how each research question parallels those in Project 1 (linear
regression). The conceptual logic is the same --- baseline model, extended
model, interaction model --- but the outcome is now binary, requiring
logistic regression instead of OLS.
:::


## Part 2: Methods --- Subjects and Variables {#sec-logistic-ex-part2}

::: {.callout-important}
## Guideline Reminder --- Part 2
Describe the **data source** and study design, the **subjects** (who, how
sampled, sample size), and provide a table listing **every variable** with its
name, type, and definition. For categorical predictors, list all categories and
the reference group. (See @sec-logistic-part2 for the full expectations.)
:::

### Study Design and Data Source

We use the same data from the **Programme for the International Assessment of
Adult Competencies** [PIAAC; @oecd2013piaac] described in @sec-linear-ex-part2.
The PIAAC provides detailed information on adult skills and labor market
outcomes, including supervisory responsibilities.

### Subjects

The analysis sample is identical to that used in Project 1. After restricting
to prime working-age adults (35--54), full-time employees, and complete cases,
the final analytic sample includes **$N$ = 1,357** Korean adults.

### Variables

Because this project reuses the same predictor variables from Project 1, we
focus on describing the **new outcome variable** and provide a complete
variable table for reference.

```{r}
#| label: logistic-load-data
#| message: false

# Load the PIAAC Korea analysis dataset (same as Project 1)
piaac <- readRDS(here::here("data", "piaac_kor.rds"))

# Ensure Skilled is the reference category (same as Ch 04)
piaac <- piaac |>
  mutate(occupation = fct_relevel(occupation, "Skilled"))
```

::: {.callout-note}
As in the linear regression example (@sec-linear-proposal), we set **Skilled** as the reference category for occupation. See Chapter 4 for the rationale.
:::

**Outcome variable.** The outcome is `manager`, a binary indicator of
supervisory status: 1 = the respondent supervises other employees as part of
their job, 0 = the respondent does not supervise others. This variable is
derived from the PIAAC item asking whether the respondent manages or supervises
other workers at their current job.

```{r}
#| label: tbl-manager-distribution
#| tbl-cap: "Distribution of the outcome variable (manager)."

# Frequency table of the outcome variable
manager_freq <- table(piaac$manager)
manager_pct  <- round(prop.table(manager_freq) * 100, 1)

manager_tbl <- tibble::tibble(
  `Supervisory Status` = c("Non-manager (0)", "Manager (1)", "Total"),
  n   = c(manager_freq[1], manager_freq[2], sum(manager_freq)),
  `%` = c(manager_pct[1], manager_pct[2], 100.0)
)

kable(manager_tbl, align = "lrr")
```

The sample contains **580 managers (42.7%)** and **777 non-managers (57.3%)**.
The outcome is not extremely rare, so logistic regression is an appropriate
modeling choice with no need for specialized rare-event methods.

**Predictor variables.** All predictor variables are identical to those used in
Project 1. @tbl-logistic-variables provides the complete variable table.

```{r}
#| label: tbl-logistic-variables
#| tbl-cap: "Variables used in the logistic regression analysis of supervisory status."

var_table <- tibble::tribble(
  ~Variable,      ~Role,        ~Type,                    ~Definition,
  "`manager`",    "Outcome",    "Binary",                 "1 = supervises other employees, 0 = does not",
  "`female`",     "Predictor",  "Binary",                 "1 = female, 0 = male (ref)",
  "`schyrs_c`",   "Predictor",  "Continuous",             "Mean-centered years of education",
  "`exper_c`",    "Predictor",  "Continuous",             "Mean-centered work experience (in decades)",
  "`exper2_c`",   "Predictor",  "Continuous",             "Squared term of mean-centered experience",
  "`occupation`", "Predictor",  "Categorical (4 levels)", "Skilled (ref), Semi-white, Semi-blue, Elementary",
  "`numeracy`",   "Predictor",  "Continuous",             "Z-scored numeracy proficiency (1 unit = 1 SD)"
)

kable(var_table, align = "lllll")
```

::: {.callout-tip}
## Writing Tip
When the same predictors are used across projects, you do not need to repeat
lengthy descriptions of every variable. Instead, reference the earlier project
and focus your writing on what has changed --- in this case, the outcome
variable and the rationale for using logistic regression.
:::


## Part 3: Methods --- Statistical Analysis {#sec-logistic-ex-part3}

::: {.callout-important}
## Guideline Reminder --- Part 3
Write out your logistic regression model(s) using proper notation. If you use a
model comparison strategy (e.g., nested models), describe each model and explain
which research question each model addresses. State your significance level.
(See @sec-logistic-part3 for the full expectations.)
:::

### Analytic Strategy

We estimate a series of **three nested logistic regression models**, mirroring
the structure used in Project 1. Each successive model adds a block of
predictors so that we can quantify the improvement in model fit and address each
research question in turn.

All analyses are conducted in R (version 4.4+). We set the significance level
at $\alpha = 0.05$ for all statistical tests. Results will be presented as
**odds ratios** with 95% confidence intervals, and we will compute **predicted
probabilities** to illustrate practical significance.

### Model 1: Baseline Human Capital Model (RQ1) {#sec-logistic-ex-model1}

The first model regresses supervisory status on gender, education, and work
experience:

$$
\text{logit}\!\big(P(\text{manager}_i = 1)\big) = \beta_0 + \beta_1 \, \text{female}_i + \beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i + \beta_4 \, \text{exper2\_c}_i
$$ {#eq-logistic-model1}

where $\text{logit}(p) = \ln\!\big(\frac{p}{1-p}\big)$.

This model addresses **RQ1**. The exponentiated coefficient $e^{\beta_2}$ gives
the **odds ratio** for a one-year increase in education, holding all other
variables constant. The intercept $\beta_0$ represents the log-odds of being a
manager for a male worker with average education and average experience (because
both predictors are mean-centered).

### Model 2: Adding Cognitive Skills and Occupation (RQ2) {#sec-logistic-ex-model2}

The second model extends Model 1 by adding occupational classification and
numeracy proficiency:

$$
\text{logit}\!\big(P(\text{manager}_i = 1)\big) = \beta_0 + \beta_1 \, \text{female}_i + \beta_2 \, \text{schyrs\_c}_i + \beta_3 \, \text{exper\_c}_i + \beta_4 \, \text{exper2\_c}_i + \beta_5 \, \text{semiwhite}_i + \beta_6 \, \text{semiblue}_i + \beta_7 \, \text{elementary}_i + \beta_8 \, \text{numeracy}_i
$$ {#eq-logistic-model2}

This model addresses **RQ2**. We compare Model 1 and Model 2 using a
**likelihood ratio test** (the logistic regression analogue of the incremental
$F$-test) to determine whether the additional predictors significantly improve
model fit.

Key coefficients of interest:

- $e^{\beta_8}$ (numeracy): the multiplicative change in the odds of being a
  manager for a one-SD increase in numeracy, holding all other variables
  constant.
- $e^{\beta_5}$, $e^{\beta_6}$, $e^{\beta_7}$: the odds of being a manager in
  each occupation category relative to the skilled reference group.

### Model 3: Gender Interactions (RQ3) {#sec-logistic-ex-model3}

The third model extends Model 2 by adding interaction terms between gender and
two key predictors --- education and numeracy:

$$
\text{logit}\!\big(P(\text{manager}_i = 1)\big) = \beta_0 + \cdots + \beta_9 \, (\text{female} \times \text{schyrs\_c})_i + \beta_{10} \, (\text{female} \times \text{numeracy})_i
$$ {#eq-logistic-model3}

where "$\cdots$" represents all terms from Model 2.

This model addresses **RQ3**. The interaction coefficients tell us whether the
effects of education and numeracy on the log-odds of being a manager differ
between men and women:

- $\beta_9 > 0$ would indicate that the effect of education on the odds of
  being a manager is *stronger* for women than for men.
- $\beta_{10} > 0$ would indicate that the effect of numeracy is *stronger*
  for women than for men.

We test the significance of the interaction block using a likelihood ratio
test comparing Model 2 and Model 3.

### Summary of Model--Research Question Mapping

| Model | Predictors Added | Research Question |
|-------|------------------|-------------------|
| Model 1 | `female`, `schyrs_c`, `exper_c`, `exper2_c` | RQ1: Education and experience effects |
| Model 2 | + `occupation` dummies, `numeracy` | RQ2: Added value of skills and occupation |
| Model 3 | + `female` $\times$ `schyrs_c`, `female` $\times$ `numeracy` | RQ3: Gender differences in effects |

: Summary of the three nested logistic regression models. {#tbl-logistic-model-summary}

### Model Evaluation

For logistic regression, we will assess model quality using the following:

1. **Overall model fit:** Likelihood ratio test for each model against the null
   (intercept-only) model.
2. **Model comparison:** Likelihood ratio tests comparing nested models
   (analogous to the incremental $F$-test in linear regression).
3. **Pseudo-$R^2$:** McFadden's $R^2$ to quantify the proportion of
   log-likelihood explained by the model.
4. **Multicollinearity:** Variance inflation factors (VIF $< 5$ as a
   guideline), computed from an equivalent linear probability model.

::: {.callout-tip}
## Writing Tip
Logistic regression does not have the same residual-based assumption checks as
linear regression (no normality or homoscedasticity of residuals). Instead,
focus on overall model fit statistics, the likelihood ratio test, and
predicted probabilities. Mentioning your evaluation plan in the proposal
shows methodological awareness.
:::


### R Setup for the Analysis

The code below shows the R commands we will use to fit the three models. The
results will be reported and interpreted in @sec-logistic-results.

```{r}
#| label: logistic-model-setup
#| eval: false

# --- Model 1: Baseline human capital model (RQ1) ---
logit1 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c,
              data = piaac, family = binomial)

# --- Model 2: + Occupation and numeracy (RQ2) ---
logit2 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                occupation + numeracy,
              data = piaac, family = binomial)

# --- Model 3: + Gender interactions (RQ3) ---
logit3 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                occupation + numeracy +
                female:schyrs_c + female:numeracy,
              data = piaac, family = binomial)

# --- Likelihood ratio tests ---
anova(logit1, logit2, test = "Chisq")   # Tests occupation + numeracy block
anova(logit2, logit3, test = "Chisq")   # Tests gender interaction block
```

::: {.callout-note}
## What Comes Next
In @sec-logistic-results, we will fit these three models, report the odds ratios
in a combined table, compute predicted probabilities for key comparisons,
interpret the results in the context of each research question, and evaluate
model fit.
:::
