# Logistic Regression Example: Results & Discussion {#sec-logistic-results}

{{< include ../_common.qmd >}}

::: {.callout-note}
## About This Chapter
This chapter demonstrates **Parts 4--6** of the logistic regression project
(Descriptive Statistics, Results, and Discussion). This is the main analysis
chapter for Project 2. Every code block runs on the PIAAC Korea dataset and
produces real output, so you can follow along step by step. The focus is on
**odds ratios**, **predicted probabilities**, and proper interpretation of
logistic regression output.
If you copy code into your own project, start from the complete script in
@sec-r-script or the appendix templates; isolated chunks may depend on earlier
objects such as `analysis_df`, fitted models, helper functions, and
`ber640_colors`.
:::


## Part 4: Descriptive Analysis {#sec-logistic-ex-part4}

::: {.callout-important}
## Guideline Reminder --- Part 4
Before running any regressions, describe your data. For logistic regression
with a binary outcome, cross-tabulations and grouped comparisons are
especially informative. Report the event and non-event distribution, check
for sparse outcome cells in categorical predictors, summarize continuous
predictors by outcome status, and include visualizations that preview the
relationships you will model.
:::

### Sample Overview {#sec-logistic-ex-sample-overview}

We begin by loading the analysis dataset and confirming the sample size and
the distribution of the modeled event, `manager = 1`.

```{r}
#| label: logistic-load-data-results
#| message: false

analysis_vars <- c(
  "manager", "female", "schyrs", "schyrs_c", "exper", "exper_c",
  "exper2_c", "occupation", "numeracy"
)

analysis_df <- readRDS(here::here("data", "piaac_kor.rds")) |>
  mutate(
    occupation = fct_relevel(occupation, "Skilled"),
    Status = factor(
      manager,
      levels = c(0, 1),
      labels = c("Non-manager (0, non-event)", "Manager (1, event)")
    ),
    Gender = factor(
      female,
      levels = c(0, 1),
      labels = c("Male", "Female")
    )
  ) |>
  drop_na(all_of(analysis_vars))

tibble(
  final_n = nrow(analysis_df),
  missing_analysis_values = sum(is.na(analysis_df[analysis_vars])),
  modeled_event = sum(analysis_df$manager == 1),
  non_event = sum(analysis_df$manager == 0),
  occupation_reference = levels(analysis_df$occupation)[1]
)
```

The analytic sample contains **N = 1,357** Korean full-time workers aged
35--54. Among them, **580 (42.7%)** hold supervisory positions and **777
(57.3%)** do not. The complete-case check reports zero missing values on the
planned analysis variables. This reasonably balanced split is well-suited for
standard logistic regression --- no rare-event corrections are needed.

@tbl-logistic-outcome presents the event and non-event distribution.

```{r}
#| label: tbl-logistic-outcome
#| tbl-cap: "Distribution of the modeled event: supervisory status."

analysis_df |>
  count(Status) |>
  mutate(`%` = round(n / sum(n) * 100, 1)) |>
  kable(align = "lrr")
```


### Cross-Tabulation: Manager Status by Gender {#sec-logistic-ex-xtab-gender}

A natural first step is to examine whether supervisory status varies by
gender. @tbl-xtab-gender presents counts and the event rate by gender.

```{r}
#| label: tbl-xtab-gender
#| tbl-cap: "Supervisory status by gender."

analysis_df |>
  group_by(Gender) |>
  summarise(
    `Non-manager n` = sum(manager == 0),
    `Manager n` = sum(manager == 1),
    Total = n(),
    `Manager %` = round(mean(manager == 1) * 100, 1),
    .groups = "drop"
  ) |>
  kable(align = "lrrrr")
```

The cross-tabulation reveals a clear gender difference: 55.5% of male workers
are managers, compared with 23.6% of female workers. This raw difference
foreshadows the gender gap we will examine in the regression analysis.

```{r}
#| label: chisq-gender

# Chi-square test: manager x female
chisq.test(table(analysis_df$manager, analysis_df$female))
```

The chi-square test confirms that the association between supervisory status
and gender is statistically significant ($p < .001$).


### Cross-Tabulation: Manager Status by Occupation {#sec-logistic-ex-xtab-occ}

@tbl-xtab-occ presents the cross-tabulation of supervisory status by
occupational classification, including the event rate within each occupation
category.

```{r}
#| label: tbl-xtab-occ
#| tbl-cap: "Supervisory status by occupational classification."

occupation_tbl <- analysis_df |>
  group_by(occupation) |>
  summarise(
    `Non-manager n` = sum(manager == 0),
    `Manager n` = sum(manager == 1),
    Total = n(),
    `Manager %` = round(mean(manager == 1) * 100, 1),
    .groups = "drop"
  )

kable(occupation_tbl, align = "lrrrr")
```

```{r}
#| label: chisq-occupation

# Chi-square test: manager x occupation
chisq.test(table(analysis_df$manager, analysis_df$occupation))
```

There is a strong association between occupation and supervisory status
($p < .001$). Skilled workers have the highest manager rate (59.8%), followed
by semi-white workers (45.8%), semi-blue workers (28.6%), and elementary
workers (8.5%). This makes intuitive sense: managerial responsibilities tend to
be concentrated in higher-status occupational categories.

The cross-tabulations also serve as a sparse-cell check. Every gender and
occupation group has both events and non-events, so there is no obvious
complete separation problem before modeling. The elementary category has only
12 managers, however, so its odds-ratio estimate should be interpreted with
some caution.


### Continuous Predictors by Manager Status {#sec-logistic-ex-continuous}

@tbl-logistic-desc-continuous presents the means of key continuous variables
(years of education, work experience, and numeracy) separately for managers
and non-managers.

```{r}
#| label: tbl-logistic-desc-continuous
#| tbl-cap: "Continuous predictors by supervisory status."

analysis_df |>
  group_by(Status) |>
  summarise(
    `Education, M (SD)` = sprintf("%.2f (%.2f)", mean(schyrs), sd(schyrs)),
    `Experience, M (SD)` = sprintf("%.2f (%.2f)", mean(exper), sd(exper)),
    `Numeracy, M (SD)` = sprintf("%.2f (%.2f)", mean(numeracy), sd(numeracy)),
    .groups = "drop"
  ) |>
  kable(align = "llll")
```

Managers have more education, more work experience, and higher numeracy scores
than non-managers in the unadjusted summaries. The education and numeracy
differences are meaningful descriptively, but the regression analysis will show
whether they remain important after gender, experience, and occupation are
included in the same model.


### Visualizations {#sec-logistic-ex-visualizations}

#### Bar Chart: Manager Status by Gender

@fig-logistic-bar-gender shows the proportion of managers and non-managers
within each gender group.

```{r}
#| label: fig-logistic-bar-gender
#| fig-cap: "Supervisory Status by Gender"
#| fig-alt: "Grouped bar chart showing the proportion of managers and non-managers by gender. About 55 percent of male workers are managers compared with about 24 percent of female workers, revealing a substantial gender gap in supervisory status."
#| fig-height: 4

analysis_df |>
  count(Gender, Status) |>
  group_by(Gender) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = Gender, y = prop, fill = Status)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(
    "Non-manager (0, non-event)" = unname(ber640_colors["primary"]),
    "Manager (1, event)" = unname(ber640_colors["accent"])
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Gender",
    y = "Proportion",
    fill = "Status",
    title = "Supervisory Status by Gender",
    subtitle = "PIAAC Korea sample (N = 1,357)"
  )
```

The bar chart clearly shows the gender gap in supervisory status: male
workers are more likely to be managers than female workers.

#### Bar Chart: Manager Status by Occupation

@fig-logistic-bar-occupation shows the manager rate by occupational
classification and previews the strong occupational gradient in supervisory
status.

```{r}
#| label: fig-logistic-bar-occupation
#| fig-cap: "Manager rate by occupational classification."
#| fig-alt: "Bar chart of the percentage of workers who are managers within each occupation category. Skilled and semi-white workers have the highest manager rates, while elementary workers have the lowest manager rate."
#| fig-height: 4

occupation_tbl |>
  ggplot(aes(x = occupation, y = `Manager %` / 100)) +
  geom_col(fill = unname(ber640_colors["secondary"]), alpha = 0.85) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Occupation",
    y = "Manager rate",
    title = "Manager Rate by Occupation",
    subtitle = "Modeled event: manager = 1"
  )
```

#### Boxplots: Continuous Predictors by Manager Status

@fig-logistic-boxplots presents the distributions of years of education,
work experience, and numeracy by supervisory status.

```{r}
#| label: fig-logistic-boxplots
#| fig-cap: "Distribution of Continuous Predictors by Supervisory Status"
#| fig-alt: "Three-panel boxplot comparing managers and non-managers on education, experience, and numeracy. Managers have higher education, more work experience, and higher numeracy scores on average, although the distributions overlap."
#| fig-height: 4.5

p_edu <- analysis_df |>
  ggplot(aes(x = Status, y = schyrs, fill = Status)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Non-manager (0, non-event)" = unname(ber640_colors["primary"]),
    "Manager (1, event)" = unname(ber640_colors["accent"])
  )) +
  labs(x = NULL, y = "Years of education", title = "Education")

p_exp <- analysis_df |>
  ggplot(aes(x = Status, y = exper, fill = Status)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Non-manager (0, non-event)" = unname(ber640_colors["primary"]),
    "Manager (1, event)" = unname(ber640_colors["accent"])
  )) +
  labs(x = NULL, y = "Experience (years)", title = "Experience")

p_num <- analysis_df |>
  ggplot(aes(x = Status, y = numeracy, fill = Status)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Non-manager (0, non-event)" = unname(ber640_colors["primary"]),
    "Manager (1, event)" = unname(ber640_colors["accent"])
  )) +
  labs(x = NULL, y = "Numeracy (z-score)", title = "Numeracy")

p_edu + p_exp + p_num
```

Key observations from the boxplots:

- **Education:** Managers have slightly more years of education on average,
  but the distributions overlap substantially.
- **Experience:** Managers tend to have more work experience, which is
  consistent with the expectation that supervisory roles require seniority.
- **Numeracy:** Managers have higher numeracy scores on average, but the two
  distributions overlap enough that we should not interpret the unadjusted
  pattern as an adjusted regression effect.

::: {.callout-tip}
## Writing Tip
For a binary outcome, bar charts (for categorical predictors) and boxplots
(for continuous predictors) are the most informative exploratory
visualizations. They help you preview which predictors are likely to be
important before fitting the logistic regression.
:::


## Part 5: Regression Analysis {#sec-logistic-ex-part5}

::: {.callout-important}
## Guideline Reminder --- Part 5
Fit the planned logistic model sequence, present results as **odds ratios**
with 95% confidence intervals, compare nested models using **likelihood ratio
tests**, and report model fit statistics. For logistic regression, the key
substantive deliverable is a predicted-probability display, which we build in
the next section.
:::

### Fit the Planned Model Sequence

We begin by fitting the three nested models from the proposal. All three models
use `analysis_df` and `family = binomial`, matching the logistic regression
labs.

```{r}
#| label: fit-logistic-models

# --- Model 1: Baseline human capital model ---
logit1 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c,
              data = analysis_df, family = binomial)

# --- Model 2: Add occupation and numeracy ---
logit2 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                occupation + numeracy,
              data = analysis_df, family = binomial)

logit2_no_occ <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                       numeracy,
                     data = analysis_df, family = binomial)

# --- Model 3: Add gender interactions ---
logit3 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                occupation + numeracy +
                female:schyrs_c + female:numeracy,
              data = analysis_df, family = binomial)

tibble(
  model = c("Model 1", "Model 2", "Model 3"),
  n = c(nobs(logit1), nobs(logit2), nobs(logit3))
)
```

### Model 1: Baseline Human-Capital Model {#sec-logistic-ex-results-m1}

Model 1 estimates the baseline association of gender, education, and
experience with the odds of the modeled event, `manager = 1`.

```{r}
#| label: params-logit1

model_params(logit1, exponentiate = TRUE)
```

**Interpretation of Model 1:**

- **Intercept (OR):** The baseline odds of being a manager for a *male*
  worker with *average* education and *average* experience. Because the
  intercept is on the log-odds scale, the exponentiated value gives the
  odds (not probability) at the reference point.
- **Female (OR $\approx$ 0.32, $p < .001$):** In the baseline model, women's
  odds of holding a supervisory position are about one-third of men's odds
  among workers with the same education and experience. This interpretation is
  for Model 1 only; later models add occupation and numeracy.
- **Education (OR $\approx$ 1.26, $p < .001$):** Each additional year of
  education is associated with a 26% increase in the odds of being a
  manager, holding gender and experience constant.
- **Experience (OR $> 1$):** Work experience in decades is associated with
  higher odds of being a manager, consistent with seniority-based promotion
  patterns.
- **Experience squared:** The negative coefficient (OR $< 1$) on the squared
  term indicates diminishing returns --- the promotion advantage of
  experience levels off at higher tenure.


### Model 2: Adding Occupation and Numeracy {#sec-logistic-ex-results-m2}

Model 2 extends Model 1 by adding the `occupation` factor and numeracy
proficiency. `Skilled` is the occupation reference category.

```{r}
#| label: params-logit2

model_params(logit2, exponentiate = TRUE)
```

**Interpretation of Model 2:**

- **Occupation effects** (all relative to **Skilled** occupations, the
  reference category):
  - Semi-white-collar workers have about 14% lower odds of being managers
    compared with skilled workers (OR $\approx$ 0.86).
  - Semi-blue-collar workers have about 76% lower odds compared with
    skilled workers (OR $\approx$ 0.24).
  - Elementary workers have about 89% lower odds compared with skilled
    workers (OR $\approx$ 0.12). Equivalently, skilled workers have about
    8.7 times the odds of elementary workers.
  - This clear occupational gradient shows that broad occupational category is
    strongly associated with whether a worker supervises others.
- **Numeracy (OR $\approx$ 1, not significant):** Numeracy proficiency is
  **not** a significant predictor of supervisory status after controlling for
  education, experience, and occupation. This is an important contrast with
  the linear regression results, where numeracy was highly significant for
  predicting earnings.
- **Female:** The gender odds ratio remains substantial (women still have
  much lower odds of being managers) even after controlling for occupation
  and cognitive skills.
- **Education:** The education OR decreases somewhat after controlling for
  occupation, suggesting that part of the education effect in Model 1
  operated through occupational sorting.

::: {.callout-tip}
## Key Contrast with Linear Regression
In the earnings analysis (Project 1), numeracy was a significant predictor:
what workers *know* mattered for how much they *earn*. Here, numeracy is
**not** significant for managerial status. This suggests that cognitive
skills are associated with earnings but do not independently distinguish who
holds supervisory roles after occupation and other covariates are included.
:::


### Model 3: Gender Interactions {#sec-logistic-ex-results-m3}

Model 3 extends Model 2 by adding interaction terms between gender and two
key predictors: education and numeracy. These interactions test whether the
education and numeracy associations with supervisory status differ by gender.

```{r}
#| label: params-logit3

model_params(logit3, exponentiate = TRUE)
```

**Interpretation of the interaction terms:**

- **Female $\times$ Education:** The interaction is not statistically
  significant. We do not have evidence that the education-supervisory-status
  association differs between men and women.
- **Female $\times$ Numeracy:** This interaction is also not significant.
  The already weak numeracy-supervisory-status association does not appear to
  vary by gender.

These null interaction results indicate that the gender gap in managerial
status operates primarily as a **level difference** (the main effect of
`female`) rather than through differential returns to education or cognitive
skills.


### Publication-Style Odds Ratio Table {#sec-logistic-ex-or-table}

@tbl-logistic-models presents all three models side by side in a
publication-ready format with exponentiated coefficients (odds ratios).

```{r}
#| label: tbl-logistic-models
#| tbl-cap: "Nested Logistic Regression Models Predicting Supervisory Status (Odds Ratios)"

base_labels <- list(
  female ~ "Female (vs male)",
  schyrs_c ~ "Education (years, centered)",
  exper_c ~ "Experience (decades, centered)",
  exper2_c ~ "Experience squared"
)

extended_labels <- c(
  base_labels,
  list(
    occupation ~ "Occupation",
    numeracy ~ "Numeracy (z-score)"
  )
)

interaction_labels <- c(
  extended_labels,
  list(
    `female:schyrs_c` ~ "Female x education",
    `female:numeracy` ~ "Female x numeracy"
  )
)

t1 <- tbl_regression(
  logit1,
  exponentiate = TRUE,
  label = base_labels
) |> bold_p(t = 0.05)

t2 <- tbl_regression(
  logit2,
  exponentiate = TRUE,
  label = extended_labels
) |> bold_p(t = 0.05)

t3 <- tbl_regression(
  logit3,
  exponentiate = TRUE,
  label = interaction_labels
) |> bold_p(t = 0.05)

tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
)
```

::: {.callout-tip}
## Writing Tip
For logistic regression, always present **odds ratios** (not raw log-odds
coefficients) in the publication table. Set `exponentiate = TRUE` in
`tbl_regression()`. The confidence intervals are also exponentiated, so they
are on the OR scale. An OR whose 95% CI includes 1.00 roughly corresponds to
weak evidence at $\alpha = .05$, although confidence intervals and p-values can
differ slightly depending on the method used.
:::

::: {.callout-note}
## Interaction Odds Ratios
Exponentiated interaction terms are ratios of odds ratios, not ordinary group
odds ratios. Use them to test whether associations differ by group, then use
predicted probabilities to explain the pattern on the probability scale.
:::


### Likelihood Ratio Tests {#sec-logistic-ex-lr-tests}

We compare nested models using **likelihood ratio (LR) tests**, the logistic
regression analogue of the incremental $F$-test used in linear regression.

#### LR Test: Model 1 vs. Model 2

```{r}
#| label: lr-test-m1-m2

# LR test: Does adding occupation + numeracy improve fit?
anova(logit1, logit2, test = "Chisq")
```

The likelihood ratio test is highly significant ($p < .001$), showing that the
added occupation/numeracy block improves model fit beyond the baseline
human-capital variables. The individual odds ratios and the occupation block
test below suggest that this improvement is driven mainly by occupation.

#### Overall LR Test for Occupation

Because `occupation` is a multi-category predictor, we also test the occupation
factor as a block while holding numeracy and the baseline covariates in the
model.

```{r}
#| label: lr-test-occupation

# LR test: Does the occupation factor improve fit after numeracy is included?
anova(logit2_no_occ, logit2, test = "Chisq")
```

The occupation block is statistically significant, $\chi^2(3) = 90.75$,
$p < .001$. This tells us that the occupation factor as a whole improves model
fit, even though one individual contrast (Semi-white versus Skilled) is not
statistically significant.

#### LR Test: Model 2 vs. Model 3

```{r}
#| label: lr-test-m2-m3

# LR test: Do gender interactions improve fit?
anova(logit2, logit3, test = "Chisq")
```

The likelihood ratio test for the two interaction terms is **not
significant** ($p > .05$), indicating that the gender interaction block does
not meaningfully improve model fit. For the single research question, this
means we do not find strong evidence that the education or numeracy
associations with supervisory status differ by gender.


### Collinearity Diagnostics {#sec-logistic-ex-vif}

@tbl-logistic-vif reports multicollinearity diagnostics from
`performance::check_collinearity()` for the full model. The VIF check belongs
near the model table because it helps readers assess coefficient stability.

```{r}
#| label: tbl-logistic-vif
#| tbl-cap: "Variance Inflation Factors for Logistic Model 3"

collinearity_table <- performance::check_collinearity(logit3, ci = FALSE) |>
  as_tibble() |>
  select(
    Term,
    VIF,
    `SE inflation factor` = SE_factor,
    Tolerance
  ) |>
  round_numeric(digits = 2)

kable(collinearity_table, align = "lrrr")
```

All VIF values are well below 5, indicating that multicollinearity is not a
concern. The interaction terms have modest VIFs because the continuous
predictors were mean-centered before the interactions were estimated.

::: {.callout-note}
## VIF Screening Convention
Use the same rule-of-thumb convention as Project 1: VIF < 5 is usually not
concerning, VIF 5--10 deserves a brief check, and VIF > 10 suggests serious
multicollinearity. Treat these as screening guidelines, not automatic rules for
dropping predictors.
:::

### Other Logistic Diagnostic Checks {#sec-logistic-ex-other-diagnostics}

Logistic regression diagnostics are different from linear regression
diagnostics. We do not check residual normality or constant residual variance.
Instead, we ask whether the fitted model is stable and interpretable for the
modeled event, `manager = 1`.

For this worked example, the main checks are:

- **Sparse cells and separation:** The Part 4 cross-tabulations showed that
  every gender and occupation group has both manager and non-manager cases.
  The elementary occupation category has only 12 manager events, so that
  contrast should still be interpreted cautiously.
- **Convergence and coefficient stability:** The `glm()` models converged
  without warnings, and @tbl-logistic-vif shows no problematic
  multicollinearity.
- **Linearity in the logit:** Education and numeracy are modeled with simple
  linear logit terms, while experience is modeled with a centered quadratic
  term. @tbl-logistic-linearity-check shows that adding a simple set of extra
  nonlinear terms does not improve fit in this worked example.
- **Fit and prediction diagnostics:** @tbl-logistic-fit reports AIC, BIC,
  log-likelihood, Tjur's $R^2$, and RMSE. @tbl-logistic-stability-check gives
  a compact same-sample stability check. Classification tables, ROC/AUC, and
  calibration plots are optional unless your research question is explicitly
  about prediction or classification.

```{r}
#| label: tbl-logistic-linearity-check
#| tbl-cap: "Exploratory Linearity-in-the-Logit Check for Model 2"

logit2_nonlinear <- update(
  logit2,
  . ~ . + I(schyrs_c^2) + I(numeracy^2) + I(exper_c^3)
)

linearity_check <- anova(logit2, logit2_nonlinear, test = "Chisq") |>
  as_tibble() |>
  dplyr::slice(2) |>
  transmute(
    `Added terms` = "Education squared, numeracy squared, experience cubed",
    df = Df,
    `Chi-square` = Deviance,
    `p-value` = format_p(`Pr(>Chi)`)
  ) |>
  round_numeric(digits = 3)

kable(linearity_check, align = "lrrr")
```

The exploratory nonlinear-term check is not statistically clear
($p = .232$). This does not prove the linear logit specification is perfect,
but it provides a simple check that the current functional form is reasonable
for the worked example.

```{r}
#| label: tbl-logistic-stability-check
#| tbl-cap: "Same-Sample Stability and Influence Checks for Logistic Model 2"

fitted_probs <- fitted(logit2)
cooks_values <- cooks.distance(logit2)
cooks_cutoff <- 4 / nobs(logit2)

stability_check <- tibble(
  Check = c(
    "Minimum fitted probability",
    "Maximum fitted probability",
    "Fitted probabilities below .05",
    "Fitted probabilities above .95",
    "Maximum Cook's distance",
    "Cook's distance values above 4/n"
  ),
  Value = c(
    sprintf("%.3f", min(fitted_probs)),
    sprintf("%.3f", max(fitted_probs)),
    as.character(sum(fitted_probs < .05)),
    as.character(sum(fitted_probs > .95)),
    sprintf("%.3f", max(cooks_values)),
    paste0(sum(cooks_values > cooks_cutoff), " (cutoff = ",
           sprintf("%.3f", cooks_cutoff), ")")
  )
)

kable(stability_check, align = "lr")
```

The fitted probabilities do not pile up at 1.00 (maximum = 0.886). The common
$4/n$ Cook's-distance rule is a sensitive screening rule, especially in larger
samples; here it flags 50 observations even though the maximum Cook's distance
is only 0.020. This calls for a sensitivity check, not automatic deletion or a
conclusion that the model failed. @tbl-logistic-influence-sensitivity compares
the focal odds ratios before and after removing the flagged observations.

```{r}
#| label: tbl-logistic-influence-sensitivity
#| tbl-cap: "Influence Sensitivity Check for Focal Odds Ratios in Logistic Model 2"

flagged_cases <- cooks_values > cooks_cutoff

logit2_without_flagged <- update(
  logit2,
  data = analysis_df[!flagged_cases, ]
)

focal_terms <- c(
  "female", "schyrs_c", "numeracy",
  "occupationSemi-white", "occupationSemi-blue", "occupationElementary"
)

term_labels <- c(
  "female" = "Female",
  "schyrs_c" = "Education",
  "numeracy" = "Numeracy",
  "occupationSemi-white" = "Semi-white vs Skilled",
  "occupationSemi-blue" = "Semi-blue vs Skilled",
  "occupationElementary" = "Elementary vs Skilled"
)

influence_sensitivity <- bind_rows(
  broom::tidy(logit2, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(Sample = "Full sample"),
  broom::tidy(logit2_without_flagged, exponentiate = TRUE, conf.int = TRUE) |>
    mutate(Sample = "Without flagged cases")
) |>
  filter(term %in% focal_terms) |>
  mutate(
    Term = term_labels[term],
    Term = factor(Term, levels = unname(term_labels[focal_terms]))
  ) |>
  arrange(Term, Sample) |>
  transmute(
    Term,
    Sample,
    OR = round(estimate, 2),
    `95% CI` = paste0("[", round(conf.low, 2), ", ", round(conf.high, 2), "]"),
    `p-value` = format_p(p.value)
  )

kable(influence_sensitivity, align = "llrll")
```

The sensitivity check does not change the substantive interpretation: the
gender and occupation associations remain large, education remains positive,
and numeracy remains weak after flagged cases are removed. Therefore, the
flagged Cook's-distance cases should be acknowledged, but they do not overturn
the main explanatory conclusions in this worked example.


### Predicted Probabilities and Marginal Effects {#sec-logistic-ex-predicted}

Odds ratios are useful for compact model tables, but the main substantive
deliverable in a logistic regression project is usually on the **probability
scale**. Predicted probabilities answer questions like "How likely is this
worker to be a manager?" Average marginal effects answer questions like "How
many percentage points does the predicted probability change, on average?"

Because the interaction block in Model 3 did not improve fit, the main
predicted-probability displays use Model 2. If an interaction model is central
to your own project, use the interaction model for the probability-scale
figures and explain why.

#### Predicted Probability Graph: Education by Gender

@fig-logistic-predicted shows the predicted probability of holding a
supervisory position as education changes, separately for men and women. The
profile holds occupation at Skilled, experience at its centered mean, and
numeracy at its centered mean.

```{r}
#| label: fig-logistic-predicted
#| fig-cap: "Predicted probability of being a manager by education and gender."
#| fig-alt: "Predicted probability curves showing that the probability of being a manager increases with education for both men and women. Men have higher predicted probabilities than women across the education range."
#| fig-height: 5

gender_colors <- c(
  "0" = unname(ber640_colors["primary"]),
  "1" = unname(ber640_colors["secondary"])
)

gender_profile <- marginaleffects::datagrid(
  model = logit2,
  exper_c = 0,
  exper2_c = 0,
  occupation = factor("Skilled", levels = levels(analysis_df$occupation)),
  numeracy = 0
)

plot_predictions(
  logit2,
  condition = list("schyrs_c" = seq(-4, 4, by = 1), "female"),
  newdata = gender_profile,
  type = "response"
) +
  labs(
    x = "Years of education (centered)",
    y = "Predicted probability of being a manager",
    color = "Gender",
    fill = "Gender",
    title = "Predicted Probability of Supervisory Status",
    subtitle = "By education level and gender (Model 2)"
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(
    values = gender_colors,
    breaks = c("0", "1"),
    labels = c("0" = "Male", "1" = "Female")
  ) +
  scale_fill_manual(
    values = gender_colors,
    breaks = c("0", "1"),
    labels = c("0" = "Male", "1" = "Female")
  )
```

The graph translates the adjusted gender odds ratio into a more intuitive
message: among skilled workers with average experience and numeracy, women
have lower predicted probabilities of supervisory status than otherwise
comparable men. Education is positively associated with the probability of
being a manager for both groups.

#### Predicted Probability Graph: Education by Occupation

@fig-logistic-predicted-occ shows predicted probabilities across education
levels for each occupational category. To isolate the occupational gradient,
the profile holds gender at male and holds experience and numeracy at their
centered means.

```{r}
#| label: fig-logistic-predicted-occ
#| fig-cap: "Predicted probability of being a manager by education and occupation."
#| fig-alt: "Predicted probability curves for four occupational categories. Skilled workers have the highest predicted probability at every education level, followed by semi-white, semi-blue, and elementary workers."
#| fig-height: 5.5

occupation_colors <- setNames(
  unname(ber640_colors[1:4]),
  levels(analysis_df$occupation)
)

occupation_profile <- marginaleffects::datagrid(
  model = logit2,
  female = 0,
  exper_c = 0,
  exper2_c = 0,
  numeracy = 0
)

plot_predictions(
  logit2,
  condition = list("schyrs_c" = seq(-4, 4, by = 1), "occupation"),
  newdata = occupation_profile,
  type = "response"
) +
  labs(
    x = "Years of education (centered)",
    y = "Predicted probability of being a manager",
    color = "Occupation",
    fill = "Occupation",
    title = "Predicted Probability of Supervisory Status",
    subtitle = "By education level and occupation (Model 2)"
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(values = occupation_colors, breaks = levels(analysis_df$occupation)) +
  scale_fill_manual(values = occupation_colors, breaks = levels(analysis_df$occupation))
```

The occupational gradient is larger than the education gradient in this
example. Among men with average experience and numeracy, skilled workers have
the highest predicted probabilities, while elementary workers have the lowest
predicted probabilities at comparable values of the other predictors.

#### Predicted Probabilities at Selected Values

@tbl-logistic-selected-probabilities presents predicted probabilities for
skilled workers at three education levels. Experience and numeracy are held at
their centered means.

```{r}
#| label: tbl-logistic-selected-probabilities
#| tbl-cap: "Predicted probabilities of supervisory status at selected education and gender values."

selected_probabilities <- marginaleffects::predictions(
  logit2,
  newdata = marginaleffects::datagrid(
    female   = c(0, 1),
    schyrs_c = c(-4, 0, 4),
    exper_c  = 0,
    exper2_c = 0,
    occupation = factor("Skilled", levels = levels(analysis_df$occupation)),
    numeracy  = 0
  ),
  type = "response"
)

mean_education <- mean(analysis_df$schyrs)

selected_probabilities |>
  as.data.frame() |>
  mutate(
    Gender = if_else(female == 0, "Male", "Female"),
    Education = case_when(
      schyrs_c == -4 ~ sprintf("Mean - 4 years (%.1f years)", mean_education - 4),
      schyrs_c ==  0 ~ sprintf("Mean (%.1f years)", mean_education),
      schyrs_c ==  4 ~ sprintf("Mean + 4 years (%.1f years)", mean_education + 4)
    ),
    `Predicted probability` = scales::percent(estimate, accuracy = 0.1),
    `95% CI` = paste0(
      "[", scales::percent(conf.low, accuracy = 0.1), ", ",
      scales::percent(conf.high, accuracy = 0.1), "]"
    )
  ) |>
  select(Gender, Education, `Predicted probability`, `95% CI`) |>
  kable(align = "llcc")
```

At mean education, average experience, average numeracy, and the Skilled
occupation reference category, the predicted probability is about **70%** for
men and **35%** for women. This is the same model as the odds-ratio table, but
the probability scale makes the practical size of the gap easier to
communicate.

#### Average Marginal Effects

Average marginal effects summarize Model 2 on the probability scale.
@tbl-logistic-ame reports the average change in predicted probability for
each focal predictor.

```{r}
#| label: tbl-logistic-ame
#| tbl-cap: "Average marginal effects from Model 2 on the probability scale."

format_pp <- function(x) {
  sprintf("%.1f pp", 100 * x)
}

ame_table <- marginaleffects::avg_slopes(
  logit2,
  variables = c("female", "schyrs_c", "occupation", "numeracy"),
  type = "response"
) |>
  as.data.frame() |>
  mutate(
    Term = case_when(
      term == "female" ~ "Gender",
      term == "schyrs_c" ~ "Education",
      term == "occupation" ~ "Occupation",
      term == "numeracy" ~ "Numeracy"
    ),
    Contrast = case_when(
      term == "female" ~ "Female - male",
      term == "schyrs_c" ~ "One additional year",
      term == "numeracy" ~ "One SD increase",
      TRUE ~ stringr::str_squish(contrast)
    ),
    `Average change` = format_pp(estimate),
    `95% CI` = paste0("[", format_pp(conf.low), ", ", format_pp(conf.high), "]"),
    `p-value` = format_p(p.value),
    Sort = case_when(
      Term == "Gender" ~ 1,
      Contrast == "Semi-white - Skilled" ~ 2,
      Contrast == "Semi-blue - Skilled" ~ 3,
      Contrast == "Elementary - Skilled" ~ 4,
      Term == "Education" ~ 5,
      Term == "Numeracy" ~ 6,
      TRUE ~ 99
    )
  ) |>
  arrange(Sort) |>
  select(Term, Contrast, `Average change`, `95% CI`, `p-value`)

kable(ame_table, align = "llccc")
```

The average marginal effects tell the same story as the odds ratios, but in
percentage points. Holding the model specification fixed, women have an
average predicted probability about **28 percentage points lower** than men.
Each additional year of education is associated with about a **2 percentage
point increase** in predicted probability. Numeracy is small and statistically
uncertain after the other predictors are included.

::: {.callout-tip}
## Writing Tip
Use odds ratios to summarize the model compactly, then use predicted
probabilities and average marginal effects to explain practical significance.
Readers usually understand percentage-point changes faster than odds-ratio
changes.
:::

#### Interaction Check on the Probability Scale

The interaction terms in Model 3 are difficult to interpret directly as odds
ratios because exponentiated interactions are ratios of odds ratios.
@tbl-logistic-interaction-slopes shows a simpler probability-scale check:
education and numeracy slopes estimated separately for men and women.

```{r}
#| label: tbl-logistic-interaction-slopes
#| tbl-cap: "Average education and numeracy slopes by gender from Model 3."

interaction_slopes <- marginaleffects::avg_slopes(
  logit3,
  variables = c("schyrs_c", "numeracy"),
  by = "female",
  type = "response"
) |>
  as.data.frame() |>
  mutate(
    Gender = if_else(female == 0, "Male", "Female"),
    Predictor = case_when(
      term == "schyrs_c" ~ "Education",
      term == "numeracy" ~ "Numeracy"
    ),
    `Average slope` = format_pp(estimate),
    `95% CI` = paste0("[", format_pp(conf.low), ", ", format_pp(conf.high), "]"),
    `p-value` = format_p(p.value)
  ) |>
  arrange(Predictor, Gender) |>
  select(Gender, Predictor, `Average slope`, `95% CI`, `p-value`)

kable(interaction_slopes, align = "llccc")
```

The education slopes are similar for men and women, and the numeracy slopes
are not statistically clear for either group. Small apparent differences in
the group-specific slopes should not be overinterpreted because the
interaction block did not improve model fit. This probability-scale check
supports the LR-test conclusion that the gender interaction block does not add
much to this example.


### Model Fit Summary {#sec-logistic-ex-fit}

@tbl-logistic-fit summarizes the overall fit of each model using $N$,
log-likelihood, AIC, BIC, Tjur's $R^2$, and RMSE. Model $N$ and
log-likelihood come from base R helpers (`nobs()` and `logLik()`), while AIC,
BIC, Tjur's $R^2$, and RMSE come from the course `model_fit()` helper.

Remember that Tjur's $R^2$ is a logistic model fit index, not the same thing as
ordinary least-squares $R^2$. It is useful for comparing models fit to the same
outcome and sample, but it should not be described as "percent of variance
explained."

```{r}
#| label: tbl-logistic-fit
#| tbl-cap: "Model Fit Statistics for the Three Logistic Regression Models"

fit_table <- tibble(
  Model = c("Model 1", "Model 2", "Model 3"),
  N = c(nobs(logit1), nobs(logit2), nobs(logit3)),
  `Log likelihood` = c(logLik(logit1), logLik(logit2), logLik(logit3)) |>
    as.numeric()
) |>
  bind_cols(
    bind_rows(
      model_fit(logit1) |> as_tibble(),
      model_fit(logit2) |> as_tibble(),
      model_fit(logit3) |> as_tibble()
    )
  ) |>
  select(Model, N, `Log likelihood`, AIC, BIC, `Tjur R2` = R2_Tjur, RMSE) |>
  round_numeric(digits = 3)

kable(fit_table, align = "lrrrrrr")
```

Model 2 shows a substantial improvement over Model 1 (lower AIC and BIC,
higher Tjur's $R^2$), consistent with the added occupation/numeracy block
improving fit. Model 3 slightly improves log-likelihood and Tjur's $R^2$, but
worsens AIC and BIC, so it adds complexity without a meaningful fit gain.


## Part 6: Discussion {#sec-logistic-ex-part6}

::: {.callout-important}
## Guideline Reminder --- Part 6
Summarize your key findings by answering the single logistic research
question directly, explain practical significance on the probability scale,
discuss model adequacy, note at least 2--3 limitations, and use association
language unless the study design supports causal inference.
:::

### Summary of Key Findings {#sec-logistic-ex-findings}

This study examined how gender, education, work experience, occupation, and
numeracy are associated with the probability of holding a supervisory position
among Korean full-time workers. The analysis used a nested series of logistic
regression models applied to PIAAC Korea data ($N = 1{,}357$). The results
answer the single research question in three parts. Taken together, Model 2 is
the preferred explanatory summary: it improves substantially over the baseline
model, while the added interaction block in Model 3 does not improve fit.

**Baseline human-capital variables are associated with supervisory status.**
In Model 1, education and work experience are positively associated with the
odds of being a manager, and experience follows a concave pattern. Each
additional year of education is associated with about 26% higher odds in the
baseline model. After occupation and numeracy are added in Model 2, the
education association is smaller but still meaningful: the odds ratio is about
1.12, and the average marginal effect is about a 2 percentage-point increase
in predicted probability per additional year of education
(@tbl-logistic-models; @tbl-logistic-ame).

**Occupation adds strong evidence; numeracy does not add much after controls.**
Model 2 improves substantially over Model 1, and the occupation block is
statistically clear, $\chi^2(3) = 90.75$, $p < .001$. Compared with Skilled
workers, Semi-blue workers and Elementary workers have much lower adjusted
probabilities of being managers. On the probability scale, the average
marginal effects are about -27.7 percentage points for Semi-blue workers and
-38.3 percentage points for Elementary workers relative to Skilled workers
(@tbl-logistic-ame). Numeracy, however, is small and statistically uncertain
after education, experience, gender, and occupation are included.

**The adjusted gender gap is large, but the interaction evidence is weak.**
Women have substantially lower adjusted odds of being managers than otherwise
similar men with the same observed covariate values included in the model. In
Model 2, the female odds ratio is about 0.23, which corresponds to about 77%
lower adjusted odds. On the probability scale, the
average marginal effect is about -28 percentage points (@tbl-logistic-ame).
The selected Skilled-worker profile in @tbl-logistic-selected-probabilities
shows a larger profile-specific contrast: about 70% for men and 35% for women
at mean education, average experience, and average numeracy. Model 3 does not
provide strong evidence that the education or numeracy associations differ by
gender; the interaction block does not improve fit over Model 2.

### The Persistent Gender Gap {#sec-logistic-ex-gender-gap}

One of the most striking findings is the large and persistent gender gap in
supervisory status. In the adjusted models, women have approximately **77%
lower odds** of being managers compared with men with the same observed
covariate values included in the model. The probability-scale results make
the size of the gap easier to communicate: the average adjusted gap is about
28 percentage points, and the selected Skilled-worker profile shows a 35
percentage-point gap at mean education, average experience, and average
numeracy.

This finding parallels the large gender earnings gap documented in Project 1
and is consistent with the well-documented underrepresentation of women in
leadership positions in South Korea. The fact that the gap persists after
controlling for measured human capital, occupation, and numeracy suggests
that unmeasured structural factors may be relevant. Possible contributors
include organizational culture, promotion practices, work-family constraints,
tenure, firm size, and industry sorting. This model cannot distinguish among
those mechanisms.

### Key Contrast: Linear vs. Logistic Regression Findings {#sec-logistic-ex-contrast}

@tbl-project-comparison compares the results across the two projects and
highlights an important substantive insight:

| Finding | Project 1 (Earnings) | Project 2 (Manager Status) |
|---------|---------------------|---------------------------|
| Education | Significant | Significant |
| Experience | Significant (concave) | Significant (concave) |
| Occupation | Strong earnings gradient | Strong managerial gradient |
| **Numeracy** | **Significant** ($p < .001$) | **Small and statistically uncertain after controls** |
| Gender gap | ~33--35% lower earnings | OR about 0.23; AME about -28 pp |
| Gender interactions | Mixed or weak evidence | Interaction block not supported |

: Comparison of linear and logistic regression findings. {#tbl-project-comparison}

The most notable contrast is that **numeracy is clearly associated with
earnings in Project 1 but is not statistically distinguishable from zero for
supervisory status after adjustment in Project 2**. This pattern is
consistent with the idea that measured cognitive skills may be more closely
linked to wages than to broad supervisory status, while promotion into
managerial roles may depend more on occupational context and unobserved
factors such as seniority, organizational fit, or network ties.

### Limitations {#sec-logistic-ex-limitations}

Several limitations should be noted when interpreting these results.

1. **Cross-sectional design.** Because PIAAC is a cross-sectional survey,
   we cannot determine whether education and experience *cause* promotion
   to supervisory roles. The observed associations may reflect selection
   effects (e.g., workers who seek managerial positions may also invest more
   in education).

2. **Single-country sample.** Our analysis focuses on South Korea, which
   has distinctive labor market institutions, a strong seniority system, and
   specific cultural norms about gender and leadership. The findings may not
   generalize to other countries.

3. **Teaching subset, survey design, and numeracy simplification.** This
   worked example analyzes an unweighted PIAAC Korea teaching subset as a
   complete-case analytic sample. The standard errors, confidence intervals,
   likelihood-ratio tests, chi-square tests, p-values, predicted probabilities,
   and average marginal effects are ordinary model-based quantities; they do not
   incorporate PIAAC sampling weights, replicate weights, complex-survey
   variance estimation, or the full set of plausible values for numeracy.
   Results should therefore be framed as sample associations in the teaching
   file, not as nationally representative estimates.

4. **Binary simplification.** The `manager` variable is a simple binary
   indicator of supervisory responsibility. It does not capture the *level*
   of management (e.g., front-line supervisor vs. senior executive) or the
   *number* of subordinates, which would provide a more nuanced picture.

5. **Omitted variables.** Important predictors such as tenure at the
   current firm, industry, firm size, and personality traits are not
   available in the PIAAC data. These factors likely influence promotion
   decisions and could confound our estimates.

6. **Model specification.** Education and numeracy are modeled with simple
   linear logit terms. The experience pattern is modeled with a quadratic
   term, but other nonlinearities or interactions could exist. Additional
   checks such as binned logit plots, spline terms, or sensitivity models
   would be useful in a longer analysis.

7. **Sparse subgroup estimate.** The elementary occupation category has only
   12 manager events. The overall occupation block is strong, but the exact
   elementary contrast should be interpreted cautiously because it is based on
   a small number of events.

### Model Adequacy and Confidence {#sec-logistic-ex-confidence}

The diagnostics support using Model 2 as the main explanatory summary for this
worked example. Adding occupation and numeracy improved fit over Model 1, LR
$\chi^2(4) = 95.09$, $p < .001$, while the gender interaction block did not
improve fit, LR $\chi^2(2) = 0.67$, $p = .714$. The descriptive
cross-tabulations did not reveal complete separation, the models converged
without warnings, and @tbl-logistic-vif shows that multicollinearity is not a
concern. @tbl-logistic-fit shows the same pattern: Model 2 improves
substantially over Model 1, while Model 3 adds complexity without improving
fit.

At the same time, the elementary occupation category has relatively few manager
events, and this analysis does not include a full calibration or classification
analysis. Those checks would be appropriate if the project goal were to build a
prediction tool. Here, the strongest conclusions are the broad association
patterns that are stable across the model sequence and the influence
sensitivity check: the persistent gender gap, the occupational gradient, and
the limited added value of numeracy for supervisory status after controls.

### Implications {#sec-logistic-ex-implications}

Despite these limitations, the findings have practical implications:

- The **nonsignificance of numeracy** for supervisory status suggests that
  measured cognitive skills alone do not explain who holds supervisory roles
  after education, experience, gender, and occupation are included.
  Organizational factors and social capital may be relevant for career
  advancement, but they are not directly observed in this model.
- The **strong occupational gradient** highlights the importance of
  occupational location. Workers in skilled occupations have much higher
  adjusted probabilities of supervisory status than workers in lower-status
  occupational categories, but this association should not be interpreted as
  the causal effect of occupational assignment.
- The **persistent gender gap** in managerial status, even after extensive
  controls, points to potential structural concerns that merit further study,
  such as mentoring access, leadership development opportunities, promotion
  practices, and implicit bias. This observational model does not estimate
  the effect of any specific intervention.


## Submission Checklist {#sec-logistic-ex-checklist}

::: {.callout-important}
## Project 2 Submission Checklist
Before submitting your Logistic Regression Results & Discussion, confirm
that you have completed all of the following:

- [ ] **Cross-tabulations** of the outcome with key categorical predictors,
  with counts, event rates, sparse-cell or separation notes, and optional
  chi-square tests as descriptive context
- [ ] **Descriptive statistics** for continuous variables by outcome group,
  with clear written interpretation
- [ ] **Exploratory visualizations** (bar charts, boxplots) with written
  interpretations
- [ ] **Logistic regression models** fitted with `glm(..., family = binomial)`
  on the same complete-case analysis dataset and inspected with
  `model_params(..., exponentiate = TRUE)`
- [ ] **Odds ratio table** using `tbl_regression(exponentiate = TRUE)`, merged
  into a single publication-style table with clear labels and reference groups
- [ ] **Focal odds-ratio interpretation** that describes changes in odds, not
  changes in probability
- [ ] **Likelihood ratio tests** comparing nested models to evaluate
  incremental contributions, including important predictor blocks
- [ ] **Overall test for any multi-category predictor** reported in addition
  to individual factor contrasts
- [ ] **Predicted probability graph** showing the probability of the modeled
  event, `manager = 1`, with the held-constant profile clearly stated; see
  @sec-template-logistic-prob-curve for a template
- [ ] **Predicted probability table** at selected predictor values for
  concrete interpretation, with 95% confidence intervals; see
  @sec-template-logistic-prob-table for a template
- [ ] **Average marginal effects or equivalent summary** using `avg_slopes()`
  or another probability-scale summary, written in percentage-point language; see
  @sec-template-logistic-ame for a template
- [ ] **VIF check** using `performance::check_collinearity()`
- [ ] **Model fit summary** using `model_fit()` with required fit evidence such
  as AIC, log-likelihood, and pseudo-$R^2$; BIC, Tjur's $R^2$, and RMSE are
  recommended when available
- [ ] **Logistic diagnostic note** addressing sparse cells/separation,
  convergence, collinearity, linearity in the logit, influence, and any optional
  classification or calibration checks used; see
  @sec-template-logistic-diagnostics for templates
- [ ] **Discussion** that answers the single research question, compares
  findings with the linear regression project if the two projects use the same
  dataset or topic, explains practical significance on the probability scale,
  frames results as sample associations when appropriate, notes limitations,
  discusses model adequacy, and uses non-causal language
- [ ] **Reproducible source file** runs from a clean session, uses relative
  paths or `here::here()`, loads required packages, and recreates every
  reported table, figure, model, diagnostic, prediction, and probability-scale
  summary
- [ ] All major figures and tables are labeled, captioned, and
  cross-referenced
:::
