# Prediction and Marginal-Effect Templates {#sec-prediction-templates}

{{< include ../_common.qmd >}}

::: {.callout-note}
## Why This Appendix Exists
Regression tables are not always the easiest way to communicate substantive
meaning. Adjusted predictions, predicted probabilities, probability
differences, and average marginal effects translate model results into the
scale readers understand. Use these templates after you have fit the model and
checked that it is adequate for your research question.
:::

All code below is set to `eval: false`. Replace model names, variable names,
labels, and held-constant values with your own.

::: {.callout-tip}
## Course Review Links
If the interpretation feels fuzzy, review [Interactions](https://joonho112.github.io/jlee-regression-course/notes/chapters/05-interactions.html), [Nonlinear Relationships](https://joonho112.github.io/jlee-regression-course/notes/chapters/06-nonlinear-relationships.html), [Multiple Logistic Regression: Predicted Probabilities](https://joonho112.github.io/jlee-regression-course/notes/chapters/10-multiple-logistic-regression.html#sec-ch10-predicted), and the matching labs: [Lab 06](https://joonho112.github.io/jlee-regression-course/labs/labs/lab06-interactions.html), [Lab 07](https://joonho112.github.io/jlee-regression-course/labs/labs/lab07-model-building.html), [Lab 10](https://joonho112.github.io/jlee-regression-course/labs/labs/lab10-logistic-extended.html), and optional [Lab 12](https://joonho112.github.io/jlee-regression-course/labs/labs/lab12-logistic-advanced.html).
:::


## General Rules for Predictions {#sec-prediction-rules}

Before making prediction or marginal-effect tables:

1. Use the same complete-case `analysis_df` used for the model.
2. Use the primary model that answers the research question.
3. State which predictors vary and which are held constant.
4. If the model includes a squared term, keep the original and squared
   variables consistent in prediction profiles. For example, if `exper_c = 0`,
   then `exper2_c = 0`; if `exper_c = 2`, then `exper2_c = 4`.
5. Use the outcome scale: predicted outcome for linear regression, predicted
   probability for logistic regression.
6. For Project 2, name the modeled event every time.
7. When you copy a template into your own Quarto report, give the final output
   chunk a `fig-` or `tbl-` label and a clear caption.

```{r}
#| label: prediction-template-setup
#| eval: false

library(dplyr)
library(ggplot2)
library(scales)
library(ggeffects)
library(marginaleffects)
library(knitr)

format_pp <- function(x) {
  sprintf("%.1f pp", 100 * x)
}

format_p <- function(p) {
  dplyr::if_else(
    is.na(p),
    NA_character_,
    dplyr::if_else(p < .001, "< .001", sprintf("%.3f", p))
  )
}
```

::: {.callout-tip}
## Which Package Should I Use?
For Project 1 linear adjusted predictions and interaction plots,
`ggeffects::predict_response()` matches the worked example and lab workflow.
For Project 2 logistic probability curves, probability tables, and AMEs, use
`marginaleffects` as the default [@arel-bundock2024]. Within one table or
figure, use one package grammar and keep the code easy to explain.
:::


## Template 1: Linear Adjusted Prediction Table {#sec-template-linear-pred-table}

Use this template when an interaction or nonlinear term is easier to interpret
with predicted values than with coefficients.

```{r}
#| label: template-linear-pred-table
#| eval: false

linear_predictions <- ggeffects::predict_response(
  model3,
  terms = c("numeracy [-2,-1,0,1,2]", "female"),
  condition = list(
    schyrs_c = 0,
    exper_c = 0,
    exper2_c = 0,
    occupation = "Skilled"
  )
)

linear_predictions |>
  as_tibble() |>
  mutate(
    Gender = if_else(as.character(group) == "0", "Male", "Female"),
    `Numeracy (z-score)` = x,
    `Predicted outcome` = round(predicted, 3),
    `95% CI` = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]")
  ) |>
  select(Gender, `Numeracy (z-score)`, `Predicted outcome`, `95% CI`) |>
  knitr::kable(align = "lrrc")
```

**Use in report:** State the model, the varying predictors, and the
held-constant profile. For a log-income outcome, tell the reader whether the
predicted values are still on the log scale.


## Template 2: Linear Interaction Plot {#sec-template-linear-interaction-plot}

Use this template to show how a continuous predictor's association with the
outcome differs across groups.

```{r}
#| label: template-linear-interaction-plot
#| eval: false

linear_predictions <- ggeffects::predict_response(
  model3,
  terms = c("numeracy [-2:2 by=0.5]", "female"),
  condition = list(
    schyrs_c = 0,
    exper_c = 0,
    exper2_c = 0,
    occupation = "Skilled"
  )
)

linear_predictions |>
  as_tibble() |>
  mutate(Gender = if_else(as.character(group) == "0", "Male", "Female")) |>
  ggplot(aes(x = x, y = predicted, color = Gender, fill = Gender)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15,
              linewidth = 0) +
  geom_line(linewidth = 1) +
  labs(
    x = "Numeracy (z-score)",
    y = "Predicted outcome",
    color = "Gender",
    fill = "Gender",
    title = "Adjusted Predictions from the Linear Model"
  )
```

**Use in report:** If the interaction test is not statistically clear, describe
the pattern cautiously. Do not imply that different-looking lines prove a real
group difference.


## Template 3: Linear Conditional Slope Table {#sec-template-linear-slopes}

Use this template when an interaction is central and you need a compact table
of group-specific slopes.

```{r}
#| label: template-linear-slopes
#| eval: false

marginaleffects::avg_slopes(
  model3,
  variables = "numeracy",
  by = "female"
) |>
  as.data.frame() |>
  mutate(
    Gender = if_else(as.character(female) == "0", "Male", "Female"),
    `Slope for numeracy` = round(estimate, 3),
    `95% CI` = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]"),
    `p-value` = format_p(p.value)
  ) |>
  select(Gender, `Slope for numeracy`, `95% CI`, `p-value`) |>
  knitr::kable(align = "lrcc")
```

**Use in report:** Interpret conditional slopes only after explaining the
interaction term or planned interaction block. For an additive linear model with
no interaction, the coefficient table is usually enough.


## Template 4: Logistic Predicted-Probability Curve {#sec-template-logistic-prob-curve}

Use this template for the required Project 2 predicted-probability figure.
Always name the modeled event in the y-axis.

```{r}
#| label: template-logistic-prob-curve
#| eval: false

gender_profile <- marginaleffects::datagrid(
  model = logit2,
  exper_c = 0,
  exper2_c = 0,
  occupation = factor("Skilled", levels = levels(analysis_df$occupation)),
  numeracy = 0
)

marginaleffects::plot_predictions(
  logit2,
  condition = list("schyrs_c" = seq(-4, 4, by = 1), "female"),
  newdata = gender_profile,
  type = "response"
) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_discrete(labels = c("0" = "Male", "1" = "Female")) +
  scale_fill_discrete(labels = c("0" = "Male", "1" = "Female")) +
  labs(
    x = "Years of education (centered)",
    y = "Predicted probability of manager = 1",
    color = "Gender",
    fill = "Gender",
    title = "Predicted Probability of the Modeled Event"
  )
```

**Use in report:** Write a probability sentence, not an odds-ratio sentence.
For example, "At the held-constant profile, the predicted probability of
`manager = 1` is higher for men than for women across the education range."


## Template 5: Logistic Predicted-Probability Table {#sec-template-logistic-prob-table}

Use this template for the required Project 2 predicted-probability table at
selected values.

```{r}
#| label: template-logistic-prob-table
#| eval: false

selected_probabilities <- marginaleffects::predictions(
  logit2,
  newdata = marginaleffects::datagrid(
    female = c(0, 1),
    schyrs_c = c(-4, 0, 4),
    exper_c = 0,
    exper2_c = 0,
    occupation = factor("Skilled", levels = levels(analysis_df$occupation)),
    numeracy = 0
  ),
  type = "response"
)

selected_probabilities |>
  as.data.frame() |>
  mutate(
    Gender = if_else(female == 0, "Male", "Female"),
    Education = case_when(
      schyrs_c == -4 ~ "Mean - 4 years",
      schyrs_c == 0 ~ "Mean",
      schyrs_c == 4 ~ "Mean + 4 years"
    ),
    `Predicted P(manager = 1)` = scales::percent(estimate, accuracy = 0.1),
    `95% CI` = paste0(
      "[", scales::percent(conf.low, accuracy = 0.1), ", ",
      scales::percent(conf.high, accuracy = 0.1), "]"
    )
  ) |>
  select(Gender, Education, `Predicted P(manager = 1)`, `95% CI`) |>
  knitr::kable(align = "llcc")
```

**Use in report:** State what values were held constant. The table should make
the practical size of the association easier to see than the odds-ratio table.


## Template 6: Logistic Average Marginal Effects Table {#sec-template-logistic-ame}

Use this template to summarize logistic regression results in percentage-point
language.

```{r}
#| label: template-logistic-ame
#| eval: false

ame_table <- marginaleffects::avg_slopes(
  logit2,
  variables = c("female", "schyrs_c", "occupation", "numeracy"),
  type = "response"
)

ame_table |>
  as.data.frame() |>
  mutate(
    Term = case_when(
      term == "female" ~ "Gender",
      term == "schyrs_c" ~ "Education",
      term == "occupation" ~ "Occupation",
      term == "numeracy" ~ "Numeracy",
      TRUE ~ term
    ),
    Contrast = case_when(
      term == "female" ~ "Female - male",
      term == "schyrs_c" ~ "One additional year",
      term == "numeracy" ~ "One SD increase",
      TRUE ~ contrast
    ),
    `Average change` = format_pp(estimate),
    `95% CI` = paste0("[", format_pp(conf.low), ", ", format_pp(conf.high), "]"),
    `p-value` = format_p(p.value)
  ) |>
  arrange(Term, Contrast) |>
  select(Term, Contrast, `Average change`, `95% CI`, `p-value`) |>
  knitr::kable(align = "llccc")
```

**Use in report:** Say "percentage points," not "percent," when interpreting
AMEs. For example, "-28 pp" means a 28 percentage-point lower predicted
probability, not 28% lower odds.


## Template 7: Logistic Conditional Effect Table {#sec-template-logistic-conditional}

Use this template when a logistic interaction model is central to the research
question.

```{r}
#| label: template-logistic-conditional
#| eval: false

conditional_slopes <- marginaleffects::avg_slopes(
  logit3,
  variables = c("schyrs_c", "numeracy"),
  by = "female",
  type = "response"
)

conditional_slopes |>
  as.data.frame() |>
  mutate(
    Gender = if_else(as.character(female) == "0", "Male", "Female"),
    Predictor = case_when(
      term == "schyrs_c" ~ "Education",
      term == "numeracy" ~ "Numeracy",
      TRUE ~ term
    ),
    `Average slope` = format_pp(estimate),
    `95% CI` = paste0("[", format_pp(conf.low), ", ", format_pp(conf.high), "]"),
    `p-value` = format_p(p.value)
  ) |>
  arrange(Predictor, Gender) |>
  select(Gender, Predictor, `Average slope`, `95% CI`, `p-value`) |>
  knitr::kable(align = "llccc")
```

**Use in report:** Conditional probability-scale effects are useful only if the
interaction is part of the planned research question or diagnostic follow-up.
They supplement, but do not replace, the likelihood-ratio test for the
interaction block. If the interaction block does not improve model fit, say that
clearly.


## Prediction Writing Checklist {#sec-template-prediction-checklist}

Before including a prediction, probability, or marginal-effect result, check:

- [ ] The model used for the result is named.
- [ ] The outcome scale is clear.
- [ ] For Project 2, the modeled event is named.
- [ ] The values held constant are stated.
- [ ] The varying predictors are labeled clearly.
- [ ] Confidence intervals are shown when the result is a table or figure.
- [ ] Logistic probability differences and AMEs are interpreted in percentage
  points.
- [ ] Copied Quarto chunks use `fig-` or `tbl-` labels and clear captions.
- [ ] The prose explains the substantive meaning, not only statistical
  significance.
