# Complete R Script Template {#sec-r-script}

{{< include ../_common.qmd >}}

::: {.callout-note}
## A Clean Starting Point
This appendix gives a student-ready R script template for the final project.
It is not meant to be a clever programming exercise. It is meant to help you
run a clear regression analysis from a clean R session, using the same simple
package workflow taught in the labs and worked examples.
:::

::: {.callout-important}
## How to Use This Template
1. Copy the complete Setup section first, then copy the relevant project
   sections into a new `.qmd` or `.R` file.
2. Replace the dataset, variable names, labels, and model formulas with your
   own.
3. Within each project, keep one complete-case analysis dataset and use it for
   descriptives, models, diagnostics, tables, and figures.
4. Run the file from top to bottom in a clean R session before submitting.
5. Avoid copying isolated chunks by themselves. Later chunks often depend on
   objects created earlier, such as `analysis_df`, fitted models, helper
   functions, and plotting settings.

All code chunks below use `eval: false`, so this appendix is a template rather
than executed book code.

This worked example uses one shared `analysis_df` because both projects use the
same PIAAC sample and the same predictor set. If your Project 1 and Project 2
use different outcomes or predictor sets, create separate objects such as
`analysis_df_linear` and `analysis_df_logistic`. The important rule is that all
nested models within a project use the same rows.
:::


## Setup {#sec-script-setup}

```{r}
#| label: script-setup
#| eval: false

# ============================================================
# BER 640 Final Project -- R Script Template
# ============================================================
# Replace data, variables, labels, and model formulas with your own.
# Run this file from top to bottom in a clean R session before submitting.
# ============================================================

# --- Install/load packages ---
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  dplyr, tidyr, forcats, tibble,
  ggplot2, scales, patchwork,
  parameters, performance, sandwich,
  marginaleffects, ggeffects,
  gtsummary, gt,
  here, knitr
)

# --- Small helper functions used below ---
# These helper defaults mirror project/_common.R. The helper layer stays small:
# display helpers plus thin course shortcuts around parameters/performance.
round_numeric <- function(x, digits = 3) {
  dplyr::mutate(
    x,
    dplyr::across(dplyr::where(is.numeric), \(col) round(col, digits))
  )
}

format_p <- function(p) {
  dplyr::if_else(
    is.na(p),
    NA_character_,
    dplyr::if_else(p < .001, "< .001", sprintf("%.3f", p))
  )
}

model_params <- function(model, exponentiate = FALSE, ...) {
  parameters::model_parameters(model, exponentiate = exponentiate, ...)
}

model_fit <- function(model, metrics = "common", ...) {
  performance::model_performance(model, metrics = metrics, ...)
}

compare_models <- function(..., metrics = "common", rank = FALSE) {
  performance::compare_performance(..., metrics = metrics, rank = rank)
}

# --- Plot theme ---
theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
)

# Optional course-style colors used in the worked-example chapters.
# Keeping this palette here helps copied plotting chunks run in a standalone script.
ber640_colors <- c(
  primary   = "#0B3D66",
  secondary = "#A8431E",
  accent    = "#2C6E91",
  success   = "#3E8E8A",
  warning   = "#C9A227",
  info      = "#6B3E7A"
)
```


## Load Data and Build Analysis Dataset(s) {#sec-script-data}

```{r}
#| label: script-data
#| eval: false

# ============================================================
# DATA LOADING AND ANALYSIS DATASET
# ============================================================

# Replace this line with your own data import.
raw_data <- readRDS(here::here("data", "piaac_kor.rds"))

# For the worked example, one shared complete-case dataset is intentional
# because both projects use the same sample and predictors. If your two
# projects use different variables, create separate analysis_vars_linear and
# analysis_vars_logistic vectors instead.
#
# List every variable needed by the largest planned model and by the
# descriptive tables. Using the largest model prevents same-sample problems
# within a project.
analysis_vars <- c(
  "ln_inc", "manager",
  "female", "schyrs", "schyrs_c", "exper", "exper_c", "exper2_c",
  "occupation", "numeracy"
)

prepared_data <- raw_data |>
  mutate(
    occupation = forcats::fct_relevel(occupation, "Skilled"),
    Gender = factor(
      female,
      levels = c(0, 1),
      labels = c("Male", "Female")
    ),
    Status = factor(
      manager,
      levels = c(0, 1),
      labels = c("Non-manager (0, non-event)", "Manager (1, event)")
    )
  )

# Report the starting N and complete-case exclusions before dropping rows.
prepared_data |>
  summarise(
    starting_n = n(),
    rows_with_missing_analysis_values =
      sum(!complete.cases(pick(all_of(analysis_vars)))),
    complete_case_n = sum(complete.cases(pick(all_of(analysis_vars)))),
    occupation_reference = levels(occupation)[1],
    modeled_event_before_drop = sum(manager == 1, na.rm = TRUE),
    non_event_before_drop = sum(manager == 0, na.rm = TRUE)
  )

analysis_df <- prepared_data |>
  tidyr::drop_na(dplyr::all_of(analysis_vars))

# Confirm sample size, missingness, reference category, and modeled event.
analysis_df |>
  summarise(
    final_n = n(),
    missing_analysis_values = sum(is.na(pick(all_of(analysis_vars)))),
    occupation_reference = levels(occupation)[1],
    modeled_event = sum(manager == 1),
    non_event = sum(manager == 0),
    event_rate = mean(manager == 1)
  )
```

::: {.callout-tip}
## Adaptation Tip
For your own project, define the event in a sentence before fitting the model.
For example: "The modeled event is `manager = 1`, meaning that the worker holds
a supervisory position." Use that event definition consistently in tables,
figures, and interpretation.

The models below use `female` as a 0/1 contrast (`0 = Male`, `1 = Female`) and
use `Gender` only for readable tables and plots. If you prefer to model a
factor directly, replace `female` with `Gender` and check the reference level
with `levels(analysis_df$Gender)`.
:::


## Project 1: Linear Regression {#sec-script-linear}

### Descriptive Tables and Figures {#sec-script-linear-desc}

```{r}
#| label: script-linear-descriptives
#| eval: false

# ============================================================
# PROJECT 1 -- DESCRIPTIVE ANALYSIS
# ============================================================

# Continuous variables: mean, SD, minimum, maximum.
analysis_df |>
  select(ln_inc, schyrs, exper, numeracy) |>
  gtsummary::tbl_summary(
    statistic = all_continuous() ~ "{mean} ({sd}); [{min}, {max}]",
    digits = all_continuous() ~ 2,
    label = list(
      ln_inc ~ "Log monthly income",
      schyrs ~ "Years of education",
      exper ~ "Work experience (years)",
      numeracy ~ "Numeracy (z-score)"
    ),
    missing = "no"
  ) |>
  gtsummary::bold_labels()

# Categorical variables: counts and percentages.
analysis_df |>
  select(Gender, occupation) |>
  gtsummary::tbl_summary(
    statistic = all_categorical() ~ "{n} ({p}%)",
    label = list(
      Gender ~ "Gender",
      occupation ~ "Occupation"
    ),
    missing = "no"
  ) |>
  gtsummary::bold_labels()

# Outcome distribution.
ggplot(analysis_df, aes(x = ln_inc)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    x = "Log monthly income",
    y = "Count",
    title = "Distribution of Log Monthly Income"
  )

# Scatterplots for continuous predictors.
p_edu <- ggplot(analysis_df, aes(x = schyrs, y = ln_inc)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Years of education", y = "Log income", title = "Education")

p_exp <- ggplot(analysis_df, aes(x = exper, y = ln_inc)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Work experience", y = "Log income", title = "Experience")

p_num <- ggplot(analysis_df, aes(x = numeracy, y = ln_inc)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Numeracy", y = "Log income", title = "Numeracy")

p_edu + p_exp + p_num

# Boxplots for categorical predictors.
ggplot(analysis_df, aes(x = occupation, y = ln_inc, fill = occupation)) +
  geom_boxplot(alpha = 0.75, show.legend = FALSE) +
  labs(x = "Occupation", y = "Log income",
       title = "Log Income by Occupation")
```


### Fit Linear Models {#sec-script-linear-models}

```{r}
#| label: script-linear-models
#| eval: false

# ============================================================
# PROJECT 1 -- LINEAR REGRESSION MODELS
# ============================================================

# Model 1: baseline human-capital model.
model1 <- lm(
  ln_inc ~ female + schyrs_c + exper_c + exper2_c,
  data = analysis_df
)

# Model 2: add occupation and numeracy.
model2 <- lm(
  ln_inc ~ female + schyrs_c + exper_c + exper2_c +
    occupation + numeracy,
  data = analysis_df
)

# Model 3: add gender interactions.
model3 <- lm(
  ln_inc ~ female + schyrs_c + exper_c + exper2_c +
    occupation + numeracy +
    female:schyrs_c + female:numeracy,
  data = analysis_df
)

# Same-sample check for nested model comparisons.
tibble(
  model = c("Model 1", "Model 2", "Model 3"),
  n = c(nobs(model1), nobs(model2), nobs(model3))
)

stopifnot(
  nobs(model1) == nobs(model2),
  nobs(model2) == nobs(model3)
)

# Compact coefficient tables for learning and interpretation.
model_params(model1)
model_params(model2)
model_params(model3)

# Planned incremental F-tests.
anova(model1, model2)
anova(model2, model3)
```


### Linear Regression Tables and Fit {#sec-script-linear-tables}

```{r}
#| label: script-linear-tables
#| eval: false

# ============================================================
# PROJECT 1 -- REPORT TABLES AND FIT
# ============================================================

linear_base_labels <- list(
  female ~ "Female (vs male)",
  schyrs_c ~ "Education (years, centered)",
  exper_c ~ "Experience (decades, centered)",
  exper2_c ~ "Experience squared"
)

linear_extended_labels <- c(
  linear_base_labels,
  list(
    occupation ~ "Occupation",
    numeracy ~ "Numeracy (z-score)"
  )
)

linear_interaction_labels <- c(
  linear_extended_labels,
  list(
    `female:schyrs_c` ~ "Female x education",
    `female:numeracy` ~ "Female x numeracy"
  )
)

t1 <- gtsummary::tbl_regression(
  model1,
  intercept = TRUE,
  label = linear_base_labels
) |>
  gtsummary::bold_p(t = 0.05)

t2 <- gtsummary::tbl_regression(
  model2,
  intercept = TRUE,
  label = linear_extended_labels
) |>
  gtsummary::bold_p(t = 0.05)

t3 <- gtsummary::tbl_regression(
  model3,
  intercept = TRUE,
  label = linear_interaction_labels
) |>
  gtsummary::bold_p(t = 0.05)

# The template calls tbl_regression() directly so you can see the table options
# and adapt them without learning an extra wrapper.
gtsummary::tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
)

# Fit summary for the three models.
bind_rows(
  model_fit(model1) |> as_tibble() |> mutate(Model = "Model 1"),
  model_fit(model2) |> as_tibble() |> mutate(Model = "Model 2"),
  model_fit(model3) |> as_tibble() |> mutate(Model = "Model 3")
) |>
  select(Model, n_Obs, R2, R2_adjusted, RMSE, Sigma, AIC, BIC) |>
  round_numeric(digits = 3) |>
  knitr::kable(align = "lrrrrrrr")

# Optional easystats comparison table.
compare_models(model1, model2, model3)
```


### Linear Diagnostics and Adjusted Predictions {#sec-script-linear-diagnostics}

```{r}
#| label: script-linear-diagnostics
#| eval: false

# ============================================================
# PROJECT 1 -- DIAGNOSTICS AND ADJUSTED PREDICTIONS
# ============================================================

# Collinearity.
performance::check_collinearity(model3, ci = FALSE) |>
  as_tibble() |>
  select(Term, VIF, `SE inflation factor` = SE_factor, Tolerance) |>
  round_numeric(digits = 2) |>
  knitr::kable(align = "lrrr")

# Residual and influence dashboard. Include the key plots in the appendix.
performance::check_model(
  model3,
  check = c("linearity", "homogeneity", "qq", "outliers")
)

# Robust-SE sensitivity check if heteroscedasticity is a concern.
model_params(model3, vcov = "HC3")

# Cook's distance screen.
tibble(
  cutoff_4_over_n = 4 / nobs(model3),
  max_cooks = max(cooks.distance(model3)),
  flagged_cooks = sum(cooks.distance(model3) > 4 / nobs(model3))
)

# Adjusted predictions for an interaction, if your model includes one.
linear_predictions <- ggeffects::predict_response(
  model3,
  terms = c("numeracy [-2,-1,0,1,2]", "female"),
  condition = list(
    schyrs_c = 0,
    exper_c = 0,
    exper2_c = 0,
    occupation = "Skilled"
  )
)

linear_predictions |>
  as_tibble() |>
  mutate(Gender = if_else(as.character(group) == "0", "Male", "Female")) |>
  ggplot(aes(x = x, y = predicted, color = Gender, fill = Gender)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15,
              linewidth = 0) +
  geom_line(linewidth = 1) +
  labs(
    x = "Numeracy (z-score)",
    y = "Predicted log monthly income",
    color = "Gender",
    fill = "Gender",
    title = "Adjusted Predictions from the Linear Model"
  )
```


## Project 2: Logistic Regression {#sec-script-logistic}

### Logistic Descriptives and Sparse-Cell Checks {#sec-script-logistic-desc}

```{r}
#| label: script-logistic-descriptives
#| eval: false

# ============================================================
# PROJECT 2 -- DESCRIPTIVE ANALYSIS
# ============================================================

# Event/non-event distribution. The modeled event is manager = 1.
analysis_df |>
  count(Status) |>
  mutate(percent = scales::percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(align = "lr")

# Predictors by outcome status.
analysis_df |>
  select(Status, Gender, occupation, schyrs, exper, numeracy) |>
  gtsummary::tbl_summary(
    by = Status,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      Gender ~ "Gender",
      occupation ~ "Occupation",
      schyrs ~ "Years of education",
      exper ~ "Work experience (years)",
      numeracy ~ "Numeracy (z-score)"
    ),
    missing = "no"
  ) |>
  gtsummary::add_overall() |>
  gtsummary::add_p() |>
  gtsummary::bold_labels()

# Sparse-cell check for a categorical predictor.
gtsummary::tbl_cross(
  data = analysis_df,
  row = occupation,
  col = Status,
  percent = "row"
)

# Event-rate plot by gender.
analysis_df |>
  group_by(Gender) |>
  summarise(event_rate = mean(manager == 1), .groups = "drop") |>
  ggplot(aes(x = Gender, y = event_rate, fill = Gender)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = NULL,
    y = "Percent with manager = 1",
    title = "Event Rate by Gender"
  )
```


### Fit Logistic Models {#sec-script-logistic-models}

```{r}
#| label: script-logistic-models
#| eval: false

# ============================================================
# PROJECT 2 -- LOGISTIC REGRESSION MODELS
# ============================================================

# Stage 1: baseline human-capital model.
logit1 <- glm(
  manager ~ female + schyrs_c + exper_c + exper2_c,
  data = analysis_df,
  family = binomial
)

# Stage 2: add occupation and numeracy.
logit2 <- glm(
  manager ~ female + schyrs_c + exper_c + exper2_c +
    occupation + numeracy,
  data = analysis_df,
  family = binomial
)

# Stage 2 comparison model for the occupation block.
logit2_no_occ <- glm(
  manager ~ female + schyrs_c + exper_c + exper2_c + numeracy,
  data = analysis_df,
  family = binomial
)

# Stage 3: add gender interactions.
logit3 <- glm(
  manager ~ female + schyrs_c + exper_c + exper2_c +
    occupation + numeracy +
    female:schyrs_c + female:numeracy,
  data = analysis_df,
  family = binomial
)

# Same-sample check.
tibble(
  model = c("Stage 1", "Stage 2", "Stage 2 no occupation", "Stage 3"),
  n = c(nobs(logit1), nobs(logit2), nobs(logit2_no_occ), nobs(logit3))
)

stopifnot(
  nobs(logit1) == nobs(logit2),
  nobs(logit2_no_occ) == nobs(logit2),
  nobs(logit2) == nobs(logit3)
)

# Odds ratios for learning and interpretation.
model_params(logit1, exponentiate = TRUE)
model_params(logit2, exponentiate = TRUE)
model_params(logit3, exponentiate = TRUE)

# Planned likelihood-ratio tests.
anova(logit1, logit2, test = "Chisq")
anova(logit2_no_occ, logit2, test = "Chisq")
anova(logit2, logit3, test = "Chisq")
```


### Logistic Odds-Ratio Tables {#sec-script-logistic-table}

```{r}
#| label: script-logistic-table
#| eval: false

# ============================================================
# PROJECT 2 -- REPORT TABLES
# ============================================================

logit_base_labels <- list(
  female ~ "Female (vs male)",
  schyrs_c ~ "Education (years, centered)",
  exper_c ~ "Experience (decades, centered)",
  exper2_c ~ "Experience squared"
)

logit_extended_labels <- c(
  logit_base_labels,
  list(
    occupation ~ "Occupation",
    numeracy ~ "Numeracy (z-score)"
  )
)

logit_interaction_labels <- c(
  logit_extended_labels,
  list(
    `female:schyrs_c` ~ "Female x education",
    `female:numeracy` ~ "Female x numeracy"
  )
)

t1 <- gtsummary::tbl_regression(
  logit1,
  exponentiate = TRUE,
  label = logit_base_labels
) |>
  gtsummary::bold_p(t = 0.05)

t2 <- gtsummary::tbl_regression(
  logit2,
  exponentiate = TRUE,
  label = logit_extended_labels
) |>
  gtsummary::bold_p(t = 0.05)

t3 <- gtsummary::tbl_regression(
  logit3,
  exponentiate = TRUE,
  label = logit_interaction_labels
) |>
  gtsummary::bold_p(t = 0.05)

gtsummary::tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Stage 1**", "**Stage 2**", "**Stage 3**")
)
```


### Predicted Probabilities and Marginal Effects {#sec-script-logistic-predicted}

For modular prediction, probability-table, AME, and conditional-effect
templates, see @sec-prediction-templates.

```{r}
#| label: script-logistic-predicted
#| eval: false

# ============================================================
# PROJECT 2 -- PROBABILITY-SCALE RESULTS
# ============================================================

# Predicted probability graph. Always name the modeled event in the y-axis.
gender_profile <- marginaleffects::datagrid(
  model = logit2,
  exper_c = 0,
  exper2_c = 0,
  occupation = factor("Skilled", levels = levels(analysis_df$occupation)),
  numeracy = 0
)

marginaleffects::plot_predictions(
  logit2,
  condition = list("schyrs_c" = seq(-4, 4, by = 1), "female"),
  newdata = gender_profile,
  type = "response"
) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = "Years of education (centered)",
    y = "Predicted probability of manager = 1",
    color = "Gender",
    fill = "Gender",
    title = "Predicted Probability of the Modeled Event"
  ) +
  scale_color_discrete(labels = c("0" = "Male", "1" = "Female")) +
  scale_fill_discrete(labels = c("0" = "Male", "1" = "Female"))

# Predicted probability table at selected values.
selected_probabilities <- marginaleffects::predictions(
  logit2,
  newdata = marginaleffects::datagrid(
    female = c(0, 1),
    schyrs_c = c(-4, 0, 4),
    exper_c = 0,
    exper2_c = 0,
    occupation = factor("Skilled", levels = levels(analysis_df$occupation)),
    numeracy = 0
  ),
  type = "response"
)

selected_probabilities |>
  as.data.frame() |>
  mutate(
    Gender = if_else(female == 0, "Male", "Female"),
    Education = case_when(
      schyrs_c == -4 ~ "Mean - 4 years",
      schyrs_c == 0 ~ "Mean",
      schyrs_c == 4 ~ "Mean + 4 years"
    ),
    `Predicted P(manager = 1)` = scales::percent(estimate, accuracy = 0.1),
    `95% CI` = paste0(
      "[", scales::percent(conf.low, accuracy = 0.1), ", ",
      scales::percent(conf.high, accuracy = 0.1), "]"
    )
  ) |>
  select(Gender, Education, `Predicted P(manager = 1)`, `95% CI`) |>
  knitr::kable(align = "llcc")

# Average marginal effects in probability points.
format_pp <- function(x) {
  sprintf("%.1f pp", 100 * x)
}

marginaleffects::avg_slopes(
  logit2,
  variables = c("female", "schyrs_c", "occupation", "numeracy"),
  type = "response"
) |>
  as.data.frame() |>
  mutate(
    `Average change` = format_pp(estimate),
    `95% CI` = paste0("[", format_pp(conf.low), ", ", format_pp(conf.high), "]"),
    `p-value` = format_p(p.value)
  ) |>
  select(term, contrast, `Average change`, `95% CI`, `p-value`) |>
  knitr::kable(align = "llccc")
```


### Logistic Diagnostics and Fit {#sec-script-logistic-diagnostics}

```{r}
#| label: script-logistic-diagnostics
#| eval: false

# ============================================================
# PROJECT 2 -- DIAGNOSTICS AND MODEL FIT
# ============================================================

# Collinearity.
performance::check_collinearity(logit3, ci = FALSE) |>
  as_tibble() |>
  select(Term, VIF, `SE inflation factor` = SE_factor, Tolerance) |>
  round_numeric(digits = 2) |>
  knitr::kable(align = "lrrr")

# Linearity-in-the-logit sensitivity check for continuous predictors.
logit2_linearity <- glm(
  manager ~ female + schyrs_c + I(schyrs_c^2) +
    exper_c + exper2_c + I(exper_c^3) +
    occupation + numeracy + I(numeracy^2),
  data = analysis_df,
  family = binomial
)

stopifnot(nobs(logit2) == nobs(logit2_linearity))
anova(logit2, logit2_linearity, test = "Chisq")

# Influence and extreme fitted probabilities.
logit2_fitted <- fitted(logit2)
logit2_cooks <- cooks.distance(logit2)

tibble(
  max_fitted_probability = max(logit2_fitted),
  min_fitted_probability = min(logit2_fitted),
  near_zero_or_one = sum(logit2_fitted < .05 | logit2_fitted > .95),
  cooks_cutoff_4_over_n = 4 / nobs(logit2),
  max_cooks = max(logit2_cooks),
  flagged_cooks = sum(logit2_cooks > 4 / nobs(logit2))
)

# Model fit summary. N and log-likelihood come from base R; AIC, BIC,
# Tjur's R2, and RMSE come from model_fit().
fit_table <- tibble(
  Model = c("Stage 1", "Stage 2", "Stage 3"),
  N = c(nobs(logit1), nobs(logit2), nobs(logit3)),
  `Log likelihood` = c(logLik(logit1), logLik(logit2), logLik(logit3)) |>
    as.numeric()
) |>
  bind_cols(
    bind_rows(
      model_fit(logit1) |> as_tibble(),
      model_fit(logit2) |> as_tibble(),
      model_fit(logit3) |> as_tibble()
    )
  ) |>
  select(Model, N, `Log likelihood`, AIC, BIC, `Tjur R2` = R2_Tjur, RMSE) |>
  round_numeric(digits = 3)

knitr::kable(fit_table, align = "lrrrrrr")

# Optional only if the project is about prediction:
# - ROC/AUC
# - classification table at a stated cutoff
# - calibration plot
```


## Clean-Session Submission Checklist {#sec-script-clean-session}

Before submitting, restart R and run the file from top to bottom. Then confirm
the checklist below.

For diagnostic placement, troubleshooting, and Quarto/R script guidance, see
@sec-diagnostics-reproducibility-templates.

- [ ] Every package used in the script is loaded in the setup section.
- [ ] The source file uses relative paths or `here::here()`.
- [ ] Each project has a complete-case analysis dataset before descriptives and
  models.
- [ ] Factor reference categories are set intentionally and checked with
  `levels()`.
- [ ] Project 2 names the modeled event and non-event.
- [ ] Descriptive tables and nested models within each project use the same
  complete-case dataset.
- [ ] Nested comparisons have matching `nobs()` values.
- [ ] Linear results use `model_params()`, `model_fit()`,
  `performance::check_collinearity()`, and report-ready tables.
- [ ] Logistic results use `model_params(..., exponentiate = TRUE)`,
  `tbl_regression(..., exponentiate = TRUE)`, predicted probabilities, and
  average marginal effects or another probability-scale summary.
- [ ] Diagnostics are summarized in words, not merely printed.
- [ ] Every reported table and figure can be recreated by the submitted source
  file.

```{r}
#| label: script-session-info
#| eval: false

# Optional reproducibility record.
sessionInfo()
```
