# Complete R Script {#sec-r-script}

{{< include ../_common.qmd >}}

::: {.callout-note}
## All Code in One Place
This appendix collects all R code from the worked examples (Chapters 4--7)
into a single, annotated script. You can copy this script and adapt it for
your own project by replacing our data, variables, and labels with yours.
:::

::: {.callout-important}
## How to Use This Script
1. **Copy** the entire script into a new `.R` or `.qmd` file.
2. **Replace** the dataset, variable names, and labels with your own.
3. **Delete** sections that do not apply (e.g., if your project is only
   linear regression, remove the logistic regression section).
4. **Run** the script from top to bottom to verify it works before
   submitting.

All code below is set to `eval: false` --- it is a reference template,
not executed as part of this book.
:::


## Setup {#sec-script-setup}

```{r}
#| label: script-setup
#| eval: false

# ============================================================
# BER 640 Final Project -- Complete R Script Template
# ============================================================
# Instructions:
#   1. Replace variable names and labels with your own
#   2. Delete the project section you don't need
#   3. Run from top to bottom before submitting
# ============================================================

# --- Load packages ---
pacman::p_load(
  dplyr, tidyr, forcats, stringr, purrr,
  ggplot2, patchwork, GGally, ggrepel,
  broom, car, lmtest, sandwich,
  parameters, performance, effectsize,
  see, modelbased, correlation, report,
  datawizard, insight,
  marginaleffects,
  gtsummary, gt, knitr, here
)

# --- Set ggplot theme ---
theme_set(
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey40"),
      axis.title = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
)
```


## Data Loading & Cleaning {#sec-script-data}

```{r}
#| label: script-data
#| eval: false

# ============================================================
# DATA LOADING & CLEANING
# ============================================================

# --- Load data ---
# Replace with your own data file
piaac <- readRDS(here::here("data", "piaac_kor.rds"))

# --- Inspect ---
dim(piaac)
str(piaac)
summary(piaac)

# --- Check for missing data ---
colSums(is.na(piaac))

# --- Select and clean variables ---
# (Adapt variable names to your own dataset)
df <- piaac |>
  select(
    # Outcome variables
    ln_inc, manager,
    # Predictors
    female, schyrs, schyrs_c, exper, exper_c, exper2_c,
    semiwhite, semiblue, elementary, occupation,
    numeracy, fem_schyrs_c, fem_numeracy
  ) |>
  drop_na()

cat("Analytic sample size:", nrow(df), "\n")
```


## Project 1: Linear Regression {#sec-script-linear}

### Descriptive Statistics {#sec-script-linear-desc}

```{r}
#| label: script-linear-descriptives
#| eval: false

# ============================================================
# PROJECT 1: LINEAR REGRESSION -- DESCRIPTIVE STATISTICS
# ============================================================

# --- Continuous variables: summary statistics ---
df |>
  select(ln_inc, schyrs, exper, numeracy) |>
  tbl_summary(
    type = everything() ~ "continuous",
    statistic = all_continuous() ~ "{mean} ({sd})",
    label = list(
      ln_inc   ~ "Log monthly income",
      schyrs   ~ "Years of education",
      exper    ~ "Work experience (years)",
      numeracy ~ "Numeracy (z-score)"
    ),
    digits = everything() ~ 2
  ) |>
  bold_labels()

# --- Categorical variables: frequency tables ---
df |>
  select(female, occupation) |>
  mutate(
    female = factor(female, levels = c(0, 1),
                    labels = c("Male", "Female"))
  ) |>
  tbl_summary(
    label = list(
      female     ~ "Gender",
      occupation ~ "Occupation"
    )
  ) |>
  bold_labels()

# --- Correlation matrix (continuous predictors) ---
df |>
  select(ln_inc, schyrs, exper, numeracy) |>
  ggpairs(
    lower = list(continuous = wrap("smooth", alpha = 0.3)),
    diag  = list(continuous = wrap("densityDiag"))
  )

# --- Histogram of outcome variable ---
ggplot(df, aes(x = ln_inc)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white",
                 alpha = 0.8) +
  geom_vline(xintercept = mean(df$ln_inc), linetype = "dashed") +
  labs(x = "Log monthly income", y = "Count",
       title = "Distribution of Log Monthly Income")

# --- Scatterplots of continuous predictors vs. outcome ---
p1 <- ggplot(df, aes(x = schyrs, y = ln_inc)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "blue",
              linetype = "dashed") +
  labs(x = "Years of education", y = "Log income",
       title = "Education")

p2 <- ggplot(df, aes(x = exper, y = ln_inc)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "blue",
              linetype = "dashed") +
  labs(x = "Work experience (years)", y = "Log income",
       title = "Experience")

p3 <- ggplot(df, aes(x = numeracy, y = ln_inc)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "blue",
              linetype = "dashed") +
  labs(x = "Numeracy (z-score)", y = "Log income",
       title = "Numeracy")

p1 + p2 + p3 +
  plot_annotation(subtitle = "Red = linear fit; Blue dashed = LOESS smooth")

# --- Boxplots by categorical predictors ---
p_gender <- df |>
  mutate(Gender = factor(female, levels = c(0, 1),
                         labels = c("Male", "Female"))) |>
  ggplot(aes(x = Gender, y = ln_inc, fill = Gender)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = NULL, y = "Log income", title = "Gender")

p_occ <- ggplot(df, aes(x = occupation, y = ln_inc, fill = occupation)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = NULL, y = "Log income", title = "Occupation")

p_gender + p_occ
```

### Regression Models {#sec-script-linear-models}

```{r}
#| label: script-linear-models
#| eval: false

# ============================================================
# PROJECT 1: LINEAR REGRESSION -- MODELS
# ============================================================

# --- Model 1: Traditional Mincer model (RQ1) ---
model1 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c,
             data = df)
summary(model1)

# --- Model 2: + Occupation and numeracy (RQ2) ---
model2 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               semiwhite + semiblue + elementary + numeracy,
             data = df)
summary(model2)

# --- Model 3: + Gender interactions (RQ3) ---
model3 <- lm(ln_inc ~ female + schyrs_c + exper_c + exper2_c +
               semiwhite + semiblue + elementary + numeracy +
               fem_schyrs_c + fem_numeracy,
             data = df)
summary(model3)

# --- Incremental F-tests ---
anova(model1, model2)  # Tests occupation + numeracy block
anova(model2, model3)  # Tests gender interaction block
```

### Publication-Style Regression Table {#sec-script-linear-table}

```{r}
#| label: script-linear-table
#| eval: false

# ============================================================
# PROJECT 1: LINEAR REGRESSION -- PUBLICATION TABLE
# ============================================================

t1 <- tbl_regression(
  model1, intercept = TRUE,
  label = list(
    female   ~ "Female",
    schyrs_c ~ "Education (years, centered)",
    exper_c  ~ "Experience (decades, centered)",
    exper2_c ~ "Experience squared"
  )
) |> bold_p(t = 0.05)

t2 <- tbl_regression(
  model2, intercept = TRUE,
  label = list(
    female     ~ "Female",
    schyrs_c   ~ "Education (years, centered)",
    exper_c    ~ "Experience (decades, centered)",
    exper2_c   ~ "Experience squared",
    semiwhite  ~ "Semi-white-collar",
    semiblue   ~ "Semi-blue-collar",
    elementary ~ "Elementary",
    numeracy   ~ "Numeracy (z-score)"
  )
) |> bold_p(t = 0.05)

t3 <- tbl_regression(
  model3, intercept = TRUE,
  label = list(
    female       ~ "Female",
    schyrs_c     ~ "Education (years, centered)",
    exper_c      ~ "Experience (decades, centered)",
    exper2_c     ~ "Experience squared",
    semiwhite    ~ "Semi-white-collar",
    semiblue     ~ "Semi-blue-collar",
    elementary   ~ "Elementary",
    numeracy     ~ "Numeracy (z-score)",
    fem_schyrs_c ~ "Female x Education",
    fem_numeracy ~ "Female x Numeracy"
  )
) |> bold_p(t = 0.05)

# --- Merge into one table ---
tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
)

# --- Coefficient plot (optional) ---
model3 |>
  tidy(conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Estimated Coefficient", y = NULL,
       title = "Coefficient Plot (Model 3)")
```

### Assumption Checks {#sec-script-linear-diagnostics}

```{r}
#| label: script-linear-diagnostics
#| eval: false

# ============================================================
# PROJECT 1: LINEAR REGRESSION -- DIAGNOSTICS
# ============================================================

# --- VIF (multicollinearity) ---
car::vif(model3) |>
  as.data.frame() |>
  setNames("VIF") |>
  knitr::kable(digits = 2)

# --- Residual diagnostics ---
resid_df <- data.frame(
  residuals = residuals(model3),
  fitted    = fitted(model3),
  std_resid = rstandard(model3)
)

# Histogram of residuals
p_hist <- ggplot(resid_df, aes(x = residuals)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white",
                 alpha = 0.8) +
  labs(x = "Residuals", y = "Count", title = "Histogram of Residuals")

# Q-Q plot
p_qq <- ggplot(resid_df, aes(sample = std_resid)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line(color = "red") +
  labs(x = "Theoretical quantiles", y = "Standardized residuals",
       title = "Q-Q Plot")

p_hist + p_qq

# Residuals vs. fitted values
ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(x = "Fitted values", y = "Residuals",
       title = "Residuals vs. Fitted Values")

# --- Comprehensive diagnostics with performance ---
performance::check_model(
  model3,
  check = c("linearity", "homogeneity", "qq", "outliers")
)
```


## Project 2: Logistic Regression {#sec-script-logistic}

### Descriptive Statistics {#sec-script-logistic-desc}

```{r}
#| label: script-logistic-descriptives
#| eval: false

# ============================================================
# PROJECT 2: LOGISTIC REGRESSION -- DESCRIPTIVE STATISTICS
# ============================================================

# --- Outcome distribution ---
table(df$manager) |> addmargins()
prop.table(table(df$manager)) |> round(3)

# --- Cross-tabulation: manager by gender ---
df |>
  mutate(
    Gender = factor(female, levels = c(0, 1),
                    labels = c("Male", "Female")),
    Status = factor(manager, levels = c(0, 1),
                    labels = c("Non-manager", "Manager"))
  ) |>
  count(Gender, Status) |>
  group_by(Gender) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup() |>
  select(Gender, Status, pct) |>
  tidyr::pivot_wider(names_from = Status, values_from = pct) |>
  kable(align = "lrr",
        col.names = c("Gender", "Non-manager (%)", "Manager (%)"))

# Chi-square test
chisq.test(table(df$manager, df$female))

# --- Cross-tabulation: manager by occupation ---
df |>
  mutate(
    Status = factor(manager, levels = c(0, 1),
                    labels = c("Non-manager", "Manager"))
  ) |>
  count(occupation, Status) |>
  group_by(occupation) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup() |>
  mutate(label = paste0(n, " (", pct, "%)")) |>
  select(occupation, Status, label) |>
  tidyr::pivot_wider(names_from = Status, values_from = label) |>
  kable(align = "lrr")

chisq.test(table(df$manager, df$occupation))

# --- Continuous variables by manager status ---
df |>
  mutate(
    Status = factor(manager, levels = c(0, 1),
                    labels = c("Non-manager", "Manager"))
  ) |>
  select(Status, schyrs, exper, numeracy) |>
  tbl_summary(
    by = Status,
    type = everything() ~ "continuous",
    statistic = all_continuous() ~ "{mean} ({sd})",
    label = list(
      schyrs   ~ "Years of education",
      exper    ~ "Work experience (years)",
      numeracy ~ "Numeracy (z-score)"
    ),
    digits = everything() ~ 2
  ) |>
  add_overall() |>
  add_p() |>
  bold_labels() |>
  bold_p(t = 0.05)

# --- Bar chart: manager status by gender ---
df |>
  mutate(
    Gender = factor(female, levels = c(0, 1),
                    labels = c("Male", "Female")),
    Status = factor(manager, levels = c(0, 1),
                    labels = c("Non-manager", "Manager"))
  ) |>
  count(Gender, Status) |>
  group_by(Gender) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = Gender, y = prop, fill = Status)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Gender", y = "Proportion", fill = "Status",
       title = "Supervisory Status by Gender")

# --- Boxplots: continuous predictors by manager status ---
p_edu <- df |>
  mutate(Status = factor(manager, levels = c(0, 1),
                         labels = c("Non-manager", "Manager"))) |>
  ggplot(aes(x = Status, y = schyrs, fill = Status)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = NULL, y = "Years of education", title = "Education")

p_exp <- df |>
  mutate(Status = factor(manager, levels = c(0, 1),
                         labels = c("Non-manager", "Manager"))) |>
  ggplot(aes(x = Status, y = exper, fill = Status)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = NULL, y = "Experience (years)", title = "Experience")

p_num <- df |>
  mutate(Status = factor(manager, levels = c(0, 1),
                         labels = c("Non-manager", "Manager"))) |>
  ggplot(aes(x = Status, y = numeracy, fill = Status)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = NULL, y = "Numeracy (z-score)", title = "Numeracy")

p_edu + p_exp + p_num
```

### Logistic Models {#sec-script-logistic-models}

```{r}
#| label: script-logistic-models
#| eval: false

# ============================================================
# PROJECT 2: LOGISTIC REGRESSION -- MODELS
# ============================================================

# --- Model 1: Baseline human capital model (RQ1) ---
logit1 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c,
              data = df, family = binomial)
summary(logit1)

# Odds ratios with 95% CI
exp(cbind(OR = coef(logit1), confint(logit1))) |> round(3)

# --- Model 2: + Occupation and numeracy (RQ2) ---
logit2 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                semiwhite + semiblue + elementary + numeracy,
              data = df, family = binomial)
summary(logit2)
exp(cbind(OR = coef(logit2), confint(logit2))) |> round(3)

# --- Model 3: + Gender interactions (RQ3) ---
logit3 <- glm(manager ~ female + schyrs_c + exper_c + exper2_c +
                semiwhite + semiblue + elementary + numeracy +
                fem_schyrs_c + fem_numeracy,
              data = df, family = binomial)
summary(logit3)
exp(cbind(OR = coef(logit3), confint(logit3))) |> round(3)

# --- Likelihood ratio tests ---
anova(logit1, logit2, test = "Chisq")  # Tests occupation + numeracy block
anova(logit2, logit3, test = "Chisq")  # Tests gender interaction block
```

### Publication-Style Odds Ratio Table {#sec-script-logistic-table}

```{r}
#| label: script-logistic-table
#| eval: false

# ============================================================
# PROJECT 2: LOGISTIC REGRESSION -- PUBLICATION TABLE
# ============================================================

t1 <- tbl_regression(
  logit1, exponentiate = TRUE,
  label = list(
    female   ~ "Female",
    schyrs_c ~ "Education (years, centered)",
    exper_c  ~ "Experience (decades, centered)",
    exper2_c ~ "Experience squared"
  )
) |> bold_p(t = 0.05)

t2 <- tbl_regression(
  logit2, exponentiate = TRUE,
  label = list(
    female     ~ "Female",
    schyrs_c   ~ "Education (years, centered)",
    exper_c    ~ "Experience (decades, centered)",
    exper2_c   ~ "Experience squared",
    semiwhite  ~ "Semi-white-collar",
    semiblue   ~ "Semi-blue-collar",
    elementary ~ "Elementary",
    numeracy   ~ "Numeracy (z-score)"
  )
) |> bold_p(t = 0.05)

t3 <- tbl_regression(
  logit3, exponentiate = TRUE,
  label = list(
    female       ~ "Female",
    schyrs_c     ~ "Education (years, centered)",
    exper_c      ~ "Experience (decades, centered)",
    exper2_c     ~ "Experience squared",
    semiwhite    ~ "Semi-white-collar",
    semiblue     ~ "Semi-blue-collar",
    elementary   ~ "Elementary",
    numeracy     ~ "Numeracy (z-score)",
    fem_schyrs_c ~ "Female x Education",
    fem_numeracy ~ "Female x Numeracy"
  )
) |> bold_p(t = 0.05)

# --- Merge into one table ---
tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
)
```

### Predicted Probabilities {#sec-script-logistic-predicted}

```{r}
#| label: script-logistic-predicted
#| eval: false

# ============================================================
# PROJECT 2: LOGISTIC REGRESSION -- PREDICTED PROBABILITIES
# ============================================================

# --- Predicted probability graph: education by gender ---
plot_predictions(logit2,
                 condition = list("schyrs_c", "female"),
                 type = "response") +
  labs(
    x = "Years of Education (centered)",
    y = "Predicted P(Manager)",
    color = "Gender", fill = "Gender",
    title = "Predicted Probability of Supervisory Status",
    subtitle = "By education level and gender (Model 2)"
  )

# --- Predicted probability table at selected values ---
pred_grid <- predictions(
  logit2,
  newdata = datagrid(
    female    = c(0, 1),
    schyrs_c  = c(-4, 0, 4),
    exper_c   = 0,
    exper2_c  = 0,
    semiwhite = 0,
    semiblue  = 0,
    elementary = 0,
    numeracy  = 0
  ),
  type = "response"
)

pred_grid |>
  as.data.frame() |>
  mutate(
    Gender    = ifelse(female == 0, "Male", "Female"),
    Education = case_when(
      schyrs_c == -4 ~ "Mean - 4 yrs",
      schyrs_c ==  0 ~ "Mean",
      schyrs_c ==  4 ~ "Mean + 4 yrs"
    ),
    `P(Manager)` = round(estimate, 3),
    `95% CI` = paste0("[", round(conf.low, 3), ", ",
                      round(conf.high, 3), "]")
  ) |>
  select(Gender, Education, `P(Manager)`, `95% CI`) |>
  kable(align = "llrr")

# --- Average marginal effects ---
avg_slopes(logit2)
```

### Model Fit & Diagnostics {#sec-script-logistic-diagnostics}

```{r}
#| label: script-logistic-diagnostics
#| eval: false

# ============================================================
# PROJECT 2: LOGISTIC REGRESSION -- FIT & DIAGNOSTICS
# ============================================================

# --- VIF (multicollinearity) ---
car::vif(logit3) |>
  as.data.frame() |>
  setNames("VIF") |>
  knitr::kable(digits = 2)

# --- Model fit summary ---
logit_null <- glm(manager ~ 1, data = df, family = binomial)
ll_null <- logLik(logit_null)

fit_table <- tibble::tibble(
  Model = c("Model 1", "Model 2", "Model 3"),
  AIC = c(AIC(logit1), AIC(logit2), AIC(logit3)),
  BIC = c(BIC(logit1), BIC(logit2), BIC(logit3)),
  `Log-Lik` = c(as.numeric(logLik(logit1)),
                as.numeric(logLik(logit2)),
                as.numeric(logLik(logit3))),
  `McFadden R2` = c(
    1 - as.numeric(logLik(logit1)) / as.numeric(ll_null),
    1 - as.numeric(logLik(logit2)) / as.numeric(ll_null),
    1 - as.numeric(logLik(logit3)) / as.numeric(ll_null)
  )
) |>
  mutate(across(AIC:BIC, ~ round(.x, 1)),
         `Log-Lik` = round(`Log-Lik`, 1),
         `McFadden R2` = round(`McFadden R2`, 3))

kable(fit_table, align = "lrrrr")
```
