# Common Pitfalls to Avoid {#sec-pitfalls}

{{< include ../_common.qmd >}}

::: {.callout-warning}
## Learn from Others' Mistakes
This appendix collects the mistakes that most often cost points on the final
project. Use it while drafting, not only at the end. The safest workflow is to
build one clear analysis dataset, fit planned models, use the course package
helpers for tables and diagnostics, and translate results into substantive
language.
:::

Items marked **Logistic-specific** apply only to Project 2. Items that mention
Project 1 or Project 2 are project-specific; the remaining items apply to both
projects. For the full expectations, see the linear guidelines in
@sec-guidelines-linear and the logistic guidelines in @sec-guidelines-logistic.
Use the submission checklists in @sec-linear-checklist and
@sec-logistic-checklist before submitting.


## Part 1: Background and Research Question {#sec-pitfalls-part1}

1. **Describing the data before motivating the question.** Part 1 should
   explain why the topic matters. Save dataset details, sample restrictions,
   variable names, and missing-data rules for Part 2.

2. **Writing a vague research question.** Bad: "What affects income?" Better:
   "Among Korean full-time workers, how are education, experience, occupation,
   numeracy, and gender associated with monthly earnings?" Name the outcome,
   focal predictors, and population.

3. **Logistic-specific: writing three separate RQs when the assignment asks
   for one.** Project 2 uses one focused logistic research question. You may
   still fit staged models, but those model stages support one overarching
   question rather than three disconnected numbered RQs.

4. **Stating only formal null hypotheses.** The project asks for substantive
   research questions. You may test hypotheses later, but Part 1 should tell
   the reader what relationship matters and why.

::: {.callout-caution}
## Critical Pitfall
If the research question does not specify the outcome and at least one
predictor, the reader cannot tell what regression you plan to run. For Project
2, also name the modeled event, such as `manager = 1`.
:::


## Part 2: Subjects and Variables {#sec-pitfalls-part2}

5. **Not defining categories for categorical variables.** List all categories
   for every categorical predictor. The reader should not have to infer them
   from a model table.

6. **Not specifying the reference category.** Every factor or dummy-coded
   predictor needs a reference group. Without a reference group, the
   coefficient or odds ratio cannot be interpreted.

7. **Letting R choose the reference category accidentally.** R uses the first
   factor level as the reference. If you create a factor from character values,
   the first level may be alphabetical unless you set it. Set the reference
   intentionally:

```{r}
#| eval: false

analysis_df <- raw_data |>
  mutate(occupation = forcats::fct_relevel(occupation, "Skilled"))

levels(analysis_df$occupation)
```

8. **Logistic-specific: not naming the event and non-event.** A binary outcome
   must be described as `1 = event` and `0 = non-event`. Every odds ratio,
   predicted probability, marginal effect, ROC/AUC value, sensitivity, and
   specificity refers to the event.

9. **Omitting the variable table.** A compact table with variable name, role,
   type, coding, and definition prevents confusion and saves space.

10. **Not defining the analytic sample.** Report the starting sample,
    exclusions, missing-data rule, final analytic $N$, and, for Project 2,
    event/non-event counts.


## Part 3: Statistical Analysis Plan {#sec-pitfalls-part3}

11. **Models that do not match the research question.** Each model should
    support the research question. For Project 1, that may mean multiple RQs
    and multiple planned model comparisons. For Project 2, it usually means a
    staged model sequence supporting one logistic RQ.

12. **Getting the regression direction backward.** You regress the outcome on
    predictors. In a formula, the outcome goes on the left: `y ~ x1 + x2`.

13. **Not stating the model clearly.** State the model in equation form or in
    precise words. The reader should be able to identify the outcome,
    predictors, interactions, and error or link function.

14. **Using different samples across nested models.** Model comparisons require
    the same rows. Build the complete-case analysis dataset before fitting the
    models:

```{r}
#| eval: false

analysis_vars <- c("y", "x1", "x2", "x3", "x4")

analysis_df <- raw_data |>
  tidyr::drop_na(dplyr::all_of(analysis_vars))
```

    Before a nested `anova()` or likelihood-ratio test, confirm that the
    compared models have the same `nobs()`.

15. **Adding terms just to search for significance.** Interactions,
    transformations, and polynomial terms should be motivated by the research
    question, theory, or clear diagnostic evidence.

16. **Writing a complex coding plan when a simple package helper is available.**
    The course workflow is intentionally compact: fit models with `lm()` or
    `glm()`, inspect estimates with `model_params()`, summarize fit with
    `model_fit()`, check collinearity with `performance::check_collinearity()`,
    build report tables with `gtsummary::tbl_regression()` or
    `modelsummary::modelsummary()`, use `ggeffects::predict_response()` for
    Project 1 adjusted predictions, and use `marginaleffects` for Project 2
    predicted probabilities, AMEs, and conditional effects.


## Part 4: Descriptive Analysis {#sec-pitfalls-part4}

17. **Using a different sample for descriptives and models.** Descriptives
    should be based on the same complete-case `analysis_df` used in the
    regression models.

18. **Reporting means and SDs for categorical variables.** For categorical
    variables, report counts and percentages. For a 0/1 logistic outcome, the
    mean is the event rate, but event/non-event counts and event percentages
    are clearer in the report. For continuous variables, report means and SDs,
    often by group.

19. **Logistic-specific: skipping sparse-cell and separation checks.** Before
    fitting a logistic model, check whether each important group has both
    events and non-events. Sparse cells can produce unstable estimates.

20. **Dropping figures without interpretation.** Every figure needs a sentence
    that says what the reader should notice. A plot is evidence only after you
    explain it.

21. **Using a visualization that does not match the outcome.** For continuous
    outcomes, histograms, scatterplots, and boxplots are often useful. For
    binary outcomes, bar charts of event rates and grouped summaries are often
    clearer.


## Part 5: Regression Results {#sec-pitfalls-part5}

::: {.callout-caution}
## Critical Pitfall
Do not paste raw model output into the report and call it interpretation. Every
model section needs a clean table, a fit or diagnostic statement, and plain
language explaining direction, size, uncertainty, and substantive meaning.
:::

22. **Pasting raw `summary()` output.** Use `model_params()` for inspection and
    `gtsummary::tbl_regression()` or `modelsummary::modelsummary()` for the
    report table. Raw console output belongs in your scratch work, not the
    final report. See the table templates in
    @sec-template-linear-single, @sec-template-linear-multi,
    @sec-template-logistic-single, and @sec-template-logistic-multi.

23. **Manually extracting results that packages already format.** Avoid
    hand-building coefficient, OR, or CI tables when `model_params()`,
    `tbl_regression()`, `tbl_merge()`, `model_fit()`, or `avg_slopes()` already
    gives the needed output.

24. **Not interpreting the final model.** Interpret every non-intercept
    coefficient that matters for the research question in the primary/final
    model. Interpret the intercept only when the reference and zero values have
    a meaningful baseline interpretation.

25. **Confusing statistical and practical significance.** A small coefficient
    can be statistically clear but practically minor. A large estimate can be
    uncertain. Discuss both size and uncertainty.

26. **Skipping diagnostics or treating them as a checkbox.** Diagnostics should
    lead to a short conclusion about model adequacy. For Project 1, address
    residual patterns, heteroscedasticity, normality, influence, and VIF. For
    Project 2, address sparse cells/separation, convergence, VIF, linearity in
    the logit, influence screening, and fit statistics.

27. **Forgetting the overall test for a multi-category predictor.** If a factor
    has three or more categories, interpret individual contrasts, but also
    report an overall block test such as a likelihood-ratio test.

28. **Misreading interaction terms.** Once an interaction is in the model, main
    effects are conditional. Use predicted values, predicted probabilities, or
    conditional-effect tables to explain the pattern.

29. **Logistic-specific: reporting log-odds as the main result.** Main tables
    should report odds ratios. Use `exponentiate = TRUE` in
    `model_params()` and `tbl_regression()`.

30. **Logistic-specific: interpreting odds ratios as probabilities.** An OR of
    2.0 does not mean "twice as likely." It means the odds are twice as large.
    Translate focal results to the probability scale with predicted
    probabilities, probability differences in percentage points, or AMEs.

31. **Logistic-specific: forgetting the event definition in results.** Write
    "predicted probability of `manager = 1`" or "odds of the modeled event,"
    not just "probability" or "odds."

32. **Logistic-specific: omitting the predicted-probability deliverables.**
    Project 2 requires a predicted-probability figure and a predicted
    probability table. AMEs from `avg_slopes()` are often the cleanest
    probability-scale summary.
    For copy-and-adapt code, see @sec-prediction-templates, especially
    @sec-template-logistic-prob-curve, @sec-template-logistic-prob-table, and
    @sec-template-logistic-ame.

33. **Logistic-specific: treating classification diagnostics as always
    required.** ROC/AUC, classification tables, sensitivity/specificity, and
    calibration plots are useful when the research question is about
    prediction. For explanatory projects, they are optional; if included, say
    what cutoff was used and what the diagnostic can and cannot show.


## Part 6: Discussion {#sec-pitfalls-part6}

34. **Drawing causal conclusions from observational data.** Unless the design
    supports causal inference, write "is associated with" rather than "causes,"
    "leads to," or "proves."

35. **Not answering the research question directly.** For Project 1, revisit
    each RQ. For Project 2, answer the single logistic RQ and explain how the
    model stages support that answer.

36. **Repeating the results table in prose.** The discussion should focus on
    meaning, practical significance, limitations, and implications. Do not
    re-list every coefficient.

37. **Discussing odds ratios when probability-scale results are clearer.**
    For Project 2, use predicted probabilities, probability differences, or
    AMEs to explain practical significance.

38. **Listing vague or incomplete limitations.** Every project needs 2--3
    concrete limitations tied to the actual analysis. Common limitations
    include cross-sectional design, omitted variables, measurement error,
    model specification, sample restrictions, and sparse subgroup estimates.
    For complex survey data or a supplied teaching subset, state whether
    sampling weights, replicate weights, plausible values, or design-based
    variance estimation were used. If not, frame ordinary standard errors,
    confidence intervals, p-values, $F$-tests, likelihood-ratio tests, and
    predicted probabilities as model-based summaries, and limit population
    claims accordingly.

39. **Ignoring diagnostics in the discussion.** If diagnostics are clean, say
    they support confidence in the main patterns. If a diagnostic flags a
    concern, describe how it limits the conclusion.


## Appendix and Reproducibility {#sec-pitfalls-appendix}

For appendix expectations, see @sec-linear-part7, @sec-logistic-part7, and
@sec-diagnostics-reproducibility-templates.

40. **Submitting code that only works in your current R session.** Restart R
    and run the script or render the Quarto document from top to bottom before
    submitting.

41. **Using absolute local paths.** Use project-relative paths or `here::here()`
    so the source file can run on another machine.

42. **Not loading packages in the source file.** The source file should load
    every package it uses. Do not rely on packages that happened to be loaded
    earlier in your session. If you use the course datasets, make sure your
    `regdatasets` installation matches the course setup.

43. **Leaving objects hidden in the global environment.** Every object used in
    a table, figure, model, prediction, or diagnostic should be created in the
    submitted source file.

44. **Not annotating the script.** Short comments should identify major blocks:
    load packages, construct `analysis_df`, descriptives, models, diagnostics,
    predictions, tables, figures, and session checks.

45. **Letting appendices contradict the main report.** The appendix script,
    table templates, and diagnostic templates should use the same model names,
    reference categories, event definitions, and package workflow as the main
    chapters.


## Quick Self-Check Before Submission {#sec-pitfalls-checklist}

Use this checklist before submitting each project:

- [ ] Research question(s) are specific and clearly stated
- [ ] Project 1 states 2--3 numbered research questions
- [ ] Project 2 states one focused binary-outcome research question
- [ ] Data source, sample restrictions, final analytic $N$, and missing-data
  rule are reported
- [ ] Every variable is defined, including categories and reference groups
- [ ] Project 2 names the modeled event and non-event
- [ ] Survey-design status is stated when using complex survey data or a
  supplied teaching subset
- [ ] Model(s) are stated clearly in equation form or precise words
- [ ] Descriptives use the same complete-case analysis dataset as the models
- [ ] Descriptive tables and exploratory figures are interpreted in prose
- [ ] Model code follows the planned sequence and uses the same sample
- [ ] Nested model comparisons use matching `nobs()` values
- [ ] Report tables are formatted with package helpers, not raw console output
- [ ] `model_params()`, `model_fit()`, `performance::check_collinearity()`,
  `gtsummary`, `ggeffects`, and `marginaleffects` are used where they simplify
  the work
- [ ] Every focal coefficient or odds ratio is interpreted substantively
- [ ] Linear assumptions or logistic adequacy diagnostics are checked and
  summarized, not merely printed
- [ ] Discussion answers the research question(s) directly
- [ ] Practical significance is explained, not only statistical significance
- [ ] At least 2--3 concrete limitations are stated, including survey-design
  or teaching-subset limits when relevant
- [ ] Code runs from a clean session using relative paths or `here::here()`
- [ ] **Logistic only:** Odds ratios, not raw log-odds, appear in the main
  table
- [ ] **Logistic only:** Odds and probabilities are not confused
- [ ] **Logistic only:** Predicted probability figure, predicted probability
  table, and probability-scale interpretation are included
