# Publication-Style Table Templates {#sec-table-templates}

{{< include ../_common.qmd >}}

::: {.callout-note}
## Ready-to-Use Table Templates
This appendix provides copy-and-adapt table templates for the final project.
The default reporting package is `gtsummary`, because it can produce clean
descriptive tables, cross-tabulations, regression tables, and merged model
tables with short code. Within each project, use these templates with the same
complete-case `analysis_df` used in your regression models.
:::

All code below is set to `eval: false`. Replace variable names, model names,
labels, captions, and reference categories with your own.


## General Rules for Tables {#sec-template-rules}

Before using any template:

1. Create one complete-case analysis dataset for the project.
2. Set factor reference categories before fitting models.
3. Fit the models first; then pass model objects to table functions.
4. Use readable labels so the reader does not see raw variable names.
5. Interpret the table in prose. A table is not a substitute for explanation.

```{r}
#| label: template-table-setup
#| eval: false

library(dplyr)
library(forcats)
library(gtsummary)
library(gt)
library(knitr)
library(modelsummary)
library(performance)
library(tibble)

round_numeric <- function(x, digits = 3) {
  dplyr::mutate(
    x,
    dplyr::across(dplyr::where(is.numeric), \(col) round(col, digits))
  )
}

model_fit <- function(model, metrics = "common", ...) {
  performance::model_performance(model, metrics = metrics, ...)
}

analysis_vars <- c("outcome", "predictor1", "predictor2", "occupation")

prepared_data <- raw_data |>
  mutate(occupation = fct_relevel(occupation, "Skilled"))

analysis_df <- prepared_data |>
  tidyr::drop_na(dplyr::all_of(analysis_vars))

levels(analysis_df$occupation)
```

::: {.callout-tip}
## Reference Categories
Avoid putting `factor(variable)` directly inside the model formula for final
tables. It is clearer to create a factor in the dataset, set the reference
category intentionally, check `levels()`, and then use the factor variable in
the model.
:::


## Template A: Analysis Sample Summary {#sec-template-sample}

Use this compact table in Part 2 to report the starting sample, complete-case
rule, final analytic sample, reference categories, and event counts for Project
2.

```{r}
#| label: template-sample-summary
#| eval: false

sample_summary <- tibble(
  `Starting N` = nrow(prepared_data),
  `Rows excluded for missing analysis variables` =
    sum(!complete.cases(prepared_data[analysis_vars])),
  `Final analytic N` = nrow(analysis_df),
  `Occupation reference` = levels(analysis_df$occupation)[1],
  `Event count` = sum(analysis_df$binary_outcome == 1),
  `Non-event count` = sum(analysis_df$binary_outcome == 0),
  `Event rate` = scales::percent(mean(analysis_df$binary_outcome == 1), accuracy = 0.1)
)

knitr::kable(
  sample_summary,
  caption = "Analysis sample and modeled-event summary"
)
```

**Use in report:** For Project 1, remove the event/non-event columns. For
Project 2, replace `binary_outcome` with your actual 0/1 outcome and name the
modeled event in the surrounding text.


## Template B: Variable and Coding Table {#sec-template-variables}

Use this table in Part 2 so the reader can see every variable, its role, type,
coding, reference category, and definition.

```{r}
#| label: template-variable-table
#| eval: false

variable_table <- tibble::tribble(
  ~Variable, ~Role, ~Type, ~Coding, ~Definition,
  "`outcome`", "Outcome", "Continuous", "Original units",
  "Brief definition of the outcome",
  "`predictor1`", "Predictor", "Continuous", "Mean-centered",
  "Brief definition of predictor 1",
  "`occupation`", "Predictor", "Categorical",
  "Skilled (ref), Semi-white, Semi-blue, Elementary",
  "Broad occupational classification",
  "`binary_outcome`", "Outcome", "Binary",
  "1 = event; 0 = non-event",
  "Brief definition of the modeled event"
)

knitr::kable(
  variable_table,
  align = "lllll",
  caption = "Variables used in the regression analysis"
)
```

**Use in report:** Delete rows that do not apply. For every categorical
predictor, name the reference category.


## Template 1: Descriptive Statistics {#sec-template-descriptives}

Use this template for the main descriptive table in Project 1 or Project 2.
It reports means and SDs for continuous variables and counts/percentages for
categorical variables.

```{r}
#| label: template-descriptives
#| eval: false

analysis_df |>
  select(outcome_var, continuous_pred1, continuous_pred2,
         categorical_pred1, categorical_pred2) |>
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd}); [{min}, {max}]",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      outcome_var ~ "Outcome Variable Name",
      continuous_pred1 ~ "Continuous Predictor 1",
      continuous_pred2 ~ "Continuous Predictor 2",
      categorical_pred1 ~ "Categorical Predictor 1",
      categorical_pred2 ~ "Categorical Predictor 2"
    ),
    missing = "no"
  ) |>
  bold_labels() |>
  modify_caption("**Table 1.** Descriptive statistics for the analysis sample")
```

**Use in report:** State that the table uses the complete-case `analysis_df`.
Mention final $N$ and any important distributional features in prose.


## Template 2: Descriptives by Outcome or Group {#sec-template-by-group}

Use this template when descriptives are clearer by group. For Project 2, the
group is usually the binary outcome status, such as `manager = 1` versus
`manager = 0`.

```{r}
#| label: template-by-group
#| eval: false

analysis_df |>
  select(group_var, predictor1, predictor2, predictor3, categorical_pred1) |>
  tbl_summary(
    by = group_var,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      predictor1 ~ "Predictor 1 Name",
      predictor2 ~ "Predictor 2 Name",
      predictor3 ~ "Predictor 3 Name",
      categorical_pred1 ~ "Categorical Predictor Name"
    ),
    missing = "no"
  ) |>
  add_overall() |>
  bold_labels() |>
  modify_caption("**Table 2.** Descriptive statistics by group")
```

::: {.callout-tip}
## Optional Group Tests
If you include simple group-comparison p-values, add `add_p()` after
`add_overall()`. Treat those p-values as descriptive context, not as a
replacement for regression results.
:::


## Template 3: Logistic Cross-Tabulation and Sparse-Cell Check {#sec-template-cross}

Use this template in Project 2 to check whether a categorical predictor has
both events and non-events in every important category.

```{r}
#| label: template-cross
#| eval: false

analysis_df |>
  tbl_cross(
    row = categorical_predictor,
    col = binary_outcome_status,
    percent = "row"
  ) |>
  bold_labels() |>
  modify_caption(
    "**Table 3.** Event and non-event counts by categorical predictor"
  )
```

**Use in report:** Look for categories with very small counts or categories
that contain only events or only non-events. Those patterns can create unstable
logistic estimates.


## Template 4: Linear Regression Table (Single Model) {#sec-template-linear-single}

Use this template for one multiple linear regression model.

```{r}
#| label: template-linear-single
#| eval: false

linear_model <- lm(
  outcome ~ predictor1 + predictor2 + categorical_predictor,
  data = analysis_df
)

tbl_regression(
  linear_model,
  intercept = TRUE,
  label = list(
    predictor1 ~ "Predictor 1 Name",
    predictor2 ~ "Predictor 2 Name",
    categorical_predictor ~ "Categorical Predictor"
  )
) |>
  add_global_p(include = c(categorical_predictor), keep = TRUE) |>
  modify_column_unhide(columns = std.error) |>
  add_glance_source_note(
    include = c(nobs, r.squared, adj.r.squared, statistic, p.value, sigma)
  ) |>
  bold_labels() |>
  bold_p(t = 0.05) |>
  modify_caption("**Table 4.** Multiple linear regression results")
```

**Use in report:** Interpret non-intercept coefficients that matter for the
research question. Interpret the intercept only if the reference and zero
values are meaningful.


## Template 5: Linear Regression Table (Multiple Models) {#sec-template-linear-multi}

Use this template for planned nested linear models. Fit all models on the same
complete-case `analysis_df`, check matching `nobs()`, then merge the tables.

```{r}
#| label: template-linear-multi
#| eval: false

model1 <- lm(outcome ~ predictor1 + predictor2, data = analysis_df)

model2 <- lm(
  outcome ~ predictor1 + predictor2 + categorical_predictor,
  data = analysis_df
)

model3 <- lm(
  outcome ~ predictor1 + predictor2 + categorical_predictor +
    predictor1:moderator,
  data = analysis_df
)

model_n <- tibble(
  Model = c("Model 1", "Model 2", "Model 3"),
  N = c(nobs(model1), nobs(model2), nobs(model3))
)

knitr::kable(model_n, caption = "Model sample sizes")

stopifnot(
  nobs(model1) == nobs(model2),
  nobs(model2) == nobs(model3)
)

labels_base <- list(
  predictor1 ~ "Predictor 1",
  predictor2 ~ "Predictor 2"
)

labels_extended <- c(
  labels_base,
  list(categorical_predictor ~ "Categorical Predictor")
)

labels_interaction <- c(
  labels_extended,
  list(`predictor1:moderator` ~ "Predictor 1 x Moderator")
)

t1 <- tbl_regression(model1, intercept = TRUE, label = labels_base) |>
  bold_p(t = 0.05)

t2 <- tbl_regression(model2, intercept = TRUE, label = labels_extended) |>
  add_global_p(include = c(categorical_predictor), keep = TRUE) |>
  bold_p(t = 0.05)

t3 <- tbl_regression(model3, intercept = TRUE, label = labels_interaction) |>
  add_global_p(include = c(categorical_predictor), keep = TRUE) |>
  bold_p(t = 0.05)

tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
) |>
  modify_caption("**Table 5.** Nested linear regression models")

# Planned nested-model comparisons to report in text.
anova(model1, model2)
anova(model2, model3)
```

**Use in report:** Use the table to compare coefficient patterns across models.
Use the incremental F-tests in prose to summarize planned model comparisons.


## Template 6: Logistic Regression Table (Single Model, Odds Ratios) {#sec-template-logistic-single}

Use this template for one logistic regression model. The table reports odds
ratios, not raw log-odds.

```{r}
#| label: template-logistic-single
#| eval: false

logit_model <- glm(
  binary_outcome ~ predictor1 + predictor2 + categorical_predictor,
  data = analysis_df,
  family = binomial
)

tbl_regression(
  logit_model,
  exponentiate = TRUE,
  label = list(
    predictor1 ~ "Predictor 1 Name",
    predictor2 ~ "Predictor 2 Name",
    categorical_predictor ~ "Categorical Predictor"
  )
) |>
  add_global_p(include = c(categorical_predictor), keep = TRUE) |>
  add_glance_source_note(
    include = c(nobs, AIC, BIC, logLik)
  ) |>
  bold_labels() |>
  bold_p(t = 0.05) |>
  modify_caption("**Table 6.** Logistic regression results (odds ratios)")
```

::: {.callout-caution}
## Logistic Table Reminder
Always use `exponentiate = TRUE` for final logistic regression tables. Then
translate focal results into predicted probabilities, probability differences,
or AMEs in the text or in the prediction templates.
:::


## Template 7: Staged Logistic Regression Table (Odds Ratios) {#sec-template-logistic-multi}

Use this template for Project 2's staged model sequence. The stages support
one focused logistic research question; they are not separate RQ1/RQ2/RQ3
questions.

```{r}
#| label: template-logistic-multi
#| eval: false

logit1 <- glm(
  binary_outcome ~ predictor1 + predictor2,
  data = analysis_df,
  family = binomial
)

logit2 <- glm(
  binary_outcome ~ predictor1 + predictor2 + categorical_predictor,
  data = analysis_df,
  family = binomial
)

logit3 <- glm(
  binary_outcome ~ predictor1 + predictor2 + categorical_predictor +
    predictor1:moderator,
  data = analysis_df,
  family = binomial
)

stage_map <- tibble::tribble(
  ~Stage, ~Predictors_added, ~Purpose,
  "Stage 1", "Baseline predictors", "Establish baseline adjusted association",
  "Stage 2", "Add predictor block", "Test whether the block improves fit",
  "Stage 3", "Add interaction", "Check whether the focal association differs by group"
)

knitr::kable(stage_map, caption = "Purpose of the staged logistic model sequence")

logit_n <- tibble(
  Stage = c("Stage 1", "Stage 2", "Stage 3"),
  N = c(nobs(logit1), nobs(logit2), nobs(logit3))
)

knitr::kable(logit_n, caption = "Logistic model sample sizes")

stopifnot(
  nobs(logit1) == nobs(logit2),
  nobs(logit2) == nobs(logit3)
)

logit_labels_base <- list(
  predictor1 ~ "Predictor 1",
  predictor2 ~ "Predictor 2"
)

logit_labels_extended <- c(
  logit_labels_base,
  list(categorical_predictor ~ "Categorical Predictor")
)

logit_labels_interaction <- c(
  logit_labels_extended,
  list(`predictor1:moderator` ~ "Predictor 1 x Moderator")
)

t1 <- tbl_regression(
  logit1,
  exponentiate = TRUE,
  label = logit_labels_base
) |>
  bold_p(t = 0.05)

t2 <- tbl_regression(
  logit2,
  exponentiate = TRUE,
  label = logit_labels_extended
) |>
  add_global_p(include = c(categorical_predictor), keep = TRUE) |>
  bold_p(t = 0.05)

t3 <- tbl_regression(
  logit3,
  exponentiate = TRUE,
  label = logit_labels_interaction
) |>
  add_global_p(include = c(categorical_predictor), keep = TRUE) |>
  bold_p(t = 0.05)

tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Stage 1**", "**Stage 2**", "**Stage 3**")
) |>
  modify_caption("**Table 7.** Staged logistic regression models (odds ratios)")

# Planned likelihood-ratio tests to report in text.
anova(logit1, logit2, test = "Chisq")
anova(logit2, logit3, test = "Chisq")
```

**Use in report:** Name the modeled event in the table title or table note, for
example "odds ratios for `manager = 1`."


## Template 8: Linear Model Fit Summary Table {#sec-template-fit}

Use this template when linear model fit statistics are clearer in a separate
table.

```{r}
#| label: template-fit-table
#| eval: false

fit_table <- bind_rows(
  model_fit(model1) |> as_tibble() |> mutate(Model = "Model 1"),
  model_fit(model2) |> as_tibble() |> mutate(Model = "Model 2"),
  model_fit(model3) |> as_tibble() |> mutate(Model = "Model 3")
) |>
  select(Model, n_Obs, R2, R2_adjusted, RMSE, AIC, BIC) |>
  round_numeric(digits = 3)

knitr::kable(
  fit_table,
  align = "lrrrrrr",
  caption = "Model fit statistics for the planned model sequence"
)
```


## Template 9: Logistic Model Fit Summary Table {#sec-template-logistic-fit}

Use this template for the Project 2 fit table. It reports required fit evidence
such as N, log-likelihood, AIC, and pseudo-$R^2$, plus recommended indices such
as BIC, Tjur's R2, and RMSE when available from `model_fit()`.

```{r}
#| label: template-logistic-fit-table
#| eval: false

logistic_fit_table <- tibble(
  Model = c("Stage 1", "Stage 2", "Stage 3"),
  N = c(nobs(logit1), nobs(logit2), nobs(logit3)),
  `Log likelihood` = c(logLik(logit1), logLik(logit2), logLik(logit3)) |>
    as.numeric()
) |>
  bind_cols(
    bind_rows(
      model_fit(logit1) |> as_tibble(),
      model_fit(logit2) |> as_tibble(),
      model_fit(logit3) |> as_tibble()
    )
  ) |>
  select(Model, N, `Log likelihood`, AIC, BIC, `Tjur R2` = R2_Tjur, RMSE) |>
  round_numeric(digits = 3)

knitr::kable(
  logistic_fit_table,
  align = "lrrrrrr",
  caption = "Model fit statistics for the staged logistic model sequence"
)
```


## Template 10: Optional `modelsummary()` Comparison Table {#sec-template-modelsummary}

`gtsummary` is the default final-project table package. `modelsummary()` is an
acceptable optional alternative for compact side-by-side model tables.

```{r}
#| label: template-modelsummary
#| eval: false

model_list <- list(
  "Model 1" = model1,
  "Model 2" = model2,
  "Model 3" = model3
)

coef_map <- c(
  "(Intercept)" = "Intercept",
  "predictor1" = "Predictor 1",
  "predictor2" = "Predictor 2",
  "categorical_predictorLevel 2" = "Category: Level 2 vs Reference",
  "predictor1:moderator" = "Predictor 1 x Moderator"
)

modelsummary::modelsummary(
  model_list,
  coef_map = coef_map,
  gof_map = c("nobs", "r.squared", "adj.r.squared", "rmse", "aic", "bic"),
  statistic = c("SE = {std.error}", "p = {p.value}"),
  fmt = 3,
  output = "markdown"
)
```

**Use in report:** If you use `modelsummary()`, check the row labels carefully.
Interaction and factor-contrast names must match the exact term names in your
model.


## Tips for Better Tables {#sec-template-tips}

::: {.callout-tip}
## Table Formatting Best Practices

1. **Use the analysis sample.** Descriptive and regression tables should use
   the same complete-case dataset.
2. **Label variables.** Replace raw variable names with readable labels.
3. **State reference groups.** The table or surrounding text must identify the
   reference category for each categorical predictor.
4. **Use term-level tests for factors.** For categorical predictors with three
   or more categories, add `add_global_p()` or report a planned nested-model
   test in the text.
5. **Do not overinterpret bold text.** `bold_p()` is a visual aid, not a
   substitute for substantive interpretation.
6. **Keep logistic tables on the odds-ratio scale.** Then explain practical
   significance with probability-scale results.
:::

::: {.callout-note}
## Where Predictions and AMEs Go
Predicted probabilities, adjusted predictions, probability tables, and average
marginal effects are important, but they are handled in the prediction and
marginal-effect templates in @sec-prediction-templates. Appendix B also
includes working examples of predicted-probability and AME tables. Keep this
appendix focused on descriptive, regression, and model-fit tables.
:::
