# Publication-Style Table Templates {#sec-table-templates}

{{< include ../_common.qmd >}}

::: {.callout-note}
## Ready-to-Use Templates
This appendix provides `gtsummary` code templates for the most common tables
in a regression project. Copy, paste, and adapt for your own data. Replace
`your_data`, variable names, and labels with your own.
:::

All code below is set to `eval: false` --- these are templates, not executed
code. Each template includes a brief description, the code, and a note about
what the output should look like.


## Template 1: Descriptive Statistics (Continuous + Categorical) {#sec-template-descriptives}

Use this template when you need a single table that summarizes both
continuous variables (mean, SD) and categorical variables (count, %).

```{r}
#| label: template-descriptives
#| eval: false

your_data |>
  select(outcome_var, continuous_pred1, continuous_pred2,
         categorical_pred1, categorical_pred2) |>
  tbl_summary(
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      outcome_var       ~ "Outcome Variable Name",
      continuous_pred1  ~ "Continuous Predictor 1",
      continuous_pred2  ~ "Continuous Predictor 2",
      categorical_pred1 ~ "Categorical Predictor 1",
      categorical_pred2 ~ "Categorical Predictor 2"
    ),
    missing = "no"
  ) |>
  bold_labels() |>
  modify_header(label = "**Variable**") |>
  modify_caption("**Table 1.** Descriptive Statistics")
```

**Expected output:** A single table with one row per continuous variable
(showing mean and SD) and one row per *level* of each categorical variable
(showing count and percentage). Continuous and categorical variables are
automatically grouped.

::: {.callout-tip}
## Tip
To add min/max for continuous variables, change the statistic to
`"{mean} ({sd}); [{min}, {max}]"`. To show median and IQR instead, use
`"{median} ({p25}, {p75})"`.
:::


## Template 2: Descriptive Statistics by Group {#sec-template-by-group}

Use this template when you want to compare descriptive statistics across
groups (e.g., male vs. female, manager vs. non-manager). This is especially
useful for the logistic regression project, where you compare continuous
predictors across outcome groups.

```{r}
#| label: template-by-group
#| eval: false

your_data |>
  select(group_var, predictor1, predictor2, predictor3) |>
  tbl_summary(
    by = group_var,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      predictor1 ~ "Predictor 1 Name",
      predictor2 ~ "Predictor 2 Name",
      predictor3 ~ "Predictor 3 Name"
    )
  ) |>
  add_overall() |>
  add_p() |>
  bold_labels() |>
  bold_p(t = 0.05) |>
  modify_caption("**Table 2.** Descriptive Statistics by Group")
```

**Expected output:** A table with one column per group, plus an "Overall"
column and a *p*-value column. Continuous variables show *t*-test or
Wilcoxon *p*-values; categorical variables show chi-square *p*-values.
Significant *p*-values are bolded.


## Template 3: Linear Regression Table (Single Model) {#sec-template-linear-single}

Use this template for a single multiple linear regression model. Includes
the intercept, all predictor coefficients with 95% CIs, and model fit
statistics in a footnote.

```{r}
#| label: template-linear-single
#| eval: false

lm(outcome ~ predictor1 + predictor2 + predictor3 + factor(categorical_var),
   data = your_data) |>
  tbl_regression(
    intercept = TRUE,
    label = list(
      predictor1             ~ "Predictor 1 Name",
      predictor2             ~ "Predictor 2 Name",
      predictor3             ~ "Predictor 3 Name",
      `factor(categorical_var)` ~ "Categorical Variable"
    )
  ) |>
  bold_labels() |>
  bold_p(t = 0.05) |>
  add_glance_source_note(
    include = c(r.squared, adj.r.squared, statistic, p.value, nobs)
  ) |>
  modify_caption("**Table 3.** Multiple Linear Regression Results")
```

**Expected output:** A table with columns for Variable, Beta (coefficient),
95% CI, and *p*-value. The intercept is included. Below the table, a
footnote reports $R^2$, adjusted $R^2$, the overall $F$-statistic, its
*p*-value, and $N$.

::: {.callout-tip}
## Tip
If your categorical variable is already a factor in the data frame, you do
not need `factor()` in the formula, and the label key should match the
variable name directly (e.g., `occupation ~ "Occupation"`).
:::

> **Note:** If your categorical variable has a specific reference category, set it with `forcats::fct_relevel(var, "reference_level")` before fitting the model.


## Template 4: Linear Regression Table (Multiple Models Side-by-Side) {#sec-template-linear-multi}

Use this template to compare nested linear regression models. This is the
standard format for presenting a model-building strategy.

```{r}
#| label: template-linear-multi
#| eval: false

# --- Fit models ---
model1 <- lm(outcome ~ predictor1 + predictor2, data = your_data)
model2 <- lm(outcome ~ predictor1 + predictor2 + predictor3,
             data = your_data)
model3 <- lm(outcome ~ predictor1 + predictor2 + predictor3 +
               predictor1:moderator, data = your_data)

# --- Create individual tables ---
t1 <- tbl_regression(
  model1, intercept = TRUE,
  label = list(
    predictor1 ~ "Predictor 1",
    predictor2 ~ "Predictor 2"
  )
) |> bold_p(t = 0.05)

t2 <- tbl_regression(
  model2, intercept = TRUE,
  label = list(
    predictor1 ~ "Predictor 1",
    predictor2 ~ "Predictor 2",
    predictor3 ~ "Predictor 3"
  )
) |> bold_p(t = 0.05)

t3 <- tbl_regression(
  model3, intercept = TRUE,
  label = list(
    predictor1              ~ "Predictor 1",
    predictor2              ~ "Predictor 2",
    predictor3              ~ "Predictor 3",
    `predictor1:moderator`  ~ "Predictor 1 x Moderator"
  )
) |> bold_p(t = 0.05)

# --- Merge into one table ---
tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
) |>
  modify_caption("**Table 4.** Nested Linear Regression Models")
```

**Expected output:** A wide table with three sets of columns (Beta, 95% CI,
*p*), one set per model. Variables that appear in later models but not
earlier ones have blank cells in the earlier model columns. This format
makes it easy to see how coefficients change as predictors are added.

::: {.callout-tip}
## Tip
To include $R^2$ and $N$ at the bottom of the merged table, add
`add_glance_source_note()` to each individual `tbl_regression()` *before*
merging. The footnotes will appear below each model column.
:::


## Template 5: Logistic Regression Table with Odds Ratios {#sec-template-logistic-single}

Use this template for a single logistic regression model. The key difference
from the linear regression template is `exponentiate = TRUE`, which
converts log-odds coefficients to odds ratios.

```{r}
#| label: template-logistic-single
#| eval: false

glm(binary_outcome ~ predictor1 + predictor2 + predictor3 +
      factor(categorical_var),
    data = your_data, family = binomial) |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      predictor1             ~ "Predictor 1 Name",
      predictor2             ~ "Predictor 2 Name",
      predictor3             ~ "Predictor 3 Name",
      `factor(categorical_var)` ~ "Categorical Variable"
    )
  ) |>
  bold_labels() |>
  bold_p(t = 0.05) |>
  add_glance_source_note(
    include = c(AIC, BIC, logLik, nobs)
  ) |>
  modify_caption("**Table 5.** Logistic Regression Results (Odds Ratios)")
```

**Expected output:** A table with columns for Variable, OR (odds ratio),
95% CI, and *p*-value. Odds ratios greater than 1 indicate increased odds;
less than 1 indicate decreased odds. CIs that include 1.00 correspond to
nonsignificant effects. The footnote reports AIC, BIC, log-likelihood,
and $N$.

::: {.callout-caution}
## Common Mistake
Forgetting `exponentiate = TRUE` will produce a table of log-odds
coefficients, which are much harder to interpret. Always use
`exponentiate = TRUE` for logistic regression tables in your final report.
:::


## Template 6: Logistic Regression Table (Multiple Models Side-by-Side) {#sec-template-logistic-multi}

Use this template to compare nested logistic regression models, all
reported as odds ratios.

```{r}
#| label: template-logistic-multi
#| eval: false

# --- Fit models ---
logit1 <- glm(binary_outcome ~ predictor1 + predictor2,
              data = your_data, family = binomial)
logit2 <- glm(binary_outcome ~ predictor1 + predictor2 + predictor3,
              data = your_data, family = binomial)
logit3 <- glm(binary_outcome ~ predictor1 + predictor2 + predictor3 +
                predictor1:moderator,
              data = your_data, family = binomial)

# --- Create individual tables ---
t1 <- tbl_regression(
  logit1, exponentiate = TRUE,
  label = list(
    predictor1 ~ "Predictor 1",
    predictor2 ~ "Predictor 2"
  )
) |> bold_p(t = 0.05)

t2 <- tbl_regression(
  logit2, exponentiate = TRUE,
  label = list(
    predictor1 ~ "Predictor 1",
    predictor2 ~ "Predictor 2",
    predictor3 ~ "Predictor 3"
  )
) |> bold_p(t = 0.05)

t3 <- tbl_regression(
  logit3, exponentiate = TRUE,
  label = list(
    predictor1              ~ "Predictor 1",
    predictor2              ~ "Predictor 2",
    predictor3              ~ "Predictor 3",
    `predictor1:moderator`  ~ "Predictor 1 x Moderator"
  )
) |> bold_p(t = 0.05)

# --- Merge into one table ---
tbl_merge(
  tbls = list(t1, t2, t3),
  tab_spanner = c("**Model 1**", "**Model 2**", "**Model 3**")
) |>
  modify_caption("**Table 6.** Nested Logistic Regression Models (Odds Ratios)")

# --- Likelihood ratio tests (report in text) ---
anova(logit1, logit2, test = "Chisq")
anova(logit2, logit3, test = "Chisq")
```

**Expected output:** A wide table with three sets of columns (OR, 95% CI,
*p*), one set per model. The format is identical to the linear regression
multi-model table except that coefficients are odds ratios instead of
unstandardized betas.


## Tips for Better Tables {#sec-template-tips}

::: {.callout-tip}
## Table Formatting Best Practices

1. **Always label variables.** Use the `label` argument to replace raw
   variable names with readable labels. Your reader should never see
   `schyrs_c` --- they should see "Education (years, centered)."
2. **Round appropriately.** Two decimal places for most statistics; three
   for correlations and *p*-values.
3. **Bold key results.** Use `bold_p()` to highlight significant predictors
   and `bold_labels()` to make variable names stand out.
4. **Add model fit notes.** Use `add_glance_source_note()` to include $R^2$,
   AIC, or sample size below the table.
5. **Number your tables.** Use `modify_caption()` to add "Table 1.",
   "Table 2.", etc. Consistent numbering helps the reader navigate your
   report.
6. **Be consistent.** Use the same formatting conventions across all tables
   in your project --- same number of decimal places, same label style, same
   significance threshold.
:::

::: {.callout-note}
## Why `gtsummary`?
The `gtsummary` package produces tables that render cleanly in HTML, PDF,
and Word output from Quarto. It also standardizes the format across
different model types (linear, logistic, ordinal, etc.), so once you learn
the template for one type, you can easily adapt it for others.
:::
