# Diagnostics and Reproducibility Templates {#sec-diagnostics-reproducibility-templates}

{{< include ../_common.qmd >}}

::: {.callout-note}
## Why This Appendix Exists
Diagnostics and reproducibility are required, but they should not take over the
main report. Use the main text to state the diagnostic conclusion, use the
appendix for bulky plots and supporting checks, and submit the complete source
file that recreates your work from a clean R session.
:::

::: {.callout-tip}
## Course Review Links
For linear diagnostic concepts, review [Model Diagnostics](https://joonho112.github.io/jlee-regression-course/notes/chapters/08-model-diagnostics.html) and [Lab 08](https://joonho112.github.io/jlee-regression-course/labs/labs/lab08-diagnostics.html). For logistic adequacy, sparse cells, separation, and optional prediction diagnostics, review [Logistic Regression Diagnostics](https://joonho112.github.io/jlee-regression-course/notes/chapters/11-logistic-diagnostics.html) and [Lab 11](https://joonho112.github.io/jlee-regression-course/labs/labs/lab11-logistic-diagnostics.html).
:::


## What Goes Where? {#sec-diagnostics-placement}

Use this table when deciding whether a diagnostic belongs in the main text,
appendix, or optional extra material.

| Topic | Project | Required? | Main text | Appendix |
|---|---|---:|---|---|
| Analysis sample and missingness | Both | Yes | Final analytic $N$ and missing-data rule | Code that constructs `analysis_df` |
| Same-sample nested models | Both | Yes | State that compared models use the same rows | `nobs()` checks before model comparisons |
| Collinearity/VIF | Both | Yes | Short VIF conclusion and any flagged predictors | VIF table from `performance::check_collinearity()` |
| Residuals vs fitted | Linear | Yes | Brief conclusion about linearity and spread | Plot and, if needed, sensitivity notes |
| Q-Q plot or residual distribution | Linear | Yes | Brief conclusion about residual normality limits | Q-Q plot or histogram |
| Influence/Cook's distance | Linear | Yes | State whether conclusions are sensitive to influential cases | Top cases or influence plot |
| Heteroscedasticity or robust-SE sensitivity | Linear | As needed | State whether inference changes | Test output or HC3 sensitivity table |
| Independence/design concerns | Linear | As needed | State clustering, repeated measures, or dependence limits | Notes or sensitivity checks if relevant |
| Event/non-event counts | Logistic | Yes | Event definition and event rate | Outcome table and sparse-cell checks |
| Sparse cells/separation | Logistic | Yes | State whether any sparse cells threaten estimation | Cross-tabs and event counts by key groups |
| Convergence | Logistic | Yes | State whether the model converged | `model$converged` checks |
| Linearity in the logit | Logistic | Yes | Brief conclusion for continuous predictors | Squared-term, binned, or other planned check |
| Influence/Cook's distance | Logistic | Yes | State whether flagged cases change conclusions | Cook's-distance summary or sensitivity table |
| Model fit summary | Logistic | Yes | AIC, log-likelihood, pseudo-$R^2$, or LR-test conclusion; BIC/Tjur's $R^2$/RMSE when useful | Fit table and nested model comparison output |
| Classification, ROC/AUC, calibration | Logistic | Optional | Include only if the question is predictive | Supporting output and cutoff justification |

::: {.callout-tip}
## One-Sentence Rule
Every diagnostic in the appendix should have a one-sentence conclusion in the
main report. A plot or table without interpretation does not show that you used
the diagnostic.
:::

Put the conclusion in the report; put the evidence trail in the appendix. In
the report, write 2--4 sentences saying whether diagnostics support the model
and naming any concern. In the appendix, include the plots, VIF table, influence
summary, sparse-cell checks, and source code that produced them.


## Setup for Diagnostic Templates {#sec-diagnostic-template-setup}

These templates assume that you already created `analysis_df` and fit the
models used in the report. Replace model names and variable names with your own.

```{r}
#| label: diagnostic-template-setup
#| eval: false

options(repos = c(CRAN = "https://cloud.r-project.org"))

library(dplyr)
library(ggplot2)
library(knitr)
library(performance)
library(scales)
library(tibble)

round_numeric <- function(x, digits = 3) {
  dplyr::mutate(
    x,
    dplyr::across(dplyr::where(is.numeric), \(col) round(col, digits))
  )
}

format_p <- function(p) {
  dplyr::if_else(
    is.na(p),
    NA_character_,
    dplyr::if_else(p < .001, "< .001", sprintf("%.3f", p))
  )
}
```


## Template 0: Same-Sample Checks {#sec-template-same-sample-checks}

Run this before nested model comparisons. If the `nobs()` values differ, rebuild
`analysis_df` using every variable needed by the largest planned model.

```{r}
#| label: template-same-sample-check
#| eval: false

tibble(
  Model = c("Model 1", "Model 2", "Model 3"),
  N = c(nobs(model1), nobs(model2), nobs(model3))
) |>
  knitr::kable(
    align = "lr",
    caption = "Model sample-size check before nested comparisons"
  )

stopifnot(
  nobs(model1) == nobs(model2),
  nobs(model2) == nobs(model3)
)
```


## Template 1: Linear Diagnostics {#sec-template-linear-diagnostics}

Use the final or primary linear model for diagnostic checks. If you fit staged
models, the diagnostic conclusion usually belongs to the final model.

```{r}
#| label: template-linear-vif
#| eval: false

# Collinearity table.
performance::check_collinearity(model3, ci = FALSE) |>
  as_tibble() |>
  select(Term, VIF, `SE inflation factor` = SE_factor, Tolerance) |>
  round_numeric(digits = 2) |>
  knitr::kable(
    align = "lrrr",
    caption = "Collinearity diagnostics for the final linear model"
  )
```

```{r}
#| label: template-linear-residual-plot
#| eval: false

linear_diagnostics <- tibble(
  fitted = fitted(model3),
  residual = residuals(model3),
  standardized_residual = rstandard(model3),
  cooks_distance = cooks.distance(model3)
)

ggplot(linear_diagnostics, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey50") +
  geom_point(alpha = 0.55) +
  geom_smooth(se = FALSE, color = "steelblue") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  )
```

```{r}
#| label: template-linear-influence
#| eval: false

# This chunk repeats the diagnostic data so it can be copied on its own.
linear_diagnostics <- tibble(
  fitted = fitted(model3),
  residual = residuals(model3),
  standardized_residual = rstandard(model3),
  cooks_distance = cooks.distance(model3)
)

linear_diagnostics |>
  arrange(desc(cooks_distance)) |>
  slice_head(n = 10) |>
  round_numeric(digits = 3) |>
  knitr::kable(
    align = "rrrr",
    caption = "Largest Cook's distance values for the final linear model"
  )
```

```{r}
#| label: template-linear-check-model
#| eval: false

# Optional dashboard for the appendix.
performance::check_model(
  model3,
  check = c("linearity", "homogeneity", "qq", "outliers")
)
```

**Write-up template:** "The linear diagnostics were
`[acceptable/questionable/a serious concern]`. VIF values were below
`[cutoff]`, the residual plots showed `[pattern]`, and Cook's distance
`[did/did not]` suggest that a small number of cases dominated the model.
Therefore, I treat the main coefficients as `[reasonably stable/limited by ...]`."


## Template 2: Logistic Diagnostics {#sec-template-logistic-diagnostics}

Logistic diagnostics are not the same as linear diagnostics. Do not check
residual normality or constant residual variance for logistic regression.

```{r}
#| label: template-logistic-event-count
#| eval: false

# Always name the event: here, manager = 1.
analysis_df |>
  count(manager, name = "n") |>
  mutate(percent = scales::percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(
    align = "lrr",
    caption = "Event and non-event counts for the logistic outcome"
  )
```

```{r}
#| label: template-logistic-sparse-cells
#| eval: false

# Sparse-cell check for a key categorical predictor.
analysis_df |>
  count(occupation, manager, name = "n") |>
  arrange(occupation, manager) |>
  knitr::kable(
    align = "llr",
    caption = "Sparse-cell check by occupation"
  )
```

```{r}
#| label: template-logistic-convergence-vif
#| eval: false

# Convergence check.
logit2$converged

# Collinearity check.
performance::check_collinearity(logit2, ci = FALSE) |>
  as_tibble() |>
  select(Term, VIF, `SE inflation factor` = SE_factor, Tolerance) |>
  round_numeric(digits = 2) |>
  knitr::kable(
    align = "lrrr",
    caption = "Collinearity diagnostics for the logistic model"
  )
```

```{r}
#| label: template-logistic-linearity-logit
#| eval: false

# Simple planned check for nonlinearity in continuous predictors.
logit2_nonlinear <- glm(
  manager ~ female + schyrs_c + I(schyrs_c^2) +
    exper_c + exper2_c +
    occupation + numeracy + I(numeracy^2),
  data = analysis_df,
  family = binomial
)

stopifnot(nobs(logit2) == nobs(logit2_nonlinear))

anova(logit2, logit2_nonlinear, test = "Chisq")
```

```{r}
#| label: template-logistic-influence
#| eval: false

logistic_influence <- tibble(
  fitted_probability = fitted(logit2),
  cooks_distance = cooks.distance(logit2),
  leverage = hatvalues(logit2)
)

logistic_influence |>
  arrange(desc(cooks_distance)) |>
  slice_head(n = 10) |>
  round_numeric(digits = 3) |>
  knitr::kable(
    align = "rrr",
    caption = "Largest Cook's distance values for the logistic model"
  )

logistic_influence |>
  summarise(
    min_fitted = min(fitted_probability),
    median_fitted = median(fitted_probability),
    max_fitted = max(fitted_probability)
  ) |>
  round_numeric(digits = 3) |>
  knitr::kable(
    align = "rrr",
    caption = "Range of fitted probabilities for the logistic model"
  )
```

**Write-up template:** "The logistic model converged, VIF values were below the
course threshold, and sparse-cell checks did/did not identify a major separation
concern. The linearity-in-the-logit check did/did not suggest that the simple
linear logit specification is adequate for the continuous predictors."

Expanded version: "The logistic diagnostics were
`[acceptable/questionable/a serious concern]`. Sparse-cell checks showed
`[pattern]`, the model `[did/did not]` converge without warnings, VIF values
were `[low/high]`, influence checks `[did/did not]` change the focal
conclusions, and fitted probabilities `[did/did not]` pile up near 0 or
1. Classification, ROC/AUC, and calibration were `[not used because the project
is explanatory/used because the project is predictive]`."


## Template 3: Optional Prediction Diagnostics {#sec-template-optional-prediction-diagnostics}

Classification and ROC/AUC are optional unless your research question is about
prediction or classification. If you include them, justify the cutoff and
remember that good classification does not prove causal or explanatory validity.

```{r}
#| label: template-logistic-classification-optional
#| eval: false

threshold <- 0.50

classification_df <- tibble(
  observed = analysis_df$manager,
  predicted_probability = fitted(logit2),
  predicted_class = if_else(predicted_probability >= threshold, 1, 0)
)

classification_df |>
  count(observed, predicted_class, name = "n") |>
  knitr::kable(
    align = "lrr",
    caption = "Optional classification table at the chosen cutoff"
  )
```

**Use only when:** your question asks about prediction, screening,
classification, or decision rules. For explanatory projects, a short note that
classification diagnostics are not central is usually enough.


## Clean-Session Checklist {#sec-clean-session-checklist}

Before submitting, restart R and run the file from top to bottom. If you use
Quarto, render the `.qmd` file rather than only running chunks interactively.

- [ ] The source file loads every package it uses.
- [ ] The source file loads the data without an absolute path tied to one
  computer.
- [ ] All cleaning, recoding, factor reference levels, and `analysis_df`
  construction happen in the submitted file.
- [ ] Descriptive tables, models, diagnostics, predictions, tables, and figures
  are recreated by the submitted file.
- [ ] Nested model comparisons use the same rows and pass `nobs()` checks.
- [ ] Every table or figure in the report is produced by code, not manually
  edited after rendering.
- [ ] Project 2 names the modeled event in code comments, tables, figures, and
  prose.
- [ ] The file runs after clearing the Environment pane or restarting R.
- [ ] The final render has no unresolved cross-references, missing figures, or
  chunk errors.
- [ ] `sessionInfo()` or `sessioninfo::session_info()` is included only if your
  instructor asks for a package/version record.


## Quarto or R Script? {#sec-quarto-r-script-checklist}

| If you submit... | Do this | Avoid this |
|---|---|---|
| Quarto report (`.qmd`) | Use chunk labels, captions, and cross-references for final tables and figures | Relying on objects created interactively but not in the file |
| R script (`.R`) plus separate report | Use comments that match the report sections and save/export every final output | Pasting unexplained console output into the report |
| Either format | Load packages, load data, create `analysis_df`, fit models, run diagnostics, and recreate outputs in order | Absolute paths, hidden global objects, raw `summary()` dumps, or code that only works on your computer |

::: {.callout-tip}
## Simple Source-File Order
Use the same order as the project: packages, data loading, cleaning and
complete-case construction, descriptives, models, diagnostics, tables, figures,
predictions, and session check.
:::

For a plain R script, a minimal package setup can be as simple as:

```{r}
#| label: template-minimal-package-setup
#| eval: false

options(repos = c(CRAN = "https://cloud.r-project.org"))

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  dplyr, tidyr, forcats, tibble,
  ggplot2, scales,
  parameters, performance,
  marginaleffects,
  gtsummary, gt,
  here, knitr
)
```

After constructing `analysis_df`, use a small preflight block:

```{r}
#| label: template-analysis-preflight
#| eval: false

names(analysis_df)
nrow(analysis_df)

analysis_df |>
  summarise(
    missing_values = sum(is.na(pick(everything()))),
    final_n = n()
  )
```


## Troubleshooting Notes {#sec-troubleshooting-notes}

| Symptom | Common cause | First thing to check |
|---|---|---|
| `object not found` | The object was created interactively but not in the source file | Restart R and run from the top |
| `could not find function` | Package not loaded | Add `library(package)` near the top |
| `file not found` | Absolute path or wrong working directory | Use a project-relative path or `here::here()` |
| `undefined columns selected` | Variable is not in the data frame being used | Check spelling with `names(analysis_df)` |
| Nested `anova()` fails | Models use different rows | Build `analysis_df` with all model variables before fitting models |
| Factor reference category is wrong | R chose alphabetical order | Use `forcats::fct_relevel()` and check `levels()` |
| `glm.fit: fitted probabilities numerically 0 or 1 occurred` | Sparse cells or near-separation | Check event counts by key categories and interpret cautiously |
| Very large logistic standard errors | Sparse categories or near-separation | Inspect cross-tabs and consider a simpler coding scheme |
| VIF is high | Redundant predictors, uncentered interactions, or duplicate coding | Check whether variables encode the same information |
| Cross-reference appears as `?@...` | Missing or duplicate label, or section not in the book | Check the label spelling and render the full book |
| Plot/table appears in the appendix but not the report | Output was created but not interpreted | Add a one-sentence diagnostic conclusion in Part 5 or Part 6 |


## Appendix Placement Guide {#sec-appendix-placement-guide}

Keep the main report readable. Put evidence where it helps the reader most.

| Put in the main report | Put in the appendix |
|---|---|
| Final descriptive tables and key exploratory figures | Long supporting tables, sparse-cell checks, and supplementary descriptives |
| Publication-style regression table | Complete source file |
| Model comparison and fit summary | Full diagnostic plots or dashboards |
| Short diagnostic conclusion | Influence tables and sensitivity checks |
| Predicted probability figure/table required for Project 2 | Optional classification/ROC/calibration output, if used |

Do not include every failed attempt, every raw console printout, or every
intermediate object. The appendix should show reproducibility and support the
report, not become a code dump.
